### 2023-11-22 iterative_monomorphisation.tex
basic outline

\documentclass{article}

\setlength{\parindent}{0pt}

\begin{document}

TODO:
    \begin{itemize}
        \item contact Sascha Boehme about authorship
        \item are there any implementation of this (or comparable) algorithm in that have been evaluated?
        \item finish implementation
        \item test and evaluate implementation
        \item (can wait) merge implementation (review code with active contributor beforehand)
        \item find pretty inline setup for pseudo code
        \item (in order) write high/low level algo
        \item write implementation and evaluation
        \item write background
        \item write related work
        \item write conclusion
        \item write intro
        \item reread, spellcheck ...
    \end{itemize}

% target page number: 2-2.5
\section{Introduction}
\begin{itemize}
    \item context and motivation (automatic theorem proving, better performance)
    \item what we are trying to do (monomorphisation)
    \item basic idea of the algorithm 
    \item overview of the paper
\end{itemize}

% target page number: 0.5-1.5
\section{Background}

\begin{itemize}
    \item Language, TF1 (might be TH1 or something else)
\end{itemize}

% target page number: 2-4
\section{High level algorithm}

Mathematical description of the algorithm. Set notation and implicit loops.

% target page number: 2-4
\section{Low level algorithm}

Precise algo description (pseudo code). Explicit loops, and bounds.

% target page number: 2-4
\section{Implementation and evaluation}

Introduce context of zipperposition (do it in intro or background?).

Explain how the final bounds were chosen (i liked the way those numbers looked)

Present evaluation method (which problems were used, options used, server characteristics ect...)

Compare the following:
    \begin{itemize}
        \item no call to e (no monomorphisation)
        \item call e on start no monomorphisation
        \item call e (default) no monomorphisation
        \item call e on start monomorphisation
        \item call e (default) monomorphisation
        \item maybe compare a few different bounding heuristics?
    \end{itemize}

Compare performance with implementation of a similar algorithm (if any).

Maybe compare with implementation of other kinds of monomorphisation algorithms (if appropriate).

% target page number: 0.5-1
\section{Related work}

\begin{itemize}
    \item that part of Sascha Boehme's thesis (Proving Theorems of Higher-Order Logic with SMT Solvers)
    \item How to Encode Polymorphic Types Safely and Efﬁciently
    \item Encoding Monomorphic and Polymorphic Types
\end{itemize}

% target page number: 0.5-1
\section{Conclusion}

\end{document}

### 2024-01-08 iterative_monomorphisation.tex
started writting intro

@@
 \documentclass{article}
+
+\usepackage{soul}
 
@@
     \begin{itemize}
-        \item contact Sascha Boehme about authorship
+       \item \st{contact Sascha Boehme about authorship}
         \item are there any implementation of this (or comparable) algorithm in that have been evaluated?
-        \item finish implementation
-        \item test and evaluate implementation
-        \item (can wait) merge implementation (review code with active contributor beforehand)
+        \item finish implementation (mostly there)
+        \item test and evaluate implementation (well on the way)
+        \item (can wait) merge implementation (review code with active contributor(s) beforehand)
         \item find pretty inline setup for pseudo code
@@
     \item context and motivation (automatic theorem proving, better performance)
+       Automatic theorem provers have many applications, and are particularly useful in assisting in the elaboration of formal proofs.
+Speed is one of the main limitations of the use of theorem provers in this application.
+For instance, when used as a backend to a proof assistant a prover can often be expected to run for no more than a few seconds for a given instance of a problem.
     \item what we are trying to do (monomorphisation)
+       This paper presents a method, first introduced in [sacha's thesis] [as far as i know], that allows for higher order polymorphic problems to be approximated monomorphically.
+This algorithm aims to acheive the following:
+         \begin{itemize}
+            \item allow for existing higher order monomorphic theorem provers to be extended [albeit very crudely] to higher order polymorphic languages
+            \item improve the performance of existing higher order polymorphic theorem provers
+         \end{itemize}
+
+         It is therefore aimed at being practically useful, as a result it is not complete [ref to example in the paper where we explain that?].
+
     \item basic idea of the algorithm 
@@
 
-Mathematical description of the algorithm. Set notation and implicit loops.
+We consider a problem \(P\) to be a set of clauses \(C_1, \dots, C_n\).
+
+Algorithm :
+\begin{itemize}
+    \item Let \(F\) be the set of all function symbols in \(P\).
+    \item We consider \(M\) a map from the function symbols of \(F\) to the set of monomorphic type arguments with which the function symbol is instantiated \(P\) [unclear]
+
+    For instance if \(f^{a \rightarrow b}\) occurs in \(P\) (with \(a\) and \(b\) monomorphic types) then, \(a \rightarrow b \in M(f) \).
+    \item We consider \(P\) a map from the function symbols of \(F\) to the set of non-monomorphic type arguments with which the function symbol is instantiated in \(P\).
+[still somewhat unclear] This map is additionally paramatrised by the clause in which the function symbol is instantiated.
+
+    This is important because type arguments are quantified at the clause level. 
+\end{itemize}
 

### 2024-01-15 iterative_monomorphisation.tex
continued working on draft

--- 2023-11-22 iterative_monomorphisation.tex
@@
 \documentclass{article}
+
+\usepackage{soul}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{amsmath}
 
@@
 \begin{itemize}
-    \item context and motivation (automatic theorem proving, better performance)
-    \item what we are trying to do (monomorphisation)
-    \item basic idea of the algorithm 
+    \item context and motivation (automatic theorem proving, better performance)\\
+       Automatic theorem provers have many applications, and are particularly useful as backends to proof assistants.
+Speed is one of the main limitations of the use of theorem provers in this setting.
+Indeed, a prover can often be expected to run for no more than a few seconds for a given instance of a problem.
+    \item what we are trying to do (monomorphisation)\\
+       This paper presents a method, first introduced in [sacha's thesis] [as far as i know], that allows for higher order polymorphic problems to be approximated monomorphically.
+This algorithm aims to acheive the following:
+         \begin{itemize}
+            \item allow for existing higher order monomorphic theorem provers to be extended [albeit very crudely] to higher order polymorphic languages
+            \item improve the performance of existing higher order polymorphic theorem provers on polymorphic problems
+         \end{itemize}
+
+         It is therefore aimed at being practically useful, to this end it forgoes completeness, this is the main theoretical weakness of this algorithm [the real main theoretical weakness would be that it is unsound, but i've done my best to make sure it isn't].
+[ref to example in the paper where we explain that?].
+
+    \item basic idea of the algorithm\\
+    A given problem is seen as a set of polymorphic clauses.
+Type variables of the problem are quantified at the clause level.
+The type arguments of the problem are collected and the uninstantiated type arguments are matched against those that contain no type variables.
+This yields new type arguments, both uninstantitated and ground, the process is then repeated making use of the newly derived type arguments.
+
+    The new type arguments we have derived are used to create new clauses, without type variables.
+This can lead to the same clause being duplicated as one type variable may be instantiated in a number of different ways.
+    A final step is required as the newly generated problem, whilst it does not contain type variables, still makes use of type constructors.
+Those can be transparently erased through the mangling method presented in [relevant paper].
     \item overview of the paper
+
+       This paper will first give some general background and set the theoretical framework that will be considered.
+Then a high-level mathematical description of the algorithm will be presented.
+A more detailed low-level description of the algorithm corresponding to its actual implementation will follow.
+       Finally the implementation and evaluation of the algorithm in zippeposition [the award winning higher order theorem prover] [introduce zipperposition before] will be discussed.
+
+       [issue with intro: focuses too much on theoretical aspect bearly mentions zipp or implementation]
 \end{itemize}
@@
 
+We want our starting logic to be:
+\begin{itemize}
+    \item higher order, we want lambdas, function symbols and the rest
+    \item polymorphic, we want atomic types, type constructors and type variables
+\end{itemize}
+and our target logic to be the same thing without polymorphism make sure to point out that the mono logic is a subset of the poly logic
+
+Definitions we need:
+\begin{itemize}
+   \item type arguments (probably part of logic)
+   \item ground and uninstantiated types 
+   \item vaguely define what matching is
+   \item define what a term, literal and clause are
+\end{itemize}
+Need to define ground types: without type variables and uninstantiated types, with type variables
+
+DO NOT use monomorphic to mean ground and polymorphic to mean uninstantiated, everything is polymorphic until we mangle, then everything is monomorphic
+
+Unsure of which terms to use: ground and non-ground, or instantiated and uninstantiated or both
+
 % target page number: 2-4
@@
 
-Mathematical description of the algorithm. Set notation and implicit loops.
+We consider a problem \(P\) to be a set of clauses \(C_1, \dots, C_n\).
+
+Algorithm :
+\begin{itemize}
+    \item Let \(F\) be the set of all function symbols in \(P\).
+    \item Given a function symbol \(f \in F\), we define \(I(f)\), the set of all ground type arguments that are passed to \(f\) is instantiated in \(P\).
+
+    For instance if \(f^{a \rightarrow b}\) occurs in \(P\) (with \(a\) and \(b\) ground types) then, \(a \rightarrow b \in I(f) \).
+ \item Given a clause \(C \in P\) and a function symbol \(f \in C\), we define \(U(C, f)\), the set of uninstantiated type arguments passed to \(f\) in \(C\).
+
+    It is important to paramatrise \(U\) with the clause of \(f\) because type variables are quantified at the clause level.
+This is not necessary for \(I\) because its type arguments do not contain type variables.
+[seems a sufficient explanation, if not clear expand further]
+
+\end{itemize}
+
+A single iteration of the algorithm will do the following:
+\begin{itemize}
+   \item For each clause \(C_i \in P\) and each function symbol \(f \in C_i\), we consider the uninstantiated type arguments \(U(C_i, f)\) these type type arguments are matched against all ground type arguments of \(I(f)\)
+
+      [We have an issue: we don't define matching, don't think we need to go into a lot of detail]
+
+      Each such matching may return a substitution which instantiates one or more type variables of \(C_i\) (that would also be in the type arguments of \(f\))
+
+   \item Once all such substiutions have been computed.
+We apply each substitutions to the clause they were generated from.
+These newly created clauses are added to the problem and the newly generated type arguments (both ground and uninstantiated) will be added to \(U\) and \(I\) to be used for the next iteration.
+\end{itemize}
+
+After a set number of iterations, we take all clauses with no type variables and mangle their types [see relevant paper].
+The resulting set of clauses forms a problem \(P'\) with the following properties:
+\begin{itemize}
+   \item \(P'\) is monomorphic [rewrite that as saying \(P'\) is in the monomorphic fragment of the original logic]
+   \item if a contradiction can be derived from \(P'\), then a contradiction can be derived from \(P\) [soundness]
+   \item we do not however have the converse [completeness]
+\end{itemize}
+
+[Add 2 or 3 examples to show how it works and explain why we have the different steps]
+
+\subsection{Incompleteness}
+
+This algorithm is incomplete, this is because not all possible instantiations of the type variables are generated.
+Consider the following problem: 
+\begin{align*}
+   P = \{ & \forall \alpha. f \langle \alpha \rightarrow int \rangle \\
+          & \forall \alpha. f \langle nat \rightarrow \alpha \rangle \}
+\end{align*}
+
+The algorithm only matches against ground type arguments.
+This is the case because polymorphic type unification is undecidable (quick ref) and whilst interweaving phases of unification with the current algorithm would be theoretically possible, it would most likely not be of any practical use.
+Indeed, the current implementation is already very explosive and generating new substitutions with a costly unification procedure would most likely not result in any improvements.
+
+Therefore, the previous problem would never generate the clause \(f \langle nat \rightarrow int \rangle\).
+[having trouble finishing the example, we would need for the nat int clause to allow finding a contradiction
+without allowing the other two to generate this one]
 
@@
 
+This section presents a more detailed description of the algorithm, it follows closely the implementation in zipperposition.
+
+The main departures from the high-level description of the algorithm are the following:
+\begin{itemize}
+   \item the substitutions are directly applied to the type arguments, instead of the clauses, this also dispenses us from having to re-extract the type arguments from the clauses at each iteration
+   \item as a result the new clauses are generated only once all substitutions have been computed and iteratively applied to the various type arguments
+   \item some operations are performed in a different order to avoid useless computations [which ones?]
+   \item bounds are used at almost all steps of the process as the various enumerations are highly explosive and require bounds in order to be useful in a practical setting
+   \item the various sets are implemented with maps
+\end{itemize}
+
 Precise algo description (pseudo code). Explicit loops, and bounds.
+
+[note that this is all very ugly and will be made beautiful once the first draft (of the algorithm) is done]
+
+\begin{algorithm}
+\begin{algorithmic}[1]
+\Function{Type arguments}{Literal: lit}
+
+   \State let \(t_1\), \(t_2\) be the terms of \(lit\)
+   \State recursively collect the function symbols and associated type arguments of \(t_1\) and \(t_2\) into a list of pairs
+
+   \State \Comment Going a bit fast here, probably need to reformulate, but don't go into details, it's not important
+
+   \State \Return the list of (function symbol, type argument) pairs
+
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+\begin{algorithmic}[1]
+\Function{Initialisation}{Problem: P}
+   \ForAll{clause \(C \in P\)}
+      \ForAll{literal \(lit \in C\)}
+         \State let fun\_sym\_ty\_args = Type arguments(lit)
+         \ForAll{(fun\_sym, ty\_args) in fun\_sym\_ty\_args}
+         \If{\( \forall ty\_arg \in ty\_args. is\_ground(ty\_arg)\) }
+
+            \State \(I(fun\_sym) = I(fun\_sym) \cup \{ty\_args\}\)
+
+         \Else
+
+            \State \(U(C, fun\_sym) = U(C, fun\_sym) \cup \{ty\_args\}\)
+
+         \EndIf
+         \EndFor
+      \EndFor
+   \EndFor
+
+   \State \Return (I, U)
+
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+\begin{algorithmic}[1]
+\Function{Iteration}{Problem: P, U, I, S}
+   \ForAll {clause \(C \in P\)}
+      \ForAll {\(fun\_sym, ty\_args \in U(C)\)}
+      \ForAll {\( ty\_args' \in I(fun\_sym) \)}
+
+            \State \(subst = match(ty\_args, ty\_arg')\)
+
+            \State \Comment the match may fail in which case we simply continue with the loop
+
+            \State \(S(C) = S(C) \cup \{subst\}\)
+         \EndFor
+      \EndFor
+
+   \EndFor
+
+   \ForAll {clause \(C \in P\)}
+      \ForAll {\(subst \in S(C)\)}
+         \ForAll {\(fun\_symbol, ty\_args \in U(C)\)}
+            \State \(ty\_args' = apply(subst, ty\_args)\)
+
+            \If {\( \forall ty\_arg \in ty\_args'. is\_ground(ty\_arg)\) }
+
+               \State \(I(fun\_sym) = I(fun\_sym) \cup \{ty\_args'\}\)
+            \Else
+
+               \State \(U(C, fun\_sym) = U(C, fun\_sym) \cup \{ty\_args'\}\)
+            \EndIf
+
+         \EndFor
+      \EndFor
+   \EndFor
+
+   \State \Return (I, U, S)
+
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+\begin{algorithmic}[1]
+\Function{Generate clauses}{Problem: P, Substitutions: S}
+
+\ForAll {clause \(C \in P\)}
+   \ForAll {\(subst \in S(C)\)}
+      \State \(C' = apply(subst, C)\)
+
+      \If {\(C'\) is ground}
+
+         \State \(P = P \cup \{C'\}\)
+      \EndIf
+
+   \EndFor
+\EndFor
+
+\State \Return \(P\)
+
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
 
@@
     \begin{itemize}
-        \item no call to e (no monomorphisation)
-        \item call e on start no monomorphisation
-        \item call e (default) no monomorphisation
-        \item call e on start monomorphisation
-        \item call e (default) monomorphisation
-        \item maybe compare a few different bounding heuristics?
+        \item no call to e (no monomorphisation) (using portfolio mode, this serves as reference)
+        \item call e on start
+        \item call e (default) monomorphisation (this is too late)
+        \item call e at 0.05 * timeout
+        \item call e at 0.1 * timeout (may still be too late)
+        \item call e at some time before 0.05 * timeout ?
+        \item different bounds worth trying: give up on the 2000 clause maximum
+              add a floor for polymorphic type args and also a ceiling 
+              messa around with multipliers a little bit
+         \item perhaps look into some dynamic bounds
+         \item consider calling a different first order prover
     \end{itemize}
@@
 
-Maybe compare with implementation of other kinds of monomorphisation algorithms (if appropriate).
+Maybe compare with implementation of other kinds of monomorphisation algorithms (if appropriate). This would require implementing a guards or tags method which could take a while
+
+Comments on results:
+So far from the testing i've done:
+\begin{itemize}
+   \item we gain very few problems compared to portfolio mode, this is very bad
+   \item these can easily be lost with small bound adjustments, this isn't great but not catastriphic
+   \item almost all newly gained problems are solved with the 0.0 e call, this one is catastrophic
+\end{itemize}
 

### 2024-01-15 iterative_monomorphisation.tex
continued expanding on draft focusing mostly on the detailed description of the algorihtm

@@
       \item \st{contact Sascha Boehme about authorship}
+
+      \item it seems that dissentangling the implemntation and low-level description is quite difficult and may not be appropriate, it might be best to instead fuse low-level description and implementation and then split the section into substitution generation and clause generation
+
       \item are there any implementation of this (or comparable) algorithm in that have been evaluated?
+      \item read mangling paper in detail
       \item finish implementation (mostly there)
@@
 
+\subsection{Soundness}
+[this isn't an actual proof, more of a justification, mostly a joke in its present form]
+Soundness of the algorithm follows from the fact that the resulting clauses are obtained by directly instantiating universally quantified type variable of the initial clauses.
+Then we rely upon the soundness of mangling [include reference].
+
+\subsection{Multiple type arguments}
+Comment that a function symbol with multiple type arguments will often see them instantiated one at a time per iteration.
+Conclude that it may be beneficial to remove clauses with function symbols with too many uninstantiated type arguments.
+Nuance that if the more there are such symbols, the more likely we are to generate substitutions that instantiate several type arguments at a time.
+Note that in practice setting a type variable limit changes nothing.
+[more tests needed for showing that
+[need to add a couple of tests to show that]
+[maybe this section doesn't belong here, probably better in the low level section]
+
 \subsection{Incompleteness}
@@
 This algorithm is incomplete, this is because not all possible instantiations of the type variables are generated.
-Consider the following problem: 
+Consider the following problem [note that we need a consistent notation for type arguments (notably to distinguish them from normal arguments) also this might not be an issue but they are terms not literals ]:
 \begin{align*}
    P = \{ & \forall \alpha. f \langle \alpha \rightarrow int \rangle \\
-          & \forall \alpha. f \langle nat \rightarrow \alpha \rangle \}
+          & \forall \alpha. f \langle nat \rightarrow \alpha \rangle \\
+          & \forall \alpha \beta. \lnot f \langle \alpha \rightarrow \beta \rangle \}
 \end{align*}
@@
 
-Therefore, the previous problem would never generate the clause \(f \langle nat \rightarrow int \rangle\).
-[having trouble finishing the example, we would need for the nat int clause to allow finding a contradiction
-without allowing the other two to generate this one]
+Therefore, the previous problem would never instantiate a clause, despite the problem becoming trvially contradictory if the the last clause and any of the first two are instantiated.
 
@@
 \end{algorithm}
+In the zipperposition implemntation this iteration procedure is limited by several bounds.
+The number of new type arguments that can be generated for each function symbol is limited by user-defined bounds.
+These bounds can be defined proportionally.
+Additionally, the number of new type arguments can be limited at the clause level.
+In both cases, the bounds for the new ground type arguments and non-ground type arguments can be set independently.
+
+A consequences of these bounds is that in order to avoid unecessary computations, the program will iteratively apply substitutions until either the ground or non-ground bound is reached.
+Then substitutions can be discarded depending on whether they instantiate or not a given set of type arguments
+
+Additionally, there are bounds for the number of substitutions that are generated.
+
+[So far, after more or less blind experimentation, I've settled on severely limiting new non-ground type arguments because they used to balloon out of control, whilst the growth in ground type arguments is more manageable]
+[It's also worth suggesting the possibility of a more sophisticated implementation that could allow for dynamic bounds]
+
+\begin{algorithm}
+\begin{algorithmic}[1]
+\Function{Instantiate clause}{Clause: C, Substitutions S}
+
+   \If{\(C\) is ground}
+
+      \State \Return \(\{C\}\)
+
+   \Else
+
+      \State let \(v\) be a type variable in \(C\)
+      \State let new\_clauses be an empty set of clauses
+
+      \ForAll{subst \(\in S(C)\)}
+
+         \If{subst instantiates \(v\)}
+
+            \State \(C' = apply(subst, C)\)
+
+            \State let v\_clauses = Instantiate clause(C', S)
+
+         \EndIf
+
+         \State \( new\_clauses = new\_clauses \cup v\_clauses\)
+
+      \EndFor
+
+      \State \Return new\_clauses
+   \EndIf
+
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
+This function differs from the implementation in a significant way, indeed, through experimentation it was found that the generated clauses were far more likely to be solved by E if the substitutions that are combined to instantiate the clause were derived from the same iteration.
+In practice, we therefore store the iteration at which each substition is generated and at the instantiation phase only apply substitutions from the same iteration.
+
+[need to make a handful of tests to check that]
+
+What the previous function does is essentially going through the tree of possible substitutions that instantiate the clause and collecting all possible instantiations.
+This is extremely explosive, and a hard cap on the number of clauses we want to generate is almost indispensable for any implementation.
 
@@
 \ForAll {clause \(C \in P\)}
-   \ForAll {\(subst \in S(C)\)}
-      \State \(C' = apply(subst, C)\)
-
-      \If {\(C'\) is ground}
-
-         \State \(P = P \cup \{C'\}\)
-      \EndIf
-
-   \EndFor
+   \State let new\_clauses = Instantiate clause(C, S)
+   \State \(P = P \cup new\_clauses\)
 \EndFor
@@
 Present evaluation method (which problems were used, options used, server characteristics ect...)
+
+Provide complete list of problems explain which ones we excluded and why: for some it was lack of axiom files, a handful of others were simply too large
 
@@
 \begin{itemize}
-   \item we gain very few problems compared to portfolio mode, this is very bad
+   \item we gain very few problems compared to portfolio mode, this is very bad especially since we only look at polymorphic problems, where this algorithm should shine at least in proportional terms
    \item these can easily be lost with small bound adjustments, this isn't great but not catastriphic
@@
 
+% summary 
+In conclusion, this paper has given a theoretical and practical description of an iterative monomorphisation algorithm.
+The practical description gives some insights on how it may be efficiently implemented.
+Additionally, its implementation in zipperposition demonstrates its practical feasibility.
+
+However, the evaluation of this implementation against the TPTP benchmarks show dissapointing results in light of the efficiency of similar ideas in other frameworks.
+Indeed, zipperposition can only solve a handful of additional problems thanks to monomorphisation.
+Despite these issues, a refinment of the existing implementation paired with a better choice of bounds and options may allow for an improvement in results [decent enough to warrant having spent so much time on it]
+
+% Future avenues
+
 \end{document}

### 2024-01-22 iterative_monomorphisation.tex
updated doc

@@
 \usepackage{soul}
+\usepackage{booktabs}
+\usepackage{multirow}
 \usepackage{algorithm}
@@
    \begin{itemize}
-      \item \st{contact Sascha Boehme about authorship}
 
-      \item it seems that dissentangling the implemntation and low-level description is quite difficult and may not be appropriate, it might be best to instead fuse low-level description and implementation and then split the section into substitution generation and clause generation
-
-      \item are there any implementation of this (or comparable) algorithm in that have been evaluated?
+      \item split the problems in two sets one for refining parameters and bounds, the other for overall evaluation
       \item read mangling paper in detail
+      \item set up tables
       \item finish implementation (mostly there)
-      \item test and evaluate implementation (well on the way)
+      \item test and evaluate implementation 
       \item (can wait) merge implementation (review code with active contributor(s) beforehand)
@@
 
-    \item basic idea of the algorithm\\
+    \item basic idea of the algorithm
+
     A given problem is seen as a set of polymorphic clauses.
@@
 \section{Background}
+
+[Start out by mentionning popular type encodings, such as tags and gards, explain the interest of this method]
+Why we didn't implement the types / guards method: in part because it relies on inferences on types that are given in Isabelle but we would have had to do ourselves
 
@@
 
-Therefore, the previous problem would never instantiate a clause, despite the problem becoming trvially contradictory if the the last clause and any of the first two are instantiated.
+The previous problem would therefore never instantiate a clause, despite the problem becoming trvially contradictory if the the last clause and any of the first two are instantiated.
 
+Furthermore, the finite number of iterations of the algorithm consitutes an additional source of incompleteness in practice.
+[find example that would require a list list list int to be solved but only allow such a type to be generated after 3 iterations]
 % target page number: 2-4
@@
 
-\begin{algorithm}
+\begin{algorithm}[tbh]
 \begin{algorithmic}[1]
@@
 
-\begin{algorithm}
+This function is very basic and simply gathers all the type arguments of the various function symbols of the problem and sorts them depending on whether they are ground or not.
+
+\begin{algorithm}[tbh]
 \begin{algorithmic}[1]
@@
 A consequences of these bounds is that in order to avoid unecessary computations, the program will iteratively apply substitutions until either the ground or non-ground bound is reached.
-Then substitutions can be discarded depending on whether they instantiate or not a given set of type arguments
+Then substitutions can be discarded depending on whether they instantiate a given set of type arguments or not.
 
@@
 
-\State \Return \(P\)
+\State \(P' = \emptyset \)
+
+\ForAll {clause \(C \in P\)}
+   \If{\(C\) is ground}
+      \State \(P' = P' \cup \{mangle(C)\}\)
+   \EndIf
+\EndFor
+
+\State \Return \(P'\)
 
@@
 % target page number: 2-4
-\section{Implementation and evaluation}
+\section{Evaluation}
 
@@
     \begin{itemize}
-        \item no call to e (no monomorphisation) (using portfolio mode, this serves as reference)
-        \item call e on start
-        \item call e (default) monomorphisation (this is too late)
-        \item call e at 0.05 * timeout
-        \item call e at 0.1 * timeout (may still be too late)
-        \item call e at some time before 0.05 * timeout ?
-        \item different bounds worth trying: give up on the 2000 clause maximum
-              add a floor for polymorphic type args and also a ceiling 
-              messa around with multipliers a little bit
-         \item perhaps look into some dynamic bounds
-         \item consider calling a different first order prover
-    \end{itemize}
+        \item e on original portfolio mode
+        \item e with original portfolio mode and monomorphisation
+        \item adjusted portforlio mode
 
-Compare performance with implementation of a similar algorithm (if any).
+        \item when e is called
+        \item using zipp only as monomorphiser (passing the result to E, vampire, satallax, leo3)
 
-Maybe compare with implementation of other kinds of monomorphisation algorithms (if appropriate). This would require implementing a guards or tags method which could take a while
-
-Comments on results:
-So far from the testing i've done:
-\begin{itemize}
-   \item we gain very few problems compared to portfolio mode, this is very bad especially since we only look at polymorphic problems, where this algorithm should shine at least in proportional terms
-   \item these can easily be lost with small bound adjustments, this isn't great but not catastriphic
-   \item almost all newly gained problems are solved with the 0.0 e call, this one is catastrophic
+        \item test different bounds
     \end{itemize}
@@
 \begin{itemize}
-    \item that part of Sascha Boehme's thesis (Proving Theorems of Higher-Order Logic with SMT Solvers)
-    \item How to Encode Polymorphic Types Safely and Efﬁciently
+    \item part of Sascha Boehme's thesis (Proving Theorems of Higher-Order Logic with SMT Solvers)
+    \item How to Encode Polymorphic Types Safely and Efﬁciently, section 4
+
+    \item Expressing polymorphic types in a many-sorted language
     \item Encoding Monomorphic and Polymorphic Types
@@
 % summary
-In conclusion, this paper has given a theoretical and practical description of an iterative monomorphisation algorithm.
-The practical description gives some insights on how it may be efficiently implemented.
-Additionally, its implementation in zipperposition demonstrates its practical feasibility.
-
-However, the evaluation of this implementation against the TPTP benchmarks show dissapointing results in light of the efficiency of similar ideas in other frameworks.
-Indeed, zipperposition can only solve a handful of additional problems thanks to monomorphisation.
-Despite these issues, a refinment of the existing implementation paired with a better choice of bounds and options may allow for an improvement in results [decent enough to warrant having spent so much time on it]
 

### 2024-01-23 iterative_monomorphisation.tex
started adding tables

@@
 \usepackage{soul}
+\usepackage{caption}
 \usepackage{booktabs}
@@
 
+\break
+
 % target page number: 2-4
@@
 
+\break
+
+\begin{center}
+\captionof{table}{Bounds and options to test}
+\begin{tabular}{ccccc}
+   \toprule
+   \multicolumn{5}{c}{monomorphisation point (steps)} \\
+   0 & 25 & 50 & 100 & 200 \\
+   \midrule
+   \multicolumn{4}{c}{new clauses multiplier} \\
+   1.0 & 1.5 & 2.0 & 3.0 \\
+   \midrule
+   \multicolumn{3}{c}{total new clauses cap} \\
+   2000 & 5000 & none\\
+   \midrule
+   \multicolumn{4}{c}{loop number} \\
+   2 & 3 & 4 & 5 \\
+   \midrule
+   \multicolumn{5}{c}{monomorphisation point (steps)} \\
+   0 & 25 & 50 & 100 & 200 \\
+
+   \bottomrule
+\end{tabular}
+\end{center}
+
+\begin{center}
+\captionof{table}{Monomorphisation to the service of Zipperposition?}
+\begin{tabular}{lccc}
+   \toprule
+   &portfolio classic & portfolio monomorphisation & custom portfolio \\
+   \midrule
+   bounds  & -  & - & - \\
+   bounds2 & - & - & - \\
+   bounds3 & - & - & - \\
+   bounds4 & - & - & - \\
+   bounds5 & - & - & - \\
+
+   \bottomrule
+\end{tabular}
+\end{center}
+
+\bigskip
+\bigskip
+\bigskip
+
+\begin{center}
+\captionof{table}{Monomorphisation as an alternative to native polymorphism?}
+\begin{tabular}{lccccccccc}
+   \toprule
+   & \multicolumn{2}{c}{E}   & \multicolumn{2}{c}{Vampire} &\multicolumn{2}{c}{Leo3} & \multicolumn{2}{c}{Satallax} \\
+   \cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
+   & native & mono & native & mono & native & mono & native & mono\\
+   \midrule
+   bounds  & - & - & - & - & - & - & - & - \\
+   bounds' & - & - & - & - & - & - & - & - \\
+   bounds* & - & - & - & - & - & - & - & - \\
+   bounds? & - & - & - & - & - & - & - & - \\
+   \bottomrule
+\end{tabular}
+\end{center}
+
+\break
+
 % target page number: 0.5-1

### 2024-01-23 iterative_monomorphisation.tex
started considering potential bounds to explore

@@
 \captionof{table}{Bounds and options to test}
-\begin{tabular}{ccccc}
+\begin{tabular}{cccccc}
    \toprule
    \multicolumn{5}{c}{monomorphisation point (steps)} \\
-   0 & 25 & 50 & 100 & 200 \\
+   0 & 25 & 50 & 100 & (200) \\
    \midrule
    \multicolumn{4}{c}{new clauses multiplier} \\
-   1.0 & 1.5 & 2.0 & 3.0 \\
+   1.0 & 1.5 & 2.0 & (3.0) \\
    \midrule
@@
    \multicolumn{5}{c}{monomorphisation point (steps)} \\
-   0 & 25 & 50 & 100 & 200 \\
+   0 & 25 & 50 & 100 & (200) \\
+   \midrule
+   \multicolumn{5}{c}{e-timeout} \\
+   2.0 & 5.0 & 10.0 & 15.0 & no timeout \\
+   \midrule
+   \multicolumn{3}{c}{new poly ty arg limit per fun sym multiplier} \\
+   0 & 0.5 & 1.0  \\
+   \midrule
+   \multicolumn{3}{c}{new poly ty arg limit per fun sym floor} \\
+   0 & 2 & 5  \\
+   \midrule
+   \multicolumn{3}{c}{new poly ty arg limit per fun sym cap} \\
+   10 & 20 & 50  \\
+   \midrule
+   \multicolumn{5}{c}{new mono ty arg limit per fun sym multiplier} \\
+   0..5 & 1.0 & 1.5 & 2.0 & (3.0)\\
+   \midrule
+   \multicolumn{5}{c}{new mono ty arg limit per fun sym floor} \\
+   0 & 2 & 5 & 7 & 10\\
+   \midrule
+   \multicolumn{4}{c}{new mono ty arg limit per fun sym cap} \\
+   10 & 20 & 50 & 100  \\
+   \midrule
+   \multicolumn{3}{c}{monomorphising subst per clause} \\
+   2 & 5 & 7  \\
+   \midrule
+   \multicolumn{5}{c}{subst per clause (total, split 50/50)} \\
+   1000 & 2000 & 5000 & 10000 & 15000  \\
 
@@
 \end{center}
+
+We ignore the following bounds: ty var limit per clause, subst per ty var, limiting the new type args per clause
 

### 2024-01-23 iterative_monomorphisation.tex
split

--- 2024-01-23 iterative_monomorphisation.tex
@@
 
-\setlength{\parindent}{0pt}
-
 \begin{document}
-
-TODO:
-   \begin{itemize}
-
-      \item split the problems in two sets one for refining parameters and bounds, the other for overall evaluation
-      \item read mangling paper in detail
-      \item set up tables
-      \item finish implementation (mostly there)
-      \item test and evaluate implementation 
-      \item (can wait) merge implementation (review code with active contributor(s) beforehand)
-      \item find pretty inline setup for pseudo code
-      \item (in order) write high/low level algo
-      \item write implementation and evaluation
-      \item write background
-      \item write related work
-      \item write conclusion
-      \item write intro
-      \item reread, spellcheck ...
-    \end{itemize}
 
@@
 
+Implementation in Zipperposition.
+
 \break
@@
 
-\begin{center}
-\captionof{table}{Bounds and options to test}
-\begin{tabular}{ccccc}
-   \toprule
-   \multicolumn{5}{c}{monomorphisation point (steps)} \\
-   0 & 25 & 50 & 100 & 200 \\
-   \midrule
-   \multicolumn{4}{c}{new clauses multiplier} \\
-   1.0 & 1.5 & 2.0 & 3.0 \\
-   \midrule
-   \multicolumn{3}{c}{total new clauses cap} \\
-   2000 & 5000 & none\\
-   \midrule
-   \multicolumn{4}{c}{loop number} \\
-   2 & 3 & 4 & 5 \\
-   \midrule
-   \multicolumn{5}{c}{monomorphisation point (steps)} \\
-   0 & 25 & 50 & 100 & 200 \\
-
-   \bottomrule
-\end{tabular}
-\end{center}
+We ignore the following bounds: ty var limit per clause, subst per ty var, limiting the new type args per clause
 

### 2024-01-23 notes_to_selves.tex
split

\documentclass{article}

\usepackage{soul}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}

\begin{document}

\section{TODOs}

TODO:
   \begin{itemize}

      \item split the problems in two sets one for refining parameters and bounds, the other for overall evaluation
      \item read mangling paper in detail
      \item set up tables
      \item finish implementation (mostly there)
      \item test and evaluate implementation 
      \item (can wait) merge implementation (review code with active contributor(s) beforehand)
      \item find pretty inline setup for pseudo code
      \item (in order) write high/low level algo
      \item write implementation and evaluation
      \item write background
      \item write related work
      \item write conclusion
      \item write intro
      \item reread, spellcheck ...
    \end{itemize}

\begin{center}
\captionof{table}{Bounds and options to test}
\begin{tabular}{cccccc}
   \toprule
   \multicolumn{5}{c}{monomorphisation point (steps)} \\
   0 & 25 & 50 & 100 & (200) \\
   \midrule
   \multicolumn{4}{c}{new clauses multiplier} \\
   1.0 & 1.5 & 2.0 & (3.0) \\
   \midrule
   \multicolumn{3}{c}{total new clauses cap} \\
   2000 & 5000 & none\\
   \midrule
   \multicolumn{4}{c}{loop number} \\
   2 & 3 & 4 & 5 \\
   \midrule
   \multicolumn{5}{c}{monomorphisation point (steps)} \\
   0 & 25 & 50 & 100 & (200) \\
   \midrule
   \multicolumn{5}{c}{e-timeout} \\
   2.0 & 5.0 & 10.0 & 15.0 & no timeout \\
   \midrule
   \multicolumn{3}{c}{new poly ty arg limit per fun sym multiplier} \\
   0 & 0.5 & 1.0  \\
   \midrule
   \multicolumn{3}{c}{new poly ty arg limit per fun sym floor} \\
   0 & 2 & 5  \\
   \midrule
   \multicolumn{3}{c}{new poly ty arg limit per fun sym cap} \\
   10 & 20 & 50  \\
   \midrule
   \multicolumn{5}{c}{new mono ty arg limit per fun sym multiplier} \\
   0..5 & 1.0 & 1.5 & 2.0 & (3.0)\\
   \midrule
   \multicolumn{5}{c}{new mono ty arg limit per fun sym floor} \\
   0 & 2 & 5 & 7 & 10\\
   \midrule
   \multicolumn{4}{c}{new mono ty arg limit per fun sym cap} \\
   10 & 20 & 50 & 100  \\
   \midrule
   \multicolumn{3}{c}{monomorphising subst per clause} \\
   2 & 5 & 7  \\
   \midrule
   \multicolumn{5}{c}{subst per clause (total, split 50/50)} \\
   1000 & 2000 & 5000 & 10000 & 15000  \\

   \bottomrule
\end{tabular}
\end{center}

\end{document}

### 2024-01-23 iterative_monomorphisation.tex
table

@@
 
-\begin{center}
-\captionof{table}{Monomorphisation to the service of Zipperposition?}
-\begin{tabular}{lccc}
+\subsection{Parameter optimisation}
+
+\subsection{Monomorphisation as preprocessor}
+
+\begin{table}[h]
+\caption{Evaluation of native polymorphism vs.\ monomorphisation}
+\centering\begin{tabular}{@{}lccc@{}}
    \toprule
-   &portfolio classic & portfolio monomorphisation & custom portfolio \\
+   & Native & Mono & Union \\
    \midrule
-   bounds  & -  & - & - \\
-   bounds2 & - & - & - \\
-   bounds3 & - & - & - \\
-   bounds4 & - & - & - \\
-   bounds5 & - & - & - \\
-
+   E  &   & 0 & 0 \\
+   Leo-III & 0 & 0 & 0 \\
+   Satallax &  & 0 & 0 \\
+   Vampire & 0 & 0 & 0 \\
+   Zipperposition & 0 & 0 & 0 \\[1.5\jot]
+   Total & 0 & 0 & 0 \\
    \bottomrule
 \end{tabular}
-\end{center}
+\end{table}
 
-\bigskip
-\bigskip
-\bigskip
+\subsection{E as Zipperposition backend}
 
-\begin{center}
-\captionof{table}{Monomorphisation as an alternative to native polymorphism?}
-\begin{tabular}{lccccccccc}
+\begin{table}[h]
+\caption{Evaluation of Zipperposition without E vs. with E}
+\centering\begin{tabular}{@{}lccc@{}}
    \toprule
-   & \multicolumn{2}{c}{E}   & \multicolumn{2}{c}{Vampire} &\multicolumn{2}{c}{Leo3} & \multicolumn{2}{c}{Satallax} \\
-   \cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
-   & native & mono & native & mono & native & mono & native & mono\\
+   & without E & with E & Union \\
    \midrule
-   bounds  & - & - & - & - & - & - & - & - \\
-   bounds' & - & - & - & - & - & - & - & - \\
-   bounds* & - & - & - & - & - & - & - & - \\
-   bounds? & - & - & - & - & - & - & - & - \\
+   Zipperposition & 0 & 0 & 0 \\
    \bottomrule
 \end{tabular}
-\end{center}
+\end{table}
 

### 2024-01-24 iterative_monomorphisation.tex
added option test tables

@@
 \usepackage{booktabs}
+\usepackage{array}
 \usepackage{multirow}
@@
 \usepackage{amsmath}
+\usepackage{authblk}
 
@@
 
+\title{Iterative Monomorphisation}
+\author{Jasmin Blanchette, Sascha B\"ohme, Tanguy Bozec}
+\affil{[add affiliation]}
+\date{[date]}
+\maketitle
+\begin{abstract}
+   [Write abstract]
+\end{abstract}
+
 % target page number: 2-2.5
 \section{Introduction}
-\begin{itemize}
-    \item context and motivation (automatic theorem proving, better performance)\\
+%context and motivation (automatic theorem proving, better performance)\\
 Automatic theorem provers have many applications, and are particularly useful as backends to proof assistants.
@@
 Indeed, a prover can often be expected to run for no more than a few seconds for a given instance of a problem.
-    \item what we are trying to do (monomorphisation)\\
-       This paper presents a method, first introduced in [sacha's thesis] [as far as i know], that allows for higher order polymorphic problems to be approximated monomorphically.
+
+%what we are trying to do (monomorphisation)\\
+This paper presents an iterative method, first introduced in [sacha's thesis], that allows for higher order polymorphic problems to be approximated monomorphically.
 This algorithm aims to acheive the following:
    \begin{itemize}
-            \item allow for existing higher order monomorphic theorem provers to be extended [albeit very crudely] to higher order polymorphic languages
+      \item allow for existing higher order monomorphic theorem provers to be extended to higher order polymorphic languages
       \item improve the performance of existing higher order polymorphic theorem provers on polymorphic problems
@@
 
-         It is therefore aimed at being practically useful, to this end it forgoes completeness, this is the main theoretical weakness of this algorithm [the real main theoretical weakness would be that it is unsound, but i've done my best to make sure it isn't].
-[ref to example in the paper where we explain that?].
+It is therefore aimed at being practically useful, to this end it forgoes completeness, this is the main theoretical weakness of this algorithm.
 
-    \item basic idea of the algorithm
+%basic idea of the algorithm
 
@@
 
-    The new type arguments we have derived are used to create new clauses, without type variables.
-This can lead to the same clause being duplicated as one type variable may be instantiated in a number of different ways.
+The new type arguments that have been derived are used to create new clauses, without type variables.
     A final step is required as the newly generated problem, whilst it does not contain type variables, still makes use of type constructors.
-Those can be transparently erased through the mangling method presented in [relevant paper].
-    \item overview of the paper
+Those can be transparently erased through the mangling method presented in [encoding monomorphic and polymorphic types].
 
-       This paper will first give some general background and set the theoretical framework that will be considered.
+%overview of the paper
+
+This paper will first give some general background and describe the logical framework.
 Then a high-level mathematical description of the algorithm will be presented.
 A more detailed low-level description of the algorithm corresponding to its actual implementation will follow.
-       Finally the implementation and evaluation of the algorithm in zippeposition [the award winning higher order theorem prover] [introduce zipperposition before] will be discussed.
-
-       [issue with intro: focuses too much on theoretical aspect bearly mentions zipp or implementation]
-\end{itemize}
+       Finally, an evaluation of the algorithm's implementation in zippeposition will be discussed.
 
@@
 
-[Start out by mentionning popular type encodings, such as tags and gards, explain the interest of this method]
-Why we didn't implement the types / guards method: in part because it relies on inferences on types that are given in Isabelle but we would have had to do ourselves
+% mention alternatives, (guards, tags) and current implementation in Isabelle
 
-\begin{itemize}
-    \item Language, TF1 (might be TH1 or something else)
-\end{itemize}
+Language: TF1  %TH1?
 
@@
 \begin{itemize}
-    \item higher order, we want lambdas, function symbols and the rest
-    \item polymorphic, we want atomic types, type constructors and type variables
+    \item higher order
+    \item polymorphic: we want atomic types, type constructors and universally quantified type variables
 \end{itemize}
-and our target logic to be the same thing without polymorphism make sure to point out that the mono logic is a subset of the poly logic
+our target logic to be the same thing except we foregoe type constructors and type variables 
+we want our target logic to be a subset of the starting logic
 
-Definitions we need:
+Definitions :
 \begin{itemize}
-   \item type arguments (probably part of logic)
-   \item ground and uninstantiated types 
-   \item vaguely define what matching is
-   \item define what a term, literal and clause are
+   \item type arguments % part of the logic
+   \item ground and uninstantiated types %monomorphic and polymorphic
+   \item matching % being overly precise probably uncecessary
+   \item term, literal, clause  % also part logic
 \end{itemize}
-Need to define ground types: without type variables and uninstantiated types, with type variables
-
-DO NOT use monomorphic to mean ground and polymorphic to mean uninstantiated, everything is polymorphic until we mangle, then everything is monomorphic
-
-Unsure of which terms to use: ground and non-ground, or instantiated and uninstantiated or both
 
 % target page number: 2-4
+% needs to be re-written
 \section{High level algorithm}
@@
 \subsection{Soundness}
-[this isn't an actual proof, more of a justification, mostly a joke in its present form]
 Soundness of the algorithm follows from the fact that the resulting clauses are obtained by directly instantiating universally quantified type variable of the initial clauses.
-Then we rely upon the soundness of mangling [include reference].
 
-\subsection{Multiple type arguments}
-Comment that a function symbol with multiple type arguments will often see them instantiated one at a time per iteration.
-Conclude that it may be beneficial to remove clauses with function symbols with too many uninstantiated type arguments.
-Nuance that if the more there are such symbols, the more likely we are to generate substitutions that instantiate several type arguments at a time.
-Note that in practice setting a type variable limit changes nothing.
-[more tests needed for showing that
-[need to add a couple of tests to show that]
-[maybe this section doesn't belong here, probably better in the low level section]
+Then we rely upon the soundness of mangling [encoding mono and poly ty].
 
@@
 Furthermore, the finite number of iterations of the algorithm consitutes an additional source of incompleteness in practice.
-[find example that would require a list list list int to be solved but only allow such a type to be generated after 3 iterations]
+%find example that would require a list list list int to be solved but only allow such a type to be generated after 3 iterations
+
 % target page number: 2-4
@@
 
-This section presents a more detailed description of the algorithm, it follows closely the implementation in zipperposition.
+This section presents a more detailed description of the algorithm, it follows closely the implementation in Zipperposition.
 
@@
 \end{itemize}
-
-Precise algo description (pseudo code). Explicit loops, and bounds.
-
-[note that this is all very ugly and will be made beautiful once the first draft (of the algorithm) is done]
 
@@
 \end{algorithm}
-In the zipperposition implemntation this iteration procedure is limited by several bounds.
+In the Zipperposition implemntation this iteration procedure is limited by several bounds.
 The number of new type arguments that can be generated for each function symbol is limited by user-defined bounds.
@@
 Additionally, there are bounds for the number of substitutions that are generated.
-
-[So far, after more or less blind experimentation, I've settled on severely limiting new non-ground type arguments because they used to balloon out of control, whilst the growth in ground type arguments is more manageable]
-[It's also worth suggesting the possibility of a more sophisticated implementation that could allow for dynamic bounds]
 
@@
 
-[need to make a handful of tests to check that]
-
 What the previous function does is essentially going through the tree of possible substitutions that instantiate the clause and collecting all possible instantiations.
@@
 
-\break
-
 % target page number: 2-4
@@
 
-Introduce context of zipperposition (do it in intro or background?).
-
-Explain how the final bounds were chosen (i liked the way those numbers looked)
-
-Present evaluation method (which problems were used, options used, server characteristics ect...)
-
-Provide complete list of problems explain which ones we excluded and why: for some it was lack of axiom files, a handful of others were simply too large
+% Provide complete list of problems explain which ones were excluded and why
 
@@
     \begin{itemize}
-        \item e on original portfolio mode
-        \item e with original portfolio mode and monomorphisation
+        \item E on original portfolio mode
+        \item E with original portfolio mode and monomorphisation
         \item adjusted portforlio mode
 
-        \item when e is called
-        \item using zipp only as monomorphiser (passing the result to E, vampire, satallax, leo3)
+        \item when E is called
+        \item using zipp only as monomorphiser (passing the result to E, Vampire, Satallax, Leo-III)
 
@@
 
+Considering the total number of parameters, it would be impractical to exhaustively test all option combinations.
+Consequently, options have been grouped together into sets of 2 or 3 options that will be more thouroughly tested.
+Two options will be part of the same group if it is deemed likely that a change in one of the options will significantly impact how much a change in the second option will affect the overall results.
+
+The first two groups of options are natural as they affect how many new monomorphic and polymorphic type arguments are generated respectively for each function symbol at each iteration.
+These options are closely linked and must be set together if they are to be relevant.
+
+During an iteration, for each function symbol (for each clause in the polymorphic case), the number of newly generated type arguments is determined by this formula:
+\( \min(\text{cap}, \max(\text{floor}, \text{multiplier} \times \text{number of type arguments})) \)
+
+The multiplier allows for a controlled increase of the number of type arguments, relative to the size of the problem whilst the floor prevents some function symbols seeing no new type arguments if the multiplier is too low.
+The cap acts as a final limit in the case the explosion is too fast despite the multiplier bound.
+
+\begin{table}[H]
+\caption{Monomorphic type argument generation}
+\centering\begin{tabular}{@{}llc*{9}{>{\centering\arraybackslash}p{1.5em}}@{}}
+   && \multicolumn{3}{c}{floor 1} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 9}\\
+   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
+   && \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier}\\
+   && 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0\\
+   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
+   \multirow{4}{1em}{cap}
+   & \multicolumn{1}{l|}{25} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & \multicolumn{1}{l|}{50} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   &\multicolumn{1}{l|}{100} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   &\multicolumn{1}{l|}{none}& 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+\end{tabular}
+\end{table}
+
+\bigskip
+\bigskip
+
+If the floor and multiplier are set to 0 and 0.0, no new polymorphic type arguments will be generated, as a result the cap will be irrelevant.
+
+\begin{table}[H]
+\caption{Polymorphic type argument generation}
+\centering\begin{tabular}{@{}llc*{9}{>{\centering\arraybackslash}p{1.5em}}@{}}
+   && \multicolumn{3}{c}{floor 0} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 6}\\
+   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
+   && \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier}\\
+   && 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0\\
+   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
+   \multirow{4}{1em}{cap}
+   & \multicolumn{1}{l|}{10} & \multirow{4}{1em}{0} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & \multicolumn{1}{l|}{20}  && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   &\multicolumn{1}{l|}{50}  && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   &\multicolumn{1}{l|}{none} && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+\end{tabular}
+\end{table}
+
+This table groups two bounds that directly affect the number of clauses generated, a first one that allows at most a number of new clauses equal to ``multiplier" times the original number of clauses.
+And the cap that gives an absolute bound over the total number of clauses generated.
+
+The choice to also include e-timeout in this table is due to the fact that E being successful in solving a monomorphised problem in a certain time limit is highly dependent on the number of clauses it is given.
+
+\begin{table}[H]
+\caption{Clause generation}
+\centering\begin{tabular}{@{}llc*{9}{>{\centering\arraybackslash}p{1.5em}}@{}}
+   && \multicolumn{3}{c}{500 cap} & \multicolumn{3}{c}{2000 cap} & \multicolumn{3}{c}{no cap}\\
+   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
+   && \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier}\\
+   && 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0\\
+   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
+   \multirow{4}{4.2em}{e-timeout}
+   & \multicolumn{1}{l|}{1 s} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & \multicolumn{1}{l|}{5 s} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   &\multicolumn{1}{l|}{10 s} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   &\multicolumn{1}{l|}{none} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\caption{Substitution generation}
+\centering\begin{tabular}{@{}llc*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
+   \toprule
+   & & \multicolumn{4}{c}{mono subst per clause}\\
+   & & 2 & 5 & 7 & 10\\
+   \cmidrule(lr){3-6}
+   \multirow{4}{4.1em}{subst cap}
+   & \multicolumn{1}{l|}{50}   & 0 & 0 & 0 & 0\\
+   & \multicolumn{1}{l|}{100}  & 0 & 0 & 0 & 0\\
+   & \multicolumn{1}{l|}{500}  & 0 & 0 & 0 & 0\\
+   & \multicolumn{1}{l|}{none} & 0 & 0 & 0 & 0\\
+
+   \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\caption{Substitution generation}
+\centering\begin{tabular}{@{}llc*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
+   \toprule
+   & & \multicolumn{4}{c}{e call step}\\
+   & & 0 & 15 & 45 & 90\\
+   \cmidrule(lr){3-6}
+   \multirow{4}{3.3em}{loop nb}
+   & \multicolumn{1}{l|}{2} & 0 & 0 & 0 & 0\\
+   & \multicolumn{1}{l|}{3} & 0 & 0 & 0 & 0\\
+   & \multicolumn{1}{l|}{4} & 0 & 0 & 0 & 0\\
+   & \multicolumn{1}{l|}{5} & 0 & 0 & 0 & 0\\
+
+   \bottomrule
+\end{tabular}
+\end{table}
+
 \subsection{Monomorphisation as preprocessor}
 
-\begin{table}[h]
+\begin{table}[ht]
 \caption{Evaluation of native polymorphism vs.\ monomorphisation}
@@
 
-\begin{table}[h]
+\begin{table}[ht]
 \caption{Evaluation of Zipperposition without E vs. with E}

### 2024-01-29 iterative_monomorphisation.tex
continued working on tables

@@
    \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
-   \multirow{4}{1em}{cap}
+   \multirow{4}{1.5em}{cap}
    & \multicolumn{1}{l|}{25}   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    & \multicolumn{1}{l|}{50}   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   &\multicolumn{1}{l|}{100} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   &\multicolumn{1}{l|}{none}& 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & \multicolumn{1}{l|}{100}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & \multicolumn{1}{l|}{none} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{tabular}
@@
    \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
-   \multirow{4}{1em}{cap}
-   & \multicolumn{1}{l|}{10} & \multirow{4}{1em}{0} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   \multirow{4}{1.5em}{cap}
+   & \multicolumn{1}{l|}{10} 
+   & \multirow{4}{1em}{0}       & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    & \multicolumn{1}{l|}{20}   && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   &\multicolumn{1}{l|}{50}  && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   &\multicolumn{1}{l|}{none} && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & \multicolumn{1}{l|}{50}   && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & \multicolumn{1}{l|}{none} && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{tabular}
@@
 
-The choice to also include e-timeout in this table is due to the fact that E being successful in solving a monomorphised problem in a certain time limit is highly dependent on the number of clauses it is given.
+The choice to also include e-timeout in this table is due to the fact that E being successful in solving a monomorphised problem in a certain time limit has been observed to be highly dependent on the number of clauses it is given.
 
@@
    & \multicolumn{1}{l|}{5 s}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   &\multicolumn{1}{l|}{10 s} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   &\multicolumn{1}{l|}{none} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & \multicolumn{1}{l|}{10 s} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & \multicolumn{1}{l|}{none} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{tabular}
@@
 
+The next table gathers the bounds linked to substitutions generation, the first one limits the number of monomorphising substitutions that are generated per clause during the final phase.
+The second one gives an absolute cap on the overall number of substitutions generated at each clause during the type argument generation.
+
 \begin{table}[H]
 \caption{Substitution generation}
-\centering\begin{tabular}{@{}llc*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
+\centering\begin{tabular}{@{}lc*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
    \toprule
-   & & \multicolumn{4}{c}{mono subst per clause}\\
-   & & 2 & 5 & 7 & 10\\
+   && \multicolumn{4}{c}{mono subst}\\
+   && 2 & 5 & 7 & 10\\
    \cmidrule(lr){3-6}
@@
 
+This last table attempts to determine which are the best bounds related to the depth at which the monomorphisation problem is run as well as how long zipperposition is allowed to run before monomorphisation (and the subsequent call of E) is ran.
+
 \begin{table}[H]
-\caption{Substitution generation}
-\centering\begin{tabular}{@{}llc*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
+% not super happy with caption
+\caption{Problem depth}
+\centering\begin{tabular}{@{}lc*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
    \toprule
-   & & \multicolumn{4}{c}{e call step}\\
-   & & 0 & 15 & 45 & 90\\
+   && \multicolumn{4}{c}{E call step}\\
+   && 0 & 15 & 45 & 90\\
    \cmidrule(lr){3-6}
@@
 
+   \bottomrule
+\end{tabular}
+\end{table}
+
+% We have two options, either use the current tables and make a tiered old/new test structure, or test everyting at once, which would be possible (32 options vs 16 with the tiered approach) however the results would most likely not be very pleasant to present, a highlight would have to be selected.
+
+% final tests to make sure everything plays well together
+\begin{table}[H]
+\caption{Problem size refinement}
+\centering\begin{tabular}{@{}llcccc@{}}
+   \toprule
+   && \multicolumn{4}{c}{substitution generation} \\
+   && \multicolumn{2}{c}{old} & \multicolumn{2}{c}{new} \\
+   \cmidrule(lr){3-4} \cmidrule(lr){5-6}
+   && \multicolumn{2}{c}{problem depth} &\multicolumn{2}{c}{problem depth}\\
+   && old & new & old & new \\
+   \cmidrule(lr){3-4} \cmidrule(lr){5-6}
+   \multirow{2}{7.5em}{clause generation}
+   &\multicolumn{1}{l|}{old} & 0  & 0 & 0 & 0 \\
+   &\multicolumn{1}{l|}{new} & 0  & 0 & 0 & 0\\
+   \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\caption{Type argument generation refinement}
+\centering\begin{tabular}{@{}llcc@{}}
+   \toprule
+   && \multicolumn{2}{c}{mono bounds} \\
+   && old & new\\
+   \cmidrule(r){3-4}
+   \multirow{2}{5.4em}{poly bounds}
+   &\multicolumn{1}{l|}{old} & 0  & 0 \\
+   &\multicolumn{1}{l|}{new} & 0  & 0 \\
+   \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\caption{Final bounds refinement}
+\centering\begin{tabular}{@{}llcc@{}}
+   \toprule
+   && \multicolumn{2}{c}{type argument bounds} \\
+   && old & new\\
+   \cmidrule(r){3-4}
+   \multirow{2}{5.4em}{problem size bounds}
+   &\multicolumn{1}{l|}{old} & 0  & 0 \\
+   &\multicolumn{1}{l|}{new} & 0  & 0 \\
    \bottomrule

### 2024-01-30 iterative_monomorphisation.tex
switched styles and improved tables

@@
-\documentclass{article}
+\documentclass[]{ceurart}
+
+\sloppy
 
@@
 \usepackage{booktabs}
+\usepackage{listings}
 \usepackage{array}
@@
 \usepackage{amsmath}
-\usepackage{authblk}
+
+\lstset{breaklines=true}
 
@@
 
+\copyrightyear{2024}
+\copyrightclause{standard copyright text (cc logo is placeholder)}
+
+\conference{not a conference}
+
 \title{Iterative Monomorphisation}
-\author{Jasmin Blanchette, Sascha B\"ohme, Tanguy Bozec}
-\affil{[add affiliation]}
-\date{[date]}
+
+\author[1]{Jasmin Blanchette}[%
+email=jasmin.blanchette@ifi.lmu.de,
+url=https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette_de.html,
+]
+\address[1]{lmu}
+\author{Sascha B\"ohme}[%
+email=sascha.boehme@posteo.de,
+]
+\author[1,2]{Tanguy Bozec}[%
+email=tanguy.bozec@ens-paris-saclay.fr,
+]
+\address[2]{ens paris-saclay}
+
+\begin{abstract}
+   Monomorphisation methods allows the extension of some higher order theorem provers to polymorphic logics.
+This paper presents an iterative approach to monomorphisation with two main goals: creating an efficient method for extending higher order provers to polymorphic logics and improving the performance of existing provers which natively support polymorphism.
+
+   We first present the iterative monomorphisation algorithm, before discussing its implementation in Zipperposition.
+Finally, the implementation is evaluated as an alternative to native polymorphism and as a means to strenghen Zipperposition.
+\end{abstract}
+
+\begin{keywords}
+   monomorphisation\sep
+   higher order theorem proving \sep
+   automatic theorem proving
+\end{keywords}
+
 \maketitle
-\begin{abstract}
-   [Write abstract]
-\end{abstract}
 
@@
 \section{Introduction}
-%context and motivation (automatic theorem proving, better performance)\\
+
+This paper will first give some general background and describe the logical framework.
+Then a high-level description of the algorithm will be presented.
+A more detailed low-level description of the algorithm corresponding to its implementation will follow.
+
+Finally, the evaluation of the implementation in Zipperposition will be discussed.
+
+\subsection{Context and motivation}
 Automatic theorem provers have many applications, and are particularly useful as backends to proof assistants.
@@
 This paper presents an iterative method, first introduced in [sacha's thesis], that allows for higher order polymorphic problems to be approximated monomorphically.
+This is done through the generation of increasingly large monomorphic type arguments which are used to instantiate the type variables of the original problem.
 This algorithm aims to acheive the following:
@@
       \item allow for existing higher order monomorphic theorem provers to be extended to higher order polymorphic languages
-      \item improve the performance of existing higher order polymorphic theorem provers on polymorphic problems
+      \item improve the performance of existing native polymorphic theorem provers on polymorphic problems
    \end{itemize}
 
-It is therefore aimed at being practically useful, to this end it forgoes completeness, this is the main theoretical weakness of this algorithm.
+It is therefore aimed at being practically useful, to this end it forgoes completeness, this is the main theoretical weakness of this approach.
 
 %basic idea of the algorithm
-
+\subsection{Algorithm overview}
 A given problem is seen as a set of polymorphic clauses.
 Type variables of the problem are quantified at the clause level.
-The type arguments of the problem are collected and the uninstantiated type arguments are matched against those that contain no type variables.
-This yields new type arguments, both uninstantitated and ground, the process is then repeated making use of the newly derived type arguments.
+The type arguments of the problem are collected and the polymorphic type arguments are matched against the monomorphic ones.
+This yields new type arguments, both polymorphic and monomorphic, the process is then repeated making use of the newly derived type arguments.
 
-The new type arguments that have been derived are used to create new clauses, without type variables.
+After a set number of iterations, the newly derived type arguments are used to create new monomorphic clauses, by instantiating their type variables.
+
 A final step is required as the newly generated problem, whilst it does not contain type variables, still makes use of type constructors.
-Those can be transparently erased through the mangling method presented in [encoding monomorphic and polymorphic types].
+Those can be erased through the mangling method presented in [encoding monomorphic and polymorphic types].
 
-%overview of the paper
-
-This paper will first give some general background and describe the logical framework.
-Then a high-level mathematical description of the algorithm will be presented.
-A more detailed low-level description of the algorithm corresponding to its actual implementation will follow.
-       Finally, an evaluation of the algorithm's implementation in zippeposition will be discussed.
+\subsection{Evaluation results}
 
@@
    \item type arguments % part of the logic
-   \item ground and uninstantiated types %monomorphic and polymorphic
+   \item monomorphic and polymorphic types
    \item matching % being overly precise probably uncecessary
@@
 
-\begin{algorithm}[tbh]
+\begin{algorithm}[th]
 \begin{algorithmic}[1]
@@
 
-\begin{table}[H]
+\begin{table}[th]
 \caption{Monomorphic type argument generation}
-\centering\begin{tabular}{@{}llc*{9}{>{\centering\arraybackslash}p{1.5em}}@{}}
-   && \multicolumn{3}{c}{floor 1} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 9}\\
-   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
-   && \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier}\\
-   && 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0\\
-   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
-   \multirow{4}{1.5em}{cap}
-   & \multicolumn{1}{l|}{25}   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   & \multicolumn{1}{l|}{50}   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   & \multicolumn{1}{l|}{100}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   & \multicolumn{1}{l|}{none} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
+   \toprule
+   & \multicolumn{3}{c}{floor 1} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 9}\\
+   \cmidrule(l){2-10}
+   \multirow{2}{2em}{cap} & \multicolumn{9}{c}{multiplier} \\
+      & 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0\\
+    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
+    25   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    50   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    100  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    \bottomrule
 \end{tabular}
@@
 
-\begin{table}[H]
+\begin{table}[th]
 \caption{Polymorphic type argument generation}
-\centering\begin{tabular}{@{}llc*{9}{>{\centering\arraybackslash}p{1.5em}}@{}}
-   && \multicolumn{3}{c}{floor 0} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 6}\\
-   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
-   && \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier}\\
-   && 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0\\
-   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
-   \multirow{4}{1.5em}{cap}
-   & \multicolumn{1}{l|}{10} 
-   & \multirow{4}{1em}{0}       & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   & \multicolumn{1}{l|}{20}   && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   & \multicolumn{1}{l|}{50}   && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   & \multicolumn{1}{l|}{none} && 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
+   \toprule
+   & \multicolumn{3}{c}{floor 0} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 6}\\
+   \cmidrule(l){2-10}
+   \multirow{2}{2em}{cap} & \multicolumn{9}{c}{multiplier} \\
+      & 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0 \\
+    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
+    10   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    20   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    50   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    \bottomrule
 \end{tabular}
@@
 
-\begin{table}[H]
+\begin{table}[th]
 \caption{Clause generation}
-\centering\begin{tabular}{@{}llc*{9}{>{\centering\arraybackslash}p{1.5em}}@{}}
-   && \multicolumn{3}{c}{500 cap} & \multicolumn{3}{c}{2000 cap} & \multicolumn{3}{c}{no cap}\\
-   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
-   && \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier} & \multicolumn{3}{c}{multiplier}\\
-   && 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0\\
-   \cmidrule(lr){3-5} \cmidrule(lr){6-8} \cmidrule(lr){9-11}
-   \multirow{4}{4.2em}{e-timeout}
-   & \multicolumn{1}{l|}{1 s}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   & \multicolumn{1}{l|}{5 s}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   & \multicolumn{1}{l|}{10 s} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-   & \multicolumn{1}{l|}{none} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
+   \toprule
+   & \multicolumn{3}{c}{cap 500} & \multicolumn{3}{c}{cap 2000} & \multicolumn{3}{c}{no cap}\\
+   \cmidrule(l){2-10}
+   \multirow{2}{4em}{E timeout} & \multicolumn{9}{c}{multiplier} \\
+      & 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0 \\
+    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
+    2s   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    5s   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    10s  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    \bottomrule
 \end{tabular}
 \end{table}
-
 The next table gathers the bounds linked to substitutions generation, the first one limits the number of monomorphising substitutions that are generated per clause during the final phase.
@@
 
-\begin{table}[H]
+\begin{table}[th]
 \caption{Substitution generation}
-\centering\begin{tabular}{@{}lc*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
+\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
    \toprule
-   && \multicolumn{4}{c}{mono subst}\\
-   && 2 & 5 & 7 & 10\\
-   \cmidrule(lr){3-6}
-   \multirow{4}{4.1em}{subst cap}
-   & \multicolumn{1}{l|}{50}   & 0 & 0 & 0 & 0\\
-   & \multicolumn{1}{l|}{100}  & 0 & 0 & 0 & 0\\
-   & \multicolumn{1}{l|}{500}  & 0 & 0 & 0 & 0\\
-   & \multicolumn{1}{l|}{none} & 0 & 0 & 0 & 0\\
+   \multirow{2}{4em}{subst cap} & \multicolumn{4}{c}{mono subst}\\
+   & 2 & 5 & 7 & 10\\
+   \midrule
+   50   & 0 & 0 & 0 & 0\\
+   100  & 0 & 0 & 0 & 0\\
+   500  & 0 & 0 & 0 & 0\\
+   none & 0 & 0 & 0 & 0\\
 
@@
 
-\begin{table}[H]
+\begin{table}[th]
 % not super happy with caption
 \caption{Problem depth}
-\centering\begin{tabular}{@{}lc*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
+\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
    \toprule
-   && \multicolumn{4}{c}{E call step}\\
-   && 0 & 15 & 45 & 90\\
-   \cmidrule(lr){3-6}
-   \multirow{4}{3.3em}{loop nb}
-   & \multicolumn{1}{l|}{2} & 0 & 0 & 0 & 0\\
-   & \multicolumn{1}{l|}{3} & 0 & 0 & 0 & 0\\
-   & \multicolumn{1}{l|}{4} & 0 & 0 & 0 & 0\\
-   & \multicolumn{1}{l|}{5} & 0 & 0 & 0 & 0\\
+   \multirow{2}{3em}{loop nb} & \multicolumn{4}{c}{E call step}\\
+   & 0 & 15 & 45 & 90\\
+   \midrule
+   2 & 0 & 0 & 0 & 0\\
+   3 & 0 & 0 & 0 & 0\\
+   4 & 0 & 0 & 0 & 0\\
+   5 & 0 & 0 & 0 & 0\\
 
@@
 % final tests to make sure everything plays well together
-\begin{table}[H]
+\begin{table}[ht]
 \caption{Problem size refinement}
-\centering\begin{tabular}{@{}llcccc@{}}
+\centering\begin{tabular}{@{}lcccc@{}}
    \toprule
-   && \multicolumn{4}{c}{substitution generation} \\
-   && \multicolumn{2}{c}{old} & \multicolumn{2}{c}{new} \\
-   \cmidrule(lr){3-4} \cmidrule(lr){5-6}
-   && \multicolumn{2}{c}{problem depth} &\multicolumn{2}{c}{problem depth}\\
-   && old & new & old & new \\
-   \cmidrule(lr){3-4} \cmidrule(lr){5-6}
-   \multirow{2}{7.5em}{clause generation}
-   &\multicolumn{1}{l|}{old} & 0  & 0 & 0 & 0 \\
-   &\multicolumn{1}{l|}{new} & 0  & 0 & 0 & 0\\
+   & \multicolumn{4}{c}{substitution generation} \\
+   & \multicolumn{2}{c}{old} & \multicolumn{2}{c}{new} \\
+   \cmidrule(lr){2-5} 
+   \multirow{2}{4.5em}{clause gen} & \multicolumn{4}{c}{problem depth} \\
+   & old & new & old & new \\
+   \cmidrule(lr){2-3} \cmidrule(lr){4-5}
+   old & 0  & 0 & 0 & 0 \\
+   new & 0  & 0 & 0 & 0\\
    \bottomrule
@@
 
-\begin{table}[H]
+\begin{table}[ht]
 \caption{Type argument generation refinement}
@@
    \multirow{2}{5.4em}{poly bounds}
-   &\multicolumn{1}{l|}{old} & 0  & 0 \\
-   &\multicolumn{1}{l|}{new} & 0  & 0 \\
+   &\multicolumn{1}{l}{old} & 0  & 0 \\
+   &\multicolumn{1}{l}{new} & 0  & 0 \\
    \bottomrule
@@
 
-\begin{table}[H]
+\begin{table}[ht]
 \caption{Final bounds refinement}
@@
    \multirow{2}{5.4em}{problem size bounds}
-   &\multicolumn{1}{l|}{old} & 0  & 0 \\
-   &\multicolumn{1}{l|}{new} & 0  & 0 \\
+   &\multicolumn{1}{l}{old} & 0  & 0 \\
+   &\multicolumn{1}{l}{new} & 0  & 0 \\
    \bottomrule

### 2024-01-31 iterative_monomorphisation.tex
changes

@@
 \copyrightyear{2024}
-\copyrightclause{standard copyright text (cc logo is placeholder)}
+\copyrightclause{Copyright for this paper by its authors. Use permitted under Creative Commons License Attribution 4.0 International (CC BY 4.0)}
 
-\conference{not a conference}
+\conference{Submitted draft} % TODO update
 
@@
 
-\author[1]{Jasmin Blanchette}[%
-email=jasmin.blanchette@ifi.lmu.de,
-url=https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette_de.html,
-]
-\address[1]{lmu}
-\author{Sascha B\"ohme}[%
-email=sascha.boehme@posteo.de,
-]
 \author[1,2]{Tanguy Bozec}[%
@@
 ]
-\address[2]{ens paris-saclay}
+\author[2]{Jasmin Blanchette}[%
+email=jasmin.blanchette@ifi.lmu.de,
+%url=https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette_de.html,
+]
+\address[1]{Departement d'informatique, École normale supérieure Paris-Saclay, France} % TODO dpt info
+\address[2]{Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}
 
@@
 \begin{keywords}
-   monomorphisation\sep
-   higher order theorem proving \sep
+   polymorphism\sep
+   higher order logic\sep
    automatic theorem proving

### 2024-01-31 iterative_monomorphisation.tex
applied some of the changes discussed with Jasmin

@@
 ]
-\address[1]{Departement d'informatique, École normale supérieure Paris-Saclay, France} % TODO dpt info
+\address[1]{ENS Paris-Saclay, Université Paris-Saclay, France}
 \address[2]{Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}
@@
 
-This paper will first give some general background and describe the logical framework.
-Then a high-level description of the algorithm will be presented.
-A more detailed low-level description of the algorithm corresponding to its implementation will follow.
-
-Finally, the evaluation of the implementation in Zipperposition will be discussed.
-
-\subsection{Context and motivation}
 Automatic theorem provers have many applications, and are particularly useful as backends to proof assistants.
@@
 %basic idea of the algorithm
-\subsection{Algorithm overview}
 A given problem is seen as a set of polymorphic clauses.
@@
 Those can be erased through the mangling method presented in [encoding monomorphic and polymorphic types].
-
-\subsection{Evaluation results}
 
@@
 \begin{table}[th]
-\caption{Monomorphic type argument generation}
+\caption{Evaluation of bounds for monomorphic type argument generation}
 \centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
    \toprule
-   & \multicolumn{3}{c}{floor 1} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 9}\\
+   & &&& \multicolumn{3}{c}{floor} \\
+   & \multicolumn{3}{c}{1} & \multicolumn{3}{c}{3} & \multicolumn{3}{c}{9}\\
    \cmidrule(l){2-10}
-   \multirow{2}{2em}{cap} & \multicolumn{9}{c}{multiplier} \\
-      & 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0 & 0.0 & 1.0 & 2.0\\
+   & &&& \multicolumn{3}{c}{multiplier} \\
+   \multirow{1}{2em}{cap} & 0 & 1 & 2 & 0 & 1 & 2 & 0 & 1 & 2\\
     \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
@@
     100  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    \(\infty\) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
     \bottomrule
@@
 
-If the floor and multiplier are set to 0 and 0.0, no new polymorphic type arguments will be generated, as a result the cap will be irrelevant.
+If the floor and multiplier are set to 0 and 0, no new polymorphic type arguments will be generated, as a result the cap will be irrelevant.
 
 \begin{table}[th]
-\caption{Polymorphic type argument generation}
+\caption{Evaluation of bounds for polymorphic type argument generation}
 \centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
    \toprule
-   & \multicolumn{3}{c}{floor 0} & \multicolumn{3}{c}{floor 3} & \multicolumn{3}{c}{floor 6}\\
+   & &&& \multicolumn{3}{c}{floor} \\
+   & \multicolumn{3}{c}{0} & \multicolumn{3}{c}{3} & \multicolumn{3}{c}{6}\\
    \cmidrule(l){2-10}
-   \multirow{2}{2em}{cap} & \multicolumn{9}{c}{multiplier} \\
-      & 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0 & 0.0 & 0.5 & 1.0 \\
+   & &&& \multicolumn{3}{c}{multiplier} \\
+    \multirow{1}{2em}{cap} & 0 & 0.5 & 1 & 0 & 0.5 & 1 & 0 & 0.5 & 1 \\
     \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
@@
     20   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    50   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    40   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    \(\infty\) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[th]
+\caption{Evaluation of bounds for substitution generation}
+\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
+   \toprule
+   & \multicolumn{4}{c}{mono subst}\\
+   \multirow{1}{4em}{subst cap} & 2 & 5 & 7 & 10\\
+   \midrule
+   50   & 0 & 0 & 0 & 0\\
+   100  & 0 & 0 & 0 & 0\\
+   500  & 0 & 0 & 0 & 0\\
+   \(\infty\) & 0 & 0 & 0 & 0\\
+
    \bottomrule
@@
 \begin{table}[th]
-\caption{Clause generation}
+\caption{Evaluation of bounds directly related to the size of the output problem}
 \centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
    \toprule
-   & \multicolumn{3}{c}{cap 500} & \multicolumn{3}{c}{cap 2000} & \multicolumn{3}{c}{no cap}\\
+   & &&& \multicolumn{3}{c}{cap} \\
+   & \multicolumn{3}{c}{500} & \multicolumn{3}{c}{2000} & \multicolumn{3}{c}{\(\infty\)}\\
    \cmidrule(l){2-10}
-   \multirow{2}{4em}{E timeout} & \multicolumn{9}{c}{multiplier} \\
-      & 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0 & 1.0 & 2.0 & 3.0 \\
+   & &&& \multicolumn{3}{c}{multiplier} \\
+    \multirow{1}{5.4em}{E timeout (s)} & 1 & 2 & 3 & 1 & 2 & 3 & 1 & 2 & 3 \\
     \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
-    2s   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    5s   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    10s  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    none & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    2   &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    5   &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    10  &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    \(\infty\)& 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
     \bottomrule
@@
 
-\begin{table}[th]
-\caption{Substitution generation}
-\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
-   \toprule
-   \multirow{2}{4em}{subst cap} & \multicolumn{4}{c}{mono subst}\\
-   & 2 & 5 & 7 & 10\\
-   \midrule
-   50   & 0 & 0 & 0 & 0\\
-   100  & 0 & 0 & 0 & 0\\
-   500  & 0 & 0 & 0 & 0\\
-   none & 0 & 0 & 0 & 0\\
-
-   \bottomrule
-\end{tabular}
-\end{table}
-
 This last table attempts to determine which are the best bounds related to the depth at which the monomorphisation problem is run as well as how long zipperposition is allowed to run before monomorphisation (and the subsequent call of E) is ran.
@@
 % not super happy with caption
-\caption{Problem depth}
+\caption{Evaluation of parameters indirectly related to the size of the output problem}
 \centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
    \toprule
-   \multirow{2}{3em}{loop nb} & \multicolumn{4}{c}{E call step}\\
-   & 0 & 15 & 45 & 90\\
+   & \multicolumn{4}{c}{loop nb} \\
+   \multirow{1}{4.5em}{E call step} & 2 & 3 & 4 & 5\\
    \midrule
-   2 & 0 & 0 & 0 & 0\\
-   3 & 0 & 0 & 0 & 0\\
-   4 & 0 & 0 & 0 & 0\\
-   5 & 0 & 0 & 0 & 0\\
+   0 & 0 & 0 & 0 & 0\\
+   15 & 0 & 0 & 0 & 0\\
+   45 & 0 & 0 & 0 & 0\\
+   90 & 0 & 0 & 0 & 0\\
 
-   \bottomrule
-\end{tabular}
-\end{table}
-
-% We have two options, either use the current tables and make a tiered old/new test structure, or test everyting at once, which would be possible (32 options vs 16 with the tiered approach) however the results would most likely not be very pleasant to present, a highlight would have to be selected.
-
-% final tests to make sure everything plays well together
-\begin{table}[ht]
-\caption{Problem size refinement}
-\centering\begin{tabular}{@{}lcccc@{}}
-   \toprule
-   & \multicolumn{4}{c}{substitution generation} \\
-   & \multicolumn{2}{c}{old} & \multicolumn{2}{c}{new} \\
-   \cmidrule(lr){2-5} 
-   \multirow{2}{4.5em}{clause gen} & \multicolumn{4}{c}{problem depth} \\
-   & old & new & old & new \\
-   \cmidrule(lr){2-3} \cmidrule(lr){4-5}
-   old & 0  & 0 & 0 & 0 \\
-   new & 0  & 0 & 0 & 0\\
-   \bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}[ht]
-\caption{Type argument generation refinement}
-\centering\begin{tabular}{@{}llcc@{}}
-   \toprule
-   && \multicolumn{2}{c}{mono bounds} \\
-   && old & new\\
-   \cmidrule(r){3-4}
-   \multirow{2}{5.4em}{poly bounds}
-   &\multicolumn{1}{l}{old} & 0  & 0 \\
-   &\multicolumn{1}{l}{new} & 0  & 0 \\
-   \bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}[ht]
-\caption{Final bounds refinement}
-\centering\begin{tabular}{@{}llcc@{}}
-   \toprule
-   && \multicolumn{2}{c}{type argument bounds} \\
-   && old & new\\
-   \cmidrule(r){3-4}
-   \multirow{2}{5.4em}{problem size bounds}
-   &\multicolumn{1}{l}{old} & 0  & 0 \\
-   &\multicolumn{1}{l}{new} & 0  & 0 \\
    \bottomrule

### 2024-02-09 iterative_monomorphisation.tex
added references
re-wrote a fair bit of the introduction

@@
 
-Automatic theorem provers have many applications, and are particularly useful as backends to proof assistants.
-Speed is one of the main limitations of the use of theorem provers in this setting.
-Indeed, a prover can often be expected to run for no more than a few seconds for a given instance of a problem.
+\cite{mono-trans}
+\cite{expr-poly-types}
+\cite{sb-phd}
 
 %what we are trying to do (monomorphisation)\\
-This paper presents an iterative method, first introduced in [sacha's thesis], that allows for higher order polymorphic problems to be approximated monomorphically.
-This is done through the generation of increasingly large monomorphic type arguments which are used to instantiate the type variables of the original problem.
-This algorithm aims to acheive the following:
+This paper presents an iterative method, first introduced in \cite{sb-phd} (section 2.2.1), that allows for higher order polymorphic problems to be approximated monomorphically.
+This algorithm aims to achieve the following:
    \begin{itemize}
-      \item allow for existing higher order monomorphic theorem provers to be extended to higher order polymorphic languages
-      \item improve the performance of existing native polymorphic theorem provers on polymorphic problems
+      \item extend higher order theorem provers to polymorphic languages
+      \item improve the performance of existing native polymorphic theorem provers
    \end{itemize}
 
-It is therefore aimed at being practically useful, to this end it forgoes completeness, this is the main theoretical weakness of this approach.
+This algorithm is implemented in Zipperpostion, a higher order polymorphic automatic theorem prover written in OCaml.
+This implementation allows for the E prover, which does not support polymorphism, to be used as a backend to Zipperposition.
+The viability of this approach is then evaluated, both as an improvement to Zipperposition and as a means to increase the performance of other provers.
 
+\medskip
 %basic idea of the algorithm
-A given problem is seen as a set of polymorphic clauses.
-Type variables of the problem are quantified at the clause level.
-The type arguments of the problem are collected and the polymorphic type arguments are matched against the monomorphic ones.
-This yields new type arguments, both polymorphic and monomorphic, the process is then repeated making use of the newly derived type arguments.
+The algorithm works as follows:
+we assume problems to be sets of clauses.
+All type arguments of the problem are collected and the polymorphic type arguments are matched against the monomorphic ones.
+This yields new type arguments, both polymorphic and monomorphic.
+The process is then repeated making use of the newly derived type arguments.
 
-After a set number of iterations, the newly derived type arguments are used to create new monomorphic clauses, by instantiating their type variables.
+These iterations are required to generate all
 
-A final step is required as the newly generated problem, whilst it does not contain type variables, still makes use of type constructors.
-Those can be erased through the mangling method presented in [encoding monomorphic and polymorphic types].
+After a set number of iterations, the new type arguments are used to instantiate type variables, generating new monomorphic clauses.
 
+Finally, we ``mangle'' the types in the output clauses using the method presented in \cite{mono-trans}, in order to remove type constructors from the output problem.
+
+This approach sacrifices completeness for the sake of performance.
+The loss of completeness occurs at the conceptual level due to the finite number of iterations.
+It also occurs on the practical level, due to the explosive nature of the procedure, various bounds limit the number of generated clauses.
 % target page number: 0.5-1.5
@@
     \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
-    10   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    20   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    40   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    10 &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    20 &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    40 &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
     \(\infty\) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
@@
     \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
-    2   &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    5   &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    10  &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    2   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    5   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    10  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
     \(\infty\)& 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
@@
 
-\begin{itemize}
-    \item part of Sascha Boehme's thesis (Proving Theorems of Higher-Order Logic with SMT Solvers)
-    \item How to Encode Polymorphic Types Safely and Efﬁciently, section 4
-
-    \item Expressing polymorphic types in a many-sorted language
-    \item Encoding Monomorphic and Polymorphic Types
-\end{itemize}
-
 % target page number: 0.5-1
@@
 % Future avenues
+\bibliography{citations}
 

### 2024-02-14 iterative_monomorphisation.tex
continued working on introduction

@@
 
-\cite{mono-trans}
-\cite{expr-poly-types}
-\cite{sb-phd}
+%\cite{mono-trans}
+%\cite{expr-poly-types}
+%\cite{sb-phd}
 
@@
 
+\smallskip
+
+[sumarry of results]
+
 \medskip
@@
 
-These iterations are required to generate all
+\smallskip
 
-After a set number of iterations, the new type arguments are used to instantiate type variables, generating new monomorphic clauses.
+These iterations are required in order to account for inductively defined types.
+Consider the type constructor \verb|list| which takes a type argument, if we obtain \verb|list(alpha)| where \verb|alpha| is a type variable, it may be possible to generate the types \verb|list(int)|, \verb|list(list(int))|, ...
+However because new types are obtained through matching, \verb|list(list(int))| can only be obtained once \verb|list(int)| has already been added to the set of monomorphic types.
+This example also makes it clear that the set of types that can be generated is infinite, this explains why the number of iterations is bounded.
+
+\smallskip
+
+After a set number of iterations, the new type arguments are used to instantiate the type variables in the problem's clauses, generating new monomorphic clauses.
 
@@
 This approach sacrifices completeness for the sake of performance.
-The loss of completeness occurs at the conceptual level due to the finite number of iterations.
-It also occurs on the practical level, due to the explosive nature of the procedure, various bounds limit the number of generated clauses.
+The loss of completeness occurs at the level of the algorithm; a finite number of iterations and at the implementation level; the explosive nature of the procedures requires various bounds which limit the number of type arguments and clauses that are generated.
+
 % target page number: 0.5-1.5

### 2024-02-14 iterative_monomorphisation.tex
added everything?
abstrac

@@
 
-\sloppy
+%\sloppy
 
@@
 \begin{abstract}
-   Monomorphisation methods allows the extension of some higher order theorem provers to polymorphic logics.
-This paper presents an iterative approach to monomorphisation with two main goals: creating an efficient method for extending higher order provers to polymorphic logics and improving the performance of existing provers which natively support polymorphism.
-
-   We first present the iterative monomorphisation algorithm, before discussing its implementation in Zipperposition.
-Finally, the implementation is evaluated as an alternative to native polymorphism and as a means to strenghen Zipperposition.
+Monomorphisation makes it possible to extend monormophic provers to support
+polymorphic logics. We propose an iterative approach, which is necessarily
+incomplete but which works well in practice. It is implemented in the
+Zipperposition prover, where it can be used to translate away polymorphism
+before invoking a monomorphic prover. Our empirical evaluation reveals that
+iterative monorphisation is a viable alternative to native polymorphism.
 \end{abstract}
@@
 This algorithm is implemented in Zipperpostion, a higher order polymorphic automatic theorem prover written in OCaml.
-This implementation allows for the E prover, which does not support polymorphism, to be used as a backend to Zipperposition.
+As a result, the E prover, which does not support polymorphism, can be used as a backend to Zipperposition.
 The viability of this approach is then evaluated, both as an improvement to Zipperposition and as a means to increase the performance of other provers.

### 2024-02-14 iterative_monomorphisation.tex
intro

@@
 \lstset{breaklines=true}
+
+\newcommand\ty[1]{\textsf{#1}}
+\newcommand\sym[1]{\textsf{#1}}
+\newcommand\var[1]{\mathit{#1}}
 
@@
 \begin{abstract}
-Monomorphisation makes it possible to extend monormophic provers to support
-polymorphic logics. We propose an iterative approach, which is necessarily
-incomplete but which works well in practice. It is implemented in the
-Zipperposition prover, where it can be used to translate away polymorphism
-before invoking a monomorphic prover. Our empirical evaluation reveals that
-iterative monorphisation is a viable alternative to native polymorphism.
+Monomorphisation makes it possible to extend monormophic provers to support polymorphic logics.
+We propose an iterative approach, which is necessarily incomplete but which works well in practice.
+It is implemented in the Zipperposition prover, where it can be used to translate away polymorphism before invoking a monomorphic prover.
+Our empirical evaluation reveals that iterative monorphisation is a viable alternative to native polymorphism.
 \end{abstract}
@@
 \begin{keywords}
-   polymorphism\sep
-   higher order logic\sep
+   Polymorphism\sep
+   monomorphism\sep
    automatic theorem proving
@@
 
-%\cite{mono-trans}
-%\cite{expr-poly-types}
-%\cite{sb-phd}
+{
+\obeylines
 
-%what we are trying to do (monomorphisation)\\
-This paper presents an iterative method, first introduced in \cite{sb-phd} (section 2.2.1), that allows for higher order polymorphic problems to be approximated monomorphically.
-This algorithm aims to achieve the following:
-   \begin{itemize}
-      \item extend higher order theorem provers to polymorphic languages
-      \item improve the performance of existing native polymorphic theorem provers
-   \end{itemize}
+  * apps are rank-1 polymorphic, but provers are mono
+  * there's a gap
+  * one solution is to extend the provers to support polymorphism;
+    this was done e.g. for Vampire \cite{xxx}
+  * but this is a lot of work, and needs to be done for every prover
 
-This algorithm is implemented in Zipperpostion, a higher order polymorphic automatic theorem prover written in OCaml.
-As a result, the E prover, which does not support polymorphism, can be used as a backend to Zipperposition.
-The viability of this approach is then evaluated, both as an improvement to Zipperposition and as a means to increase the performance of other provers.
+\bigskip
 
-\smallskip
+  * the alternative is to translate polymorphic problems to monomorphic problems
+    * two main approaches:
+      * complete encoding of polymorphism, using guards or tags,
+        as in Blanchette et al.~\cite{xxx} and Bobot et al.~\cite{xxx}
+      * incomplete encoding of polymorphism based on iterative monomorphisation
+        such as introduced by Böhme \cite[Section 2.2.1]{sb-phd}, whereby type
+        variables are heuristically instantiated by concrete types
 
-[sumarry of results]
+\bigskip
 
-\medskip
-%basic idea of the algorithm
-The algorithm works as follows:
-we assume problems to be sets of clauses.
-All type arguments of the problem are collected and the polymorphic type arguments are matched against the monomorphic ones.
+  * By a type version of Herbrand's theorem, if a set of polymorphic
+    formulas is unsatisfiable, then there exists a finite set of monomorphic
+    instances of the formulas that is also unsatisfiable. However, this
+    finite set cannot be computed \cite{xxx}; hence any approach based on
+    instantiation of type variables is incomplete.
+
+\bigskip
+
+  * Böhme's iterative approach is implemented as part of the SMT integration
+    \cite{xxx} in Isabelle/HOL. This implementation is also used by
+    Sledgehammer \cite{xxx} to interface automatic theorem provers based on
+    superposition.
+
+\bigskip
+
+  * The main drawback of Böhme's approach is that it is only briefly documented,
+    as a single paragraph in his PhD thesis \cite[Section 2.2.1]{sb-phd}
+  * In this paper, we present an algorithm inspired by his description and
+    implementation (Section~\ref{sec:high-level-algorithm}).
+    * we aimed at mathematical precision, so that people who want to reimplement
+      the algorithm can follow this blueprint
+    * we also show how to bound the algorithm to avoid explosions in practice
+      (Section~\ref{sec:low-level-algorithm}).
+      ``the explosive nature of the procedures requires various bounds which
+      limit the number of type arguments and clauses that are generated.''
+
+}
+
+The algorithm works as follows.
+We assume problems to be sets of formulas.
+All type arguments of the problem are collected, and the polymorphic type arguments are matched against the monomorphic ones.
 This yields new type arguments, both polymorphic and monomorphic.
-The process is then repeated making use of the newly derived type arguments.
+The process is then repeated using the newly derived type arguments.
 
-\smallskip
+These iterations are required to account for $n$-ary type constructors.
+Consider the unary type constructor \ty{list}.
+If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, it may be possible to generate the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc.
+However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once $\ty{list}(\ty{int})$ has already been added to the set of monomorphic types.
+This example also makes it clear that the set of types that can be generated is infinite.
 
-These iterations are required in order to account for inductively defined types.
-Consider the type constructor \verb|list| which takes a type argument, if we obtain \verb|list(alpha)| where \verb|alpha| is a type variable, it may be possible to generate the types \verb|list(int)|, \verb|list(list(int))|, ...
-However because new types are obtained through matching, \verb|list(list(int))| can only be obtained once \verb|list(int)| has already been added to the set of monomorphic types.
-This example also makes it clear that the set of types that can be generated is infinite, this explains why the number of iterations is bounded.
+To keep the number of generated formulas finite, we limit the number of iterations.
+After the iterations are completed, the new type arguments are used to instantiate the type variables in the problem's formulas, generating new monomorphic clauses.
+Finally, because monomorphic provers generally do not support $n$-ary type constructors, types must be mangled.
+For example, the compound type $\ty{list}(\ty{int})$ could be mangled to $\ty{list\_int}$.
 
-\smallskip
+We implemented this algorithm in Zipperpostion, a higher order automatic theorem prover written in OCaml.
+Although Zipperposition is polymorphic, it uses the monomorphic prover E as a backend.
+Our algorithm makes it possible, for the first time, to use E with polymorphic problems.
+Moreover, our implementation of the algorithm in Zipperposition can be used as a preprocessor to interface with other stand-alone provers.
 
-After a set number of iterations, the new type arguments are used to instantiate the type variables in the problem's clauses, generating new monomorphic clauses.
+Our empirical evaluation on TPTP problems \cite{xxx} attempts to answer three questions (Section~\ref{sec:evaluation}):
+%
+\begin{enumerate}
+\item Is the new Zipperposition with the E backend more successful on polymorphic problems than Zipperposition without backend?
 
-Finally, we ``mangle'' the types in the output clauses using the method presented in \cite{mono-trans}, in order to remove type constructors from the output problem.
+\item How competitive are monomorphic provers on monomorphised polymorphic problems?
 
-This approach sacrifices completeness for the sake of performance.
-The loss of completeness occurs at the level of the algorithm; a finite number of iterations and at the implementation level; the explosive nature of the procedures requires various bounds which limit the number of type arguments and clauses that are generated.
+\item Is iterative monomophisation more effective than the native polymorphism implemented in polymorphic provers?
+\end{enumerate}
+
+We find that:
+%
+\begin{enumerate}
+\item Zipperposition benefits substantially from the E backend.
+
+\item E with monomorphisation comes close second to the polymorphic prover Vampire.
+
+\item For Leo-III and Vampire, we find that monomorphisation is indeed more effective than native polymorphism.
+\end{enumerate}
 
@@
 \section{Background}
+\label{sec:background}
 
@@
 \section{High level algorithm}
+\label{sec:high-level-algorithm}
 
@@
 \section{Low level algorithm}
+\label{sec:low-level-algorithm}
 
@@
 \section{Evaluation}
+\label{sec:evaluation}
 
@@
 \section{Related work}
+\label{sec:related-work}
 
@@
 \section{Conclusion}
+\label{sec:conclusion}
 

### 2024-02-14 iterative_monomorphisation.tex
more

@@
 
-  * By a type version of Herbrand's theorem, if a set of polymorphic
+  * By a type version of compactness, if a set of polymorphic
     formulas is unsatisfiable, then there exists a finite set of monomorphic
@@
 
-We implemented this algorithm in Zipperpostion, a higher order automatic theorem prover written in OCaml.
-Although Zipperposition is polymorphic, it uses the monomorphic prover E as a backend.
+We implemented this algorithm in Zipperpostion \cite{xxx}, a higher order automatic theorem prover written in OCaml.
+Although Zipperposition is polymorphic, it uses the monomorphic prover E \cite{xxx} as a backend.
 Our algorithm makes it possible, for the first time, to use E with polymorphic problems.
@@
 
-We find that:
+Our findings are as follows:
 %
@@
 
-\item For Leo-III and Vampire, we find that monomorphisation is indeed more effective than native polymorphism.
+\item For Leo-III \cite{xxx} and Vampire \cite{xxx}, we find that monomorphisation is indeed more effective than native polymorphism.
 \end{enumerate}

### 2024-02-15 iterative_monomorphisation.tex
mostly finished the introduction

@@
 \begin{abstract}
-Monomorphisation makes it possible to extend monormophic provers to support polymorphic logics.
+Monomorphisation makes it possible to extend monomorphic provers to support polymorphic logics.
 We propose an iterative approach, which is necessarily incomplete but which works well in practice.
 It is implemented in the Zipperposition prover, where it can be used to translate away polymorphism before invoking a monomorphic prover.
-Our empirical evaluation reveals that iterative monorphisation is a viable alternative to native polymorphism.
+Our empirical evaluation reveals that iterative monomorphisation is a viable alternative to native polymorphism.
 \end{abstract}
@@
 
-{
-\obeylines
+One of the main applications of automatic theorem provers is making the use of proof assistants easier.
+Many proof assistants support rank-1 polymorphism, however, many automatic theorem provers only operate on monomorphic logics.
+One approach to close this gap is extending existing provers, allowing them to natively support polymorphism, as was the case for Vampire \cite{vamp}.
+The main drawbacks of this solution are the amount of work required to extend these provers and the fact that it needs to be done from scratch for every monomorphic prover.
 
-  * apps are rank-1 polymorphic, but provers are mono
-  * there's a gap
-  * one solution is to extend the provers to support polymorphism;
-    this was done e.g. for Vampire \cite{xxx}
-  * but this is a lot of work, and needs to be done for every prover
+  % apps are rank-1 polymorphic, but provers are mono
+  % there's a gap
+  % one solution is to extend the provers to support polymorphism;
+  % this was done e.g. for Vampire \cite{xxx}
+  % but this is a lot of work, and needs to be done for every prover
+
+The alternative would be to translate polymorphic problems into monomorphic problems.
+There are two main approaches to this, the first is a complete encoding of polymorphism using guards and tags \cite{mono-trans} into a monomorphic logic.
+The second solution is an incomplete encoding of polymorphism based on iterative monomorphisation, as introduced by B\"ome \cite[Section 2.2.1]{sb-phd}.
+This method relies on heuristically instantiating type variables with concrete types.
+
+ % * the alternative is to translate polymorphic problems to monomorphic problems
+ %   * two main approaches:
+ %     * complete encoding of polymorphism, using guards or tags,
+ %       as in Blanchette et al.~\cite{mono-trans} and Bobot et al.~\cite{expr-poly-types}
+ %     * incomplete encoding of polymorphism based on iterative monomorphisation
+ %       such as introduced by Böhme \cite[Section 2.2.1]{sb-phd}, whereby type
+ %       variables are heuristically instantiated by concrete types
+
+By compactness, we have that in first order logic, given a polymorphic formula \(F\), there exist an equisatisfiable finite set of monomorphic instances of \(F\).
+However such a set cannot be computed \cite[Section 2, Theorem 1]{expr-poly-types}, consequently any monomorphisation method based on instantiation of type variables is incomplete.
+
+  %* By a type version of compactness, if a set of polymorphic
+  %  formulas is unsatisfiable, then there exists a finite set of monomorphic
+  %  instances of the formulas that is also unsatisfiable. However, this
+  %  finite set cannot be computed \cite[Section 2, Theorem 1]{expr-poly-types}; hence any approach based on
+  %  instantiation of type variables is incomplete.
+
+B\"ome's iterative approach is implemented as part of the SMT integration \cite[Chapter 2]{sb-phd} in Isabelle/HOL.
+This implementation is also used by Sledgehammer \cite{judgement, hammer} to interface with superposition based automatic theorem provers.
+However, it is is only briefly documented as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.
 
@@
 
-  * the alternative is to translate polymorphic problems to monomorphic problems
-    * two main approaches:
-      * complete encoding of polymorphism, using guards or tags,
-        as in Blanchette et al.~\cite{xxx} and Bobot et al.~\cite{xxx}
-      * incomplete encoding of polymorphism based on iterative monomorphisation
-        such as introduced by Böhme \cite[Section 2.2.1]{sb-phd}, whereby type
-        variables are heuristically instantiated by concrete types
+In this paper, we present an algorithm based on his description and implementation (Section~\ref{sec:high-level-algorithm}).
+We also provide a more detailed description to allow the algorithm to be implemented by following the paper.
+Additionally, this descriptions shows some of the ways in which an implementation can be made more efficient and avoid explosions (Section~\ref{sec:low-level-algorithm}).
 
-\bigskip
-
-  * By a type version of compactness, if a set of polymorphic
-    formulas is unsatisfiable, then there exists a finite set of monomorphic
-    instances of the formulas that is also unsatisfiable. However, this
-    finite set cannot be computed \cite{xxx}; hence any approach based on
-    instantiation of type variables is incomplete.
-
-\bigskip
-
-  * Böhme's iterative approach is implemented as part of the SMT integration
-    \cite{xxx} in Isabelle/HOL. This implementation is also used by
-    Sledgehammer \cite{xxx} to interface automatic theorem provers based on
-    superposition.
-
-\bigskip
-
-  * The main drawback of Böhme's approach is that it is only briefly documented,
-    as a single paragraph in his PhD thesis \cite[Section 2.2.1]{sb-phd}
-  * In this paper, we present an algorithm inspired by his description and
-    implementation (Section~\ref{sec:high-level-algorithm}).
-    * we aimed at mathematical precision, so that people who want to reimplement
-      the algorithm can follow this blueprint
-    * we also show how to bound the algorithm to avoid explosions in practice
-      (Section~\ref{sec:low-level-algorithm}).
-      ``the explosive nature of the procedures requires various bounds which
-      limit the number of type arguments and clauses that are generated.''
-
-}
+% * Böhme's iterative approach is implemented as part of the SMT integration \cite[Chapter 2]{sb-phd}
+%     in Isabelle/HOL. This implementation is also used by
+%    Sledgehammer \cite{judgement, hammer} to interface automatic theorem provers based on
+%    superposition.
+%
+%  * The main drawback of Böhme's approach is that it is only briefly documented,
+%    as a single paragraph in his PhD thesis \cite[Section 2.2.1]{sb-phd}
+%  * In this paper, we present an algorithm inspired by his description and
+%    implementation (Section~\ref{sec:high-level-algorithm}).
+%    * we aimed at mathematical precision, so that people who want to reimplement
+%      the algorithm can follow this blueprint
+%    * we also show how to bound the algorithm to avoid explosions in practice
+%      (Section~\ref{sec:low-level-algorithm}).
+%      ``the explosive nature of the procedures requires various bounds which
+%      limit the number of type arguments and clauses that are generated.''
 
@@
 
-We implemented this algorithm in Zipperpostion \cite{xxx}, a higher order automatic theorem prover written in OCaml.
-Although Zipperposition is polymorphic, it uses the monomorphic prover E \cite{xxx} as a backend.
+\bigskip
+
+We implemented this algorithm in Zipperpostion \cite{zipp}, a higher order automatic theorem prover written in OCaml.
+Although Zipperposition is polymorphic, it uses the monomorphic prover E \cite{e} as a backend.
 Our algorithm makes it possible, for the first time, to use E with polymorphic problems.
@@
 
-Our empirical evaluation on TPTP problems \cite{xxx} attempts to answer three questions (Section~\ref{sec:evaluation}):
+Our empirical evaluation on TPTP problems \cite{tptp} attempts to answer three questions (Section~\ref{sec:evaluation}):
 %
@@
 
-\item Is iterative monomophisation more effective than the native polymorphism implemented in polymorphic provers?
+\item Is iterative monomorphisation more effective than the native polymorphism implemented in polymorphic provers?
 \end{enumerate}
@@
 
-\item For Leo-III \cite{xxx} and Vampire \cite{xxx}, we find that monomorphisation is indeed more effective than native polymorphism.
+\item For Leo-III \cite{leo-iii} and Vampire \cite{vamp}, we find that monomorphisation is indeed more effective than native polymorphism.
 \end{enumerate}
@@
 \end{itemize}
-our target logic to be the same thing except we foregoe type constructors and type variables 
+our target logic to be the same thing except we forego type constructors and type variables 
 we want our target logic to be a subset of the starting logic
@@
    \item monomorphic and polymorphic types
-   \item matching % being overly precise probably uncecessary
+   \item matching 
    \item term, literal, clause  % also part logic
@@
 
-   \item Once all such substiutions have been computed.
+   \item Once all such substitutions have been computed.
 We apply each substitutions to the clause they were generated from.
@@
 
-The previous problem would therefore never instantiate a clause, despite the problem becoming trvially contradictory if the the last clause and any of the first two are instantiated.
+The previous problem would therefore never instantiate a clause, despite the problem becoming trivially contradictory if the last clause and any of the first two are instantiated.
 
-Furthermore, the finite number of iterations of the algorithm consitutes an additional source of incompleteness in practice.
+Furthermore, the finite number of iterations of the algorithm constitutes an additional source of incompleteness in practice.
 %find example that would require a list list list int to be solved but only allow such a type to be generated after 3 iterations

### 2024-02-15 iterative_monomorphisation.tex
started working on background

@@
 
-One of the main applications of automatic theorem provers is making the use of proof assistants easier.
+One of the main applications of automatic theorem provers is reducing the workload when using proof assistants.
 Many proof assistants support rank-1 polymorphism, however, many automatic theorem provers only operate on monomorphic logics.
 One approach to close this gap is extending existing provers, allowing them to natively support polymorphism, as was the case for Vampire \cite{vamp}.
-The main drawbacks of this solution are the amount of work required to extend these provers and the fact that it needs to be done from scratch for every monomorphic prover.
+The main drawbacks of this solution are the amount of work required to extend such provers and that it needs to be done from scratch for every monomorphic prover.
 
@@
 The alternative would be to translate polymorphic problems into monomorphic problems.
-There are two main approaches to this, the first is a complete encoding of polymorphism using guards and tags \cite{mono-trans} into a monomorphic logic.
+There are two main ideas to do this, the first is a complete encoding of polymorphism using guards or tags \cite{mono-trans} into a monomorphic logic.
 The second solution is an incomplete encoding of polymorphism based on iterative monomorphisation, as introduced by B\"ome \cite[Section 2.2.1]{sb-phd}.
@@
   %* By a type version of compactness, if a set of polymorphic
-  %  formulas is unsatisfiable, then there exists a finite set of monomorphic
-  %  instances of the formulas that is also unsatisfiable. However, this
+  %  formulae is unsatisfiable, then there exists a finite set of monomorphic
+  %  instances of the formulae that is also unsatisfiable. However, this
   %  finite set cannot be computed \cite[Section 2, Theorem 1]{expr-poly-types}; hence any approach based on
@@
 This implementation is also used by Sledgehammer \cite{judgement, hammer} to interface with superposition based automatic theorem provers.
-However, it is is only briefly documented as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.
+However, it is is only documented as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.
 
@@
 We also provide a more detailed description to allow the algorithm to be implemented by following the paper.
-Additionally, this descriptions shows some of the ways in which an implementation can be made more efficient and avoid explosions (Section~\ref{sec:low-level-algorithm}).
+Additionally, this description shows some of the ways in which an implementation can be made more efficient and avoid explosions (Section~\ref{sec:low-level-algorithm}).
 
@@
 The algorithm works as follows.
-We assume problems to be sets of formulas.
+We assume problems to be sets of formulae.
 All type arguments of the problem are collected, and the polymorphic type arguments are matched against the monomorphic ones.
@@
 
-To keep the number of generated formulas finite, we limit the number of iterations.
-After the iterations are completed, the new type arguments are used to instantiate the type variables in the problem's formulas, generating new monomorphic clauses.
+To keep the number of generated formulae finite, we limit the number of iterations.
+After the iterations are completed, the new type arguments are used to instantiate the type variables in the problem's formulae, generating new monomorphic clauses.
 Finally, because monomorphic provers generally do not support $n$-ary type constructors, types must be mangled.
-For example, the compound type $\ty{list}(\ty{int})$ could be mangled to $\ty{list\_int}$.
+For example, the compound type $\ty{list}(\ty{int})$ could be mangled to the constant $\ty{list\_int}$.
 
@@
 % mention alternatives, (guards, tags) and current implementation in Isabelle
-
-Language: TF1  %TH1?
-
-We want our starting logic to be:
-\begin{itemize}
-    \item higher order
-    \item polymorphic: we want atomic types, type constructors and universally quantified type variables
-\end{itemize}
-our target logic to be the same thing except we forego type constructors and type variables 
-we want our target logic to be a subset of the starting logic
-
 Definitions :
 \begin{itemize}
-   \item type arguments % part of the logic
+   \item our logic (TF1), our starting and target logics (with and without polymorphism)
+   \item notations
+   \item type arguments
    \item monomorphic and polymorphic types
    \item matching 
-   \item term, literal, clause  % also part logic
+   \item term, literal, clause 
 \end{itemize}
+
+% TH1 ugly for height of 1 reasons, can be fixed with case feature
+Our starting logic will be TPTP's TH1 \cite{th1}, a higher-order logic with rank-1 polymorphism. We will assume the initial problem to be TH1 formulae.
+
+types are either constants, type variables, or a type constructor of a certain arity applied to other types
+
+we consider a set of function symbols, with type arguments and arguments of a certain arity
+
+a problem is made up of a set of formulae, type variables are quantified at the formula level
+
+note that in the implementation, a problem is a conjunction of disjunctive clauses, this is only relevant when the problem is parsed when the initial type arguments are collected and at the very end when the final clauses are generated
 
@@
 This is not necessary for \(I\) because its type arguments do not contain type variables.
-[seems a sufficient explanation, if not clear expand further]
 
@@
 
-[Add 2 or 3 examples to show how it works and explain why we have the different steps]
-
-\subsection{Soundness}
-Soundness of the algorithm follows from the fact that the resulting clauses are obtained by directly instantiating universally quantified type variable of the initial clauses.
-
-Then we rely upon the soundness of mangling [encoding mono and poly ty].
-
-\subsection{Incompleteness}
-
-This algorithm is incomplete, this is because not all possible instantiations of the type variables are generated.
-Consider the following problem [note that we need a consistent notation for type arguments (notably to distinguish them from normal arguments) also this might not be an issue but they are terms not literals ]:
-\begin{align*}
-   P = \{ & \forall \alpha. f \langle \alpha \rightarrow int \rangle \\
-          & \forall \alpha. f \langle nat \rightarrow \alpha \rangle \\
-          & \forall \alpha \beta. \lnot f \langle \alpha \rightarrow \beta \rangle \}
-\end{align*}
-
-The algorithm only matches against ground type arguments.
-This is the case because polymorphic type unification is undecidable (quick ref) and whilst interweaving phases of unification with the current algorithm would be theoretically possible, it would most likely not be of any practical use.
-Indeed, the current implementation is already very explosive and generating new substitutions with a costly unification procedure would most likely not result in any improvements.
-
-The previous problem would therefore never instantiate a clause, despite the problem becoming trivially contradictory if the last clause and any of the first two are instantiated.
-
-Furthermore, the finite number of iterations of the algorithm constitutes an additional source of incompleteness in practice.
-%find example that would require a list list list int to be solved but only allow such a type to be generated after 3 iterations
+This algorithm is clearly sound as the newly derived clauses are instances of the initial clauses where universally quantified types have been replaced by concrete types.
 

### 2024-02-16 iterative_monomorphisation.tex
work on background and high level sections

@@
 \documentclass[]{ceurart}
-
-%\sloppy
 
@@
 
+\newtheorem{definition}{Definition}
+
 \begin{document}
@@
 \conference{Submitted draft} % TODO update
+
+% TODO, at the end check the following
+%   - formulas -> formulae
+%   - clause -> formula
+%   - monomorphization -> monomorphisation
+%   - higer order -> higer-order
+%   - TH1 with ugly 1
+%   - correctly spelled TPTP, Leo-III, Zipperposition, E, Vampire
+%   - use of formula vs clause
+%   - polymorphic and monomorphic
+%   - title of bits of the algorithms
+%   - eg or ex -> e.g.
 
@@
 
-% mention alternatives, (guards, tags) and current implementation in Isabelle
-Definitions :
+% TH1 ugly for height of 1 reasons, can be fixed with case feature
+%The base logic of the implementation will be TPTP's TH1 \cite{th1}, a higher-order logic with rank-1 polymorphism.
+We will assume the initial problem to be a set of TH1 formulae.
+The choice of this logic is motivated by the widespread support it enjoys among higher-order provers.
+However, iterative monomorphisation only
+The implementation require problems to be based on TPTP's TH1 (or TF1) logic \cite{th1} because it is the a widespread standard among automatic provers for higher-order rank-1 polymorphic logics.
+However, the iterative monomorphisation algorithm relies only on assumptions regarding type construction and use in the problem.
+Consequently, the precise forms of formulae, terms or literals are irrelevant.
+
+\begin{definition}[Function symbol]
+   A function symbol \(f\), has two arities which define the number of arguments and type arguments it takes.
+The type arguments are enumerated between angled brackets and the other arguments between parentheses, e.g.: \(f\langle \tau_1, ..., \tau_n\rangle (x_1, ..., x_m)\)
+
+   In a well formed problem, the number of type arguments to which a function symbol is applied is assumed to correspond to the relevant arity.
+In other words, we expect all function symbols of type argument arity \(n\) to have \(n\) type arguments.
+\end{definition}
+
+% TODO: WARNING - heavily inspired from sections 2.2-2.3 of the th1 paper
+\begin{definition}[Type]
+A type \(\tau\) can be:
 \begin{itemize}
-   \item our logic (TF1), our starting and target logics (with and without polymorphism)
-   \item notations
-   \item type arguments
-   \item monomorphic and polymorphic types
-   \item matching 
-   \item term, literal, clause 
+   \item a constant, e.g.: \(\ty{int}\)
+   \item a type variable universally quantified at the formula level, e.g.: \(\alpha\)
+   \item built from an \(n\)-ary type constructor applied to \(n\) types, e.g.: \(\ty{map}(\ty{int}, \ty{list}(\alpha))\)
 \end{itemize}
 
-% TH1 ugly for height of 1 reasons, can be fixed with case feature
-Our starting logic will be TPTP's TH1 \cite{th1}, a higher-order logic with rank-1 polymorphism. We will assume the initial problem to be TH1 formulae.
+\end{definition}
 
-types are either constants, type variables, or a type constructor of a certain arity applied to other types
-
-we consider a set of function symbols, with type arguments and arguments of a certain arity
-
-a problem is made up of a set of formulae, type variables are quantified at the formula level
-
-note that in the implementation, a problem is a conjunction of disjunctive clauses, this is only relevant when the problem is parsed when the initial type arguments are collected and at the very end when the final clauses are generated
+\begin{definition}[Monomorphic, Polymorphic types]
+A type is monomorphic if it contains no type variables, otherwise it is polymorphic.
+\end{definition}
 
 % target page number: 2-4
-% needs to be re-written
 \section{High level algorithm}
@@
 
-We consider a problem \(P\) to be a set of clauses \(C_1, \dots, C_n\).
+We consider a problem \(P\) to be a set of formulae \(F_1, \dots, F_n\).
 
-Algorithm :
+\subsection{Initialisation step}
+The initial phase of the algorithm consists in gathering the maps \(M\) and \(P\), defined as follows:
 \begin{itemize}
-    \item Let \(F\) be the set of all function symbols in \(P\).
-    \item Given a function symbol \(f \in F\), we define \(I(f)\), the set of all ground type arguments that are passed to \(f\) is instantiated in \(P\).
+    \item Given a type \(n\)-ary function symbol \(f \in P\), we define \(M(f)\), the set of all \(n\)-uples of monomorphic type arguments passed to \(f\) in \(P\).
 
-    For instance if \(f^{a \rightarrow b}\) occurs in \(P\) (with \(a\) and \(b\) ground types) then, \(a \rightarrow b \in I(f) \).
- \item Given a clause \(C \in P\) and a function symbol \(f \in C\), we define \(U(C, f)\), the set of uninstantiated type arguments passed to \(f\) in \(C\).
+    For example, if \(f\langle \tau_1, \tau_2, \tau_3\rangle\) occurs in \(P\) (with \(\tau_1, \tau_2\) and \( \tau_3\) monomorphic) then, \((\tau_1, \tau_2, \tau_3) \in M(f) \).
 
-    It is important to paramatrise \(U\) with the clause of \(f\) because type variables are quantified at the clause level.
-This is not necessary for \(I\) because its type arguments do not contain type variables.
+ \item Given a formula \(F \in P\) and a type \(n\)-ary function symbol \(f \in F\), we define \(P(C, f)\), the set \(n\)-uples of polymorphic type arguments passed to \(f\) in \(F\).
+
+    It is important to paramatrise \(P\) with the formula of \(f\) because type variables are quantified at the formula level.
+This is not necessary for \(M\) because its types do not contain type variables.
 
@@
 
+\subsection{Iteration}
 A single iteration of the algorithm will do the following:
+
 \begin{itemize}
@@
 % Provide complete list of problems explain which ones were excluded and why
-
-Compare the following:
-    \begin{itemize}
-        \item E on original portfolio mode
-        \item E with original portfolio mode and monomorphisation
-        \item adjusted portforlio mode
-
-        \item when E is called
-        \item using zipp only as monomorphiser (passing the result to E, Vampire, Satallax, Leo-III)
-
-        \item test different bounds
-    \end{itemize}
-
-\break
-
-We ignore the following bounds: ty var limit per clause, subst per ty var, limiting the new type args per clause
+% We ignore the following bounds: ty var limit per clause, subst per ty var, limiting the new type args per clause
 

### 2024-02-16 iterative_monomorphisation.tex
pass on intro

@@
 
-One of the main applications of automatic theorem provers is reducing the workload when using proof assistants.
-Many proof assistants support rank-1 polymorphism, however, many automatic theorem provers only operate on monomorphic logics.
-One approach to close this gap is extending existing provers, allowing them to natively support polymorphism, as was the case for Vampire \cite{vamp}.
-The main drawbacks of this solution are the amount of work required to extend such provers and that it needs to be done from scratch for every monomorphic prover.
+One of the main applications of automatic theorem provers is to provide automation to users of proof assistants.
+Many proof assistants, such as HOL4 \cite{xxx}, HOL Light \cite{xxx} and Isabelle/HOL \cite{xxx}, support rank-1 polymorphism, where type quantification is allowed only at the top level of a formula.
+On the other hand, many provers operate only on monomorphic logics.
+One approach to close this gap is to extend provers to natively support polymorphism, as has been done for Vampire \cite{vamp}.
+This, however, entails a lot of work that needs to be redone for every prover.
 
@@
 
-The alternative would be to translate polymorphic problems into monomorphic problems.
-There are two main ideas to do this, the first is a complete encoding of polymorphism using guards or tags \cite{mono-trans} into a monomorphic logic.
-The second solution is an incomplete encoding of polymorphism based on iterative monomorphisation, as introduced by B\"ome \cite[Section 2.2.1]{sb-phd}.
-This method relies on heuristically instantiating type variables with concrete types.
+The alternative is to translate polymorphic problems to monomorphic problems.
+One approach is to encode polymorphism using guards or tags \cite{mono-trans} in a monomorphic logic.
+Another is to encode polymorphism using {iterative monomorphisation}, as introduced by B\"ohme \cite[Section 2.2.1]{sb-phd}.
+This method relies on heuristically instantiating the formulae's type variables with concrete types.
 
@@
 
-By compactness, we have that in first order logic, given a polymorphic formula \(F\), there exist an equisatisfiable finite set of monomorphic instances of \(F\).
-However such a set cannot be computed \cite[Section 2, Theorem 1]{expr-poly-types}, consequently any monomorphisation method based on instantiation of type variables is incomplete.
+By a type version of the compactness theorem, we have that in first order logic, given a polymorphic formula \(\varphi\), there exist an equisatisfiable finite set of monomorphic instances of \(\varphi\).
+However, such a set cannot be computed \cite[Theorem 1]{expr-poly-types}.
+As a result, any monomorphisation method based on instantiation of type variables is
+bound to be incomplete.
 
@@
 
-B\"ome's iterative approach is implemented as part of the SMT integration \cite[Chapter 2]{sb-phd} in Isabelle/HOL.
+B\"ohme's iterative approach is implemented as part of the SMT (satisfiability modulo theories) integration \cite[Chapter 2]{sb-phd} in Isabelle/HOL.
 This implementation is also used by Sledgehammer \cite{judgement, hammer} to interface with superposition based automatic theorem provers.
@@
 
-\bigskip
-
 In this paper, we present an algorithm based on his description and implementation (Section~\ref{sec:high-level-algorithm}).
-We also provide a more detailed description to allow the algorithm to be implemented by following the paper.
-Additionally, this description shows some of the ways in which an implementation can be made more efficient and avoid explosions (Section~\ref{sec:low-level-algorithm}).
+We also provide a more detailed description to help future implementers.
+In addition, this description shows some of the ways in which an implementation can be made more efficient and avoid explosions (Section~\ref{sec:low-level-algorithm}).
 
@@
 We assume problems to be sets of formulae.
-All type arguments of the problem are collected, and the polymorphic type arguments are matched against the monomorphic ones.
-This yields new type arguments, both polymorphic and monomorphic.
-The process is then repeated using the newly derived type arguments.
+All symbols of the problem are collected, and the polymorphic instances of symbols are matched against the monomorphic ones.
+This yields new instances of symbols, both polymorphic and monomorphic.
+The process is then iterated
+a number of times, making use of the newly generated instances.
 
-These iterations are required to account for $n$-ary type constructors.
+%These iterations are required to account for $n$-ary type constructors.
 Consider the unary type constructor \ty{list}.
 If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, it may be possible to generate the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc.
-However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once $\ty{list}(\ty{int})$ has already been added to the set of monomorphic types.
-This example also makes it clear that the set of types that can be generated is infinite.
+However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has already been generated.
+This example also makes it clear that the set of instances that can be generated is infinite.
 
 To keep the number of generated formulae finite, we limit the number of iterations.
-After the iterations are completed, the new type arguments are used to instantiate the type variables in the problem's formulae, generating new monomorphic clauses.
-Finally, because monomorphic provers generally do not support $n$-ary type constructors, types must be mangled.
-For example, the compound type $\ty{list}(\ty{int})$ could be mangled to the constant $\ty{list\_int}$.
+After the iterations are completed, the new monomorphic symbol instances are used to instantiate the polymorphic symbols in the problem's formulae, generating new monomorphic formulae.
+Finally, because monomorphic provers generally do not support $n$-ary type constructors, types must be `mangled'; for example, the compound type $\ty{list}(\ty{int})$ could be mangled to the constant $\ty{list\_int}$.
 
-\bigskip
-
-We implemented this algorithm in Zipperpostion \cite{zipp}, a higher order automatic theorem prover written in OCaml.
+We implemented iterative monormorphisation in Zipperpostion \cite{zipp}, a higher order prover written in OCaml.
 Although Zipperposition is polymorphic, it uses the monomorphic prover E \cite{e} as a backend.
-Our algorithm makes it possible, for the first time, to use E with polymorphic problems.
+This means that E can now be used with polymorphic problems.
 Moreover, our implementation of the algorithm in Zipperposition can be used as a preprocessor to interface with other stand-alone provers.
 
-Our empirical evaluation on TPTP problems \cite{tptp} attempts to answer three questions (Section~\ref{sec:evaluation}):
+Our empirical evaluation on TPTP \cite{tptp} problems attempts to answer three questions (Section~\ref{sec:evaluation}):
 %
@@
 However, iterative monomorphisation only
-The implementation require problems to be based on TPTP's TH1 (or TF1) logic \cite{th1} because it is the a widespread standard among automatic provers for higher-order rank-1 polymorphic logics.
+The implementation require problems to be based on TPTP's TH1 (or TF1) logic \cite{th1} because it is the a widespread standard among provers for higher-order rank-1 polymorphic logics.
 However, the iterative monomorphisation algorithm relies only on assumptions regarding type construction and use in the problem.
@@
 
-This table groups two bounds that directly affect the number of clauses generated, a first one that allows at most a number of new clauses equal to ``multiplier" times the original number of clauses.
+This table groups two bounds that directly affect the number of clauses generated, a first one that allows at most a number of new clauses equal to `multiplier' times the original number of clauses.
 And the cap that gives an absolute bound over the total number of clauses generated.

### 2024-02-16 iterative_monomorphisation.tex
Sect 2 work

@@
 
-\newtheorem{definition}{Definition}
+\newdefinition{definition}{Definition}
 
@@
 One of the main applications of automatic theorem provers is to provide automation to users of proof assistants.
-Many proof assistants, such as HOL4 \cite{xxx}, HOL Light \cite{xxx} and Isabelle/HOL \cite{xxx}, support rank-1 polymorphism, where type quantification is allowed only at the top level of a formula.
-On the other hand, many provers operate only on monomorphic logics.
-One approach to close this gap is to extend provers to natively support polymorphism, as has been done for Vampire \cite{vamp}.
+Many proof assistants, such as HOL4 \cite{slind-norrish-2008}, HOL Light \cite{harrison-2009}, and
+Isabelle/HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed only at the top level of formulae.
+On the other hand, many automatic provers operate only on monomorphic logics.
+One approach to close this gap is to extend provers to natively support polymorphism, as has been done for Vampire \cite{bhayat-reger-2020}.
 This, however, entails a lot of work that needs to be redone for every prover.
@@
 The alternative is to translate polymorphic problems to monomorphic problems.
-One approach is to encode polymorphism using guards or tags \cite{mono-trans} in a monomorphic logic.
-Another is to encode polymorphism using {iterative monomorphisation}, as introduced by B\"ohme \cite[Section 2.2.1]{sb-phd}.
+One approach is to encode polymorphism using dedicated function or predicate symbols \cite{mono-trans} in a monomorphic logic.
+Another approach is to encode polymorphism using {iterative monomorphisation}, as introduced by B\"ohme \cite[Section 2.2.1]{sb-phd}.
 This method relies on heuristically instantiating the formulae's type variables with concrete types.
@@
 This implementation is also used by Sledgehammer \cite{judgement, hammer} to interface with superposition based automatic theorem provers.
-However, it is is only documented as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.
+However, it is is documented only as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.
 
-In this paper, we present an algorithm based on his description and implementation (Section~\ref{sec:high-level-algorithm}).
+In this paper, we present an algorithm based on our understanding of his description and implementation (Section~\ref{sec:high-level-algorithm}).
 We also provide a more detailed description to help future implementers.
@@
 
+{\obeylines
+
+\noindent\kern\parindent
+  * algorithm works a priori with any rank-1 polymorphic logic
+
+  * in particular, it works with polymorphic first- and higher-order logic as 
+    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-201x,th1}
+
+  * the algorithm does not need to inspect the precise structure of formulae
+  * instead, it works on monomorphic or polymorphic instances of symbols that 
+    occur in the formulae
+  * type quantification is implicit, meaning that all type variables occurring
+    in a formula are assumed to be universally quantified at the formula's top
+    level
+}
+
+\bigskip
+
 % TH1 ugly for height of 1 reasons, can be fixed with case feature
-%The base logic of the implementation will be TPTP's TH1 \cite{th1}, a higher-order logic with rank-1 polymorphism.
-We will assume the initial problem to be a set of TH1 formulae.
-The choice of this logic is motivated by the widespread support it enjoys among higher-order provers.
-However, iterative monomorphisation only
-The implementation require problems to be based on TPTP's TH1 (or TF1) logic \cite{th1} because it is the a widespread standard among provers for higher-order rank-1 polymorphic logics.
-However, the iterative monomorphisation algorithm relies only on assumptions regarding type construction and use in the problem.
-Consequently, the precise forms of formulae, terms or literals are irrelevant.
 
-\begin{definition}[Function symbol]
-   A function symbol \(f\), has two arities which define the number of arguments and type arguments it takes.
-The type arguments are enumerated between angled brackets and the other arguments between parentheses, e.g.: \(f\langle \tau_1, ..., \tau_n\rangle (x_1, ..., x_m)\)
-
-   In a well formed problem, the number of type arguments to which a function symbol is applied is assumed to correspond to the relevant arity.
-In other words, we expect all function symbols of type argument arity \(n\) to have \(n\) type arguments.
+\begin{definition}%[Type]
+A \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
+the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\)).
+If $n = 0$, we omit the parentheses (e.g.\ \(\ty{int}\)).
 \end{definition}
 
-% TODO: WARNING - heavily inspired from sections 2.2-2.3 of the th1 paper
-\begin{definition}[Type]
-A type \(\tau\) can be:
-\begin{itemize}
-   \item a constant, e.g.: \(\ty{int}\)
-   \item a type variable universally quantified at the formula level, e.g.: \(\alpha\)
-   \item built from an \(n\)-ary type constructor applied to \(n\) types, e.g.: \(\ty{map}(\ty{int}, \ty{list}(\alpha))\)
-\end{itemize}
-
+\begin{definition}%[Monomorphism and polymorphism]
+A type is \emph{monomorphic} if it contains no type variables; otherwise it is \emph{polymorphic}.
 \end{definition}
 
-\begin{definition}[Monomorphic, Polymorphic types]
-A type is monomorphic if it contains no type variables, otherwise it is polymorphic.
+\begin{definition}%[Function symbol]
+A (function or predicate) \emph{symbol} \(f\) has a type arity that defines the number of type arguments it takes.
+These arguments are listed between angle brackets: \(f\langle \tau_1, \dots, \tau_n\rangle\), where each $\tau_i$ is a type.
 \end{definition}
+
+{
+\obeylines
+
+* substitutions
+
+* matching
+}
 
@@
 
+TODO: Use $\Phi_1$ etc.
+
 \subsection{Initialisation step}
@@
 \begin{itemize}
-    \item Given a type \(n\)-ary function symbol \(f \in P\), we define \(M(f)\), the set of all \(n\)-uples of monomorphic type arguments passed to \(f\) in \(P\).
+    \item Given a type \(n\)-ary symbol \(f \in P\), we define \(M(f)\), the set of all \(n\)-uples of monomorphic type arguments passed to \(f\) in \(P\).
 
@@
 
- \item Given a formula \(F \in P\) and a type \(n\)-ary function symbol \(f \in F\), we define \(P(C, f)\), the set \(n\)-uples of polymorphic type arguments passed to \(f\) in \(F\).
+ \item Given a formula \(F \in P\) and a type \(n\)-ary symbol \(f \in F\), we define \(P(C, f)\), the set \(n\)-uples of polymorphic type arguments passed to \(f\) in \(F\).
 
@@
 \begin{itemize}
-   \item For each clause \(C_i \in P\) and each function symbol \(f \in C_i\), we consider the uninstantiated type arguments \(U(C_i, f)\) these type type arguments are matched against all ground type arguments of \(I(f)\)
+   \item For each clause \(C_i \in P\) and each symbol \(f \in C_i\), we consider the uninstantiated type arguments \(U(C_i, f)\) these type type arguments are matched against all ground type arguments of \(I(f)\)
 
@@
 This algorithm is clearly sound as the newly derived clauses are instances of the initial clauses where universally quantified types have been replaced by concrete types.
+
+\subsection{Example}
+
+TODO
 
@@
    \State let \(t_1\), \(t_2\) be the terms of \(lit\)
-   \State recursively collect the function symbols and associated type arguments of \(t_1\) and \(t_2\) into a list of pairs
+   \State recursively collect the symbols and associated type arguments of \(t_1\) and \(t_2\) into a list of pairs
 
@@
 
-   \State \Return the list of (function symbol, type argument) pairs
+   \State \Return the list of (symbol, type argument) pairs
 
@@
 
-This function is very basic and simply gathers all the type arguments of the various function symbols of the problem and sorts them depending on whether they are ground or not.
+This function is very basic and simply gathers all the type arguments of the various symbols of the problem and sorts them depending on whether they are ground or not.
 
@@
 In the Zipperposition implemntation this iteration procedure is limited by several bounds.
-The number of new type arguments that can be generated for each function symbol is limited by user-defined bounds.
+The number of new type arguments that can be generated for each symbol is limited by user-defined bounds.
 These bounds can be defined proportionally.
@@
 
-The first two groups of options are natural as they affect how many new monomorphic and polymorphic type arguments are generated respectively for each function symbol at each iteration.
+The first two groups of options are natural as they affect how many new monomorphic and polymorphic type arguments are generated respectively for each symbol at each iteration.
 These options are closely linked and must be set together if they are to be relevant.
 
-During an iteration, for each function symbol (for each clause in the polymorphic case), the number of newly generated type arguments is determined by this formula:
+During an iteration, for each symbol (for each clause in the polymorphic case), the number of newly generated type arguments is determined by this formula:
 \( \min(\text{cap}, \max(\text{floor}, \text{multiplier} \times \text{number of type arguments})) \)
 
-The multiplier allows for a controlled increase of the number of type arguments, relative to the size of the problem whilst the floor prevents some function symbols seeing no new type arguments if the multiplier is too low.
+The multiplier allows for a controlled increase of the number of type arguments, relative to the size of the problem whilst the floor prevents some symbols seeing no new type arguments if the multiplier is too low.
 The cap acts as a final limit in the case the explosion is too fast despite the multiplier bound.

### 2024-02-16 iterative_monomorphisation.tex
merged abstract

@@
 \begin{abstract}
-Monomorphisation makes it possible to extend monomorphic provers to support polymorphic logics.
+Monomorphisation can be used to extend monomorphic provers to support polymorphic logics.
 We propose an iterative approach, which is necessarily incomplete but which works well in practice.
-It is implemented in the Zipperposition prover, where it can be used to translate away polymorphism before invoking a monomorphic prover.
-Our empirical evaluation reveals that iterative monomorphisation is a viable alternative to native polymorphism.
+It is implemented in the Zipperposition prover, where it can be used to translate away polymorphism before invoking the monomorphic prover E as a backend.
+Our evaluation demonstrates that this approach increases Zipperposition's success rate.
+Moreover, we find iterative monomorphisation to be competitive with native implementations of polymorphism.
 \end{abstract}
@@
 % target page number: 0.5-1.5
-\section{Background}
-\label{sec:background}
+\section{Preliminaries}
+\label{sec:preliminaries}
 
@@
 
+  * Böhme
+  * similar algo implemented in Why3?
+
 % target page number: 0.5-1
@@
 % Future avenues
+
+\section*{Acknowledgements}
+
+* Sascha Böhme
+
 \bibliography{citations}

### 2024-02-19 iterative_monomorphisation.tex
first proper draft for preliminaries

@@
 %   - eg or ex -> e.g.
+%   - function symbol -> symbol instance (depends on context)
+%   - ... -> \dots
+%   - comma splices!!!!
+%   - ``'' -> `'
+%   - use \Phi instead of F for formulae
+%   - concrete types -> monomorphic types
+%   - use in not \in when we are not precisely dealing with set membership
+%
+%
 
@@
 
-{\obeylines
+This algorithm does not need to inspect the structure of the problem's formulae because it relies exclusively on the formulae's monomorphic and polymorphic symbol instances.
+The notion of formula is necessary as type variables are assumed to be universally quantified at a formula's top level.
+The precise form of a formula can nevertheless be left unspecified.
 
-\noindent\kern\parindent
-  * algorithm works a priori with any rank-1 polymorphic logic
+Iterative monomorphisation can therefore be used in a vast range of rank-1 polymorphic logic.
+In particular, it can operate in the polymorphic first- and higher-order logics embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}, popular among automatic provers.
 
-  * in particular, it works with polymorphic first- and higher-order logic as 
-    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-201x,th1}
+The following terms are used in the algorithm descriptions.
 
-  * the algorithm does not need to inspect the precise structure of formulae
-  * instead, it works on monomorphic or polymorphic instances of symbols that 
-    occur in the formulae
-  * type quantification is implicit, meaning that all type variables occurring
-    in a formula are assumed to be universally quantified at the formula's top
-    level
-}
+%{\obeylines
+%
+%\noindent\kern\parindent
+%  * algorithm works a priori with any rank-1 polymorphic logic
+%
+%  * in particular, it works with polymorphic first- and higher-order logic as 
+%    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}
+%
+%  * the algorithm does not need to inspect the precise structure of formulae
+%  * instead, it works on monomorphic or polymorphic instances of symbols that 
+%    occur in the formulae
+%  * type quantification is implicit, meaning that all type variables occurring
+%    in a formula are assumed to be universally quantified at the formula's top
+%    level
+%}
 
@@
 These arguments are listed between angle brackets: \(f\langle \tau_1, \dots, \tau_n\rangle\), where each $\tau_i$ is a type.
+
+A \emph{symbol instance} refers to a symbol provided with its type arguments.
 \end{definition}
 
-{
-\obeylines
+\begin{definition} % HEAVILY inspired from 2.4 in mono-trans (it is the definition of a substitution, there's bound to be some comonality between all of them) 
+A \emph{substitution} \(\sigma\) is a function mapping one or more type variables to corresponding monomorphic type.
+Substitutions are in practice mostly be applied to formulae.
+Given a formula \(\Phi\), the formula obtained by replacing each type variable \(\alpha\) in \(\Phi\) by \(\sigma\alpha\) is written \(\sigma\Phi\).
 
-* substitutions
+\end{definition}
 
-* matching
-}
+\begin{definition}
+   Given a monomorphic type \(\tau_1\) and a polymorphic type \(\tau_2\), \emph{matching} \(\tau_2\) against \(\tau_1\) will either fail or yield a substitution \(\sigma\) such that \(\sigma\tau_2 = \tau_1\).
+Note that in the second case \(\sigma\tau_2\) is monomorphic.
+\end{definition}
+
+%{
+%\obeylines
+%
+%* substitutions
+%
+%* matching
+%}
 

### 2024-02-19 iterative_monomorphisation.tex
edits prelims

--- 2024-02-16 iterative_monomorphisation.tex
@@
 
-{\obeylines
+This algorithm works independently of the structure of the problem's formulae.
+It relies exclusively on the formulae's monomorphic and polymorphic symbol instances.
+Type variables are assumed to be universally quantified at a formula's top level.
+The precise form of formulae is left unspecified.
+%
+Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic.
+In particular, it can operate in the polymorphic first- and higher-order logics embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.
 
-\noindent\kern\parindent
-  * algorithm works a priori with any rank-1 polymorphic logic
+%The following terms are used in the algorithm descriptions.
 
-  * in particular, it works with polymorphic first- and higher-order logic as 
-    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-201x,th1}
-
-  * the algorithm does not need to inspect the precise structure of formulae
-  * instead, it works on monomorphic or polymorphic instances of symbols that 
-    occur in the formulae
-  * type quantification is implicit, meaning that all type variables occurring
-    in a formula are assumed to be universally quantified at the formula's top
-    level
-}
-
-\bigskip
+%{\obeylines
+%
+%\noindent\kern\parindent
+%  * algorithm works a priori with any rank-1 polymorphic logic
+%
+%  * in particular, it works with polymorphic first- and higher-order logic as 
+%    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}
+%
+%  * the algorithm does not need to inspect the precise structure of formulae
+%  * instead, it works on monomorphic or polymorphic instances of symbols that 
+%    occur in the formulae
+%  * type quantification is implicit, meaning that all type variables occurring
+%    in a formula are assumed to be universally quantified at the formula's top
+%    level
+%}
 
@@
 
+Our abstract framework relies on the following basic definitions.
+
 \begin{definition}%[Type]
 A \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
-the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\)).
+the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\), \(\ty{map}(\ty{int},\ty{string})\)).
 If $n = 0$, we omit the parentheses (e.g.\ \(\ty{int}\)).
@@
 \begin{definition}%[Function symbol]
-A (function or predicate) \emph{symbol} \(f\) has a type arity that defines the number of type arguments it takes.
-These arguments are listed between angle brackets: \(f\langle \tau_1, \dots, \tau_n\rangle\), where each $\tau_i$ is a type.
+A (function or predicate) {symbol} \(f\) has a \emph{type arity} that defines the number of type arguments it takes.
+A \emph{symbol instance} is a symbol applied to type arguments listed between angle brackets: \(f\langle \tau_1, \dots, \tau_n\rangle\), where each $\tau_i$ is a type.
+If $n = 0$, we omit the brackets (e.g.\ $f$).
 \end{definition}
 
-{
-\obeylines
+\begin{definition} % HEAVILY inspired from 2.4 in mono-trans (it is the definition of a substitution, there's bound to be some comonality between all of them) 
+A (type) \emph{substitution} is a partial function mapping a finite number of type variables to corresponding types. Substitutions are written as
+$\sigma = \{\alpha_1\mapsto\tau_1, \dots, \alpha_n\mapsto\tau_n\}$.
+They are assumed to be lifted to formulae; thus, $\sigma(\varphi)$ yields the variant of $\varphi$ in which each $\alpha_i$ is replaced by $\tau_i$.
+%The substitution %$\{\alpha_1\mapsto\tau_1, \dots, \alpha_n\mapsto\tau_n\}$
+%is \emph{monomorphising} if the types $\tau_i$ are all monomorphic.
+\end{definition}
 
-* substitutions
+\begin{definition}
+Given two types \(\tau, \upsilon\), \emph{matching} \(\upsilon\) against \(\tau\) will either fail or yield a substitution \(\sigma\) such that \(\sigma\upsilon = \tau\).
+\end{definition}
 
-* matching
-}
+In the following sections, we will always match a polymorphic type $\upsilon$ against a monomorphic type $\tau$.
+% and, on success, obtain a monomorphising substitution.
+
+%{
+%\obeylines
+%
+%* substitutions
+%
+%* matching
+%}
 
@@
 
-TODO: Use $\Phi_1$ etc.
+TODO: Use $\Phi$ for problem and $\varphi_1$ etc.
 

### 2024-02-19 iterative_monomorphisation.tex
decent progress on the high-level algorithm description

@@
 %   - use in not \in when we are not precisely dealing with set membership
-%
-%
+%   - $\Phi$ for problem and $\varphi_1$ etc.
+%   - n-uples -> n-tuples
+%   - ize -> ise (in cas i forgot any)
+%   - remove vertical space before lists (if allowed)
+%   - add full stops to end of sentences in itemize
+%   - B\"ome -> B\"ohme
 
@@
 
-\begin{definition} % HEAVILY inspired from 2.4 in mono-trans (it is the definition of a substitution, there's bound to be some comonality between all of them) 
+\begin{definition}
 A (type) \emph{substitution} is a partial function mapping a finite number of type variables to corresponding types. Substitutions are written as
@@
 They are assumed to be lifted to formulae; thus, $\sigma(\varphi)$ yields the variant of $\varphi$ in which each $\alpha_i$ is replaced by $\tau_i$.
-%The substitution %$\{\alpha_1\mapsto\tau_1, \dots, \alpha_n\mapsto\tau_n\}$
-%is \emph{monomorphising} if the types $\tau_i$ are all monomorphic.
+\end{definition}
+
+\begin{definition}
+Two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_n\mapsto\tau_n\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_m\mapsto\upsilon_m\}\) are \emph{compatible} if for all \(i, j\): \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\).
 \end{definition}
@@
 In the following sections, we will always match a polymorphic type $\upsilon$ against a monomorphic type $\tau$.
-% and, on success, obtain a monomorphising substitution.
-
-%{
-%\obeylines
-%
-%* substitutions
-%
-%* matching
-%}
 
@@
 
-We consider a problem \(P\) to be a set of formulae \(F_1, \dots, F_n\).
+We consider an initial problem \(\Phi\) to be a set of formulae \(\varphi_1, \dots, \varphi_n\).
 
-TODO: Use $\Phi$ for problem and $\varphi_1$ etc.
+\subsection{Collection}
+The initial phase of each iteration consists in collecting the elements of the maps \(M\) and \(P\), defined as follows:
+\begin{itemize}
+    \item Given a type \(n\)-ary symbol \(f\) in \(\Phi\). \(M(f)\) is the set of all \(n\)-tuples of monomorphic type arguments to which \(f\) is applied in \(\Phi\).
 
-\subsection{Initialisation step}
-The initial phase of the algorithm consists in gathering the maps \(M\) and \(P\), defined as follows:
-\begin{itemize}
-    \item Given a type \(n\)-ary symbol \(f \in P\), we define \(M(f)\), the set of all \(n\)-uples of monomorphic type arguments passed to \(f\) in \(P\).
+    For example, if \(f\langle \tau_1, \tau_2, \tau_3\rangle\) (with \(\tau_1, \tau_2\) and \( \tau_3\) monomorphic) occurs in \(\Phi\) then, \((\tau_1, \tau_2, \tau_3) \in M(f) \).
 
-    For example, if \(f\langle \tau_1, \tau_2, \tau_3\rangle\) occurs in \(P\) (with \(\tau_1, \tau_2\) and \( \tau_3\) monomorphic) then, \((\tau_1, \tau_2, \tau_3) \in M(f) \).
+ \item Given a formula \(\varphi \in \Phi\) and a type \(n\)-ary symbol \(f\) in \(\varphi\).
+\(P(\varphi, f)\) is the set of all \(n\)-tuples of type arguments to which \(f\) is applied in \(\varphi\).
+At least one of these type arguments must be polymorphic.
 
- \item Given a formula \(F \in P\) and a type \(n\)-ary symbol \(f \in F\), we define \(P(C, f)\), the set \(n\)-uples of polymorphic type arguments passed to \(f\) in \(F\).
+    For example, if \(f\langle \tau_1, \tau_2, \tau_3\rangle\) (with at least one of \(\tau_1, \tau_2\) or \( \tau_3\) polymorphic) occurs in \(\varphi\) then, \((\tau_1, \tau_2, \tau_3) \in P(\varphi, f) \).
 
-    It is important to paramatrise \(P\) with the formula of \(f\) because type variables are quantified at the formula level.
-This is not necessary for \(M\) because its types do not contain type variables.
+    It is important to parametrise \(P\) with both \(\varphi\) and \(f\) because type variables are quantified at the formula level.
+This is not necessary for \(M\) since all the types it contains are monomorphic.
 
@@
 \subsection{Iteration}
-A single iteration of the algorithm will do the following:
+A single iteration will consist in the following:
+\begin{itemize}
 
-\begin{itemize}
-   \item For each clause \(C_i \in P\) and each symbol \(f \in C_i\), we consider the uninstantiated type arguments \(U(C_i, f)\) these type type arguments are matched against all ground type arguments of \(I(f)\)
+   \item The maps \(M\) and \(P\) are initialised with respect to \(\Phi\).
 
-      [We have an issue: we don't define matching, don't think we need to go into a lot of detail]
+   \item An empty set of formulae \(\Psi\) is created.
 
-      Each such matching may return a substitution which instantiates one or more type variables of \(C_i\) (that would also be in the type arguments of \(f\))
+   \item For each formula \(\varphi \in \Phi\), for each symbol \(f\) in \(\varphi\):
 
-   \item Once all such substitutions have been computed.
-We apply each substitutions to the clause they were generated from.
-These newly created clauses are added to the problem and the newly generated type arguments (both ground and uninstantiated) will be added to \(U\) and \(I\) to be used for the next iteration.
+      An empty set of substitutions \(\Sigma_{\varphi, f}\) is created, then, for each type argument tuple \((\tau_1, \dots, \tau_n) \in  P(\varphi, f)\) and each type argument tuple \((\upsilon_1, \dots, \upsilon_n) \in M(f)\):
+
+      Each \(\tau_i\) is matched against its corresponding \(\upsilon_i\) possibly yielding the substitution \(\sigma_i\).
+      If all \(n\) matchings are successful and the substitutions are two by two compatible, \(\sigma_1 \circ \dots \circ \sigma_n\) is added to \(\Sigma_{\varphi, f}\).
+
+   \item For each formula \(\varphi \in \Phi\), for each symbol \(f\) in \(\varphi\), for each substitution \(\sigma \in \Sigma_{\varphi, f}\):
+
+      \(\sigma(\varphi)\) is added to \(\Psi\).
+
+   \item The initial problem \(\Phi\) is replaced with \(\Phi \cup \Psi\).
+
 \end{itemize}
 
-After a set number of iterations, we take all clauses with no type variables and mangle their types [see relevant paper].
-The resulting set of clauses forms a problem \(P'\) with the following properties:
-\begin{itemize}
-   \item \(P'\) is monomorphic [rewrite that as saying \(P'\) is in the monomorphic fragment of the original logic]
-   \item if a contradiction can be derived from \(P'\), then a contradiction can be derived from \(P\) [soundness]
-   \item we do not however have the converse [completeness]
-\end{itemize}
+After a set number of iterations, all monomorphic formulae are collected from \(\Phi\) and output.
 
-This algorithm is clearly sound as the newly derived clauses are instances of the initial clauses where universally quantified types have been replaced by concrete types.
+This algorithm is clearly sound as the newly derived formulae are instances of the initial formulae where universally quantified type variables have been instantiated with monomorphic types.
 
@@
 
-TODO
+Consider a problem made up of the following formulae:
+\begin{enumerate}
+   \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha,\ty{list}(\alpha)\rangle(a, b)\)
+   \item \(\forall i: \ty{int}, j: \ty{list}(\ty{int}), f\langle \ty{int}, \ty{list}(\ty{int})\rangle(i, j)\)
+\end{enumerate}
 
@@
 \begin{itemize}
-   \item the substitutions are directly applied to the type arguments, instead of the clauses, this also dispenses us from having to re-extract the type arguments from the clauses at each iteration
-   \item as a result the new clauses are generated only once all substitutions have been computed and iteratively applied to the various type arguments
-   \item some operations are performed in a different order to avoid useless computations [which ones?]
-   \item bounds are used at almost all steps of the process as the various enumerations are highly explosive and require bounds in order to be useful in a practical setting
-   \item the various sets are implemented with maps
+   \item substitutions are directly applied to the type arguments, instead of the formulae, this also dispenses us from having to re-extract the type arguments from the clauses at each iteration
+   \item new clauses are generated only once all substitutions have been computed
+   \item bounds are used at almost all stages of the algorithm as the various enumerations are highly explosive and must be limited in a practical setting
+   \item type arguments are separated into old and new sets, thereby avoiding computing the same matchings for each iteration
+   \item once the new formulae are generated, their monomorphic types are mangled
 \end{itemize}
 
-\begin{algorithm}[th]
-\begin{algorithmic}[1]
-\Function{Type arguments}{Literal: lit}
-
-   \State let \(t_1\), \(t_2\) be the terms of \(lit\)
-   \State recursively collect the symbols and associated type arguments of \(t_1\) and \(t_2\) into a list of pairs
-
-   \State \Comment Going a bit fast here, probably need to reformulate, but don't go into details, it's not important
-
-   \State \Return the list of (symbol, type argument) pairs
-
-\EndFunction
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}[tbh]
-\begin{algorithmic}[1]
-\Function{Initialisation}{Problem: P}
-   \ForAll{clause \(C \in P\)}
-      \ForAll{literal \(lit \in C\)}
-         \State let fun\_sym\_ty\_args = Type arguments(lit)
-         \ForAll{(fun\_sym, ty\_args) in fun\_sym\_ty\_args}
-         \If{\( \forall ty\_arg \in ty\_args. is\_ground(ty\_arg)\) }
-
-            \State \(I(fun\_sym) = I(fun\_sym) \cup \{ty\_args\}\)
-
-         \Else
-
-            \State \(U(C, fun\_sym) = U(C, fun\_sym) \cup \{ty\_args\}\)
-
-         \EndIf
-         \EndFor
-      \EndFor
-   \EndFor
-
-   \State \Return (I, U)
-
-\EndFunction
-\end{algorithmic}
-\end{algorithm}
-
-This function is very basic and simply gathers all the type arguments of the various symbols of the problem and sorts them depending on whether they are ground or not.
+We assume that the maps \(P\) and \(M\) have been collected from the initial problem. This step is dependent on the precise form of the formulae which has been left undefined.
 

### 2024-02-20 iterative_monomorphisation.tex
continued working on high level description, particularly the example

@@
 \begin{definition}
-Two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_n\mapsto\tau_n\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_m\mapsto\upsilon_m\}\) are \emph{compatible} if for all \(i, j\): \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\).
+Two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_n\mapsto\tau_n\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_m\mapsto\upsilon_m\}\) are \emph{compatible} if for all \(i, j\), \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\).
 \end{definition}
@@
 \subsection{Collection}
-The initial phase of each iteration consists in collecting the elements of the maps \(M\) and \(P\), defined as follows:
-\begin{itemize}
-    \item Given a type \(n\)-ary symbol \(f\) in \(\Phi\). \(M(f)\) is the set of all \(n\)-tuples of monomorphic type arguments to which \(f\) is applied in \(\Phi\).
+The initial phase of each iteration consists in collecting the elements of the maps \(M\) and \(P\).
 
+Given a type \(n\)-ary symbol \(f\) in \(\Phi\).
+\(M(f)\) is the set of all \(n\)-tuples of monomorphic type arguments to which \(f\) is applied in \(\Phi\).
 For example, if \(f\langle \tau_1, \tau_2, \tau_3\rangle\) (with \(\tau_1, \tau_2\) and \( \tau_3\) monomorphic) occurs in \(\Phi\) then, \((\tau_1, \tau_2, \tau_3) \in M(f) \).
 
- \item Given a formula \(\varphi \in \Phi\) and a type \(n\)-ary symbol \(f\) in \(\varphi\).
-\(P(\varphi, f)\) is the set of all \(n\)-tuples of type arguments to which \(f\) is applied in \(\varphi\).
-At least one of these type arguments must be polymorphic.
-
+Given a formula \(\varphi \in \Phi\) and a type \(n\)-ary symbol \(f\) in \(\varphi\).
+\(P(\varphi, f)\) is the set of all remaining \(n\)-tuples of type arguments to which \(f\) is applied in \(\varphi\) which are not in \(M(f)\).
+Meaning that at least one the type arguments is polymorphic.
 For example, if \(f\langle \tau_1, \tau_2, \tau_3\rangle\) (with at least one of \(\tau_1, \tau_2\) or \( \tau_3\) polymorphic) occurs in \(\varphi\) then, \((\tau_1, \tau_2, \tau_3) \in P(\varphi, f) \).
@@
 
-\end{itemize}
-
 \subsection{Iteration}
-A single iteration will consist in the following:
+Each iteration of the algorithm follows these steps:
 \begin{itemize}
@@
 
-      An empty set of substitutions \(\Sigma_{\varphi, f}\) is created, then, for each type argument tuple \((\tau_1, \dots, \tau_n) \in  P(\varphi, f)\) and each type argument tuple \((\upsilon_1, \dots, \upsilon_n) \in M(f)\):
+      An empty set of substitutions \(\Sigma_{\varphi, f}\) is created, then, for every type argument tuple \((\tau_1, \dots, \tau_n) \in  P(\varphi, f)\) and every type argument tuple \((\upsilon_1, \dots, \upsilon_n) \in M(f)\):
 
@@
 
-After a set number of iterations, all monomorphic formulae are collected from \(\Phi\) and output.
+After a set number of iterations, all monomorphic formulae in \(\Phi\) are output.
 
@@
 \begin{enumerate}
-   \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha,\ty{list}(\alpha)\rangle(a, b)\)
-   \item \(\forall i: \ty{int}, j: \ty{list}(\ty{int}), f\langle \ty{int}, \ty{list}(\ty{int})\rangle(i, j)\)
+   \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
+   \item \(\forall i: \ty{int}, f\langle \ty{int}\rangle(i)\)
 \end{enumerate}
+
+\begin{table}[ht]
+\caption{State of the algorithm}
+\centering\begin{tabular}{@{}lccccccc@{}}
+   \toprule
+               & Initialisation & Iteration 1 & Iteration 2 & Iteration 3\\
+   \midrule
+   \(M\)       & \((\ty{int})\)                     & -                     & \((\ty{list}(\ty{int}))\) \\
+   \(P(1, f)\) & \((\alpha)\; (\ty{list}(\alpha))\) & -                     & - \\
+   \(P(2, f)\) & \(\emptyset\)                      & -                     & - & \(\emptyset\) \\
+   \(\Phi\)    & [1] \; [2]                         & \(\sigma(\text{[1]})\)& \(\sigma'(\text{[1]})\)& \\
+   \bottomrule
+\end{tabular}
+\end{table}
+
+On the first iteration, two matchings are attempted, that of \(\alpha\) against \(\ty{int}\) which yields \(\sigma = \{\alpha \mapsto \ty{int}\}\) and that of \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails.
+Then, \(\sigma\) is applied to [1], resulting in a new formula: \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\)
+
+At the beginning of the second iteration, the clause generated in the previous iteration is used to update \(M\).
 

### 2024-02-20 iterative_monomorphisation.tex
citations.bib

@@
    \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
-   \item \(\forall i: \ty{int}, f\langle \ty{int}\rangle(i)\)
+   \item \(f\langle \ty{int}\rangle(0)\)
 \end{enumerate}
@@
    \midrule
-   \(M\)       & \((\ty{int})\)                     & -                     & \((\ty{list}(\ty{int}))\) \\
-   \(P(1, f)\) & \((\alpha)\; (\ty{list}(\alpha))\) & -                     & - \\
-   \(P(2, f)\) & \(\emptyset\)                      & -                     & - & \(\emptyset\) \\
-   \(\Phi\)    & [1] \; [2]                         & \(\sigma(\text{[1]})\)& \(\sigma'(\text{[1]})\)& \\
+   \(M\)       & \((\ty{int})\) & - &\(+(\ty{list}(\ty{int}))\) & \(+(\ty{list}(\ty{list}(\ty{int})))\) \\
+   \(P(1, f)\) & \((\alpha)\; (\ty{list}(\alpha))\) & - & - & -\\
+   \(P(2, f)\) & \(\emptyset\) & - & - & - \\
+   \(\Phi\)    & [1] \; [2] &\(+\sigma(\text{[1]})\)&\(+\sigma'(\text{[1]})\)& \(+\sigma''(\text{[1]})\)\\
    \bottomrule
@@
 
-At the beginning of the second iteration, the clause generated in the previous iteration is used to update \(M\).
+At the beginning of the second iteration, the clause generated in the previous iteration is used to update \(M(f)\), adding \((\ty{list}(\ty{int}))\).
+\(P\) has no new elements.
+Similarly to the first iteration we attempt the following matchings:
+\begin{itemize}
+   \item \(\alpha\) against \(\ty{int}\) which succeeds resulting in \(\sigma\)
+   \item \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails
+   \item \(\alpha\) against \(\ty{list}(\ty{int})\) resulting in \(\sigma'=\{\alpha\mapsto\ty{list}(\ty{int})\}\)
+   \item \(\ty{list}(\alpha)\) against \(\ty{list}(\ty{int})\) resulting in \(\sigma\)
+\end{itemize}
+
+From these substitutions, a single new formula can be generated from the application of \(\sigma'\) to [1], giving us: \(\sigma'(\text{[1]}) = \forall a: \ty{list}(\ty{int}), b:\ty{list}(\ty{list}(\ty{int})), f\langle\ty{list}(\ty{int})\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{int}))\rangle(b)\)
+
+This new formula will be used to add \((\ty{list}(\ty{list}(\ty{int})))\) to \(M(f)\) at the beginning of the third iteration.
+Similarly to the previous iteration, a single new formula will be generated.
+It is obtained by applying the new substitution \(\sigma'' = \{\alpha\mapsto\ty{list}(\ty{list}(\ty{int}))\}\) to [1].
+
+This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
+It also shows that the high-level description is unsuitable for direct implementation.
+Indeed, many operations, from matchings to substitution applications, are unnecessarily repeated.
+Additionally, in practice input problems will be far larger and more explosive, any reasonable implementation will require bounds to be useful.
 
@@
 
-This section presents a more detailed description of the algorithm, it follows closely the implementation in Zipperposition.
+This section presents a more detailed description of a practical adaptation of iterative monomorphisation, it closely follows the implementation in Zipperposition.
 

### 2024-02-20 iterative_monomorphisation.tex
edits with Tanguy

@@
 \begin{definition}%[Type]
-A \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
+A (\emph{polymorphic}) \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
 the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\), \(\ty{map}(\ty{int},\ty{string})\)).
@@
 \begin{definition}%[Monomorphism and polymorphism]
-A type is \emph{monomorphic} if it contains no type variables; otherwise it is \emph{polymorphic}.
+A type is \emph{monomorphic} if it contains no type variables.
 \end{definition}
@@
 \begin{definition}
-A (type) \emph{substitution} is a partial function mapping a finite number of type variables to corresponding types. Substitutions are written as
+A (\emph{type}) \emph{substitution} is a partial function mapping a finite number of type variables to corresponding types. Substitutions are written as
 $\sigma = \{\alpha_1\mapsto\tau_1, \dots, \alpha_n\mapsto\tau_n\}$.
@@
 \begin{definition}
-Two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_n\mapsto\tau_n\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_m\mapsto\upsilon_m\}\) are \emph{compatible} if for all \(i, j\), \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\).
+Two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_m\mapsto\tau_m\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_n\mapsto\upsilon_n\}\) are \emph{compatible} if \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\) for all \(i, j\).
 \end{definition}
@@
 \begin{definition}
-Given two types \(\tau, \upsilon\), \emph{matching} \(\upsilon\) against \(\tau\) will either fail or yield a substitution \(\sigma\) such that \(\sigma\upsilon = \tau\).
+Given two types \(\tau, \upsilon\), \emph{matching} \(\upsilon\) against \(\tau\) will either fail or yield a substitution \(\sigma\) such that \(\sigma(\upsilon) = \tau\).
 \end{definition}
@@
 
-We consider an initial problem \(\Phi\) to be a set of formulae \(\varphi_1, \dots, \varphi_n\).
+We start with a high level description of the iterative monomorphisation algorithm.
+The algorithm takes a polymorphic problem as argument and returns a monomorphic problem.
+It operates by applying an arbitrary number of iterations.
+Each iteration takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae.
+A single iteration consists of a collection phase (Section~\ref{ssec:collection}) and a processing phase (Section~\ref{ssec:processing}).
 
 \subsection{Collection}
-The initial phase of each iteration consists in collecting the elements of the maps \(M\) and \(P\).
+\label{ssec:collection}
 
-Given a type \(n\)-ary symbol \(f\) in \(\Phi\).
-\(M(f)\) is the set of all \(n\)-tuples of monomorphic type arguments to which \(f\) is applied in \(\Phi\).
-For example, if \(f\langle \tau_1, \tau_2, \tau_3\rangle\) (with \(\tau_1, \tau_2\) and \( \tau_3\) monomorphic) occurs in \(\Phi\) then, \((\tau_1, \tau_2, \tau_3) \in M(f) \).
+The initial phase of each iteration consists of computing two maps, \(M\) and \(P\), from the input problem~$\Phi$:
 
-Given a formula \(\varphi \in \Phi\) and a type \(n\)-ary symbol \(f\) in \(\varphi\).
-\(P(\varphi, f)\) is the set of all remaining \(n\)-tuples of type arguments to which \(f\) is applied in \(\varphi\) which are not in \(M(f)\).
-Meaning that at least one the type arguments is polymorphic.
-For example, if \(f\langle \tau_1, \tau_2, \tau_3\rangle\) (with at least one of \(\tau_1, \tau_2\) or \( \tau_3\) polymorphic) occurs in \(\varphi\) then, \((\tau_1, \tau_2, \tau_3) \in P(\varphi, f) \).
+\begin{itemize}
+\item Given a symbol \(f\) occurring in \(\Phi\), the set \(M(f)\) consists of all tuples of monomorphic type arguments to which \(f\) is applied in \(\Phi\).
+For example, if \(\sym{foldl}\langle \ty{nat}, \ty{int}\rangle\) occurs in \(\Phi\), then \((\ty{nat}, \ty{int}) \in M(\sym{foldl}) \).
 
-It is important to parametrise \(P\) with both \(\varphi\) and \(f\) because type variables are quantified at the formula level.
+\item Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(P(\varphi, f)\) consists of all tuples of type arguments to which \(f\) is applied in \(\varphi\) and which contains a type variable.
+For example, if \(\sym{foldl}\langle \ty{nat}, \ty{list}(\alpha)\rangle\) occurs in \(\varphi\), then \((\ty{nat}, \ty{list}(\alpha)) \in P(\varphi, \sym{foldl}) \).
+\end{itemize}
+
+It is important to parametrise \(P\) with \(\varphi\) because type variables are quantified at the formula level.
+The formula indicates the scope of type variables.
 This is not necessary for \(M\) since all the types it contains are monomorphic.
 
-\subsection{Iteration}
+\subsection{Processing}
+\label{ssec:processing}
+
 Each iteration of the algorithm follows these steps:

### 2024-02-20 iterative_monomorphisation.tex
continued work on high-level part

@@
 They are assumed to be lifted to formulae; thus, $\sigma(\varphi)$ yields the variant of $\varphi$ in which each $\alpha_i$ is replaced by $\tau_i$.
+
+Given \(\tau\) and \(\upsilon\) two substitutions, \(\tau \circ \upsilon\) is the successive application of \(\upsilon\) then \(\tau\).
 \end{definition}
@@
 
-We start with a high level description of the iterative monomorphisation algorithm.
-The algorithm takes a polymorphic problem as argument and returns a monomorphic problem.
+%We start with a high level description of the iterative monomorphisation algorithm.
+The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem.
 It operates by applying an arbitrary number of iterations.
@@
 
-Each iteration of the algorithm follows these steps:
-\begin{itemize}
+Each iteration follows these steps:
+\begin{enumerate}
 
-   \item The maps \(M\) and \(P\) are initialised with respect to \(\Phi\).
+   \item The maps \(M\) and \(P\) are initialised with respect to \(\Phi\) as per the collection phase.
 
@@
 
-   \item For each formula \(\varphi \in \Phi\), for each symbol \(f\) in \(\varphi\):
+   \item For each formula \(\varphi \in \Phi\), for each symbol \(f\) occuring in \(\varphi\):
 
-      An empty set of substitutions \(\Sigma_{\varphi, f}\) is created, then, for every type argument tuple \((\tau_1, \dots, \tau_n) \in  P(\varphi, f)\) and every type argument tuple \((\upsilon_1, \dots, \upsilon_n) \in M(f)\):
+      An empty set of substitutions \(\Sigma_{\varphi, f}\) is created, then for every tuple \((\tau_1, \dots, \tau_n) \in  P(\varphi, f)\) and tuple \((\upsilon_1, \dots, \upsilon_n) \in M(f)\):
 
-      Each \(\tau_i\) is matched against its corresponding \(\upsilon_i\) possibly yielding the substitution \(\sigma_i\).
+      \(\tau_i\) is matched against \(\upsilon_i\), possibly yielding the substitution \(\sigma_i\), for all \(i\).
 If all \(n\) matchings are successful and the substitutions are two by two compatible, \(\sigma_1 \circ \dots \circ \sigma_n\) is added to \(\Sigma_{\varphi, f}\).
 
-   \item For each formula \(\varphi \in \Phi\), for each symbol \(f\) in \(\varphi\), for each substitution \(\sigma \in \Sigma_{\varphi, f}\):
+   \item For each formula \(\varphi \in \Phi\), for each symbol \(f\) in \(\varphi\), for each substitution \(\sigma \in \Sigma_{\varphi, f}\), the formula \(\sigma(\varphi)\) is added to \(\Psi\).
 
-      \(\sigma(\varphi)\) is added to \(\Psi\).
+   \item \(\Phi \cup \Psi\) is returned.
 
-   \item The initial problem \(\Phi\) is replaced with \(\Phi \cup \Psi\).
+\end{enumerate}
 
-\end{itemize}
-
-After a set number of iterations, all monomorphic formulae in \(\Phi\) are output.
+An arbitrary number of iterations can be performed, re-using the output of one iteration as the input for the next.
+The final step requires all monomorphic formulae returned by the last iteration to be output.
 
@@
 
-Consider a problem made up of the following formulae:
+Consider the initial problem:
 \begin{enumerate}
@@
 
+During the first iteration, the matching of \(\alpha\) against \(\ty{int}\) will allow the generation an additionnal clause, resulting in this problem:
+
+\begin{enumerate}
+   \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
+   \item \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\)
+   \item \(f\langle \ty{int}\rangle(0)\)
+\end{enumerate}
+
+The second iteration will see \(\alpha\) matched against \(\ty{list}(\ty{int})\) leading to another clause being added to the problem:
+
+\begin{enumerate}
+   \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
+   \item \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\)
+   \item \(\forall a: \ty{list}(\ty{int}), b:\ty{list}(\ty{list}(\ty{int})), f\langle\ty{list}(\ty{int})\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{int}))\rangle(b)\)
+   \item \(f\langle \ty{int}\rangle(0)\)
+\end{enumerate}
+
+Similarly the third iteration yields:
+
+\begin{enumerate}
+   \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
+   \item \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\)
+   \item \(\forall a: \ty{list}(\ty{int}), b:\ty{list}(\ty{list}(\ty{int})), f\langle\ty{list}(\ty{int})\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{int}))\rangle(b)\)
+   \item \(\forall a: \ty{list}(\ty{list}(\ty{int})), b:\ty{list}(\ty{list}(\ty{list}(\ty{int}))), f\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(b)\)
+   \item \(f\langle \ty{int}\rangle(0)\)
+\end{enumerate}
+
 \begin{table}[ht]
-\caption{State of the algorithm}
-\centering\begin{tabular}{@{}lccccccc@{}}
+\caption{New formula}
+\centering\begin{tabular}{@{}ll@{}}
    \toprule
-               & Initialisation & Iteration 1 & Iteration 2 & Iteration 3\\
+   loop & new formula\\
    \midrule
-   \(M\)       & \((\ty{int})\) & - &\(+(\ty{list}(\ty{int}))\) & \(+(\ty{list}(\ty{list}(\ty{int})))\) \\
-   \(P(1, f)\) & \((\alpha)\; (\ty{list}(\alpha))\) & - & - & -\\
-   \(P(2, f)\) & \(\emptyset\) & - & - & - \\
-   \(\Phi\)    & [1] \; [2] &\(+\sigma(\text{[1]})\)&\(+\sigma'(\text{[1]})\)& \(+\sigma''(\text{[1]})\)\\
+   1 -& \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\) \\
+   2 -& \(\forall a: \ty{list}(\ty{int}), b:\ty{list}(\ty{list}(\ty{int})), f\langle\ty{list}(\ty{int})\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{int}))\rangle(b)\) \\
+   3 -& \(\forall a: \ty{list}(\ty{list}(\ty{int})), b:\ty{list}(\ty{list}(\ty{list}(\ty{int}))), f\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(b)\) \\
    \bottomrule
@@
 \end{table}
-
-On the first iteration, two matchings are attempted, that of \(\alpha\) against \(\ty{int}\) which yields \(\sigma = \{\alpha \mapsto \ty{int}\}\) and that of \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails.
+%
+%
+%On the first iteration, two matchings are attempted, that of \(\alpha\) against \(\ty{int}\) which yields \(\sigma = \{\alpha \mapsto \ty{int}\}\) and that of \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails.
 Then, \(\sigma\) is applied to [1], resulting in a new formula: \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\)
-
-At the beginning of the second iteration, the clause generated in the previous iteration is used to update \(M(f)\), adding \((\ty{list}(\ty{int}))\).
+%
+%At the beginning of the second iteration, the clause generated in the previous iteration is used to update \(M(f)\), adding \((\ty{list}(\ty{int}))\).
 \(P\) has no new elements.
 Similarly to the first iteration we attempt the following matchings:
-\begin{itemize}
-   \item \(\alpha\) against \(\ty{int}\) which succeeds resulting in \(\sigma\)
-   \item \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails
-   \item \(\alpha\) against \(\ty{list}(\ty{int})\) resulting in \(\sigma'=\{\alpha\mapsto\ty{list}(\ty{int})\}\)
-   \item \(\ty{list}(\alpha)\) against \(\ty{list}(\ty{int})\) resulting in \(\sigma\)
-\end{itemize}
-
-From these substitutions, a single new formula can be generated from the application of \(\sigma'\) to [1], giving us: \(\sigma'(\text{[1]}) = \forall a: \ty{list}(\ty{int}), b:\ty{list}(\ty{list}(\ty{int})), f\langle\ty{list}(\ty{int})\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{int}))\rangle(b)\)
-
-This new formula will be used to add \((\ty{list}(\ty{list}(\ty{int})))\) to \(M(f)\) at the beginning of the third iteration.
+%\begin{itemize}
+%   \item \(\alpha\) against \(\ty{int}\) which succeeds resulting in \(\sigma\)
+%   \item \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails
+%   \item \(\alpha\) against \(\ty{list}(\ty{int})\) resulting in \(\sigma'=\{\alpha\mapsto\ty{list}(\ty{int})\}\)
+%   \item \(\ty{list}(\alpha)\) against \(\ty{list}(\ty{int})\) resulting in \(\sigma\)
+%\end{itemize}
+%
+%From these substitutions, a single new formula can be generated from the application of \(\sigma'\) to [1], giving us: \(\sigma'(\text{[1]}) = \forall a: \ty{list}(\ty{int}), b:\ty{list}(\ty{list}(\ty{int})), f\langle\ty{list}(\ty{int})\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{int}))\rangle(b)\)
+%
+%This new formula will be used to add \((\ty{list}(\ty{list}(\ty{int})))\) to \(M(f)\) at the beginning of the third iteration.
 Similarly to the previous iteration, a single new formula will be generated.
@@
 This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
-It also shows that the high-level description is unsuitable for direct implementation.
-Indeed, many operations, from matchings to substitution applications, are unnecessarily repeated.
-Additionally, in practice input problems will be far larger and more explosive, any reasonable implementation will require bounds to be useful.
+Additionally, input problems are far larger and more explosive in practice and any reasonable implementation requires bounds to be useful.
 

### 2024-02-21 iterative_monomorphisation.tex
continued refining high level description

@@
 It is important to parametrise \(P\) with \(\varphi\) because type variables are quantified at the formula level.
-The formula indicates the scope of type variables.
+The formula indicates the scope of [the?] type variables.
 This is not necessary for \(M\) since all the types it contains are monomorphic.
@@
 
-   \item For each formula \(\varphi \in \Phi\), for each symbol \(f\) occuring in \(\varphi\):
+   \item For each formula \(\varphi \in \Phi\) and for each symbol \(f\) occuring in \(\varphi\):
 
-      An empty set of substitutions \(\Sigma_{\varphi, f}\) is created, then for every tuple \((\tau_1, \dots, \tau_n) \in  P(\varphi, f)\) and tuple \((\upsilon_1, \dots, \upsilon_n) \in M(f)\):
+      An empty set of substitutions \(\Sigma_{\varphi, f}\) is created, then for each tuple \((\tau_1, \dots, \tau_n) \in  P(\varphi, f)\) and \((\upsilon_1, \dots, \upsilon_n) \in M(f)\):
 
-      \(\tau_i\) is matched against \(\upsilon_i\), possibly yielding the substitution \(\sigma_i\), for all \(i\).
+      For all \(i\), we match \(\tau_i\) against \(\upsilon_i\), possibly yielding the substitution \(\sigma_i\).
 If all \(n\) matchings are successful and the substitutions are two by two compatible, \(\sigma_1 \circ \dots \circ \sigma_n\) is added to \(\Sigma_{\varphi, f}\).
 
-   \item For each formula \(\varphi \in \Phi\), for each symbol \(f\) in \(\varphi\), for each substitution \(\sigma \in \Sigma_{\varphi, f}\), the formula \(\sigma(\varphi)\) is added to \(\Psi\).
+   \item For each formula \(\varphi \in \Phi\), for each symbol \(f\) in \(\varphi\) and for each substitution \(\sigma \in \Sigma_{\varphi, f}\), the formula \(\sigma(\varphi)\) is added to \(\Psi\).
 
@@
 
-An arbitrary number of iterations can be performed, re-using the output of one iteration as the input for the next.
-The final step requires all monomorphic formulae returned by the last iteration to be output.
+An arbitrary number of iterations can be executed, using the output of one iteration as the input for the next.
+The final step requires filtering all monomorphic formulae returned by the last iteration.
 
-This algorithm is clearly sound as the newly derived formulae are instances of the initial formulae where universally quantified type variables have been instantiated with monomorphic types.
+This algorithm is clearly sound as the newly generated formulae are instances of the initial problem's formulae where universally quantified type variables have been instantiated with monomorphic types.
 
@@
 \begin{enumerate}
+   \item \(f\langle \ty{int}\rangle(0)\)
    \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
-   \item \(f\langle \ty{int}\rangle(0)\)
 \end{enumerate}
@@
 \begin{enumerate}
+   \item \(f\langle \ty{int}\rangle(0)\)
    \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
    \item \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\)
-   \item \(f\langle \ty{int}\rangle(0)\)
 \end{enumerate}
@@
 \begin{enumerate}
+   \item \(f\langle \ty{int}\rangle(0)\)
    \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
@@
    \item \(\forall a: \ty{list}(\ty{int}), b:\ty{list}(\ty{list}(\ty{int})), f\langle\ty{list}(\ty{int})\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{int}))\rangle(b)\)
-   \item \(f\langle \ty{int}\rangle(0)\)
 \end{enumerate}
@@
    \item \(\forall a: \ty{list}(\ty{list}(\ty{int})), b:\ty{list}(\ty{list}(\ty{list}(\ty{int}))), f\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(b)\)
-   \item \(f\langle \ty{int}\rangle(0)\)
 \end{enumerate}
@@
 This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
-Additionally, input problems are far larger and more explosive in practice and any reasonable implementation requires bounds to be useful.
+Input problems are far larger and more explosive in practice.
+Any reasonable implementation requires bounds limiting the number of new type arguments, substitutions and formulae to be useful.
 
@@
 This section presents a more detailed description of a practical adaptation of iterative monomorphisation, it closely follows the implementation in Zipperposition.
-
 The main departures from the high-level description of the algorithm are the following:
 \begin{itemize}
-   \item substitutions are directly applied to the type arguments, instead of the formulae, this also dispenses us from having to re-extract the type arguments from the clauses at each iteration
+   \item substitutions are applied to the type arguments, instead of the formulae. This dispenses us from having to extract type arguments from the clauses at each iteration
    \item new clauses are generated only once all substitutions have been computed
-   \item bounds are used at almost all stages of the algorithm as the various enumerations are highly explosive and must be limited in a practical setting
-   \item type arguments are separated into old and new sets, thereby avoiding computing the same matchings for each iteration
-   \item once the new formulae are generated, their monomorphic types are mangled
+   \item bounds are used at several stages of the algorithm as a means to curb the highly explosive enumerations
+   \item type arguments are separated into old and new sets, thereby avoiding computing the same matchings at each iteration
+   \item once the new formulae are generated, their monomorphic types are mangled before the end result is output
 \end{itemize}
 
-We assume that the maps \(P\) and \(M\) have been collected from the initial problem. This step is dependent on the precise form of the formulae which has been left undefined.
+We assume that the maps \(P\) and \(M\) have been collected from the initial problem. This step depends on the precise form of the input formulae which has been left undefined.
 

### 2024-02-21 iterative_monomorphisation.tex
more

@@
 They are assumed to be lifted to formulae; thus, $\sigma(\varphi)$ yields the variant of $\varphi$ in which each $\alpha_i$ is replaced by $\tau_i$.
-
-Given \(\tau\) and \(\upsilon\) two substitutions, \(\tau \circ \upsilon\) is the successive application of \(\upsilon\) then \(\tau\).
+Given two substitutions \(\tau, \upsilon\), the successive application of \(\tau\) and \(\upsilon\) is denoted by \(\upsilon \circ \tau\).
 \end{definition}
@@
 A single iteration consists of a collection phase (Section~\ref{ssec:collection}) and a processing phase (Section~\ref{ssec:processing}).
+Once the iterations are completed, a final step filters out all non-monomorphic formulae returned by the last iteration.
 
@@
 It is important to parametrise \(P\) with \(\varphi\) because type variables are quantified at the formula level.
-The formula indicates the scope of [the?] type variables.
+The formula indicates the scope of type variables.
 This is not necessary for \(M\) since all the types it contains are monomorphic.
@@
 
-Each iteration follows these steps:
+Once the maps \(M\) and \(P\) are initialised, each iteration performs the following processing steps:
+
 \begin{enumerate}
 
-   \item The maps \(M\) and \(P\) are initialised with respect to \(\Phi\) as per the collection phase.
-
-   \item An empty set of formulae \(\Psi\) is created.
+   \item Create an empty set of formulae \(\Phi'\).
 
    \item For each formula \(\varphi \in \Phi\) and for each symbol \(f\) occuring in \(\varphi\):
+   \begin{enumerate}
+    \item[2.1.] For each tuple \((\tau_1, \dots, \tau_n) \in  M(f)\)        and each tuple \((\upsilon_1, \dots, \upsilon_n) \in P(\varphi, f)\),
+     for each \(i\), match \(\upsilon_i\) against \(\tau_i\), yielding the substitution \(\sigma_i\) in case of success.
 
-      An empty set of substitutions \(\Sigma_{\varphi, f}\) is created, then for each tuple \((\tau_1, \dots, \tau_n) \in  P(\varphi, f)\) and \((\upsilon_1, \dots, \upsilon_n) \in M(f)\):
+    \item[2.2.] If all \(n\) matchings are successful and the substitutions \(\sigma_i\) are pairwise compatible,
+add the formula \((\sigma_1 \circ \dots \circ \sigma_n)(\varphi)\) to \(\Phi'\).
+   \end{enumerate}
 
-      For all \(i\), we match \(\tau_i\) against \(\upsilon_i\), possibly yielding the substitution \(\sigma_i\).
-If all \(n\) matchings are successful and the substitutions are two by two compatible, \(\sigma_1 \circ \dots \circ \sigma_n\) is added to \(\Sigma_{\varphi, f}\).
-
-   \item For each formula \(\varphi \in \Phi\), for each symbol \(f\) in \(\varphi\) and for each substitution \(\sigma \in \Sigma_{\varphi, f}\), the formula \(\sigma(\varphi)\) is added to \(\Psi\).
-
-   \item \(\Phi \cup \Psi\) is returned.
+   \item Return \(\Phi \cup \Phi'\).
 
@@
 
-An arbitrary number of iterations can be executed, using the output of one iteration as the input for the next.
-The final step requires filtering all monomorphic formulae returned by the last iteration.
+An arbitrary number of iterations can be executed, using the output of one iteration as the input of the next.
 

### 2024-02-21 iterative_monomorphisation.tex
started work on section 4 notably revisiting the algorithm environment

@@
 \usepackage{multirow}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
+%\usepackage{algorithm}
+%\usepackage{algpseudocode}
+\usepackage[linesnumbered]{algorithm2e}
 \usepackage{amsmath}
@@
 
-This section presents a more detailed description of a practical adaptation of iterative monomorphisation, it closely follows the implementation in Zipperposition.
-The main departures from the high-level description of the algorithm are the following:
+This section presents a detailed description of a practical adaptation of the high-level algorithm, it closely follows the implementation in Zipperposition.
+There are several significant departures from the high-level algorithm:
 \begin{itemize}
-   \item substitutions are applied to the type arguments, instead of the formulae. This dispenses us from having to extract type arguments from the clauses at each iteration
-   \item new clauses are generated only once all substitutions have been computed
+   \item substitutions are applied to the type arguments, instead of the formulae. This dispenses from having to extract type arguments from the formulae each iteration
+   \item new formulae are generated only once all substitutions have been computed in a separate step
    \item bounds are used at several stages of the algorithm as a means to curb the highly explosive enumerations
-   \item type arguments are separated into old and new sets, thereby avoiding computing the same matchings at each iteration
+   \item type arguments are separated into old and new sets, avoiding the computation of the same matchings each iteration
    \item once the new formulae are generated, their monomorphic types are mangled before the end result is output
@@
 
-We assume that the maps \(P\) and \(M\) have been collected from the initial problem. This step depends on the precise form of the input formulae which has been left undefined.
+The maps \(P\) and \(M\) are assumed to have been initialised from the initial problem. This step depends on the precise form of the input formulae which has been left undefined.
 
-\begin{algorithm}[tbh]
-\begin{algorithmic}[1]
-\Function{Iteration}{Problem: P, U, I, S}
-   \ForAll {clause \(C \in P\)}
-      \ForAll {\(fun\_sym, ty\_args \in U(C)\)}
-      \ForAll {\( ty\_args' \in I(fun\_sym) \)}
+\begin{algorithm}[H]
+   \LinesNumbered
+   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
+   \SetKwData{Continue}{continue}\SetKwData{SubstRes}{substitution\_result}
+   \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
+   \SetKwFunction{Match}{match}
+   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
 
-            \State \(subst = match(ty\_args, ty\_arg')\)
+   \Input{\(\varphi\), \(P\), \(M\)}
+   \Output{\(S\)}
+   \BlankLine
 
-            \State \Comment the match may fail in which case we simply continue with the loop
+   \(S\leftarrow \emptyset\)\;
+   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
+      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
+         \(\sigma\leftarrow \{\}\)\;
+         \Continue \(\leftarrow\) \True\;
+         \(i \leftarrow 0\)\;
+         \While{\(i \leq n\) \And \Continue}{
+            \SubstRes \(\leftarrow\) \Match{\(\upsilon_i\), \(\tau_i\)}\;
+            \lIf{\SubstRes = \Success{\(\sigma_i\)} \And \Compatible{\(\sigma\), \(\sigma_i\)}}{
+               \(\sigma\leftarrow\sigma_i\circ\sigma\) 
+            }
+            \lElse{\Continue\(\leftarrow\)\False}
+            \(i\leftarrow i+1\)\;
+         }
+         \lIf{\Continue}{\(S\leftarrow S\cup \{\sigma\}\)}
+      }
 
-            \State \(S(C) = S(C) \cup \{subst\}\)
-         \EndFor
-      \EndFor
+   }
+   \Return \(S\)\;
 
-   \EndFor
+\caption{Substitution Generation}
+\end{algorithm}
 
-   \ForAll {clause \(C \in P\)}
-      \ForAll {\(subst \in S(C)\)}
-         \ForAll {\(fun\_symbol, ty\_args \in U(C)\)}
-            \State \(ty\_args' = apply(subst, ty\_args)\)
-
-            \If {\( \forall ty\_arg \in ty\_args'. is\_ground(ty\_arg)\) }
-
-               \State \(I(fun\_sym) = I(fun\_sym) \cup \{ty\_args'\}\)
-            \Else
-
-               \State \(U(C, fun\_sym) = U(C, fun\_sym) \cup \{ty\_args'\}\)
-            \EndIf
-
-         \EndFor
-      \EndFor
-   \EndFor
-
-   \State \Return (I, U, S)
-
-\EndFunction
-\end{algorithmic}
-\end{algorithm}
-In the Zipperposition implemntation this iteration procedure is limited by several bounds.
+%\begin{algorithm}[]
+%\begin{algorithmic}[1]
+%\Function{Substitution Generation}{Input: \(\varphi\), M, P}
+%   \State new\_substitutions = \(\emptyset\)
+%   \ForAll {\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}
+%      \ForAll {\( (\tau_1, \dots,\tau_n) \in M(f) \)}
+%         \State \(\sigma\) = \{\}
+%         \State continue = true
+%         \While {\(i\leq n\) and continue}
+%            \State subtitution\_result = match(\(\upsilon_i\), \(\tau_i\))
+%            \If {substitution\_result = Success(\(\sigma_i\)) and compatible(\(\sigma_i\), \(\sigma\))}
+%               \State \(\sigma = \sigma \circ \sigma_i\)
+%            \Else
+%               \State continue = false
+%            \EndIf
+%
+%         \EndWhile
+%
+%      \EndFor
+%   \EndFor
+%
+%   \State \Return (I, U, S)
+%
+%\EndFunction
+%\end{algorithmic}
+%\end{algorithm}
+%In the Zipperposition implemntation this iteration procedure is limited by several bounds.
 The number of new type arguments that can be generated for each symbol is limited by user-defined bounds.
@@
 In both cases, the bounds for the new ground type arguments and non-ground type arguments can be set independently.
-
-A consequences of these bounds is that in order to avoid unecessary computations, the program will iteratively apply substitutions until either the ground or non-ground bound is reached.
+%
+%A consequences of these bounds is that in order to avoid unecessary computations, the program will iteratively apply substitutions until either the ground or non-ground bound is reached.
 Then substitutions can be discarded depending on whether they instantiate a given set of type arguments or not.
-
-Additionally, there are bounds for the number of substitutions that are generated.
-
-\begin{algorithm}[tbh]
-\begin{algorithmic}[1]
-\Function{Instantiate clause}{Clause: C, Substitutions S}
-
-   \If{\(C\) is ground}
-
-      \State \Return \(\{C\}\)
-
-   \Else
-
-      \State let \(v\) be a type variable in \(C\)
-      \State let new\_clauses be an empty set of clauses
-
-      \ForAll{subst \(\in S(C)\)}
-
-         \If{subst instantiates \(v\)}
-
-            \State \(C' = apply(subst, C)\)
-
-            \State let v\_clauses = Instantiate clause(C', S)
-
-         \EndIf
-
-         \State \( new\_clauses = new\_clauses \cup v\_clauses\)
-
-      \EndFor
-
-      \State \Return new\_clauses
-   \EndIf
-
-\EndFunction
-\end{algorithmic}
-\end{algorithm}
-This function differs from the implementation in a significant way, indeed, through experimentation it was found that the generated clauses were far more likely to be solved by E if the substitutions that are combined to instantiate the clause were derived from the same iteration.
+%
+%Additionally, there are bounds for the number of substitutions that are generated.
+%
+%\begin{algorithm}[tbh]
+%\begin{algorithmic}[1]
+%\Function{Instantiate clause}{Clause: C, Substitutions S}
+%
+%   \If{\(C\) is ground}
+%
+%      \State \Return \(\{C\}\)
+%
+%   \Else
+%
+%      \State let \(v\) be a type variable in \(C\)
+%      \State let new\_clauses be an empty set of clauses
+%
+%      \ForAll{subst \(\in S(C)\)}
+%
+%         \If{subst instantiates \(v\)}
+%
+%            \State \(C' = apply(subst, C)\)
+%
+%            \State let v\_clauses = Instantiate clause(C', S)
+%
+%         \EndIf
+%
+%         \State \( new\_clauses = new\_clauses \cup v\_clauses\)
+%
+%      \EndFor
+%
+%      \State \Return new\_clauses
+%   \EndIf
+%
+%\EndFunction
+%\end{algorithmic}
+%\end{algorithm}
+%This function differs from the implementation in a significant way, indeed, through experimentation it was found that the generated clauses were far more likely to be solved by E if the substitutions that are combined to instantiate the clause were derived from the same iteration.
 In practice, we therefore store the iteration at which each substition is generated and at the instantiation phase only apply substitutions from the same iteration.
-
-What the previous function does is essentially going through the tree of possible substitutions that instantiate the clause and collecting all possible instantiations.
+%
+%
+%What the previous function does is essentially going through the tree of possible substitutions that instantiate the clause and collecting all possible instantiations.
 This is extremely explosive, and a hard cap on the number of clauses we want to generate is almost indispensable for any implementation.
-
-\begin{algorithm}[tbh]
-\begin{algorithmic}[1]
-\Function{Generate clauses}{Problem: P, Substitutions: S}
-
-\ForAll {clause \(C \in P\)}
-   \State let new\_clauses = Instantiate clause(C, S)
-   \State \(P = P \cup new\_clauses\)
-\EndFor
-
-\State \(P' = \emptyset \)
-
-\ForAll {clause \(C \in P\)}
-   \If{\(C\) is ground}
-      \State \(P' = P' \cup \{mangle(C)\}\)
-   \EndIf
-\EndFor
-
-\State \Return \(P'\)
-
-\EndFunction
-\end{algorithmic}
-\end{algorithm}
-
-Implementation in Zipperposition.
+%
+%\begin{algorithm}[tbh]
+%\begin{algorithmic}[1]
+%\Function{Generate clauses}{Problem: P, Substitutions: S}
+%
+%\ForAll {clause \(C \in P\)}
+%   \State let new\_clauses = Instantiate clause(C, S)
+%   \State \(P = P \cup new\_clauses\)
+%\EndFor
+%
+%\State \(P' = \emptyset \)
+%
+%\ForAll {clause \(C \in P\)}
+%   \If{\(C\) is ground}
+%      \State \(P' = P' \cup \{mangle(C)\}\)
+%   \EndIf
+%\EndFor
+%
+%\State \Return \(P'\)
+%
+%\EndFunction
+%\end{algorithmic}
+%\end{algorithm}
+%
+%Implementation in Zipperposition.
 

### 2024-02-21 iterative_monomorphisation.tex
more

@@
 \newdefinition{definition}{Definition}
+\newdefinition{example}[definition]{Example}
 
@@
 It relies exclusively on the formulae's monomorphic and polymorphic symbol instances.
-Type variables are assumed to be universally quantified at a formula's top level.
+Type variables are assumed to be implicitly universally quantified at a formula's top level.
 The precise form of formulae is left unspecified.
@@
 Each iteration takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae.
-A single iteration consists of a collection phase (Section~\ref{ssec:collection}) and a processing phase (Section~\ref{ssec:processing}).
+A single iteration consists of a collection phase
+% (Section~\ref{ssec:collection})
+and an instantiation phase.
+% (Section~\ref{ssec:instantiation})
 Once the iterations are completed, a final step filters out all non-monomorphic formulae returned by the last iteration.
 
-\subsection{Collection}
-\label{ssec:collection}
+%\subsection{Collection}
+%\label{ssec:collection}
 
-The initial phase of each iteration consists of computing two maps, \(M\) and \(P\), from the input problem~$\Phi$:
-
-\begin{itemize}
-\item Given a symbol \(f\) occurring in \(\Phi\), the set \(M(f)\) consists of all tuples of monomorphic type arguments to which \(f\) is applied in \(\Phi\).
+The initial phase of each iteration consists of computing two maps, \(M\) and \(P\), from the input problem~$\Phi$.
+%
+\begin{enumerate}
+\item[\labelitemi] Given a symbol \(f\) occurring in \(\Phi\), the set \(M(f)\) consists of all tuples of monomorphic type arguments to which \(f\) is applied in \(\Phi\).
 For example, if \(\sym{foldl}\langle \ty{nat}, \ty{int}\rangle\) occurs in \(\Phi\), then \((\ty{nat}, \ty{int}) \in M(\sym{foldl}) \).
 
-\item Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(P(\varphi, f)\) consists of all tuples of type arguments to which \(f\) is applied in \(\varphi\) and which contains a type variable.
+\item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(P(\varphi, f)\) consists of all tuples of type arguments to which \(f\) is applied in \(\varphi\) and which contains a type variable.
 For example, if \(\sym{foldl}\langle \ty{nat}, \ty{list}(\alpha)\rangle\) occurs in \(\varphi\), then \((\ty{nat}, \ty{list}(\alpha)) \in P(\varphi, \sym{foldl}) \).
-\end{itemize}
+\end{enumerate}
 
@@
 
-\subsection{Processing}
-\label{ssec:processing}
+%\subsection{Instantiation}
+%\label{ssec:instantiation}
 
-Once the maps \(M\) and \(P\) are initialised, each iteration performs the following processing steps:
+Once the maps \(M\) and \(P\) are initialised, each iteration performs the following steps to create new instances of formulae:
 
@@
 
-An arbitrary number of iterations can be executed, using the output of one iteration as the input of the next.
+The algorithm is clearly sound because the newly generated formulae are instances of the initial problem's formulae, where type variables have been instantiated with monomorphic types.
+It is, however, not guaranteed to be complete.
 
-This algorithm is clearly sound as the newly generated formulae are instances of the initial problem's formulae where universally quantified type variables have been instantiated with monomorphic types.
+%\subsection{Example}
 
-\subsection{Example}
-
-Consider the initial problem:
+\begin{example}Consider the following problem:
 \begin{enumerate}
-   \item \(f\langle \ty{int}\rangle(0)\)
-   \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
+   \item \(\sym{p}\langle \ty{int}\rangle(0)\)
+   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){,}\; \sym{p}\langle\alpha\rangle(a) \land \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
+\end{enumerate}
+%
+The first iteration matches \(\alpha\) against \(\ty{int}\) for $\sym{p}$, generating the formula
+%
+\begin{enumerate}
+   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){,}\; \sym{p}\langle\ty{int}\rangle(a) \land \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
+\end{enumerate}
+%
+The second iteration matches \(\alpha\) against \(\ty{list}(\ty{int})\), leading to the formula
+%
+\begin{enumerate}
+   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){,}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \land \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
+\end{enumerate}
+%
+Similarly the third iteration adds
+%
+\begin{enumerate}
+   \item[5.] \(\forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,}\; \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \land \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
 \end{enumerate}
 
-During the first iteration, the matching of \(\alpha\) against \(\ty{int}\) will allow the generation an additionnal clause, resulting in this problem:
-
-\begin{enumerate}
-   \item \(f\langle \ty{int}\rangle(0)\)
-   \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
-   \item \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\)
-\end{enumerate}
-
-The second iteration will see \(\alpha\) matched against \(\ty{list}(\ty{int})\) leading to another clause being added to the problem:
-
-\begin{enumerate}
-   \item \(f\langle \ty{int}\rangle(0)\)
-   \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
-   \item \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\)
-   \item \(\forall a: \ty{list}(\ty{int}), b:\ty{list}(\ty{list}(\ty{int})), f\langle\ty{list}(\ty{int})\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{int}))\rangle(b)\)
-\end{enumerate}
-
-Similarly the third iteration yields:
-
-\begin{enumerate}
-   \item \(f\langle \ty{int}\rangle(0)\)
-   \item \(\forall a: \alpha, b:\ty{list}(\alpha), f\langle\alpha\rangle(a) \land f\langle \ty{list}(\alpha)\rangle(b)\)
-   \item \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\)
-   \item \(\forall a: \ty{list}(\ty{int}), b:\ty{list}(\ty{list}(\ty{int})), f\langle\ty{list}(\ty{int})\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{int}))\rangle(b)\)
-   \item \(\forall a: \ty{list}(\ty{list}(\ty{int})), b:\ty{list}(\ty{list}(\ty{list}(\ty{int}))), f\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(b)\)
-\end{enumerate}
-
-\begin{table}[ht]
-\caption{New formula}
-\centering\begin{tabular}{@{}ll@{}}
-   \toprule
-   loop & new formula\\
-   \midrule
-   1 -& \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\) \\
-   2 -& \(\forall a: \ty{list}(\ty{int}), b:\ty{list}(\ty{list}(\ty{int})), f\langle\ty{list}(\ty{int})\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{int}))\rangle(b)\) \\
-   3 -& \(\forall a: \ty{list}(\ty{list}(\ty{int})), b:\ty{list}(\ty{list}(\ty{list}(\ty{int}))), f\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(b)\) \\
-   \bottomrule
-\end{tabular}
-\end{table}
 %
@@
 %On the first iteration, two matchings are attempted, that of \(\alpha\) against \(\ty{int}\) which yields \(\sigma = \{\alpha \mapsto \ty{int}\}\) and that of \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails.
-Then, \(\sigma\) is applied to [1], resulting in a new formula: \(\forall a: \ty{int}, b:\ty{list}(\ty{int}), f\langle\ty{int}\rangle(a) \land f\langle \ty{list}(\ty{int})\rangle(b)\)
+Then, \(\sigma\) is applied to [1], resulting in a new formula: \(\forall a: \ty{int}, \mathit{as}:\ty{list}(\ty{int}), \sym{p}\langle\ty{int}\rangle(a) \land \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
 %
@@
 %
-%From these substitutions, a single new formula can be generated from the application of \(\sigma'\) to [1], giving us: \(\sigma'(\text{[1]}) = \forall a: \ty{list}(\ty{int}), b:\ty{list}(\ty{list}(\ty{int})), f\langle\ty{list}(\ty{int})\rangle(a) \land f\langle \ty{list}(\ty{list}(\ty{int}))\rangle(b)\)
+%From these substitutions, a single new formula can be generated from the application of \(\sigma'\) to [1], giving us: \(\sigma'(\text{[1]}) = \forall a: \ty{list}(\ty{int}), \mathit{as}:\ty{list}(\ty{list}(\ty{int})), \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \land \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
 %
@@
 It is obtained by applying the new substitution \(\sigma'' = \{\alpha\mapsto\ty{list}(\ty{list}(\ty{int}))\}\) to [1].
+\end{example}
 
 This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
-Input problems are far larger and more explosive in practice.
+Although the example may seem artificial, similar problems frequently arise in practice.
+For example, the \sym{concat} function of Isabelle/HOL \cite{xxx}, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
 Any reasonable implementation requires bounds limiting the number of new type arguments, substitutions and formulae to be useful.
@@
 
+  * explicitly quantified formulas a la TPTP, existentials, how to eliminate
+
   * Böhme

### 2024-02-21 iterative_monomorphisation.tex
more

@@
    \item \(\sym{p}\langle \ty{int}\rangle(0)\)
-   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){,}\; \sym{p}\langle\alpha\rangle(a) \land \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
+   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){,}\; \sym{p}\langle\alpha\rangle(a) \longrightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
 \end{enumerate}
@@
 \begin{enumerate}
-   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){,}\; \sym{p}\langle\ty{int}\rangle(a) \land \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
+   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){,}\; \sym{p}\langle\ty{int}\rangle(a) \longrightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
 \end{enumerate}
@@
 \begin{enumerate}
-   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){,}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \land \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
+   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){,}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \longrightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
 \end{enumerate}
@@
 \begin{enumerate}
-   \item[5.] \(\forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,}\; \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \land \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
+   \item[5.] \(\forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,}\; \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \longrightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
 \end{enumerate}
@@
 %On the first iteration, two matchings are attempted, that of \(\alpha\) against \(\ty{int}\) which yields \(\sigma = \{\alpha \mapsto \ty{int}\}\) and that of \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails.
-Then, \(\sigma\) is applied to [1], resulting in a new formula: \(\forall a: \ty{int}, \mathit{as}:\ty{list}(\ty{int}), \sym{p}\langle\ty{int}\rangle(a) \land \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
+Then, \(\sigma\) is applied to [1], resulting in a new formula: \(\forall a: \ty{int}, \mathit{as}:\ty{list}(\ty{int}), \sym{p}\langle\ty{int}\rangle(a) \longrightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
 %
@@
 %
-%From these substitutions, a single new formula can be generated from the application of \(\sigma'\) to [1], giving us: \(\sigma'(\text{[1]}) = \forall a: \ty{list}(\ty{int}), \mathit{as}:\ty{list}(\ty{list}(\ty{int})), \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \land \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
+%From these substitutions, a single new formula can be generated from the application of \(\sigma'\) to [1], giving us: \(\sigma'(\text{[1]}) = \forall a: \ty{list}(\ty{int}), \mathit{as}:\ty{list}(\ty{list}(\ty{int})), \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \longrightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
 %

### 2024-02-21 iterative_monomorphisation.tex
more

--- 2024-02-21 iterative_monomorphisation.tex
@@
 %\usepackage{algpseudocode}
-\usepackage[linesnumbered]{algorithm2e}
+\usepackage%[linesnumbered]
+  {algorithm2e}
 \usepackage{amsmath}
@@
 \newcommand\var[1]{\mathit{#1}}
+
+%% TYPESETTING: hacks
+\newcommand\medrightarrow{\mathrel{{{\color{black}\relbar}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex}\joinrel{\color{black}\rightarrow}}}
+\newcommand\medleftarrow{\mathrel{{\color{black}\leftarrow}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex\joinrel{{\color{black}\relbar}}}}
+\newcommand\medleftrightarrow{\mathrel{\leftarrow\kern-1.685ex\rightarrow}}
+\newcommand\Medrightarrow{\mathrel{{{\color{black}\Relbar}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex}\joinrel{\color{black}\Rightarrow}}}
+\newcommand\Medleftrightarrow{\mathrel{\Leftarrow\kern-1.685ex\Rightarrow}}
 
@@
 %   - monomorphization -> monomorphisation
-%   - higer order -> higer-order
+%   - higher-order -> higher order
 %   - TH1 with ugly 1
@@
 
-In this paper, we present an algorithm based on our understanding of his description and implementation (Section~\ref{sec:high-level-algorithm}).
+In this paper, we present an algorithm based on our understanding of his description and implementation (Section~\ref{sec:high level-algorithm}).
 We also provide a more detailed description to help future implementers.
@@
 %  * In this paper, we present an algorithm inspired by his description and
-%    implementation (Section~\ref{sec:high-level-algorithm}).
+%    implementation (Section~\ref{sec:high level-algorithm}).
 %    * we aimed at mathematical precision, so that people who want to reimplement
@@
 Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic.
-In particular, it can operate in the polymorphic first- and higher-order logics embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.
+In particular, it can operate in the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.
 
@@
 %
-%  * in particular, it works with polymorphic first- and higher-order logic as 
+%  * in particular, it works with polymorphic first and higher order logic as 
 %    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}
@@
 \section{High level algorithm}
-\label{sec:high-level-algorithm}
+\label{sec:high level-algorithm}
 
@@
 
-It is important to parametrise \(P\) with \(\varphi\) because type variables are quantified at the formula level.
+These definitions depend on the precise form of the input formulae, which has been left undefined.
+It is important to parametrise \(P\) with \(\varphi\) because type variables are implicitly quantified at the formula level.
 The formula indicates the scope of type variables.
@@
    \item \(\sym{p}\langle \ty{int}\rangle(0)\)
-   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){,}\; \sym{p}\langle\alpha\rangle(a) \land \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
+   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){,}\; \sym{p}\langle\alpha\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
 \end{enumerate}
@@
 \begin{enumerate}
-   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){,}\; \sym{p}\langle\ty{int}\rangle(a) \land \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
+   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){,}\; \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
 \end{enumerate}
@@
 \begin{enumerate}
-   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){,}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \land \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
+   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){,}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
 \end{enumerate}
@@
 \begin{enumerate}
-   \item[5.] \(\forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,}\; \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \land \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
+   \item[5.] \(\forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,}\; \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
 \end{enumerate}
@@
 %On the first iteration, two matchings are attempted, that of \(\alpha\) against \(\ty{int}\) which yields \(\sigma = \{\alpha \mapsto \ty{int}\}\) and that of \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails.
-Then, \(\sigma\) is applied to [1], resulting in a new formula: \(\forall a: \ty{int}, \mathit{as}:\ty{list}(\ty{int}), \sym{p}\langle\ty{int}\rangle(a) \land \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
+Then, \(\sigma\) is applied to [1], resulting in a new formula: \(\forall a: \ty{int}, \mathit{as}:\ty{list}(\ty{int}), \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
 %
@@
 %
-%From these substitutions, a single new formula can be generated from the application of \(\sigma'\) to [1], giving us: \(\sigma'(\text{[1]}) = \forall a: \ty{list}(\ty{int}), \mathit{as}:\ty{list}(\ty{list}(\ty{int})), \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \land \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
+%From these substitutions, a single new formula can be generated from the application of \(\sigma'\) to [1], giving us: \(\sigma'(\text{[1]}) = \forall a: \ty{list}(\ty{int}), \mathit{as}:\ty{list}(\ty{list}(\ty{int})), \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
 %
@@
 Although the example may seem artificial, similar problems frequently arise in practice.
-For example, the \sym{concat} function of Isabelle/HOL \cite{xxx}, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
+For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
 Any reasonable implementation requires bounds limiting the number of new type arguments, substitutions and formulae to be useful.
@@
 
-This section presents a detailed description of a practical adaptation of the high-level algorithm, it closely follows the implementation in Zipperposition.
-There are several significant departures from the high-level algorithm:
-\begin{itemize}
-   \item substitutions are applied to the type arguments, instead of the formulae. This dispenses from having to extract type arguments from the formulae each iteration
-   \item new formulae are generated only once all substitutions have been computed in a separate step
-   \item bounds are used at several stages of the algorithm as a means to curb the highly explosive enumerations
-   \item type arguments are separated into old and new sets, avoiding the computation of the same matchings each iteration
-   \item once the new formulae are generated, their monomorphic types are mangled before the end result is output
-\end{itemize}
+The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a more low level algorithm with the following features.
+%
+\begin{enumerate}
+    \item[\labelitemi] Bounds are used at several stages of the algorithm to curb the highly explosive enumerations.
 
-The maps \(P\) and \(M\) are assumed to have been initialised from the initial problem. This step depends on the precise form of the input formulae which has been left undefined.
+    \item[\labelitemi] Type arguments are separated into old and new sets to avoid re-computing some of the same matchings in successive iterations.
 
+    \item[\labelitemi] Substitutions are directly applied to the type arguments in $M$ and $P$ instead of the formulae.
+This dispenses from having to extract the type arguments from the formulae at each iteration.
+New formulae are generated only once all iterations are complete, in a separate, penultimate step.
+
+    \item[\labelitemi] Once the new formulae are generated, their monomorphic types are mangled before the end result is output.
+\end{enumerate}
+
+\begin{quote}
 \begin{algorithm}[H]
-   \LinesNumbered
+   \SetAlgoNoLine
+%   \LinesNumbered
    \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
@@
 
-   \Input{\(\varphi\), \(P\), \(M\)}
-   \Output{\(S\)}
+   \Input{\,\(\varphi\), \(M\), \(P\)}
+   \Output{\,a set of substitutions}
    \BlankLine
@@
    \Return \(S\)\;
-
-\caption{Substitution Generation}
 \end{algorithm}
+\end{quote}
 

### 2024-02-21 iterative_monomorphisation.tex
couple of small changes

@@
 
+\definecolor{grey}{rgb}{0.8,0.8,0.8}
+\definecolor{algoColorKeyword}{named}{grey}
+\makeatletter
+% Block with a vertical line
+\renewcommand{\algocf@Vsline}[1]{%
+   \strut\par\nointerlineskip%
+   \algocf@push{\skiprule}%
+   \hbox{%
+      {\color{algoColorKeyword}\vrule}
+      \vtop{\algocf@push{\skiptext}%
+      \vtop{\algocf@addskiptotal\advance\hsize by -\skiplength #1}}%
+   }%
+   \algocf@pop{\skiprule}%
+}
+
 \begin{document}
@@
 %   - title of bits of the algorithms
+%   - TH1 ugly for height of 1 reasons, can be fixed with case feature
 %   - eg or ex -> e.g.
@@
 
-  % apps are rank-1 polymorphic, but provers are mono
-  % there's a gap
-  % one solution is to extend the provers to support polymorphism;
-  % this was done e.g. for Vampire \cite{xxx}
-  % but this is a lot of work, and needs to be done for every prover
-
 The alternative is to translate polymorphic problems to monomorphic problems.
@@
 
- % * the alternative is to translate polymorphic problems to monomorphic problems
- %   * two main approaches:
- %     * complete encoding of polymorphism, using guards or tags,
- %       as in Blanchette et al.~\cite{mono-trans} and Bobot et al.~\cite{expr-poly-types}
- %     * incomplete encoding of polymorphism based on iterative monomorphisation
- %       such as introduced by Böhme \cite[Section 2.2.1]{sb-phd}, whereby type
- %       variables are heuristically instantiated by concrete types
-
 By a type version of the compactness theorem, we have that in first order logic, given a polymorphic formula \(\varphi\), there exist an equisatisfiable finite set of monomorphic instances of \(\varphi\).
@@
 bound to be incomplete.
-
-  %* By a type version of compactness, if a set of polymorphic
-  %  formulae is unsatisfiable, then there exists a finite set of monomorphic
-  %  instances of the formulae that is also unsatisfiable. However, this
-  %  finite set cannot be computed \cite[Section 2, Theorem 1]{expr-poly-types}; hence any approach based on
-  %  instantiation of type variables is incomplete.
 
@@
 
-% * Böhme's iterative approach is implemented as part of the SMT integration \cite[Chapter 2]{sb-phd}
-%     in Isabelle/HOL. This implementation is also used by
-%    Sledgehammer \cite{judgement, hammer} to interface automatic theorem provers based on
-%    superposition.
-%
-%  * The main drawback of Böhme's approach is that it is only briefly documented,
-%    as a single paragraph in his PhD thesis \cite[Section 2.2.1]{sb-phd}
-%  * In this paper, we present an algorithm inspired by his description and
-%    implementation (Section~\ref{sec:high level-algorithm}).
-%    * we aimed at mathematical precision, so that people who want to reimplement
-%      the algorithm can follow this blueprint
-%    * we also show how to bound the algorithm to avoid explosions in practice
-%      (Section~\ref{sec:low-level-algorithm}).
-%      ``the explosive nature of the procedures requires various bounds which
-%      limit the number of type arguments and clauses that are generated.''
-
 The algorithm works as follows.
@@
 
-%These iterations are required to account for $n$-ary type constructors.
 Consider the unary type constructor \ty{list}.
@@
 Our empirical evaluation on TPTP \cite{tptp} problems attempts to answer three questions (Section~\ref{sec:evaluation}):
-%
 \begin{enumerate}
@@
 Our findings are as follows:
-%
 \begin{enumerate}
@@
 The precise form of formulae is left unspecified.
-%
 Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic.
@@
 
-%The following terms are used in the algorithm descriptions.
-
-%{\obeylines
-%
-%\noindent\kern\parindent
-%  * algorithm works a priori with any rank-1 polymorphic logic
-%
-%  * in particular, it works with polymorphic first and higher order logic as 
-%    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}
-%
-%  * the algorithm does not need to inspect the precise structure of formulae
-%  * instead, it works on monomorphic or polymorphic instances of symbols that 
-%    occur in the formulae
-%  * type quantification is implicit, meaning that all type variables occurring
-%    in a formula are assumed to be universally quantified at the formula's top
-%    level
-%}
-
-% TH1 ugly for height of 1 reasons, can be fixed with case feature
-
 Our abstract framework relies on the following basic definitions.
 
-\begin{definition}%[Type]
+\begin{definition}
 A (\emph{polymorphic}) \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
@@
 
-\begin{definition}%[Monomorphism and polymorphism]
+\begin{definition}
 A type is \emph{monomorphic} if it contains no type variables.
@@
 
-\begin{definition}%[Function symbol]
+\begin{definition}
 A (function or predicate) {symbol} \(f\) has a \emph{type arity} that defines the number of type arguments it takes.
@@
 
-%We start with a high level description of the iterative monomorphisation algorithm.
 The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem.
@@
 Each iteration takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae.
-A single iteration consists of a collection phase
-% (Section~\ref{ssec:collection})
-and an instantiation phase.
-% (Section~\ref{ssec:instantiation})
+A single iteration consists of a collection phase and an instantiation phase.
 Once the iterations are completed, a final step filters out all non-monomorphic formulae returned by the last iteration.
-
-%\subsection{Collection}
-%\label{ssec:collection}
 
@@
 This is not necessary for \(M\) since all the types it contains are monomorphic.
-
-%\subsection{Instantiation}
-%\label{ssec:instantiation}
 
@@
 It is, however, not guaranteed to be complete.
-
-%\subsection{Example}
 
@@
 
-%
-%
-%On the first iteration, two matchings are attempted, that of \(\alpha\) against \(\ty{int}\) which yields \(\sigma = \{\alpha \mapsto \ty{int}\}\) and that of \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails.
-Then, \(\sigma\) is applied to [1], resulting in a new formula: \(\forall a: \ty{int}, \mathit{as}:\ty{list}(\ty{int}), \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
-%
-%At the beginning of the second iteration, the clause generated in the previous iteration is used to update \(M(f)\), adding \((\ty{list}(\ty{int}))\).
-\(P\) has no new elements.
-Similarly to the first iteration we attempt the following matchings:
-%\begin{itemize}
-%   \item \(\alpha\) against \(\ty{int}\) which succeeds resulting in \(\sigma\)
-%   \item \(\ty{list}(\alpha)\) against \(\ty{int}\) which fails
-%   \item \(\alpha\) against \(\ty{list}(\ty{int})\) resulting in \(\sigma'=\{\alpha\mapsto\ty{list}(\ty{int})\}\)
-%   \item \(\ty{list}(\alpha)\) against \(\ty{list}(\ty{int})\) resulting in \(\sigma\)
-%\end{itemize}
-%
-%From these substitutions, a single new formula can be generated from the application of \(\sigma'\) to [1], giving us: \(\sigma'(\text{[1]}) = \forall a: \ty{list}(\ty{int}), \mathit{as}:\ty{list}(\ty{list}(\ty{int})), \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
-%
-%This new formula will be used to add \((\ty{list}(\ty{list}(\ty{int})))\) to \(M(f)\) at the beginning of the third iteration.
-Similarly to the previous iteration, a single new formula will be generated.
-It is obtained by applying the new substitution \(\sigma'' = \{\alpha\mapsto\ty{list}(\ty{list}(\ty{int}))\}\) to [1].
 \end{example}
@@
 The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a more low level algorithm with the following features.
-%
 \begin{enumerate}
@@
 \begin{algorithm}[H]
-   \SetAlgoNoLine
+   \caption{Substitution Generation}
+%   \SetAlgoNoLine
 %   \LinesNumbered
@@
 
-   \Input{\,\(\varphi\), \(M\), \(P\)}
-   \Output{\,a set of substitutions}
+   \Input{\,formula \(\varphi\), monomorphic type argument map \(M\), non-monomorphic type argument map \(P\)}
+   \Output{\,set of substitutions}
    \BlankLine
@@
             \SubstRes \(\leftarrow\) \Match{\(\upsilon_i\), \(\tau_i\)}\;
-            \lIf{\SubstRes = \Success{\(\sigma_i\)} \And \Compatible{\(\sigma\), \(\sigma_i\)}}{
-               \(\sigma\leftarrow\sigma_i\circ\sigma\) 
+            \Case{\SubstRes = \Success{\(\sigma_i\)}}{
+               \lIf{\Compatible{\(\sigma\), \(\sigma_i\)}}{
+                  \(\sigma\leftarrow\sigma_i\circ\sigma\)}
+               \lElse{\Continue\(\leftarrow\)\False}
             }
-            \lElse{\Continue\(\leftarrow\)\False}
+            \lOther{\Continue\(\leftarrow\)\False}
+
+         }
          \(i\leftarrow i+1\)\;
-         }
+
          \lIf{\Continue}{\(S\leftarrow S\cup \{\sigma\}\)}

### 2024-02-28 iterative_monomorphisation.tex
clause monomorphisation step

@@
 %\usepackage{algpseudocode}
-\usepackage%[linesnumbered]
+\usepackage%[]
    {algorithm2e}
@@
    \strut\par\nointerlineskip%
+   \algocf@bblockcode%
    \algocf@push{\skiprule}%
-   \hbox{%
-      {\color{algoColorKeyword}\vrule}
+   \hbox{{\color{algoColorKeyword}\vrule}%
       \vtop{\algocf@push{\skiptext}%
@@
    \algocf@pop{\skiprule}%
+   \algocf@bblockcode%
 }
+\makeatother
 
@@
 
+TODO:
+\begin{itemize}
+   \item substitution generation
+   \item clause generation
+   \item fitting everything together
+\end{itemize}
+
 \begin{quote}
@@
 %   \LinesNumbered
-   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
+   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
    \SetKwData{Continue}{continue}\SetKwData{SubstRes}{substitution\_result}
@@
 
-   \Input{\,formula \(\varphi\), monomorphic type argument map \(M\), non-monomorphic type argument map \(P\)}
+   \Input{\,monomorphic type argument map \(M\), non-monomorphic type argument map \(P(\varphi)\)}
    \Output{\,set of substitutions}
@@
          \Continue \(\leftarrow\) \True\;
-         \(i \leftarrow 0\)\;
+         \(i \leftarrow 1\)\;
          \While{\(i \leq n\) \And \Continue}{
@@
 
+            \(i\leftarrow i+1\)\;
          }
-         \(i\leftarrow i+1\)\;
 
-         \lIf{\Continue}{\(S\leftarrow S\cup \{\sigma\}\)}
+         \If{\Continue}{
+            \(S\leftarrow S\cup \{\sigma\}\)\;
+         }
       }
@@
 
+Notes:
+   * In the implementation, the old and new maps are actually one map with function symbols pointing to pairs of sets of type arguments
+   * In the implementation, instead of tuples we have lists of types
+
+   * In order to later have an easier time we store susbtitutions in a map that gathers them by which type variable they instantiate.
+
+   * We assume the existence of "primitives" that compose substitutions, apply them to formulae ...
+
+\begin{quote}
+\begin{algorithm}[H]
+   \caption{Clause Monomorphisation Step}
+
+   \LinesNumbered
+
+   \SetKwData{Subst}{substitutions}\SetKwData{TypeVars}{type\_variables}
+   \SetKwData{MaxMono}{max\_mono\_args}
+   \SetKwData{MaxPoly}{max\_poly\_args}
+   \SetKwData{MonoCap}{mono\_cap}\SetKwData{MonoMult}{mono\_mult}\SetKwData{MonoFloor}{mono\_floor}
+   \SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
+   \SetKwData{UsedSubst}{used\_subst}
+
+   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
+
+   \SetKwFunction{SubstGen}{substitution\_generation}\SetKwFunction{TyVars}{type\_variables}
+   \SetKwFunction{Max}{max}\SetKwFunction{Min}{min}
+   \SetKwFunction{Domain}{subst\_domain}\SetKwFunction{Len}{len}
+   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
+
+   \Input{\,monomorphic type argument map \(M, M'\), new non-monomorphic type argument map \(P(\varphi), P'(\varphi)\)}
+   \Output{\,monomorphic type argument map, non-monomorphic type argument map, set of substitutions}
+   \BlankLine
+
+   \Subst \(\leftarrow\) \SubstGen{\(M\), \(P(\varphi)\)}\;
+   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M'\), \(P(\varphi)\)}\;
+   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M\), \(P'(\varphi)\)}\;
+   \BlankLine
+
+   \TypeVars \(\leftarrow\) \TyVars{\(\varphi\)}\;
+
+   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M| \times\)\MonoMult}, \MonoCap}\;
+   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)| \times\)\PolyMult}, \PolyCap}\;
+   \BlankLine
+
+   \UsedSubst \(\leftarrow\emptyset\)\;
+   \ForEach{\(\sigma\in \Subst\)}{
+      \ForEach{\(f\) occuring in \(\varphi\) }{
+         \ForEach{\((\tau_1, \dots, \tau_n)\in P(\varphi)(f)\)}{
+            \If{\TyVars{\((\tau_1, \dots, \tau_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
+               \If{\(|M''| < \MaxMono\) \label{algo:relative_bound}}{
+                  \(M''(f)\leftarrow M''(f)\cup\{(\sigma(\tau_1),\dots,\sigma(\tau_n))\}\)\;
+                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
+               }
+            }
+            \ElseIf{\(|P''(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
+               \(P''(\varphi)(f)\leftarrow P''(\varphi)(f)\cup\{(\sigma(\tau_1),\dots,\sigma(\tau_n))\}\)\;
+               \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
+            }
+
+         }
+      }
+   }
+   \BlankLine
+
+   \Return \((M'', P''(\varphi), \UsedSubst)\)
+
+\end{algorithm}
+\end{quote}
+
+At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), if we add conditions dependent on the size of \(M''(f)\) (respectively \(P''(\varphi)(f)\)), it is possible to create more fine grained bounds.
+This may be useful if some function symbols have many type arguments relative to the overall bound thus preventing new type arguments of other function symbols from being generated.
+Then, adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments.
 %\begin{algorithm}[]

### 2024-02-29 iterative_monomorphisation.tex
continued adding algos to detailed description part

@@
 %   - comma splices!!!!
+%   - add \; at the end of each algorithm line
 %   - ``'' -> `'
@@
 This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
-Although the example may seem artificial, similar problems frequently arise in practice.
+Although the example may seem artificial [contrived?], similar problems frequently arise in practice.
 For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
@@
 
+This function computes all possible matchings between the given type argument maps and returns the resulting substitutions.
+A single symbol may take multiple type arguments, they are treated as a unit.
+The matching of individual types from two type argument tuples must yield compatible substitutions, otherwise, they are discarded.
+This approach has the advantage of having the same behaviour whether a symbol has multiple type arguments, or a single type argument which encodes multiple type arguments.
+Adding a limit to the number of substitutions generated may be necessary if substitution generation becomes a bottleneck.
+In this case, it is a simple matter of stopping the function and returning \(S\) once a certain limit is reached.
+
 Notes:
@@
 
-   * In order to later have an easier time we store susbtitutions in a map that gathers them by which type variable they instantiate.
+   * In order to later have an easier time we store substitutions in a map that gathers them by which type variable they instantiate.
 
-   * We assume the existence of "primitives" that compose substitutions, apply them to formulae ...
+   * We assume the existence of "primitives" that
+   \begin{itemize}
+      \item compose substitutions
+      \item apply substitutions to formula
+      \item match types to generate substitutions
+      \item test the compatibility of two substitutions
+      \item extract the type variables from a type tuple
+      \item return the type variables that are instantiated by a substitution
+      \item functions to compute the initial type argument maps
+   \end{itemize}
 
@@
 \begin{algorithm}[H]
-   \caption{Clause Monomorphisation Step}
+   \caption{Formula Monomorphisation Step}
 
@@
 
-   \Subst \(\leftarrow\) \SubstGen{\(M\), \(P(\varphi)\)}\;
+   \Subst \(\leftarrow\) \SubstGen{\(M'\), \(P'(\varphi)\)}\;
    \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M'\), \(P(\varphi)\)}\;
@@
 
-   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M| \times\)\MonoMult}, \MonoCap}\;
-   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)| \times\)\PolyMult}, \PolyCap}\;
+   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M\cup M'| \times\)\MonoMult}, \MonoCap}\;
+   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)\cup P'(\varphi)| \times\)\PolyMult}, \PolyCap}\;
    \BlankLine
@@
    \ForEach{\(\sigma\in \Subst\)}{
-      \ForEach{\(f\) occuring in \(\varphi\) }{
+      \ForEach{\(f\) occurring in \(\varphi\) }{
          \ForEach{\((\tau_1, \dots, \tau_n)\in P(\varphi)(f)\)}{
@@
 
+This function computes new type arguments for a single formula.
+The separation of the type arguments into "old" and "new" maps is put to use here.
+Substitutions are generated only if one of the type arguments tuples used in the matching is new, this avoids computing matchings that have already been made in previous iterations.
+
 At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), if we add conditions dependent on the size of \(M''(f)\) (respectively \(P''(\varphi)(f)\)), it is possible to create more fine grained bounds.
 This may be useful if some function symbols have many type arguments relative to the overall bound thus preventing new type arguments of other function symbols from being generated.
-Then, adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments.
-%\begin{algorithm}[]
-%\begin{algorithmic}[1]
-%\Function{Substitution Generation}{Input: \(\varphi\), M, P}
-%   \State new\_substitutions = \(\emptyset\)
-%   \ForAll {\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}
-%      \ForAll {\( (\tau_1, \dots,\tau_n) \in M(f) \)}
-%         \State \(\sigma\) = \{\}
-%         \State continue = true
-%         \While {\(i\leq n\) and continue}
-%            \State subtitution\_result = match(\(\upsilon_i\), \(\tau_i\))
-%            \If {substitution\_result = Success(\(\sigma_i\)) and compatible(\(\sigma_i\), \(\sigma\))}
-%               \State \(\sigma = \sigma \circ \sigma_i\)
-%            \Else
-%               \State continue = false
-%            \EndIf
-%
-%         \EndWhile
-%
-%      \EndFor
-%   \EndFor
-%
-%   \State \Return (I, U, S)
-%
-%\EndFunction
-%\end{algorithmic}
-%\end{algorithm}
+Then, adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments over function symbols.
+
+\begin{quote}
+\begin{algorithm}
+   \caption{Monomorphising Substitutions}
+
+   \SetKwFunction{MonoSubst}{monomorphising\_substs}
+   \SetKwFunction{Domain}{domain}
+   \SetKwData{Limit}{subst\_limit}
+   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
+   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
+
+   \Input{\,set of substitutions \(U, S\), substitution \(\sigma\), set of type variables \(V\)}
+   \Output{\,set of substitutions}
+   \BlankLine
+
+   \If{\(V = \emptyset\) \And \(|S| < \Limit\)}{
+      \Return \(S\cup\{\sigma\}\)\;
+   }
+   \Else{
+      \(\exists\alpha\in V\)\;
+      \ForEach{\(\sigma'\in U\) such that \(\sigma'\) instantiates \(\alpha\) \And \(\sigma\) compatible with \(\sigma'\)}{
+         \If{\(|S|<\Limit\)}{
+            \(S\leftarrow \MonoSubst{U, S, \(\sigma'\circ\sigma\), \(V\setminus\Domain{\(\sigma'\)}\)}\)\;
+         }
+         \lElse{\Return \(S\)}
+      }
+      \Return \(S\)\;
+   }
+
+\end{algorithm}
+\end{quote}
+
+\begin{quote}
+\begin{algorithm}
+   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
+
+   \SetKwData{Loop}{loop\_nb}\SetKwData{NewFormulae}{new\_formulae}
+   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{used\_substitution}
+
+   \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step}
+   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{monomorphising\_substs}
+
+   \caption{Iterative Monomorphisation}
+
+   \((M, P) \leftarrow\) [empty maps]\;
+   \((M', P') \leftarrow \Init{\Problem}\)\;
+
+   \AllSubst\(\leftarrow \emptyset\)\;
+
+   \For{\(1\leq i\leq\Loop\) }{
+      \(M''\leftarrow\emptyset\)\;
+      \(P''\leftarrow\emptyset\)\;
+      \ForEach{\(\varphi \in \Problem\)} {
+         \((M^*, P^*(\varphi), S)\leftarrow \FStep(M, P(\varphi), M', P'(\varphi))\) \;
+         \(\AllSubst\leftarrow\AllSubst\cup S\)\;
+         \(M''\leftarrow M''\cup M^*\)\;
+         \(P''(\varphi)\leftarrow P''(\varphi)\cup P^*(\varphi)\)\;
+      }
+
+      \((M, M')\leftarrow (M\cup M', M'')\)\;
+      \((P, P')\leftarrow (P\cup P', P'')\)\;
+
+   }
+
+   \NewFormulae \(\leftarrow\emptyset\)\;
+   \For{\(\varphi\in\Phi\)}{
+      \(V\leftarrow \TyVars{\(\varphi\)}\)\;
+      \(S\leftarrow\MonoSubst{\AllSubst, \(\emptyset\), \(\{\}\), \(V\)}\)\;
+      \ForEach{}{
+         \(\NewFormulae\leftarrow\NewFormulae\cup\)
+      }
+
+   }
+
+\end{algorithm}
+\end{quote}
+
+The first loops apply the monomorphisation steps for each formula and perform a previously set number of iterations.
+At the end of a single iteration, the old and new maps are updated, this could in theory be done during a single iteration but would result in an uneven treatment of clauses.
+Formulae processed later would benefit from the additional monomorphic type arguments, generated when processing earlier formulae.
+Another undesirable effect of this approach would be making the algorithm dependent on the order of formulae in the problem.
+
 %In the Zipperposition implemntation this iteration procedure is limited by several bounds.
@@
 %
-%\begin{algorithm}[tbh]
-%\begin{algorithmic}[1]
-%\Function{Instantiate clause}{Clause: C, Substitutions S}
-%
-%   \If{\(C\) is ground}
-%
-%      \State \Return \(\{C\}\)
-%
-%   \Else
-%
-%      \State let \(v\) be a type variable in \(C\)
-%      \State let new\_clauses be an empty set of clauses
-%
-%      \ForAll{subst \(\in S(C)\)}
-%
-%         \If{subst instantiates \(v\)}
-%
-%            \State \(C' = apply(subst, C)\)
-%
-%            \State let v\_clauses = Instantiate clause(C', S)
-%
-%         \EndIf
-%
-%         \State \( new\_clauses = new\_clauses \cup v\_clauses\)
-%
-%      \EndFor
-%
-%      \State \Return new\_clauses
-%   \EndIf
-%
-%\EndFunction
-%\end{algorithmic}
-%\end{algorithm}
 %This function differs from the implementation in a significant way, indeed, through experimentation it was found that the generated clauses were far more likely to be solved by E if the substitutions that are combined to instantiate the clause were derived from the same iteration.
@@
 %
-%\begin{algorithm}[tbh]
-%\begin{algorithmic}[1]
-%\Function{Generate clauses}{Problem: P, Substitutions: S}
-%
-%\ForAll {clause \(C \in P\)}
-%   \State let new\_clauses = Instantiate clause(C, S)
-%   \State \(P = P \cup new\_clauses\)
-%\EndFor
-%
-%\State \(P' = \emptyset \)
-%
-%\ForAll {clause \(C \in P\)}
-%   \If{\(C\) is ground}
-%      \State \(P' = P' \cup \{mangle(C)\}\)
-%   \EndIf
-%\EndFor
-%
-%\State \Return \(P'\)
-%
-%\EndFunction
-%\end{algorithmic}
-%\end{algorithm}
-%
-%Implementation in Zipperposition.
 

### 2024-03-01 iterative_monomorphisation.tex
finished first draft for all algorithms
continued working on algo description

@@
 \usepackage{soul}
+\usepackage[dvipsnames]{xcolor}
 \usepackage{caption}
@@
 \usepackage{multirow}
-%\usepackage{algorithm}
-%\usepackage{algpseudocode}
 \usepackage%[]
@@
 \usepackage{amsmath}
+
+\SetNlSty{bfseries}{\color{black}}{}
 
@@
 
-The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a more low level algorithm with the following features.
+The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a lower level algorithm with the following features.
 \begin{enumerate}
-    \item[\labelitemi] Bounds are used at several stages of the algorithm to curb the highly explosive enumerations.
+    \item[\labelitemi] Bounds are presented at several stages of the algorithm to curb potentially explosive enumerations.
 
@@
     \item[\labelitemi] Substitutions are directly applied to the type arguments in $M$ and $P$ instead of the formulae.
-This dispenses from having to extract the type arguments from the formulae at each iteration.
-New formulae are generated only once all iterations are complete, in a separate, penultimate step.
+This dispenses from having to re-extract the type arguments from the formulae at each iteration.
+New formulae are only generated once all iterations are complete, in a separate, penultimate step.
 
@@
 
-TODO:
-\begin{itemize}
-   \item substitution generation
-   \item clause generation
-   \item fitting everything together
-\end{itemize}
+These algorithms rely on the existence of what will be considered ``primitives'' whose implementation depends on the specifics of the logic used and will therefore not be expanded upon.
+Any implementation of these algorithms will require the following functions:
+\begin{enumerate}
+   \item[\labelitemi] substitution composition
+   \item[\labelitemi] substitution application to formulae
+   \item[\labelitemi] type matching
+   \item[\labelitemi] testing the compatibility of two substitutions
+   \item[\labelitemi] extracting type variables from a type tuple
+   \item[\labelitemi] returning the type variables instantiated by a substitution
+   \item[\labelitemi] computing the initial type argument maps
+\end{enumerate}
 
@@
    \SetKwData{Continue}{continue}\SetKwData{SubstRes}{substitution\_result}
+   \SetKwData{SubstLimit}{substitution\_limit}
    \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
@@
    \Input{\,monomorphic type argument map \(M\), non-monomorphic type argument map \(P(\varphi)\)}
-   \Output{\,set of substitutions}
+   \Output{\,set of substitutions \(S\)}
    \BlankLine
@@
          \If{\Continue}{
-            \(S\leftarrow S\cup \{\sigma\}\)\;
+            \textcolor{BrickRed}{
+               \If{\(|S| < \SubstLimit\)}{
+                  \textcolor{black}{\(S\leftarrow S\cup \{\sigma\}\)\;}
+               }
+               \lElse{\Return \(S\)}
+            }
          }
@@
 
-This function computes all possible matchings between the given type argument maps and returns the resulting substitutions.
-A single symbol may take multiple type arguments, they are treated as a unit.
-The matching of individual types from two type argument tuples must yield compatible substitutions, otherwise, they are discarded.
-This approach has the advantage of having the same behaviour whether a symbol has multiple type arguments, or a single type argument which encodes multiple type arguments.
+This function computes all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions.
+For each pair of type argument tuples, the non-monomrophic types are matched against their monomorphic counterparts and the resulting substitutions are composed if they are compatible.
+If the substitutions are incompatible, the matchings are discarded.
+[Note that the composition of two compatible substitutions is commutative].
+
 Adding a limit to the number of substitutions generated may be necessary if substitution generation becomes a bottleneck.
@@
 
-Notes:
-   * In the implementation, the old and new maps are actually one map with function symbols pointing to pairs of sets of type arguments
-   * In the implementation, instead of tuples we have lists of types
-
-   * In order to later have an easier time we store substitutions in a map that gathers them by which type variable they instantiate.
-
-   * We assume the existence of "primitives" that
-   \begin{itemize}
-      \item compose substitutions
-      \item apply substitutions to formula
-      \item match types to generate substitutions
-      \item test the compatibility of two substitutions
-      \item extract the type variables from a type tuple
-      \item return the type variables that are instantiated by a substitution
-      \item functions to compute the initial type argument maps
-   \end{itemize}
-
+\SetKwData{MonoCap}{mono\_cap}\SetKwData{MonoMult}{mono\_mult}\SetKwData{MonoFloor}{mono\_floor}
+\SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
 \begin{quote}
@@
    \SetKwData{MaxPoly}{max\_poly\_args}
-   \SetKwData{MonoCap}{mono\_cap}\SetKwData{MonoMult}{mono\_mult}\SetKwData{MonoFloor}{mono\_floor}
-   \SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
    \SetKwData{UsedSubst}{used\_subst}
@@
 
-   \Input{\,monomorphic type argument map \(M, M'\), new non-monomorphic type argument map \(P(\varphi), P'(\varphi)\)}
-   \Output{\,monomorphic type argument map, non-monomorphic type argument map, set of substitutions}
+   \Input{\,old and new monomorphic type argument maps \(M, M'\), old and new non-monomorphic type argument maps \(P(\varphi), P'(\varphi)\)}
+   \Output{\,monomorphic type argument map \(M''\), non-monomorphic type argument map \(P''(\varphi)\), set of substitutions \(S\)}
    \BlankLine
@@
 
+   \textcolor{BrickRed}{
    \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M\cup M'| \times\)\MonoMult}, \MonoCap}\;
    \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)\cup P'(\varphi)| \times\)\PolyMult}, \PolyCap}\;
+   }
+
    \BlankLine
@@
             \If{\TyVars{\((\tau_1, \dots, \tau_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
+               \textcolor{BrickRed}{
                \If{\(|M''| < \MaxMono\) \label{algo:relative_bound}}{
+                  \textcolor{black}{
                   \(M''(f)\leftarrow M''(f)\cup\{(\sigma(\tau_1),\dots,\sigma(\tau_n))\}\)\;
-                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
+                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
                }
+               \ElseIf{\(|P''(\varphi)| \geq \MaxPoly\)}{
+                  \Return \((M'', P''(\varphi), \UsedSubst)\)\;
+               }}
             }
+            \textcolor{BrickRed}{
             \ElseIf{\(|P''(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
+               \textcolor{black}{
                \(P''(\varphi)(f)\leftarrow P''(\varphi)(f)\cup\{(\sigma(\tau_1),\dots,\sigma(\tau_n))\}\)\;
@@
                }
+            }}
 
@@
 This function computes new type arguments for a single formula.
-The separation of the type arguments into "old" and "new" maps is put to use here.
-Substitutions are generated only if one of the type arguments tuples used in the matching is new, this avoids computing matchings that have already been made in previous iterations.
+The separation of the type arguments into ``old'' and ``new'' maps is put to use here.
+Pairs of type argument tuples must contain at least one new type argument in order to be matched and generate a substitution.
+This avoids computing matchings that have already been made in previous iterations.
 
-At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), if we add conditions dependent on the size of \(M''(f)\) (respectively \(P''(\varphi)(f)\)), it is possible to create more fine grained bounds.
-This may be useful if some function symbols have many type arguments relative to the overall bound thus preventing new type arguments of other function symbols from being generated.
-Then, adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments over function symbols.
+The limits on the number of new type monomorphic arguments has three components (similar bounds are used for the non-monomorphic):
+\begin{enumerate}
+   \item \MonoCap is an absolute limit on the total number of new type arguments
+   \item \MonoMult is used to allow the total number of type arguments to increase by a certain proportion
+   \item \MonoFloor balances out \MonoMult when the number of type arguments is too low, and guarantees that a minimum number of new type arguments
+\end{enumerate}
+
+At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), adding conditions dependent on the size of \(M''(f)\) (respectively \(P''(\varphi)(f)\)), would create finer grained bounds.
+This may be useful if some function symbols have too many type arguments and prevent new type arguments of other function symbols from being generated.
+Adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments over function symbols.
 
@@
 
-   \If{\(V = \emptyset\) \And \(|S| < \Limit\)}{
-      \Return \(S\cup\{\sigma\}\)\;
+   \If{\(V = \emptyset\)}{
+    %  \textcolor{BrickRed}{
+    %  \If{\(|S|<\Limit\) }{
+    %     \textcolor{black}{
+    \Return \(S\cup\{\sigma\}\)\;%}
+    %  }
+    %  \lElse{\Return \(S\)}}
    }
@@
       \(\exists\alpha\in V\)\;
-      \ForEach{\(\sigma'\in U\) such that \(\sigma'\) instantiates \(\alpha\) \And \(\sigma\) compatible with \(\sigma'\)}{
+      \ForEach{\(\sigma'\in U\) such that, \(\alpha\in\Domain{\(\sigma'\)}\) \And \(\sigma\) compatible with \(\sigma'\)}{
+         \textcolor{BrickRed}{
          \If{\(|S|<\Limit\)}{
-            \(S\leftarrow \MonoSubst{U, S, \(\sigma'\circ\sigma\), \(V\setminus\Domain{\(\sigma'\)}\)}\)\;
+            \textcolor{black}{
+            \(S\leftarrow \MonoSubst{U, S, \(\sigma'\circ\sigma\), \(V\setminus\Domain{\(\sigma'\)}\)}\)\;}
          }
-         \lElse{\Return \(S\)}
+         \lElse{\Return \(S\)}}
       }
@@
 
+This recursive function computes all substitutions that can instantiate a set of type variables.
+The substitutions are obtained by composition of compatible substitutions from a given set.
+The returned substitutions are said to be ``monomorphising'' with regards to the set of input type variables \(V\) because they instantiate all type variables in \(V\).
+In the Zipperposition implementation this limit was one of the most important bounds because the total number of possible monomorphising substitutions is very high but very few are necessary in practice.
+
+In the Zipperposition implementation, due the substitutions in \(U\) being selected on the basis on which type variable they instantiate, all substitutions are kept in a map which makes this selection faster.
+
+Because this function often hits the substitution limit, the order of substitutions in \(U\) (or the map in Zipperposition) can have a significant impact on the resulting monomorphising substitutions.
+This may be a relevant heuristic to tweak depending on the input problem.
+
 \begin{quote}
@@
    \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{used\_substitution}
+   \SetKwData{Limit}{total\_new\_formulae}
 
@@
 
-   \For{\(1\leq i\leq\Loop\) }{
+   \For{\(1\leq i\leq \textcolor{BrickRed}{\Loop}\) }{
       \(M''\leftarrow\emptyset\)\;
@@
       \(S\leftarrow\MonoSubst{\AllSubst, \(\emptyset\), \(\{\}\), \(V\)}\)\;
-      \ForEach{}{
-         \(\NewFormulae\leftarrow\NewFormulae\cup\)
+      \ForEach{\(\sigma\in S\)}{
+         \textcolor{BrickRed}{
+         \If{\(|\NewFormulae|<\Limit\)}{
+            \textcolor{black}{
+            \(\NewFormulae\leftarrow\NewFormulae\cup \sigma(\varphi)\)\;}
          }
-
+         \lElse{\Return \NewFormulae}
+      }}
    }
+   \Return \NewFormulae\;
 
@@
 The first loops apply the monomorphisation steps for each formula and perform a previously set number of iterations.
-At the end of a single iteration, the old and new maps are updated, this could in theory be done during a single iteration but would result in an uneven treatment of clauses.
+At the end of a single iteration, the old and new maps are updated.
+This could in theory be done during a single iteration but would result in an uneven treatment of clauses.
 Formulae processed later would benefit from the additional monomorphic type arguments, generated when processing earlier formulae.
@@
 
-%In the Zipperposition implemntation this iteration procedure is limited by several bounds.
-The number of new type arguments that can be generated for each symbol is limited by user-defined bounds.
-These bounds can be defined proportionally.
-Additionally, the number of new type arguments can be limited at the clause level.
-In both cases, the bounds for the new ground type arguments and non-ground type arguments can be set independently.
-%
-%A consequences of these bounds is that in order to avoid unecessary computations, the program will iteratively apply substitutions until either the ground or non-ground bound is reached.
-Then substitutions can be discarded depending on whether they instantiate a given set of type arguments or not.
-%
-%Additionally, there are bounds for the number of substitutions that are generated.
-%
-%This function differs from the implementation in a significant way, indeed, through experimentation it was found that the generated clauses were far more likely to be solved by E if the substitutions that are combined to instantiate the clause were derived from the same iteration.
-In practice, we therefore store the iteration at which each substition is generated and at the instantiation phase only apply substitutions from the same iteration.
-%
-%
-%What the previous function does is essentially going through the tree of possible substitutions that instantiate the clause and collecting all possible instantiations.
-This is extremely explosive, and a hard cap on the number of clauses we want to generate is almost indispensable for any implementation.
-%
+Once all type arguments have been generated, the substitutions used to create those new type arguments are then combined in order to form monomorphising substitutions that are then applied to the polymorphic clauses.
+There can be a high number of possible monomorphic clauses resulting from this process.
+This can be limited both using a fixed number of
 

### 2024-03-04 iterative_monomorphisation.tex
some work with Tanguy
minor improvements on algos
some minor styling changes, still more to be done

@@
 \usepackage{multirow}
-\usepackage%[]
+\usepackage[noend]
    {algorithm2e}
@@
 
+% algo2e package
+\SetFuncSty{textsf}
+\SetFuncArgSty{textit}
+\SetArgSty{text}
+\SetDataSty{textit}
+
+\DontPrintSemicolon
+
+\SetKwProg{Fn}{Function}{}{end}\SetKwFunction{FRecurs}{FnRecursive}%
+\newcommand{\forcond}{$i=0$ \KwTo $n$}
+
+\definecolor{ourblueviolet}{HTML}{0071BC}
 \definecolor{grey}{rgb}{0.8,0.8,0.8}
@@
 
+\SetKwFunction{IterMono}{iterative\_monomorphisation}
+\begin{figure}
 \begin{quote}
 \begin{algorithm}[H]
-   \caption{Substitution Generation}
+\Fn(){\IterMono{\(\Phi\)}}{
+   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
+
+   \SetKwData{Loop}{loop\_nb}\SetKwData{NewFormulae}{\(\Psi\)}
+   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(U\)}
+   \SetKwData{Limit}{total\_new\_formulae}
+
+   \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step}
+   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{monomorphising\_substs}
+
+   \KwData{set of formulae \(\varphi\)}
+   \KwResult{set of monomorphic formulae}
+
+   \((M, P) \leftarrow (\emptyset, \emptyset)\)\;
+   \((M^{\text{new}}, P^{\text{new}}) \leftarrow \Init{\Problem}\)\;
+
+   \AllSubst\(\leftarrow \emptyset\)\;
+
+   \For{\(1\leq i\leq \textcolor{ourblueviolet}{\Loop}\) }{
+      \((M^{\text{next}}, P^{\text{next}})\leftarrow(\emptyset,\emptyset)\)\;
+      \ForEach{\(\varphi \in \Problem\)} {
+         \((M^*, P^*(\varphi), S)\leftarrow \FStep(M, P(\varphi), M^{\text{new}}, P^{\text{new}}(\varphi))\)\;
+         \(\AllSubst\leftarrow\AllSubst\cup S\)\;
+         \(M^{\text{next}}\leftarrow M^{\text{next}}\cup M^*\)\;
+         \(P^{\text{next}}(\varphi)\leftarrow P^{\text{next}}(\varphi)\cup P^*(\varphi)\)\;
+      }
+
+      \((M, M^{\text{new}})\leftarrow (M\cup M^{\text{new}}, M^{\text{next}})\)\;
+      \((P, P^{\text{new}})\leftarrow (P\cup P^{\text{new}}, P^{\text{next}})\)\;
+
+   }
+
+   \NewFormulae \(\leftarrow\emptyset\)\;
+   \For{\(\varphi\in\Phi\)}{
+      \(V\leftarrow \TyVars{\(\varphi\)}\)\;
+      \(S\leftarrow\MonoSubst{\AllSubst, \(\emptyset\), \(\{\}\), \(V\)}\)\;
+      \ForEach{\(\sigma\in S\)}{
+         \textcolor{ourblueviolet}{
+         \eIf{\(|\NewFormulae|<\Limit\)}{
+            \textcolor{black}{
+            \(\NewFormulae\leftarrow\NewFormulae\cup \{\sigma(\varphi)\}\)\;}
+         }
+         {\Return \NewFormulae}
+      }}
+   }
+   \Return \NewFormulae\;
+
+}
+\end{algorithm}
+\end{quote}
+\caption{Pseudocode for iterative monomorphisation}
+\end{figure}
+
+The first loops apply the monomorphisation steps for each formula and perform a previously set number of iterations.
+At the end of a single iteration, the old and new maps are updated.
+This could in theory be done during a single iteration but would result in an uneven treatment of clauses.
+Formulae processed later would benefit from the additional monomorphic type arguments, generated when processing earlier formulae.
+Another undesirable effect of this approach would be making the algorithm dependent on the order of formulae in the problem.
+
+Once all type arguments have been generated, the substitutions used to create those new type arguments are then combined in order to form monomorphising substitutions that are then applied to the polymorphic clauses.
+There can be a high number of possible monomorphic clauses resulting from this process.
+This can be limited both using a fixed number of
+
+\begin{figure}
+\begin{quote}
+\begin{algorithm}[H]
+\SetKwFunction{SubstGen}{substitution\_generation}
+\Fn(){\SubstGen{\(M, P(\varphi)\)}}{
+%   \caption{Substitution Generation()}
 %   \SetAlgoNoLine
@@
    \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
-   \SetKwData{Continue}{continue}\SetKwData{SubstRes}{substitution\_result}
    \SetKwData{SubstLimit}{substitution\_limit}
@@
    \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
+   \SetKw{Break}{break}
 
-   \Input{\,monomorphic type argument map \(M\), non-monomorphic type argument map \(P(\varphi)\)}
-   \Output{\,set of substitutions \(S\)}
+   \KwData{monomorphic type argument map \(M\), non-monomorphic type argument map \(P(\varphi)\)}
+   \KwResult{set of substitutions}
    \BlankLine
@@
          \(\sigma\leftarrow \{\}\)\;
-         \Continue \(\leftarrow\) \True\;
          \(i \leftarrow 1\)\;
-         \While{\(i \leq n\) \And \Continue}{
-            \SubstRes \(\leftarrow\) \Match{\(\upsilon_i\), \(\tau_i\)}\;
-            \Case{\SubstRes = \Success{\(\sigma_i\)}}{
-               \lIf{\Compatible{\(\sigma\), \(\sigma_i\)}}{
-                  \(\sigma\leftarrow\sigma_i\circ\sigma\)}
-               \lElse{\Continue\(\leftarrow\)\False}
+         \While{\(i \leq n\)}{
+            \eIf{\(\Match{\(\upsilon_i\), \(\tau_i\)}\) has the form \Success{\(\sigma_i\)} \And \Compatible{\(\sigma\), \(\sigma_i\)}}{
+               \(\sigma\leftarrow\sigma_i\circ\sigma\)
             }
-            \lOther{\Continue\(\leftarrow\)\False}
+            {\Break}
 
@@
 
-         \If{\Continue}{
-            \textcolor{BrickRed}{
-               \If{\(|S| < \SubstLimit\)}{
+         \If{\(i > n\)}{
+            \textcolor{ourblueviolet}{
+               \eIf{\(|S| < \SubstLimit\)}{
                   \textcolor{black}{\(S\leftarrow S\cup \{\sigma\}\)\;}
                }
-               \lElse{\Return \(S\)}
+               {\Return \(S\)}
             }
@@
    \Return \(S\)\;
+
+}
 \end{algorithm}
 \end{quote}
+\caption{Pseudocode for substitution generation}
+\end{figure}
 
@@
 \SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
+\begin{figure}
 \begin{quote}
 \begin{algorithm}[H]
-   \caption{Formula Monomorphisation Step}
 
@@
    \SetKwData{MaxPoly}{max\_poly\_args}
-   \SetKwData{UsedSubst}{used\_subst}
+   \SetKwData{UsedSubst}{\(U\)}
 
@@
    \SetKwFunction{Domain}{subst\_domain}\SetKwFunction{Len}{len}
-   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
 
-   \Input{\,old and new monomorphic type argument maps \(M, M'\), old and new non-monomorphic type argument maps \(P(\varphi), P'(\varphi)\)}
-   \Output{\,monomorphic type argument map \(M''\), non-monomorphic type argument map \(P''(\varphi)\), set of substitutions \(S\)}
+   \KwData{old and new monomorphic type argument maps \(M, M^{\text{new}}\), old and new non-monomorphic type argument maps \(P(\varphi), P^{\text{new}}(\varphi)\)}
+   \KwResult{monomorphic type argument map \(M^{\text{next}}\), non-monomorphic type argument map \(P^{\text{next}}(\varphi)\), set of substitutions \(S\)}
    \BlankLine
 
-   \Subst \(\leftarrow\) \SubstGen{\(M'\), \(P'(\varphi)\)}\;
-   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M'\), \(P(\varphi)\)}\;
-   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M\), \(P'(\varphi)\)}\;
+   \Subst \(\leftarrow\) \SubstGen{\(M^{\text{new}}\), \(P^{\text{new}}(\varphi)\)}\;
+   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M^{\text{new}}\), \(P(\varphi)\)}\;
+   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M\), \(P^{\text{new}}(\varphi)\)}\;
    \BlankLine
@@
 
-   \textcolor{BrickRed}{
-   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M\cup M'| \times\)\MonoMult}, \MonoCap}\;
-   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)\cup P'(\varphi)| \times\)\PolyMult}, \PolyCap}\;
+   \textcolor{ourblueviolet}{
+   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M\cup M^{\text{new}}| \cdot \)\MonoMult}, \MonoCap}\;
+   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)\cup P^{\text{new}}(\varphi)| \cdot \)\PolyMult}, \PolyCap}\;
    }
@@
    \ForEach{\(\sigma\in \Subst\)}{
-      \ForEach{\(f\) occurring in \(\varphi\) }{
-         \ForEach{\((\tau_1, \dots, \tau_n)\in P(\varphi)(f)\)}{
-            \If{\TyVars{\((\tau_1, \dots, \tau_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
-               \textcolor{BrickRed}{
-               \If{\(|M''| < \MaxMono\) \label{algo:relative_bound}}{
+      \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
+            \eIf{\TyVars{\((\upsilon_1, \dots, \upsilon_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
+               \textcolor{ourblueviolet}{
+               \uIf{\(|M^{\text{next}}| < \MaxMono\) \label{algo:relative_bound}}{
                   \textcolor{black}{
-                  \(M''(f)\leftarrow M''(f)\cup\{(\sigma(\tau_1),\dots,\sigma(\tau_n))\}\)\;
+                  \(M^{\text{next}}(f)\leftarrow M^{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
                   \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
                }
-               \ElseIf{\(|P''(\varphi)| \geq \MaxPoly\)}{
-                  \Return \((M'', P''(\varphi), \UsedSubst)\)\;
+               \uElseIf{\(|P^{\text{next}}(\varphi)| \geq \MaxPoly\) }{
+                  blablabliblou\;
+                  \Return \((M^{\text{next}}, P^{\text{next}}(\varphi), \UsedSubst)\)\;
                }}
             }
-            \textcolor{BrickRed}{
-            \ElseIf{\(|P''(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
+            {
+               \textcolor{ourblueviolet}{
+               \If{\(|P^{\text{next}}(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
                \textcolor{black}{
-               \(P''(\varphi)(f)\leftarrow P''(\varphi)(f)\cup\{(\sigma(\tau_1),\dots,\sigma(\tau_n))\}\)\;
+               \(P^{\text{next}}(\varphi)(f)\leftarrow P^{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
                \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
+            }}
             }
-            }}
 
@@
 
-   \Return \((M'', P''(\varphi), \UsedSubst)\)
+   \Return \((M^{\text{next}}, P^{\text{next}}(\varphi), \UsedSubst)\)
 
@@
 \end{quote}
+\caption{Formula Monomorphisation Step}
+\end{figure}
 
@@
 
-At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), adding conditions dependent on the size of \(M''(f)\) (respectively \(P''(\varphi)(f)\)), would create finer grained bounds.
+At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), adding conditions dependent on the size of \(M^{\text{next}}(f)\) (respectively \(P^{\text{next}}(\varphi)(f)\)), would create finer grained bounds.
 This may be useful if some function symbols have too many type arguments and prevent new type arguments of other function symbols from being generated.
@@
 
+\begin{figure}
 \begin{quote}
-\begin{algorithm}
-   \caption{Monomorphising Substitutions}
+   \begin{algorithm}[H]
 
@@
    \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
+   \SetKw{Let}{let}
    \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
 
-   \Input{\,set of substitutions \(U, S\), substitution \(\sigma\), set of type variables \(V\)}
-   \Output{\,set of substitutions}
+   \KwData{set of substitutions \(U\) and \(S\), single substitution \(\sigma\), set of type variables \(V\)}
+   \KwResult{set of substitutions}
    \BlankLine
 
-   \If{\(V = \emptyset\)}{
-    %  \textcolor{BrickRed}{
-    %  \If{\(|S|<\Limit\) }{
-    %     \textcolor{black}{
+   \eIf{\(V = \emptyset\)}{
     \Return \(S\cup\{\sigma\}\)\;%}
-    %  }
-    %  \lElse{\Return \(S\)}}
    }
-   \Else{
-      \(\exists\alpha\in V\)\;
+   {
+      \Let \(\alpha \) such that \(\alpha\in V\)\;
       \ForEach{\(\sigma'\in U\) such that, \(\alpha\in\Domain{\(\sigma'\)}\) \And \(\sigma\) compatible with \(\sigma'\)}{
-         \textcolor{BrickRed}{
-         \If{\(|S|<\Limit\)}{
+         \textcolor{ourblueviolet}{
+         \eIf{\(|S|<\Limit\)}{
             \textcolor{black}{
@@
          }
-         \lElse{\Return \(S\)}}
+         {\Return \(S\)}}
       }
@@
 \end{quote}
+\caption{Monomorphising Substitutions}
+\end{figure}
 
@@
 This may be a relevant heuristic to tweak depending on the input problem.
-
-\begin{quote}
-\begin{algorithm}
-   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
-
-   \SetKwData{Loop}{loop\_nb}\SetKwData{NewFormulae}{new\_formulae}
-   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{used\_substitution}
-   \SetKwData{Limit}{total\_new\_formulae}
-
-   \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step}
-   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{monomorphising\_substs}
-
-   \caption{Iterative Monomorphisation}
-
-   \((M, P) \leftarrow\) [empty maps]\;
-   \((M', P') \leftarrow \Init{\Problem}\)\;
-
-   \AllSubst\(\leftarrow \emptyset\)\;
-
-   \For{\(1\leq i\leq \textcolor{BrickRed}{\Loop}\) }{
-      \(M''\leftarrow\emptyset\)\;
-      \(P''\leftarrow\emptyset\)\;
-      \ForEach{\(\varphi \in \Problem\)} {
-         \((M^*, P^*(\varphi), S)\leftarrow \FStep(M, P(\varphi), M', P'(\varphi))\) \;
-         \(\AllSubst\leftarrow\AllSubst\cup S\)\;
-         \(M''\leftarrow M''\cup M^*\)\;
-         \(P''(\varphi)\leftarrow P''(\varphi)\cup P^*(\varphi)\)\;
-      }
-
-      \((M, M')\leftarrow (M\cup M', M'')\)\;
-      \((P, P')\leftarrow (P\cup P', P'')\)\;
-
-   }
-
-   \NewFormulae \(\leftarrow\emptyset\)\;
-   \For{\(\varphi\in\Phi\)}{
-      \(V\leftarrow \TyVars{\(\varphi\)}\)\;
-      \(S\leftarrow\MonoSubst{\AllSubst, \(\emptyset\), \(\{\}\), \(V\)}\)\;
-      \ForEach{\(\sigma\in S\)}{
-         \textcolor{BrickRed}{
-         \If{\(|\NewFormulae|<\Limit\)}{
-            \textcolor{black}{
-            \(\NewFormulae\leftarrow\NewFormulae\cup \sigma(\varphi)\)\;}
-         }
-         \lElse{\Return \NewFormulae}
-      }}
-   }
-   \Return \NewFormulae\;
-
-\end{algorithm}
-\end{quote}
-
-The first loops apply the monomorphisation steps for each formula and perform a previously set number of iterations.
-At the end of a single iteration, the old and new maps are updated.
-This could in theory be done during a single iteration but would result in an uneven treatment of clauses.
-Formulae processed later would benefit from the additional monomorphic type arguments, generated when processing earlier formulae.
-Another undesirable effect of this approach would be making the algorithm dependent on the order of formulae in the problem.
-
-Once all type arguments have been generated, the substitutions used to create those new type arguments are then combined in order to form monomorphising substitutions that are then applied to the polymorphic clauses.
-There can be a high number of possible monomorphic clauses resulting from this process.
-This can be limited both using a fixed number of
 
@@
 During an iteration, for each symbol (for each clause in the polymorphic case), the number of newly generated type arguments is determined by this formula:
-\( \min(\text{cap}, \max(\text{floor}, \text{multiplier} \times \text{number of type arguments})) \)
+\( \min(\text{cap}, \max(\text{floor}, \text{multiplier} \cdot \text{number of type arguments})) \)
 

### 2024-03-05 iterative_monomorphisation.tex
continued improving and writting detailed section

@@
 % algo2e package
-\SetFuncSty{textsf}
-\SetFuncArgSty{textit}
+\SetFuncSty{textit}
+\SetFuncArgSty{text}
 \SetArgSty{text}
-\SetDataSty{textit}
+\SetDataSty{text}
 
@@
 
-These algorithms rely on the existence of what will be considered ``primitives'' whose implementation depends on the specifics of the logic used and will therefore not be expanded upon.
-Any implementation of these algorithms will require the following functions:
+These algorithms rely on ``primitives'' whose implementation depends on the specifics of the grammar and logic used.
+They will therefore not be expanded upon.
+The functions computing the following are assumed to exist:
 \begin{enumerate}
+   \item[\labelitemi] computing the initial type argument maps
+   \item[\labelitemi] extracting type variables from a type tuple
+   \item[\labelitemi] type matching
    \item[\labelitemi] substitution composition
+   \item[\labelitemi] testing the compatibility of two substitutions
+   \item[\labelitemi] returning the type variables instantiated by a substitution
    \item[\labelitemi] substitution application to formulae
-   \item[\labelitemi] type matching
-   \item[\labelitemi] testing the compatibility of two substitutions
-   \item[\labelitemi] extracting type variables from a type tuple
-   \item[\labelitemi] returning the type variables instantiated by a substitution
-   \item[\labelitemi] computing the initial type argument maps
 \end{enumerate}
@@
 \Fn(){\IterMono{\(\Phi\)}}{
-   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
+   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{To}{to}
 
    \SetKwData{Loop}{loop\_nb}\SetKwData{NewFormulae}{\(\Psi\)}
-   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(U\)}
+   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(S^{\text{used}}\)}
    \SetKwData{Limit}{total\_new\_formulae}
@@
    \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step}
-   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{monomorphising\_substs}
+   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{mono\_substs}
 
-   \KwData{set of formulae \(\varphi\)}
+   \KwData{set of formulae \(\Phi\)}
    \KwResult{set of monomorphic formulae}
 
-   \((M, P) \leftarrow (\emptyset, \emptyset)\)\;
+   \((M^{\text{old}}, P^{\text{old}}) \leftarrow (\emptyset, \emptyset)\)\;
    \((M^{\text{new}}, P^{\text{new}}) \leftarrow \Init{\Problem}\)\;
@@
 
-   \For{\(1\leq i\leq \textcolor{ourblueviolet}{\Loop}\) }{
+   \For{\(i = 1\) \To \textcolor{ourblueviolet}{\Loop}}{
       \((M^{\text{next}}, P^{\text{next}})\leftarrow(\emptyset,\emptyset)\)\;
       \ForEach{\(\varphi \in \Problem\)} {
-         \((M^*, P^*(\varphi), S)\leftarrow \FStep(M, P(\varphi), M^{\text{new}}, P^{\text{new}}(\varphi))\)\;
-         \(\AllSubst\leftarrow\AllSubst\cup S\)\;
-         \(M^{\text{next}}\leftarrow M^{\text{next}}\cup M^*\)\;
-         \(P^{\text{next}}(\varphi)\leftarrow P^{\text{next}}(\varphi)\cup P^*(\varphi)\)\;
+         \((M^{\Delta}, P^{\Delta}(\varphi), S^\Delta)\leftarrow \FStep(M^{\text{old}}, P^{\text{old}}(\varphi), M^{\text{new}}, P^{\text{new}}(\varphi))\)\;
+         \(\AllSubst\leftarrow\AllSubst\cup S^\Delta\)\;
+         \(M^{\text{next}}\leftarrow M^{\text{next}}\cup M^{\Delta}\)\;
+         \(P^{\text{next}}(\varphi)\leftarrow P^{\text{next}}(\varphi)\cup P^{\Delta}(\varphi)\)\;
       }
 
-      \((M, M^{\text{new}})\leftarrow (M\cup M^{\text{new}}, M^{\text{next}})\)\;
-      \((P, P^{\text{new}})\leftarrow (P\cup P^{\text{new}}, P^{\text{next}})\)\;
+      \((M^{\text{old}}, M^{\text{new}})\leftarrow (M^{\text{old}}\cup M^{\text{new}}, M^{\text{next}})\)\;
+      \((P^{\text{old}}, P^{\text{new}})\leftarrow (P^{\text{old}}\cup P^{\text{new}}, P^{\text{next}})\)\;
 
@@
    \NewFormulae \(\leftarrow\emptyset\)\;
-   \For{\(\varphi\in\Phi\)}{
+   \ForEach{\(\varphi\in\Phi\) polymorphic}{
       \(V\leftarrow \TyVars{\(\varphi\)}\)\;
-      \(S\leftarrow\MonoSubst{\AllSubst, \(\emptyset\), \(\{\}\), \(V\)}\)\;
-      \ForEach{\(\sigma\in S\)}{
+      \(S^{\text{mono}}\leftarrow\MonoSubst{\AllSubst, \(\emptyset\), \(\{\}\), \(V\)}\)\;
+      \ForEach{\(\sigma\in S^{\text{mono}}\)}{
          \textcolor{ourblueviolet}{
@@
 \caption{Pseudocode for iterative monomorphisation}
+\label{iter_mono}
 \end{figure}
 
-The first loops apply the monomorphisation steps for each formula and perform a previously set number of iterations.
-At the end of a single iteration, the old and new maps are updated.
-This could in theory be done during a single iteration but would result in an uneven treatment of clauses.
-Formulae processed later would benefit from the additional monomorphic type arguments, generated when processing earlier formulae.
-Another undesirable effect of this approach would be making the algorithm dependent on the order of formulae in the problem.
+The iterative monomorphisation algorithm, as implemented in Zipperposition, is described in \ref{iter_mono}.
+There are two main phases each consisting of a separate loop.
+The first loop applies a monomorphisation step to each formula for each iteration of the overall for loop until the user-set limit it reached.
+This bound is the only limit which is necessary for the algorithm to terminate.
+At the end of each iteration, the old and new maps are updated.
+%Updating the maps progressively over the course of an iteration would be feasible but would result in an uneven treatment of formulae.
+Type arguments in a formula processed later could be matched against monomorphic type arguments generated at an earlier time in the same iteration.
+As a result, the later a formula is processed the more new type arguments it could generate.
 
-Once all type arguments have been generated, the substitutions used to create those new type arguments are then combined in order to form monomorphising substitutions that are then applied to the polymorphic clauses.
-There can be a high number of possible monomorphic clauses resulting from this process.
-This can be limited both using a fixed number of
+Once all type arguments have been generated, the substitutions used to create those new type arguments are then combined in order to form monomorphising substitutions which are subsequently applied to the polymorphic formulae.
+There can be a high number of possible monomorphic formulae resulting from this process.
+This can be limited by setting a maximal number of new monomorphic formulae.
+In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected when it is given too many clauses.
+
+\SetKwData{MonoCap}{mono\_cap}\SetKwData{MonoMult}{mono\_mult}\SetKwData{MonoFloor}{mono\_floor}
+\SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
+\begin{figure}
+\begin{quote}
+\begin{algorithm}[H]
+
+   %\LinesNumbered
+   \SetKwFunction{SubstGen}{subst\_generation}\SetKwFunction{TyVars}{type\_variables}
+   \SetKwFunction{FMonoStep}{formula\_mono\_step}
+
+   \Fn(){\FMonoStep{\(M^{\text{old}}, M^{\text{new}}, P^{\text{old}}(\varphi), P^{\text{new}}(\varphi)\)}}{
+   \SetKwData{Subst}{substs}\SetKwData{TypeVars}{type\_variables}
+   \SetKwData{MaxMono}{max\_mono\_args}
+   \SetKwData{MaxPoly}{max\_poly\_args}
+   \SetKwData{UsedSubst}{\(S^{\text{used}}\)}
+
+   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
+
+   \SetKwFunction{Max}{max}\SetKwFunction{Min}{min}
+   \SetKwFunction{Domain}{subst\_domain}\SetKwFunction{Len}{len}
+
+   \KwData{
+      \hspace{-2.7em}\begin{minipage}[t]{.8\textwidth}
+      \begin{enumerate}
+         \item[] old and new monomorphic type argument maps \(M^{\text{old}}, M^{\text{new}}\)
+         \item[] old and new non-monomorphic type argument maps \(P^{\text{old}}(\varphi), P^{\text{new}}(\varphi)\)
+      \end{enumerate}
+      \end{minipage}
+   }
+   \KwResult{
+      \hspace{-2.7em}\begin{minipage}[t]{.8\textwidth}
+      \begin{enumerate}
+         \item[] monomorphic type argument map
+         \item[] non-monomorphic type argument map
+         \item[] set of substitutions
+      \end{enumerate}
+      \end{minipage}
+   }
+   \BlankLine
+
+   \Subst \(\leftarrow\) \SubstGen{\(M^{\text{new}}\), \(P^{\text{new}}(\varphi)\)}\;
+   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M^{\text{new}}\), \(P(\varphi)\)}\;
+   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M\), \(P^{\text{new}}(\varphi)\)}\;
+   \BlankLine
+
+   \TypeVars \(\leftarrow\) \TyVars{\(\varphi\)}\;
+
+   \textcolor{ourblueviolet}{
+   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M\cup M^{\text{new}}| \cdot \)\MonoMult}, \MonoCap}\;
+   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)\cup P^{\text{new}}(\varphi)| \cdot \)\PolyMult}, \PolyCap}\;
+   }
+
+   \BlankLine
+
+   \UsedSubst \(\leftarrow\emptyset\)\;
+   \ForEach{\(\sigma\in \Subst\)}{
+      \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
+            \eIf{\TyVars{\((\upsilon_1, \dots, \upsilon_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
+               \textcolor{ourblueviolet}{
+               \uIf{\(|M^{\text{next}}| < \MaxMono\) \label{algo:relative_bound}}{
+                  \textcolor{black}{
+                  \(M^{\text{next}}(f)\leftarrow M^{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
+                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
+               }
+               \uElseIf{\(|P^{\text{next}}(\varphi)| \geq \MaxPoly\) }{
+                  blablabliblou\;
+                  \(M^\text{next} \leftarrow M^{\text{next}}\setminus (M^\text{old} \cup M^\text{new})\)\;
+                  \(P^\text{next}(\varphi)\leftarrow P^{\text{next}}(\varphi)\setminus (P^\text{old}(\varphi)\cup P^\text{new}(\varphi))\)\;
+                  \Return \((M^\text{next}, P^\text{next}(\varphi), \UsedSubst)\)
+               }}
+            }
+            {
+               \textcolor{ourblueviolet}{
+               \If{\(|P^{\text{next}}(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
+               \textcolor{black}{
+               \(P^{\text{next}}(\varphi)(f)\leftarrow P^{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
+               \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
+            }}
+            }
+
+         }
+      }
+   }
+   \BlankLine
+
+   \Return \((M^{\text{next}}\setminus (M^\text{old} \cup M^\text{new}), P^{\text{next}}(\varphi)\setminus (P^\text{old}(\varphi)\cup P^\text{new}(\varphi)), \UsedSubst)\)
+}
+
+\end{algorithm}
+\end{quote}
+\caption{Formula Monomorphisation Step}
+\label{mono_step}
+\end{figure}
+
+Function \ref{mono_step} computes new type arguments for a single formula.
+The separation of the type arguments into ``old'' and ``new'' maps is exploited here.
+Pairs of type argument tuples must contain at least one new type argument in order to be matched and generate a substitution.
+This avoids computing matchings that have already been made in previous iterations.
+Unfortunately, not all such redundant computations can be avoided because different matchings with entirely different types can yield the same substitution.
+Furthermore,
+
+The limits on the number of new type monomorphic arguments has three components (similar bounds are used for the non-monomorphic):
+\begin{enumerate}
+   \item \MonoCap is an absolute limit on the total number of new type arguments
+   \item \MonoMult is used to allow the total number of type arguments to increase by a certain proportion
+   \item \MonoFloor balances out \MonoMult when the number of type arguments is too low, and guarantees that a minimum number of new type arguments
+\end{enumerate}
+
+At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), adding conditions dependent on the size of \(M^{\text{next}}(f)\) (respectively \(P^{\text{next}}(\varphi)(f)\)), would create finer grained bounds.
+This may be useful if some function symbols have too many type arguments and prevent new type arguments of other function symbols from being generated.
+Adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments over function symbols.
 
@@
 
-   \KwData{monomorphic type argument map \(M\), non-monomorphic type argument map \(P(\varphi)\)}
+   \KwData{%
+      \hspace{-2.7em}\begin{minipage}[t]{.8\textwidth}
+      \begin{enumerate}
+         \item[] monomorphic type argument map \(M\)
+         \item[] non-monomorphic type argument map \(P(\varphi)\)
+      \end{enumerate}
+      \end{minipage}
+   }
    \KwResult{set of substitutions}
@@
 In this case, it is a simple matter of stopping the function and returning \(S\) once a certain limit is reached.
-
-\SetKwData{MonoCap}{mono\_cap}\SetKwData{MonoMult}{mono\_mult}\SetKwData{MonoFloor}{mono\_floor}
-\SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
 \begin{figure}
@@
 
-   \LinesNumbered
-
-   \SetKwData{Subst}{substitutions}\SetKwData{TypeVars}{type\_variables}
-   \SetKwData{MaxMono}{max\_mono\_args}
-   \SetKwData{MaxPoly}{max\_poly\_args}
-   \SetKwData{UsedSubst}{\(U\)}
-
-   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
-
-   \SetKwFunction{SubstGen}{substitution\_generation}\SetKwFunction{TyVars}{type\_variables}
-   \SetKwFunction{Max}{max}\SetKwFunction{Min}{min}
-   \SetKwFunction{Domain}{subst\_domain}\SetKwFunction{Len}{len}
-
-   \KwData{old and new monomorphic type argument maps \(M, M^{\text{new}}\), old and new non-monomorphic type argument maps \(P(\varphi), P^{\text{new}}(\varphi)\)}
-   \KwResult{monomorphic type argument map \(M^{\text{next}}\), non-monomorphic type argument map \(P^{\text{next}}(\varphi)\), set of substitutions \(S\)}
-   \BlankLine
-
-   \Subst \(\leftarrow\) \SubstGen{\(M^{\text{new}}\), \(P^{\text{new}}(\varphi)\)}\;
-   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M^{\text{new}}\), \(P(\varphi)\)}\;
-   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M\), \(P^{\text{new}}(\varphi)\)}\;
-   \BlankLine
-
-   \TypeVars \(\leftarrow\) \TyVars{\(\varphi\)}\;
-
-   \textcolor{ourblueviolet}{
-   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M\cup M^{\text{new}}| \cdot \)\MonoMult}, \MonoCap}\;
-   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)\cup P^{\text{new}}(\varphi)| \cdot \)\PolyMult}, \PolyCap}\;
-   }
-
-   \BlankLine
-
-   \UsedSubst \(\leftarrow\emptyset\)\;
-   \ForEach{\(\sigma\in \Subst\)}{
-      \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
-            \eIf{\TyVars{\((\upsilon_1, \dots, \upsilon_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
-               \textcolor{ourblueviolet}{
-               \uIf{\(|M^{\text{next}}| < \MaxMono\) \label{algo:relative_bound}}{
-                  \textcolor{black}{
-                  \(M^{\text{next}}(f)\leftarrow M^{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
-                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
-               }
-               \uElseIf{\(|P^{\text{next}}(\varphi)| \geq \MaxPoly\) }{
-                  blablabliblou\;
-                  \Return \((M^{\text{next}}, P^{\text{next}}(\varphi), \UsedSubst)\)\;
-               }}
-            }
-            {
-               \textcolor{ourblueviolet}{
-               \If{\(|P^{\text{next}}(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
-               \textcolor{black}{
-               \(P^{\text{next}}(\varphi)(f)\leftarrow P^{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
-               \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
-            }}
-            }
-
-         }
-      }
-   }
-   \BlankLine
-
-   \Return \((M^{\text{next}}, P^{\text{next}}(\varphi), \UsedSubst)\)
-
-\end{algorithm}
-\end{quote}
-\caption{Formula Monomorphisation Step}
-\end{figure}
-
-This function computes new type arguments for a single formula.
-The separation of the type arguments into ``old'' and ``new'' maps is put to use here.
-Pairs of type argument tuples must contain at least one new type argument in order to be matched and generate a substitution.
-This avoids computing matchings that have already been made in previous iterations.
-
-The limits on the number of new type monomorphic arguments has three components (similar bounds are used for the non-monomorphic):
-\begin{enumerate}
-   \item \MonoCap is an absolute limit on the total number of new type arguments
-   \item \MonoMult is used to allow the total number of type arguments to increase by a certain proportion
-   \item \MonoFloor balances out \MonoMult when the number of type arguments is too low, and guarantees that a minimum number of new type arguments
-\end{enumerate}
-
-At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), adding conditions dependent on the size of \(M^{\text{next}}(f)\) (respectively \(P^{\text{next}}(\varphi)(f)\)), would create finer grained bounds.
-This may be useful if some function symbols have too many type arguments and prevent new type arguments of other function symbols from being generated.
-Adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments over function symbols.
-
-\begin{figure}
-\begin{quote}
-   \begin{algorithm}[H]
-
-   \SetKwFunction{MonoSubst}{monomorphising\_substs}
+   \SetKwFunction{MonoSubst}{mono\_substs}
    \SetKwFunction{Domain}{domain}
@@
 
-   \KwData{set of substitutions \(U\) and \(S\), single substitution \(\sigma\), set of type variables \(V\)}
+   \Fn(){\MonoSubst{\(S^{\text{used}}, S^\text{acc}, \sigma^\text{acc}, V\)}}{
+
+   \KwData{%
+      \hspace{-2.7em}\begin{minipage}[t]{.8\textwidth}
+      \begin{enumerate}
+         \item[] set of substitutions \(S^{\text{used}}\) and \(S^\text{acc}\)
+         \item[] single substitution \(\sigma^\text{acc}\)
+         \item[] set of type variables \(V\)
+      \end{enumerate}
+      \end{minipage}
+   }
+
    \KwResult{set of substitutions}
@@
    \eIf{\(V = \emptyset\)}{
-    \Return \(S\cup\{\sigma\}\)\;%}
+    \Return \(S^\text{acc}\cup\{\sigma^\text{acc}\}\)\;%}
    }
@@
       \Let \(\alpha \) such that \(\alpha\in V\)\;
-      \ForEach{\(\sigma'\in U\) such that, \(\alpha\in\Domain{\(\sigma'\)}\) \And \(\sigma\) compatible with \(\sigma'\)}{
+      \ForEach{\(\sigma^\text{next}\in S^{\text{used}}\) such that, \(\alpha\in\Domain{\(\sigma^\text{next}\)}\) \And \(\sigma^\text{acc}\) compatible with \(\sigma^\text{next}\)}{
          \textcolor{ourblueviolet}{
-         \eIf{\(|S|<\Limit\)}{
+         \eIf{\(|S^\text{acc}|<\Limit\)}{
             \textcolor{black}{
-            \(S\leftarrow \MonoSubst{U, S, \(\sigma'\circ\sigma\), \(V\setminus\Domain{\(\sigma'\)}\)}\)\;}
+            \(S^\text{acc}\leftarrow \MonoSubst{\(S^{\text{used}}, S^\text{acc}, \sigma^\text{next}\circ\sigma^\text{acc}, V\setminus\Domain{\(\sigma^\text{next}\)}\)}\)\;}
          }
-         {\Return \(S\)}}
+         {\Return \(S^\text{acc}\)}}
       }
-      \Return \(S\)\;
+      \Return \(S^\text{acc}\)\;
+   }
    }

### 2024-03-06 iterative_monomorphisation.tex
more work on the detailed algorithm section, corrected a certain number of spelling mistakes

@@
 
-We implemented iterative monormorphisation in Zipperpostion \cite{zipp}, a higher order prover written in OCaml.
+We implemented iterative monomorphisation in Zipperposition \cite{zipp}, a higher order prover written in OCaml.
 Although Zipperposition is polymorphic, it uses the monomorphic prover E \cite{e} as a backend.
@@
 
-   \item For each formula \(\varphi \in \Phi\) and for each symbol \(f\) occuring in \(\varphi\):
+   \item For each formula \(\varphi \in \Phi\) and for each symbol \(f\) occurring in \(\varphi\):
    \begin{enumerate}
@@
 
-Once all type arguments have been generated, the substitutions used to create those new type arguments are then combined in order to form monomorphising substitutions which are subsequently applied to the polymorphic formulae.
-There can be a high number of possible monomorphic formulae resulting from this process.
-This can be limited by setting a maximal number of new monomorphic formulae.
+Once all type arguments have been generated, the substitutions used to create these new type arguments are then combined in order to form monomorphising substitutions.
+These substitutions are subsequently applied to the input problem's polymorphic formulae.
+The possible number of new (monomorphic) formulae resulting from this process is very high but can be limited by setting a maximal number of new formulae.
 In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected when it is given too many clauses.
@@
 Function \ref{mono_step} computes new type arguments for a single formula.
+The first step in this process generates substitutions (detailed in \ref{subst_gen}) by matching non-monomorphic type arguments against monomorphic ones.
 The separation of the type arguments into ``old'' and ``new'' maps is exploited here.
-Pairs of type argument tuples must contain at least one new type argument in order to be matched and generate a substitution.
-This avoids computing matchings that have already been made in previous iterations.
-Unfortunately, not all such redundant computations can be avoided because different matchings with entirely different types can yield the same substitution.
-Furthermore,
+Pairs of type argument tuples must contain at least one new type argument in order to be matched and possibly generate a substitution.
+This avoids computing matchings having already been made in previous iterations.
+Not all such redundant computations can be avoided because the same substitution can be obtained from matchings between entirely different pairs of types.
 
-The limits on the number of new type monomorphic arguments has three components (similar bounds are used for the non-monomorphic):
+The main loop of this function, iterates over each substitution generated this way and applied them to all polymorphic type arguments.
+The total number of type arguments can increase (doubly) exponentially in the number of iterations.
+
+[If we assume there are after \(k\) iterations \(N_k\) total type arguments split evenly between monomorphic and polymorphic type arguments, all in a single formula.
+There can be up to \((\frac{N_k}{2})^2\) successful matches yielding as many substitutions.
+Each substitution is then applied to each polymorphic type argument for a total of \((\frac{N_k}{2})^3\) new type arguments.
+If half of these new type arguments are monomorphic and the other half is polymorphic, then the next iteration will begin with \(N_{k+1} = (\frac{N_k}{2})^{3}\) evenly split type arguments.
+Meaning that the total number of type arguments on the \(k^\text{th}\) iteration can reach \(N_k = (\frac{N_0}{2})^{3^k}\).
+This paragraph is wrong on several points, it can be fixed but it would be annoying, the main idea is that it can be shown relatively easily in a semi-formal way that the total number of type arguments is doubly exponential in the number of iterations.]
+
+Depending on the shape and size of the input problem, this may problematic and lead to an explosion of the number of type arguments.
+The limits on the number of new type monomorphic arguments (in \textcolor{ourblueviolet}{blue}) has three components (similar bounds are used for the non-monomorphic type arguments):
 \begin{enumerate}
-   \item \MonoCap is an absolute limit on the total number of new type arguments
-   \item \MonoMult is used to allow the total number of type arguments to increase by a certain proportion
-   \item \MonoFloor balances out \MonoMult when the number of type arguments is too low, and guarantees that a minimum number of new type arguments
+   \item \textcolor{ourblueviolet}{\MonoCap} is an absolute limit on the total number of new type arguments
+   \item \textcolor{ourblueviolet}{\MonoMult} is used to allow the total number of type arguments to increase by a certain proportion
+   \item \textcolor{ourblueviolet}{\MonoFloor} balances out \MonoMult when the number of type arguments is too low, and guarantees that a minimum number of new type arguments
 \end{enumerate}
 
-At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), adding conditions dependent on the size of \(M^{\text{next}}(f)\) (respectively \(P^{\text{next}}(\varphi)(f)\)), would create finer grained bounds.
-This may be useful if some function symbols have too many type arguments and prevent new type arguments of other function symbols from being generated.
-Adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments over function symbols.
+Additionally, it is possible to generalise or further refine these type argument bounds.
+Indeed, the maximum number of newly generated type arguments is fixed per iteration, per formula.
+This algorithm can easily be modified such that the limit on new type arguments is fixed per iteration, or conversely, be made to limit new type arguments for each function symbol of each formula for each iteration.
+The coarseness and values of these bounds can be adapted to the kinds of problems to be processed by the monomorphisation algorithm.
 
@@
 \Fn(){\SubstGen{\(M, P(\varphi)\)}}{
-%   \caption{Substitution Generation()}
 %   \SetAlgoNoLine
@@
 \caption{Pseudocode for substitution generation}
+\label{subst_gen}
 \end{figure}
 
-This function computes all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions.
-For each pair of type argument tuples, the non-monomrophic types are matched against their monomorphic counterparts and the resulting substitutions are composed if they are compatible.
+Substitution generation (\ref{subst_gen}) consists of computing all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions.
+For each pair of type argument tuples, the non-monomorphic types are matched against their monomorphic counterparts and the resulting substitutions are composed if they are compatible.
+In the pseudocode, this is checked by making sure that the while loop has iterated over all elements of the type argument tuple.
 If the substitutions are incompatible, the matchings are discarded.
-[Note that the composition of two compatible substitutions is commutative].
+[Note that the composition of two compatible substitutions is commutative and that the order of composition is therefore irrelevant].
+[The Zipperposition implementation is written in a functional style and may therefore not directly map onto the pseudocode].
 
@@
 In this case, it is a simple matter of stopping the function and returning \(S\) once a certain limit is reached.
+
+The various bounds presented are, to some extent, redundant.
+For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type arguments, making these bounds superfluous.
+Whilst, implementing all the bounds described in this section may not be necessary, all the limits presented may be useful in particular cases.
+For example, sets of problems which lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst sets of problems where substitution generation is a bottleneck might consider a limit on the number of generated substitution to be more relevant.
+Not sure about this paragraph
+
 \begin{figure}

### 2024-03-07 iterative_monomorphisation.tex
finished first draft of detailed algorithm description

--- 2024-03-05 iterative_monomorphisation.tex
@@
 %   - clause -> formula
+%   - terms in ``xxx'' -> \emph{xxx}
 %   - monomorphization -> monomorphisation
@@
 
-    \item[\labelitemi] Substitutions are directly applied to the type arguments in $M$ and $P$ instead of the formulae.
+    \item[\labelitemi] Substitutions are directly applied to the type arguments instead of the formulae.
 This dispenses from having to re-extract the type arguments from the formulae at each iteration.
-New formulae are only generated once all iterations are complete, in a separate, penultimate step.
+New formulae are only generated once all iterations are complete, in a separate, final step.
 
-    \item[\labelitemi] Once the new formulae are generated, their monomorphic types are mangled before the end result is output.
 \end{enumerate}
 
-These algorithms rely on ``primitives'' whose implementation depends on the specifics of the grammar and logic used.
+This algorithm relies on ``primitives'' whose implementation depend on the specifics of the grammar and logic used.
 They will therefore not be expanded upon.
-The functions computing the following are assumed to exist:
+Functions computing the following are assumed to be available:
 \begin{enumerate}
-   \item[\labelitemi] computing the initial type argument maps
-   \item[\labelitemi] extracting type variables from a type tuple
+   \item[\labelitemi] extracting the initial type argument maps from the input problem
+   \item[\labelitemi] gathering type variables from a type tuple
    \item[\labelitemi] type matching
+   \item[\labelitemi] returning the type variables instantiated by a substitution
+   \item[\labelitemi] testing the compatibility of two substitutions
    \item[\labelitemi] substitution composition
-   \item[\labelitemi] testing the compatibility of two substitutions
-   \item[\labelitemi] returning the type variables instantiated by a substitution
    \item[\labelitemi] substitution application to formulae
@@
 
-The iterative monomorphisation algorithm, as implemented in Zipperposition, is described in \ref{iter_mono}.
-There are two main phases each consisting of a separate loop.
-The first loop applies a monomorphisation step to each formula for each iteration of the overall for loop until the user-set limit it reached.
+The iterative monomorphisation algorithm (\ref{iter_mono}), as implemented in Zipperposition, has two main phases each consisting of a separate loop.
+The first loop applies a \emph{monomorphisation step} to each formula (\ref{mono_step}) for each iteration of the outermost for loop until the user-set limit it reached.
 This bound is the only limit which is necessary for the algorithm to terminate.
-At the end of each iteration, the old and new maps are updated.
+At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
 %Updating the maps progressively over the course of an iteration would be feasible but would result in an uneven treatment of formulae.
@@
 
-Once all type arguments have been generated, the substitutions used to create those new type arguments are then combined in order to form monomorphising substitutions which are subsequently applied to the polymorphic formulae.
-There can be a high number of possible monomorphic formulae resulting from this process.
-This can be limited by setting a maximal number of new monomorphic formulae.
-In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected when it is given too many clauses.
+Once the first phase is completed, the substitutions used to create new type arguments are combined to form \emph{monomorphising substitutions}.
+These substitutions are subsequently applied to the input problem's polymorphic formulae.
+The possible number of new (monomorphic) formulae resulting from this process is high but can be limited by setting a maximal number of new formulae.
+In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of clauses it is given.
 
@@
 
-Function \ref{mono_step} computes new type arguments for a single formula.
+Formula monomorphisation (\ref{mono_step}) computes new type arguments for a single formula, it is the core of the monomorphisation process.
+The first step generates substitutions (\ref{subst_gen}) by matching the non-monomorphic type arguments against monomorphic ones.
 The separation of the type arguments into ``old'' and ``new'' maps is exploited here.
-Pairs of type argument tuples must contain at least one new type argument in order to be matched and generate a substitution.
-This avoids computing matchings that have already been made in previous iterations.
-Unfortunately, not all such redundant computations can be avoided because different matchings with entirely different types can yield the same substitution.
-Furthermore,
+Pairs of type argument tuples must contain at least one new type argument in order to be matched.
+This avoids re-computing some matchings already processed in previous iterations.
+Not all such redundant computations can be avoided because the same substitution can be obtained from matchings between entirely different pairs of types.
 
-The limits on the number of new type monomorphic arguments has three components (similar bounds are used for the non-monomorphic):
+The set of substitutions generated thusly is then iterated over and each substitution is applied to all polymorphic type arguments.
+The application of a substitution to a polymorphic type argument will only take place if any type variable of the type argument is instantiated by the substitution.
+This condition filters out all substitution application that would have left the type arguments unchanged.
+
+%[If we assume there are after \(k\) iterations \(N_k\) total type arguments split evenly between monomorphic and polymorphic type arguments, all in a single formula.
+There can be up to \((\frac{N_k}{2})^2\) successful matches yielding as many substitutions.
+Each substitution is then applied to each polymorphic type argument for a total of \((\frac{N_k}{2})^3\) new type arguments.
+If half of these new type arguments are monomorphic and the other half is polymorphic, then the next iteration will begin with \(N_{k+1} = (\frac{N_k}{2})^{3}\) evenly split type arguments.
+Meaning that the total number of type arguments on the \(k^\text{th}\) iteration can reach \(N_k = (\frac{N_0}{2})^{3^k}\).
+This paragraph is wrong on several points, it can be fixed but it would be annoying, the main idea is that it can be shown relatively easily in a semi-formal way that the total number of type arguments is doubly exponential in the number of iterations.]
+
+The total number of type arguments can increase cubicly in the number of type arguments at each iteration and can therefore grow doubly exponentially in the number of iterations.
+Depending on the shape and size of the input problem and the number of iterations performed, this may be problematic.
+The limit on the number of new monomorphic type arguments (in \textcolor{ourblueviolet}{blue}) has three components (similar bounds are used for the non-monomorphic type arguments):
 \begin{enumerate}
-   \item \MonoCap is an absolute limit on the total number of new type arguments
-   \item \MonoMult is used to allow the total number of type arguments to increase by a certain proportion
-   \item \MonoFloor balances out \MonoMult when the number of type arguments is too low, and guarantees that a minimum number of new type arguments
+   \item \textcolor{ourblueviolet}{\MonoCap} is an absolute limit on the total number of new type arguments
+   \item \textcolor{ourblueviolet}{\MonoMult} is used to allow the total number of type arguments to increase by a certain proportion
+   \item \textcolor{ourblueviolet}{\MonoFloor} balances out \MonoMult when the number of type arguments is too low, and guarantees that a minimum number of new type arguments
 \end{enumerate}
 
-At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), adding conditions dependent on the size of \(M^{\text{next}}(f)\) (respectively \(P^{\text{next}}(\varphi)(f)\)), would create finer grained bounds.
-This may be useful if some function symbols have too many type arguments and prevent new type arguments of other function symbols from being generated.
-Adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments over function symbols.
+Additionally, it is possible to either generalise or further refine these bounds.
+Indeed, the maximum number of newly generated type arguments is fixed per formula, per iteration.
+This algorithm can easily be modified such that the limit on new type arguments is fixed per iteration, or conversely, be made to limit new type arguments per symbol, per formula, per iteration.
 
@@
 \Fn(){\SubstGen{\(M, P(\varphi)\)}}{
-%   \caption{Substitution Generation()}
-%   \SetAlgoNoLine
-%   \LinesNumbered
-   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
+   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
    \SetKwData{SubstLimit}{substitution\_limit}
@@
 \caption{Pseudocode for substitution generation}
+\label{subst_gen}
 \end{figure}
 
-This function computes all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions.
-For each pair of type argument tuples, the non-monomrophic types are matched against their monomorphic counterparts and the resulting substitutions are composed if they are compatible.
+Substitution generation (\ref{subst_gen}) consists in computing all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions.
+For each pair of type argument tuples, the non-monomorphic types are matched against their monomorphic counterparts and the resulting substitutions are composed if they are compatible.
+In the pseudocode, this is checked by making sure that the while loop has iterated over all elements of the type argument tuple.
 If the substitutions are incompatible, the matchings are discarded.
-[Note that the composition of two compatible substitutions is commutative].
+[Note that the composition of two compatible substitutions is commutative and that the order of composition is therefore irrelevant].
+[The Zipperposition implementation is written in a functional style and does not directly map onto the pseudocode].
 
-Adding a limit to the number of substitutions generated may be necessary if substitution generation becomes a bottleneck.
-In this case, it is a simple matter of stopping the function and returning \(S\) once a certain limit is reached.
+The various bounds presented here are, to some extent, redundant.
+For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type arguments, making the related bounds superfluous.
+Each bound nonetheless has its use cases.
+For example, sets of problems which lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst sets of problems where substitution generation is a bottleneck might consider a limit on the number of generated substitution to be more useful.
+
 \begin{figure}
@@
 
-   \Fn(){\MonoSubst{\(S^{\text{used}}, S^\text{acc}, \sigma^\text{acc}, V\)}}{
+   \Fn(){\MonoSubst{\(S^{\text{all}}, S^\text{acc}, \sigma^\text{acc}, V\)}}{
 
@@
       \begin{enumerate}
-         \item[] set of substitutions \(S^{\text{used}}\) and \(S^\text{acc}\)
+         \item[] set of substitutions \(S^{\text{all}}\) and \(S^\text{acc}\)
          \item[] single substitution \(\sigma^\text{acc}\)
@@
       \Let \(\alpha \) such that \(\alpha\in V\)\;
-      \ForEach{\(\sigma^\text{next}\in S^{\text{used}}\) such that, \(\alpha\in\Domain{\(\sigma^\text{next}\)}\) \And \(\sigma^\text{acc}\) compatible with \(\sigma^\text{next}\)}{
+      \ForEach{\(\sigma^\text{next}\in S^{\text{all}}\) such that, \(\alpha\in\Domain{\(\sigma^\text{next}\)}\) \And \(\sigma^\text{acc}\) compatible with \(\sigma^\text{next}\)}{
          \textcolor{ourblueviolet}{
@@
             \textcolor{black}{
-            \(S^\text{acc}\leftarrow \MonoSubst{\(S^{\text{used}}, S^\text{acc}, \sigma^\text{next}\circ\sigma^\text{acc}, V\setminus\Domain{\(\sigma^\text{next}\)}\)}\)\;}
+            \(S^\text{acc}\leftarrow \MonoSubst{\(S^{\text{all}}, S^\text{acc}, \sigma^\text{next}\circ\sigma^\text{acc}, V\setminus\Domain{\(\sigma^\text{next}\)}\)}\)\;}
          }
@@
 \caption{Monomorphising Substitutions}
+\label{mono_substs}
 \end{figure}
 
-This recursive function computes all substitutions that can instantiate a set of type variables.
-The substitutions are obtained by composition of compatible substitutions from a given set.
-The returned substitutions are said to be ``monomorphising'' with regards to the set of input type variables \(V\) because they instantiate all type variables in \(V\).
-In the Zipperposition implementation this limit was one of the most important bounds because the total number of possible monomorphising substitutions is very high but very few are necessary in practice.
+Once all monomorphisation iterations have been completed, we are left with a set of substitutions.
+The last phase uses these substitutions to instantiate the type variables in the input problem's polymorphic formuale.
 
-In the Zipperposition implementation, due the substitutions in \(U\) being selected on the basis on which type variable they instantiate, all substitutions are kept in a map which makes this selection faster.
+To monomorphise a polymorphic formula, we first compute its \emph{monomorphising substitutions} (\ref{mono_substs}).
+A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
+\ref{mono_substs} is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S^\text{all}\), it will select a substitution \(\sigma^\text{next}\) from \(S^\text{all}\) which instantiates at least one of the type variables in \(V\).
+It is important that \(\sigma^\text{next}\) be compatible with \(\sigma^\text{acc}\) so that they may be composed and the function recursively called to instantiate the remaining type variables.
 
-Because this function often hits the substitution limit, the order of substitutions in \(U\) (or the map in Zipperposition) can have a significant impact on the resulting monomorphising substitutions.
-This may be a relevant heuristic to tweak depending on the input problem.
+The Zipperposition implementation of this function makes use of a map from type variables to substitutions in place of a simple set to filter the relevant substitutions from \(S^\text{all}\) with greater ease.
+The \textcolor{ourblueviolet}{subst\_limit} exists for two main reasons:
+\begin{enumerate}
+   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{subst\_limit} new monomorphic clause per initial polymorphic clause.
+Generating an excessive number of new clauses can lead to the downstream prover being flooded.
+   \item This function is the most explosive part of the algorithm.
+If \(S^\text{all}\) contains \(n\) substitutions which each instantiate exactly one of \(v\) different type variable, up to \(n^v\) monomorphising substititions may be generated.
+This should be considered in light of the fact that the total number of type variables (used to generate these substitutions) is already doubly exponential in the number of loop iterations.
+\end{enumerate}
+
+%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S^\text{acc}\) would represent the leaves that have already been visited and \(\sigma^\text{acc}\) would represent the current node of the tree.]
+
+Once the newly generated monomorphic formulae are output, an extra post-processing step may be required.
+Indeed, monomorphic provers, such as E do not necessarily support type constructors.
+Consequently, transforming the monomorphic formulae into equivalent ones without type constructors may be necessary as it was for the Zipperposition implementation.
+This transformation is largely syntactic, it is documented in [insert names] 's paper [insert citation].
 

### 2024-03-08 iterative_monomorphisation.tex
edits with Tanguy

@@
 
-\item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(P(\varphi, f)\) consists of all tuples of type arguments to which \(f\) is applied in \(\varphi\) and which contains a type variable.
-For example, if \(\sym{foldl}\langle \ty{nat}, \ty{list}(\alpha)\rangle\) occurs in \(\varphi\), then \((\ty{nat}, \ty{list}(\alpha)) \in P(\varphi, \sym{foldl}) \).
+\item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(P(\varphi)(f)\) consists of all tuples of type arguments to which \(f\) is applied in \(\varphi\) and which contains a type variable.
+For example, if \(\sym{foldl}\langle \ty{nat}, \ty{list}(\alpha)\rangle\) occurs in \(\varphi\), then \((\ty{nat}, \ty{list}(\alpha)) \in P(\varphi)(\sym{foldl}) \).
 \end{enumerate}
@@
    \begin{enumerate}
-    \item[2.1.] For each tuple \((\tau_1, \dots, \tau_n) \in  M(f)\)        and each tuple \((\upsilon_1, \dots, \upsilon_n) \in P(\varphi, f)\),
+    \item[2.1.] For each tuple \((\tau_1, \dots, \tau_n) \in  M(f)\)        and each tuple \((\upsilon_1, \dots, \upsilon_n) \in P(\varphi)(f)\),
      for each \(i\), match \(\upsilon_i\) against \(\tau_i\), yielding the substitution \(\sigma_i\) in case of success.
@@
 \SetKwFunction{IterMono}{iterative\_monomorphisation}
-\begin{figure}
+\begin{figure}[b]
 \begin{quote}
 \begin{algorithm}[H]
-\Fn(){\IterMono{\(\Phi\)}}{
+\Fn(){\(\IterMono(\Phi)\)}{
    \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{To}{to}
 
-   \SetKwData{Loop}{loop\_nb}\SetKwData{NewFormulae}{\(\Psi\)}
-   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(S^{\text{used}}\)}
+   \SetKwData{Loop}{num\_loops}\SetKwData{NewFormulae}{\(\Psi\)}
+   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(S\)}
    \SetKwData{Limit}{total\_new\_formulae}
@@
    \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{mono\_substs}
+   \SetKwFunction{GenFormulae}{generate\_mono\_formulae}
 
-   \KwData{set of formulae \(\Phi\)}
-   \KwResult{set of monomorphic formulae}
+   \KwData{finite set \(\Phi\) of polymorphic formulae}
+   \KwResult{finite set of monomorphic formulae}
 
-   \((M^{\text{old}}, P^{\text{old}}) \leftarrow (\emptyset, \emptyset)\)\;
-   \((M^{\text{new}}, P^{\text{new}}) \leftarrow \Init{\Problem}\)\;
+   \BlankLine
+
+   \((M_{\text{old}}, P_{\text{old}}) \leftarrow (\emptyset, \emptyset)\)\;
+   \((M_{\text{new}}, P_{\text{new}}) \leftarrow \Init(\Problem)\)\;
 
@@
 
+   \BlankLine
+
    \For{\(i = 1\) \To \textcolor{ourblueviolet}{\Loop}}{
-      \((M^{\text{next}}, P^{\text{next}})\leftarrow(\emptyset,\emptyset)\)\;
+      \((M_{\text{next}}, P_{\text{next}})\leftarrow(\emptyset,\emptyset)\)\;
       \ForEach{\(\varphi \in \Problem\)} {
-         \((M^{\Delta}, P^{\Delta}(\varphi), S^\Delta)\leftarrow \FStep(M^{\text{old}}, P^{\text{old}}(\varphi), M^{\text{new}}, P^{\text{new}}(\varphi))\)\;
-         \(\AllSubst\leftarrow\AllSubst\cup S^\Delta\)\;
-         \(M^{\text{next}}\leftarrow M^{\text{next}}\cup M^{\Delta}\)\;
-         \(P^{\text{next}}(\varphi)\leftarrow P^{\text{next}}(\varphi)\cup P^{\Delta}(\varphi)\)\;
+         \((M_{\Delta}, P_{\Delta}(\varphi), S_{\Delta})\leftarrow \FStep(M_{\text{old}}, P_{\text{old}}(\varphi), M_{\text{new}}, P_{\text{new}}(\varphi))\)\;
+         \(\AllSubst\leftarrow\AllSubst\cup S_{\Delta}\)\;
+         \(M_{\text{next}}\leftarrow M_{\text{next}}\cup M_{\Delta}\)\;
+         \(P_{\text{next}}(\varphi)\leftarrow P_{\text{next}}(\varphi)\cup P_{\Delta}(\varphi)\)\;
       }
 
-      \((M^{\text{old}}, M^{\text{new}})\leftarrow (M^{\text{old}}\cup M^{\text{new}}, M^{\text{next}})\)\;
-      \((P^{\text{old}}, P^{\text{new}})\leftarrow (P^{\text{old}}\cup P^{\text{new}}, P^{\text{next}})\)\;
+      \((M_{\text{old}}, M_{\text{new}})\leftarrow (M_{\text{old}}\cup M_{\text{new}}, M_{\text{next}})\)\;
+      \((P_{\text{old}}, P_{\text{new}})\leftarrow (P_{\text{old}}\cup P_{\text{new}}, P_{\text{next}})\)\;
 
@@
 
+   \BlankLine
+
+   \Return \(\GenFormulae(\Phi, S)\)
+}
+\end{algorithm}
+\end{quote}
+\caption{Pseudocode for iterative monomorphisation}
+\label{iter_mono}
+\end{figure}
+
+\SetKwFunction{GenFormulae}{generate\_mono\_formulae}
+\begin{figure}
+\begin{quote}
+\begin{algorithm}[H]
+\Fn(){\(\GenFormulae(\Phi, S)\)}{
+   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{To}{to}
+
+   \SetKwData{Loop}{num\_loops}\SetKwData{NewFormulae}{\(\Psi\)}
+   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(S\)}
+   \SetKwData{Limit}{total\_new\_formulae}
+
+   \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step} 
+   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{mono\_substs}
+   \SetKwFunction{GenFormulae}{generate\_formulae}
+
+   \KwData{\begin{minipage}[t]{.8\textwidth}
+     \strut finite set \(\Phi\) of polymorphic formulae \\
+     finite set $S$ of substitutions\strut
+     \end{minipage}}
+   \KwResult{finite set of monomorphic formulae}
+
+   \BlankLine
+
    \NewFormulae \(\leftarrow\emptyset\)\;
-   \ForEach{\(\varphi\in\Phi\) polymorphic}{
-      \(V\leftarrow \TyVars{\(\varphi\)}\)\;
-      \(S^{\text{mono}}\leftarrow\MonoSubst{\AllSubst, \(\emptyset\), \(\{\}\), \(V\)}\)\;
-      \ForEach{\(\sigma\in S^{\text{mono}}\)}{
+
+   \BlankLine
+
+   \ForEach{\(\varphi\in\Phi\) such that $\varphi$ is non-monomorphic}{
+      \ForEach{\(\sigma\in \MonoSubst{\AllSubst, \(\TyVars{\(\varphi\)}\), \(\emptyset\), \(\{\}\)}\)}{
          \textcolor{ourblueviolet}{
@@
    }
-   \Return \NewFormulae\;
 
+   \BlankLine
+
+   \Return \NewFormulae
 }
@@
 \end{quote}
-\caption{Pseudocode for iterative monomorphisation}
-\label{iter_mono}
+\caption{Pseudocode for monomorphic formula generation}
+\label{genformulae}
 \end{figure}
@@
    %\LinesNumbered
-   \SetKwFunction{SubstGen}{subst\_generation}\SetKwFunction{TyVars}{type\_variables}
+   \SetKwFunction{SubstGen}{matches}\SetKwFunction{TyVars}{type\_variables}
    \SetKwFunction{FMonoStep}{formula\_mono\_step}
 
-   \Fn(){\FMonoStep{\(M^{\text{old}}, M^{\text{new}}, P^{\text{old}}(\varphi), P^{\text{new}}(\varphi)\)}}{
-   \SetKwData{Subst}{substs}\SetKwData{TypeVars}{type\_variables}
+   \Fn(){\FMonoStep{\(M_{\text{old}}, M_{\text{new}}, P_{\text{old}}(\varphi), P_{\text{new}}(\varphi)\)}}{
+   \SetKwData{TypeVars}{type\_variables}
    \SetKwData{MaxMono}{max\_mono\_args}
    \SetKwData{MaxPoly}{max\_poly\_args}
-   \SetKwData{UsedSubst}{\(S^{\text{used}}\)}
+   \SetKwData{UsedSubst}{\(S\)}
 
@@
 
-   \KwData{
-      \hspace{-2.7em}\begin{minipage}[t]{.8\textwidth}
-      \begin{enumerate}
-         \item[] old and new monomorphic type argument maps \(M^{\text{old}}, M^{\text{new}}\)
-         \item[] old and new non-monomorphic type argument maps \(P^{\text{old}}(\varphi), P^{\text{new}}(\varphi)\)
-      \end{enumerate}
+   \KwData{\begin{minipage}[t]{.8\textwidth}
+     \strut old and new monomorphic type argument maps \(M_{\text{old}}, M_{\text{new}}\) \\
+     old and new non-monomorphic type argument maps \(P_{\text{old}}(\varphi), P_{\text{new}}(\varphi)\)\strut
       \end{minipage}
    }
-   \KwResult{
-      \hspace{-2.7em}\begin{minipage}[t]{.8\textwidth}
-      \begin{enumerate}
-         \item[] monomorphic type argument map
-         \item[] non-monomorphic type argument map
-         \item[] set of substitutions
-      \end{enumerate}
+   \KwResult{\begin{minipage}[t]{.8\textwidth}
+      \strut monomorphic type argument map \\
+      non-monomorphic type argument map \\
+      set of substitutions\strut
       \end{minipage}
    }
-   \BlankLine
-
-   \Subst \(\leftarrow\) \SubstGen{\(M^{\text{new}}\), \(P^{\text{new}}(\varphi)\)}\;
-   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M^{\text{new}}\), \(P(\varphi)\)}\;
-   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M\), \(P^{\text{new}}(\varphi)\)}\;
    \BlankLine
@@
    \textcolor{ourblueviolet}{
-   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M\cup M^{\text{new}}| \cdot \)\MonoMult}, \MonoCap}\;
-   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)\cup P^{\text{new}}(\varphi)| \cdot \)\PolyMult}, \PolyCap}\;
+   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M_{\text{old}}\cup M_{\text{new}}| \cdot \MonoMult\)}, \MonoCap}\;
+   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P_{\text{old}}(\varphi)\cup P_{\text{new}}(\varphi)| \cdot \PolyMult\)}, \PolyCap}\;
    }
@@
    \UsedSubst \(\leftarrow\emptyset\)\;
-   \ForEach{\(\sigma\in \Subst\)}{
-      \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
+
+   \BlankLine
+
+   \ForEach{\(\sigma\in {}\)\SubstGen{\(M_{\text{new}}\), \(P_{\text{new}}(\varphi)\)}
+\(\cup\) \SubstGen{\(M_{\text{new}}\), \(P_{\text{old}}(\varphi)\)}
+\(\cup\) \SubstGen{\(M_{\text{old}}\), \(P_{\text{new}}(\varphi)\)}}{
+      \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P_{\text{old}}(\varphi)\)}{
             \eIf{\TyVars{\((\upsilon_1, \dots, \upsilon_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
                \textcolor{ourblueviolet}{
-               \uIf{\(|M^{\text{next}}| < \MaxMono\) \label{algo:relative_bound}}{
+               \uIf{\(|M_{\text{next}}| < \MaxMono\) \label{algo:relative_bound}}{
                   \textcolor{black}{
-                  \(M^{\text{next}}(f)\leftarrow M^{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
+                  \(M_{\text{next}}(f)\leftarrow M_{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
                   \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
                }
-               \uElseIf{\(|P^{\text{next}}(\varphi)| \geq \MaxPoly\) }{
-                  blablabliblou\;
-                  \(M^\text{next} \leftarrow M^{\text{next}}\setminus (M^\text{old} \cup M^\text{new})\)\;
-                  \(P^\text{next}(\varphi)\leftarrow P^{\text{next}}(\varphi)\setminus (P^\text{old}(\varphi)\cup P^\text{new}(\varphi))\)\;
-                  \Return \((M^\text{next}, P^\text{next}(\varphi), \UsedSubst)\)
+               \uElseIf{\(|P_{\text{next}}(\varphi)| \geq \MaxPoly\) }{
+                  \(M_{\text{next}} \leftarrow M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}})\)\;
+                  \(P_{\text{next}}(\varphi)\leftarrow P_{\text{next}}(\varphi)\setminus (P_{\text{old}}(\varphi)\cup P_{\text{new}}(\varphi))\)\;
+                  \Return \((M_{\text{next}}, P_{\text{next}}(\varphi), \UsedSubst)\)
                }}
@@
                \textcolor{ourblueviolet}{
-               \If{\(|P^{\text{next}}(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
+               \If{\(|P_{\text{next}}(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
                \textcolor{black}{
-               \(P^{\text{next}}(\varphi)(f)\leftarrow P^{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
+               \(P_{\text{next}}(\varphi)(f)\leftarrow P_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
                \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
@@
 
-   \Return \((M^{\text{next}}\setminus (M^\text{old} \cup M^\text{new}), P^{\text{next}}(\varphi)\setminus (P^\text{old}(\varphi)\cup P^\text{new}(\varphi)), \UsedSubst)\)
+   \Return \((M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}}), P_{\text{next}}(\varphi)\setminus (P_{\text{old}}(\varphi)\cup P_{\text{new}}(\varphi)), \UsedSubst)\)
 }
@@
 \end{quote}
-\caption{Formula Monomorphisation Step}
+\caption{Pseudocode for formula monomorphisation step}
 \label{mono_step}
@@
 \begin{algorithm}[H]
-\SetKwFunction{SubstGen}{substitution\_generation}
+\SetKwFunction{SubstGen}{matches}
 \Fn(){\SubstGen{\(M, P(\varphi)\)}}{
@@
 
-   \KwData{%
-      \hspace{-2.7em}\begin{minipage}[t]{.8\textwidth}
-      \begin{enumerate}
-         \item[] monomorphic type argument map \(M\)
-         \item[] non-monomorphic type argument map \(P(\varphi)\)
-      \end{enumerate}
+   \KwData{\begin{minipage}[t]{.8\textwidth}
+      \strut monomorphic type argument map \(M\) \\
+      non-monomorphic type argument map \(P(\varphi)\)\strut
       \end{minipage}
@@
 
-   \Fn(){\MonoSubst{\(S^{\text{all}}, S^\text{acc}, \sigma^\text{acc}, V\)}}{
+   \Fn(){\MonoSubst{\(S_{\text{all}}, S^\text{acc}, \sigma^\text{acc}, V\)}}{
 
@@
       \begin{enumerate}
-         \item[] set of substitutions \(S^{\text{all}}\) and \(S^\text{acc}\)
+         \item[] set of substitutions \(S_{\text{all}}\) and \(S^\text{acc}\)
          \item[] single substitution \(\sigma^\text{acc}\)
@@
       \Let \(\alpha \) such that \(\alpha\in V\)\;
-      \ForEach{\(\sigma^\text{next}\in S^{\text{all}}\) such that, \(\alpha\in\Domain{\(\sigma^\text{next}\)}\) \And \(\sigma^\text{acc}\) compatible with \(\sigma^\text{next}\)}{
+      \ForEach{\(\sigma_{\text{next}}\in S_{\text{all}}\) such that, \(\alpha\in\Domain{\(\sigma_{\text{next}}\)}\) \And \(\sigma^\text{acc}\) compatible with \(\sigma_{\text{next}}\)}{
          \textcolor{ourblueviolet}{
@@
             \textcolor{black}{
-            \(S^\text{acc}\leftarrow \MonoSubst{\(S^{\text{all}}, S^\text{acc}, \sigma^\text{next}\circ\sigma^\text{acc}, V\setminus\Domain{\(\sigma^\text{next}\)}\)}\)\;}
+            \(S^\text{acc}\leftarrow \MonoSubst{\(S_{\text{all}}, S^\text{acc}, \sigma_{\text{next}}\circ\sigma^\text{acc}, V\setminus\Domain{\(\sigma_{\text{next}}\)}\)}\)\;}
          }
@@
 \end{quote}
-\caption{Monomorphising Substitutions}
+\caption{Pseudocode for monomorphising substitution generation}
 \label{mono_substs}
@@
 A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
-\ref{mono_substs} is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S^\text{all}\), it will select a substitution \(\sigma^\text{next}\) from \(S^\text{all}\) which instantiates at least one of the type variables in \(V\).
-It is important that \(\sigma^\text{next}\) be compatible with \(\sigma^\text{acc}\) so that they may be composed and the function recursively called to instantiate the remaining type variables.
+\ref{mono_substs} is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S^\text{all}\), it will select a substitution \(\sigma_{\text{next}}\) from \(S^\text{all}\) which instantiates at least one of the type variables in \(V\).
+It is important that \(\sigma_{\text{next}}\) be compatible with \(\sigma^\text{acc}\) so that they may be composed and the function recursively called to instantiate the remaining type variables.
 

### 2024-03-08 iterative_monomorphisation.tex
modifs with Tanguy

@@
 \newcommand\Medleftrightarrow{\mathrel{\Leftarrow\kern-1.685ex\Rightarrow}}
+
+\def\negvthinspace{\kern-0.083333em}
+\def\vthinspace{\kern+0.083333em}
 
@@
 
+\textbf{TODO: Mention that all sets referenced in the algorithm are finite.}
+
 The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a lower level algorithm with the following features.
@@
 
-   \KwData{finite set \(\Phi\) of polymorphic formulae}
-   \KwResult{finite set of monomorphic formulae}
+   \KwData{set \(\Phi\) of polymorphic formulae}
+   \KwResult{set of monomorphic formulae}
 
@@
 
-   \((M_{\text{old}}, P_{\text{old}}) \leftarrow (\emptyset, \emptyset)\)\;
-   \((M_{\text{new}}, P_{\text{new}}) \leftarrow \Init(\Problem)\)\;
+   \((M_{\text{old}}, N_{\text{old}}) \leftarrow (\emptyset, \emptyset)\)\;
+   \((M_{\text{new}}, N_{\text{new}}) \leftarrow \Init(\Problem)\)\;
 
@@
    \For{\(i = 1\) \To \textcolor{ourblueviolet}{\Loop}}{
-      \((M_{\text{next}}, P_{\text{next}})\leftarrow(\emptyset,\emptyset)\)\;
+      \((M_{\text{next}}, N_{\text{next}})\leftarrow(\emptyset,\emptyset)\)\;
       \ForEach{\(\varphi \in \Problem\)} {
-         \((M_{\Delta}, P_{\Delta}(\varphi), S_{\Delta})\leftarrow \FStep(M_{\text{old}}, P_{\text{old}}(\varphi), M_{\text{new}}, P_{\text{new}}(\varphi))\)\;
+         \((M_{\Delta}, N_{\Delta}(\varphi), S_{\Delta})\leftarrow \FStep(M_{\text{old}}, N_{\text{old}}(\varphi), M_{\text{new}}, N_{\text{new}}(\varphi))\)\;
          \(\AllSubst\leftarrow\AllSubst\cup S_{\Delta}\)\;
          \(M_{\text{next}}\leftarrow M_{\text{next}}\cup M_{\Delta}\)\;
-         \(P_{\text{next}}(\varphi)\leftarrow P_{\text{next}}(\varphi)\cup P_{\Delta}(\varphi)\)\;
+         \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\cup N_{\Delta}(\varphi)\)\;
       }
@@
       \((M_{\text{old}}, M_{\text{new}})\leftarrow (M_{\text{old}}\cup M_{\text{new}}, M_{\text{next}})\)\;
-      \((P_{\text{old}}, P_{\text{new}})\leftarrow (P_{\text{old}}\cup P_{\text{new}}, P_{\text{next}})\)\;
+      \((N_{\text{old}}, N_{\text{new}})\leftarrow (N_{\text{old}}\cup N_{\text{new}}, N_{\text{next}})\)\;
 
@@
    \KwData{\begin{minipage}[t]{.8\textwidth}
-     \strut finite set \(\Phi\) of polymorphic formulae \\
-     finite set $S$ of substitutions\strut
+     \strut set \(\Phi\) of polymorphic formulae \\
+     set $S$ of substitutions\strut
      \end{minipage}}
-   \KwResult{finite set of monomorphic formulae}
+   \KwResult{set of monomorphic formulae}
 
@@
 
-   \ForEach{\(\varphi\in\Phi\) such that $\varphi$ is non-monomorphic}{
+   \ForEach{\(\varphi\in\Phi\) \textbf{such that} $\varphi$ is non-monomorphic}{
       \ForEach{\(\sigma\in \MonoSubst{\AllSubst, \(\TyVars{\(\varphi\)}\), \(\emptyset\), \(\{\}\)}\)}{
@@
 
-   \Fn(){\FMonoStep{\(M_{\text{old}}, M_{\text{new}}, P_{\text{old}}(\varphi), P_{\text{new}}(\varphi)\)}}{
+   \Fn(){\FMonoStep{\(\varphi, M_{\text{old}}, M_{\text{new}}, N_{\text{old}}(\varphi), N_{\text{new}}(\varphi)\)}}{
    \SetKwData{TypeVars}{type\_variables}
@@
    \KwData{\begin{minipage}[t]{.8\textwidth}
-     \strut old and new monomorphic type argument maps \(M_{\text{old}}, M_{\text{new}}\) \\
-     old and new non-monomorphic type argument maps \(P_{\text{old}}(\varphi), P_{\text{new}}(\varphi)\)\strut
+     \strut polymorphic formula $\varphi$ \\
+     old and new monomorphic type argument maps \(M_{\text{old}}, M_{\text{new}}\) \\
+     old and new non-monomorphic type argument maps \(N_{\text{old}}(\varphi), N_{\text{new}}(\varphi)\)\strut
       \end{minipage}
@@
    \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M_{\text{old}}\cup M_{\text{new}}| \cdot \MonoMult\)}, \MonoCap}\;
-   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P_{\text{old}}(\varphi)\cup P_{\text{new}}(\varphi)| \cdot \PolyMult\)}, \PolyCap}\;
+   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)| \cdot \PolyMult\)}, \PolyCap}\;
    }
@@
 
-   \ForEach{\(\sigma\in {}\)\SubstGen{\(M_{\text{new}}\), \(P_{\text{new}}(\varphi)\)}
-\(\cup\) \SubstGen{\(M_{\text{new}}\), \(P_{\text{old}}(\varphi)\)}
-\(\cup\) \SubstGen{\(M_{\text{old}}\), \(P_{\text{new}}(\varphi)\)}}{
-      \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P_{\text{old}}(\varphi)\)}{
+   \ForEach{\(\sigma\in {}\)\SubstGen{\(M_{\text{new}}\), \(N_{\text{new}}(\varphi)\)}
+\(\cup\) \SubstGen{\(M_{\text{new}}\), \(N_{\text{old}}(\varphi)\)}
+\(\cup\) \SubstGen{\(M_{\text{old}}\), \(N_{\text{new}}(\varphi)\)}}{
+      \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N_{\text{old}}(\varphi)\)}{
             \eIf{\TyVars{\((\upsilon_1, \dots, \upsilon_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
@@
                }
-               \uElseIf{\(|P_{\text{next}}(\varphi)| \geq \MaxPoly\) }{
+               \uElseIf{\(|N_{\text{next}}(\varphi)| \geq \MaxPoly\) }{
                   \(M_{\text{next}} \leftarrow M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}})\)\;
-                  \(P_{\text{next}}(\varphi)\leftarrow P_{\text{next}}(\varphi)\setminus (P_{\text{old}}(\varphi)\cup P_{\text{new}}(\varphi))\)\;
-                  \Return \((M_{\text{next}}, P_{\text{next}}(\varphi), \UsedSubst)\)
+                  \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\setminus (N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi))\)\;
+                  \Return \((M_{\text{next}}, N_{\text{next}}(\varphi), \UsedSubst)\)
                }}
@@
                \textcolor{ourblueviolet}{
-               \If{\(|P_{\text{next}}(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
+               \If{\(|N_{\text{next}}(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
                \textcolor{black}{
-               \(P_{\text{next}}(\varphi)(f)\leftarrow P_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
+               \(N_{\text{next}}(\varphi)(f)\leftarrow N_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
                \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
@@
 
-   \Return \((M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}}), P_{\text{next}}(\varphi)\setminus (P_{\text{old}}(\varphi)\cup P_{\text{new}}(\varphi)), \UsedSubst)\)
+   \Return \((M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}}), N_{\text{next}}(\varphi)\setminus (N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)), \UsedSubst)\)
 }
@@
    \(S\leftarrow \emptyset\)\;
+
+   \BlankLine
+
    \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
@@
             \eIf{\(\Match{\(\upsilon_i\), \(\tau_i\)}\) has the form \Success{\(\sigma_i\)} \And \Compatible{\(\sigma\), \(\sigma_i\)}}{
-               \(\sigma\leftarrow\sigma_i\circ\sigma\)
+               \(\strut\sigma\leftarrow\sigma_i\circ\sigma\)
             }
@@
    }
+
+   \BlankLine
+
    \Return \(S\)\;
@@
 \end{quote}
-\caption{Pseudocode for substitution generation}
+\caption{Pseudocode for match generation}
 \label{subst_gen}
@@
 
+   \SetKwFunction{Compatible}{compatible}
    \SetKwFunction{MonoSubst}{mono\_substs}
    \SetKwFunction{Domain}{domain}
-   \SetKwData{Limit}{subst\_limit}
+   \SetKwData{Limit}{max\_substs}
    \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
@@
 
-   \Fn(){\MonoSubst{\(S_{\text{all}}, S^\text{acc}, \sigma^\text{acc}, V\)}}{
+   \Fn(){\MonoSubst{\(S, V, S_{\text{res}}, \sigma\)}}{
 
-   \KwData{%
-      \hspace{-2.7em}\begin{minipage}[t]{.8\textwidth}
-      \begin{enumerate}
-         \item[] set of substitutions \(S_{\text{all}}\) and \(S^\text{acc}\)
-         \item[] single substitution \(\sigma^\text{acc}\)
-         \item[] set of type variables \(V\)
-      \end{enumerate}
+   \KwData{\begin{minipage}[t]{.8\textwidth}
+     \strut set \(S\) of substitutions \\
+     set \(V\) of type variables \\
+     \strut set \(S_{\text{res}}\) of substitutions \\
+     substitution \(\sigma\)\strut
       \end{minipage}
@@
    \eIf{\(V = \emptyset\)}{
-    \Return \(S^\text{acc}\cup\{\sigma^\text{acc}\}\)\;%}
+    \Return \(S_{\text{res}}\cup\{\sigma\}\)\;%}
    }
    {
-      \Let \(\alpha \) such that \(\alpha\in V\)\;
-      \ForEach{\(\sigma_{\text{next}}\in S_{\text{all}}\) such that, \(\alpha\in\Domain{\(\sigma_{\text{next}}\)}\) \And \(\sigma^\text{acc}\) compatible with \(\sigma_{\text{next}}\)}{
+      \Let \(\alpha \) \textbf{such that} \(\alpha\in V\)\;
+      \ForEach{\(\sigma_{\negvthinspace\Delta}\in S\) \textbf{such that} \(\alpha\in\Domain{\(\sigma_{\negvthinspace\Delta}\)}\) \And \(\Compatible(\sigma,\sigma_{\negvthinspace\Delta})\)}{
          \textcolor{ourblueviolet}{
-         \eIf{\(|S^\text{acc}|<\Limit\)}{
+         \eIf{\(|S_{\text{res}}|<\Limit\)}{
             \textcolor{black}{
-            \(S^\text{acc}\leftarrow \MonoSubst{\(S_{\text{all}}, S^\text{acc}, \sigma_{\text{next}}\circ\sigma^\text{acc}, V\setminus\Domain{\(\sigma_{\text{next}}\)}\)}\)\;}
+            \(S_{\text{res}}\leftarrow \MonoSubst{\(S, V\setminus\Domain{\(\sigma_{\negvthinspace\Delta}\)}, S_{\text{res}}, \sigma_{\negvthinspace\Delta}\circ\sigma\)}\)\;}
          }
-         {\Return \(S^\text{acc}\)}}
+         {\Return \(S_{\text{res}}\)}}
       }
-      \Return \(S^\text{acc}\)\;
+      \Return \(S_{\text{res}}\)\;
    }
@@
 A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
-\ref{mono_substs} is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S^\text{all}\), it will select a substitution \(\sigma_{\text{next}}\) from \(S^\text{all}\) which instantiates at least one of the type variables in \(V\).
-It is important that \(\sigma_{\text{next}}\) be compatible with \(\sigma^\text{acc}\) so that they may be composed and the function recursively called to instantiate the remaining type variables.
+\ref{mono_substs} is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S^\text{all}\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S^\text{all}\) which instantiates at least one of the type variables in \(V\).
+It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they may be composed and the function recursively called to instantiate the remaining type variables.
 
 The Zipperposition implementation of this function makes use of a map from type variables to substitutions in place of a simple set to filter the relevant substitutions from \(S^\text{all}\) with greater ease.
-The \textcolor{ourblueviolet}{subst\_limit} exists for two main reasons:
+The \textcolor{ourblueviolet}{max\_substs} exists for two main reasons:
 \begin{enumerate}
-   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{subst\_limit} new monomorphic clause per initial polymorphic clause.
+   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic clause per initial polymorphic clause.
 Generating an excessive number of new clauses can lead to the downstream prover being flooded.
@@
 
-%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S^\text{acc}\) would represent the leaves that have already been visited and \(\sigma^\text{acc}\) would represent the current node of the tree.]
+%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]
 

### 2024-03-11 iterative_monomorphisation.tex
some rewritting and typesetting for the pseudocode

@@
 
-\textbf{TODO: Mention that all sets referenced in the algorithm are finite.}
+\textbf{ TODO rename non-monomorphic maps from \(P\) to \(N\) in high-level description section}
 
@@
 
-This algorithm relies on ``primitives'' whose implementation depend on the specifics of the grammar and logic used.
+The data structures used in the pseudocode are based on the ones used in the high-level description.
+Instead of \(M\) a map from symbols to monomorphic type argument tuples, we have \(M_\text{old}\) and \(M_\text{new}\) which play the same role whilst also distinguishing between type arguments having already been matched against and those that have not.
+Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace \(N\) the map from formulae to symbols to non-monomorphic type argument tuples.
+Finally, a set \(S\) of the substitutions generated by the matchings is kept track of throughout the iterations, until it is used in the final phase.
+
+All sets referenced in the pseudocode are assumed to be finite.
+Additionally, this algorithm relies on ``primitives'' whose implementation depend on the specifics of the grammar and logic used.
 They will therefore not be expanded upon.
@@
 \SetKwFunction{GenFormulae}{generate\_mono\_formulae}
+\SetKw{ST}{such that}
+
 \begin{figure}
@@
 
-   \ForEach{\(\varphi\in\Phi\) \textbf{such that} $\varphi$ is non-monomorphic}{
-      \ForEach{\(\sigma\in \MonoSubst{\AllSubst, \(\TyVars{\(\varphi\)}\), \(\emptyset\), \(\{\}\)}\)}{
+   \ForEach{\(\varphi\in\Phi\) \ST $\varphi$ is non-monomorphic}{
+      \ForEach{\(\sigma\in \MonoSubst(\AllSubst, \TyVars{\(\varphi\)}, \emptyset, \{\})\)}{
          \textcolor{ourblueviolet}{
@@
 \caption{Pseudocode for monomorphic formula generation}
-\label{genformulae}
+\label{gen_formulae}
 \end{figure}
 
-The iterative monomorphisation algorithm (\ref{iter_mono}), as implemented in Zipperposition, has two main phases each consisting of a separate loop.
-The first loop applies a \emph{monomorphisation step} to each formula (\ref{mono_step}) for each iteration of the outermost for loop until the user-set limit it reached.
+The iterative monomorphisation algorithm \ref{iter_mono}, as implemented in Zipperposition, has two main phases each consisting of a separate loop.
+The first loop applies a \emph{monomorphisation step} to each formula \ref{mono_step} for each iteration of the outermost for loop until the user-set limit it reached.
 This bound is the only limit which is necessary for the algorithm to terminate.
@@
 
-Once the first phase is completed, the substitutions used to create new type arguments are combined to form \emph{monomorphising substitutions}.
+Once the first phase is completed, the substitutions used to create new type arguments are combined to form \emph{monomorphising substitutions} (\ref{gen_formulae}).
 These substitutions are subsequently applied to the input problem's polymorphic formulae.
@@
 \SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
-\begin{figure}
-\begin{quote}
-\begin{algorithm}[H]
-
-   %\LinesNumbered
-   \SetKwFunction{SubstGen}{matches}\SetKwFunction{TyVars}{type\_variables}
-   \SetKwFunction{FMonoStep}{formula\_mono\_step}
-
-   \Fn(){\FMonoStep{\(\varphi, M_{\text{old}}, M_{\text{new}}, N_{\text{old}}(\varphi), N_{\text{new}}(\varphi)\)}}{
 \SetKwData{TypeVars}{type\_variables}
@@
 \SetKwFunction{Domain}{subst\_domain}\SetKwFunction{Len}{len}
+\SetKwFunction{SubstGen}{matches}\SetKwFunction{TyVars}{type\_variables}
+\SetKwFunction{FMonoStep}{formula\_mono\_step}
+
+\begin{figure}
+\begin{quote}
+\begin{algorithm}[H]
+
+   \Fn(){\(\FMonoStep(\varphi, M_{\text{old}}, M_{\text{new}}, N_{\text{old}}(\varphi), N_{\text{new}}(\varphi)\)}{
 
@@
 
-   \TypeVars \(\leftarrow\) \TyVars{\(\varphi\)}\;
+   \(\TypeVars \leftarrow \TyVars(\varphi)\)\;
 
    \textcolor{ourblueviolet}{
-   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M_{\text{old}}\cup M_{\text{new}}| \cdot \MonoMult\)}, \MonoCap}\;
-   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)| \cdot \PolyMult\)}, \PolyCap}\;
+   \(\MaxMono \leftarrow \Max(\Min(\MonoFloor, |M_{\text{old}}\cup M_{\text{new}}| \cdot \MonoMult), \MonoCap)\)\;
+   \(\MaxPoly \leftarrow \Max(\Min(\PolyFloor, |N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)| \cdot \PolyMult), \PolyCap)\)\;
    }
@@
 
-   \ForEach{\(\sigma\in {}\)\SubstGen{\(M_{\text{new}}\), \(N_{\text{new}}(\varphi)\)}
-\(\cup\) \SubstGen{\(M_{\text{new}}\), \(N_{\text{old}}(\varphi)\)}
-\(\cup\) \SubstGen{\(M_{\text{old}}\), \(N_{\text{new}}(\varphi)\)}}{
+   \ForEach{\(\sigma\in \SubstGen(M_{\text{new}}, N_{\text{new}}(\varphi))
+   \cup \SubstGen(M_{\text{new}}, N_{\text{old}}(\varphi))
+   \cup \SubstGen(M_{\text{old}}, N_{\text{new}}(\varphi))\)}{
       \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N_{\text{old}}(\varphi)\)}{
-            \eIf{\TyVars{\((\upsilon_1, \dots, \upsilon_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
+            \eIf{\(\TyVars((\upsilon_1, \dots, \upsilon_n)) \subseteq \Domain(\sigma)\)}{
                \textcolor{ourblueviolet}{
-               \uIf{\(|M_{\text{next}}| < \MaxMono\) \label{algo:relative_bound}}{
+               \uIf{\(|M_{\text{next}}| < \MaxMono\)}{
                   \textcolor{black}{
@@
                }}
-            }
-            {
-               \textcolor{ourblueviolet}{
-               \If{\(|N_{\text{next}}(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
-               \textcolor{black}{
+            }{\textcolor{ourblueviolet}{
+               \If{\(|N_{\text{next}}(\varphi)| < \MaxPoly\)}{ \textcolor{black} {
                \(N_{\text{next}}(\varphi)(f)\leftarrow N_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
-               \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
+               \(\UsedSubst \leftarrow \UsedSubst \cup\{\sigma\}\)\;
+               }}
             }}
       }
+   }
 
-         }
-      }
-   }
    \BlankLine
@@
 The set of substitutions generated thusly is then iterated over and each substitution is applied to all polymorphic type arguments.
-The application of a substitution to a polymorphic type argument will only take place if any type variable of the type argument is instantiated by the substitution.
-This condition filters out all substitution application that would have left the type arguments unchanged.
+The application of a substitution to a tuple of polymorphic type argument only takes place if least one type variable of a type in the tuple is instantiated by the substitution.
+This condition filters out all substitution application that would have otherwise left the type arguments unchanged.
 
@@
 Depending on the shape and size of the input problem and the number of iterations performed, this may be problematic.
+Introducing bounds can help solve this potential issue.
 The limit on the number of new monomorphic type arguments (in \textcolor{ourblueviolet}{blue}) has three components (similar bounds are used for the non-monomorphic type arguments):
@@
 Indeed, the maximum number of newly generated type arguments is fixed per formula, per iteration.
-This algorithm can easily be modified such that the limit on new type arguments is fixed per iteration, or conversely, be made to limit new type arguments per symbol, per formula, per iteration.
+This algorithm can easily be modified such that the limit on new type arguments is fixed per iteration.
+Conversely, it can be made to limit new type arguments per symbol, per formula, per iteration.
 
@@
 \SetKwFunction{SubstGen}{matches}
-\Fn(){\SubstGen{\(M, P(\varphi)\)}}{
+\Fn(){\(\SubstGen(M, P(\varphi))\)}{
    \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
@@
          \While{\(i \leq n\)}{
-            \eIf{\(\Match{\(\upsilon_i\), \(\tau_i\)}\) has the form \Success{\(\sigma_i\)} \And \Compatible{\(\sigma\), \(\sigma_i\)}}{
+            \eIf{\(\Match(\upsilon_i, \tau_i)\) has the form \(\Success(\sigma_i)\) \And \(\Compatible(\sigma, \sigma_i)\)}{
                \(\strut\sigma\leftarrow\sigma_i\circ\sigma\)
@@
 Each bound nonetheless has its use cases.
-For example, sets of problems which lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst sets of problems where substitution generation is a bottleneck might consider a limit on the number of generated substitution to be more useful.
+For example, sets of problems which lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst sets of problems where substitution generation is more explosive might consider a limit on the number of generated substitution to be more useful.
 
@@
 
-   \Fn(){\MonoSubst{\(S, V, S_{\text{res}}, \sigma\)}}{
+   \Fn(){\(\MonoSubst(S, V, S_{\text{res}}, \sigma)\)}{
 
@@
    {
-      \Let \(\alpha \) \textbf{such that} \(\alpha\in V\)\;
-      \ForEach{\(\sigma_{\negvthinspace\Delta}\in S\) \textbf{such that} \(\alpha\in\Domain{\(\sigma_{\negvthinspace\Delta}\)}\) \And \(\Compatible(\sigma,\sigma_{\negvthinspace\Delta})\)}{
+      \Let \(\alpha \) \ST \(\alpha\in V\)\;
+      \ForEach{\(\sigma_{\negvthinspace\Delta}\in S\) \ST \(\alpha\in\Domain(\sigma_{\negvthinspace\Delta})\) \And \(\Compatible(\sigma,\sigma_{\negvthinspace\Delta})\)}{
          \textcolor{ourblueviolet}{
@@
             \textcolor{black}{
-            \(S_{\text{res}}\leftarrow \MonoSubst{\(S, V\setminus\Domain{\(\sigma_{\negvthinspace\Delta}\)}, S_{\text{res}}, \sigma_{\negvthinspace\Delta}\circ\sigma\)}\)\;}
+            \(S_{\text{res}}\leftarrow \MonoSubst(S, V\setminus\Domain(\sigma_{\negvthinspace\Delta}), S_{\text{res}}, \sigma_{\negvthinspace\Delta}\circ\sigma)\)\;}
          }
@@
 A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
-\ref{mono_substs} is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S^\text{all}\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S^\text{all}\) which instantiates at least one of the type variables in \(V\).
+It is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S^\text{all}\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S^\text{all}\) which instantiates at least one of the type variables in \(V\).
 It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they may be composed and the function recursively called to instantiate the remaining type variables.
 
-The Zipperposition implementation of this function makes use of a map from type variables to substitutions in place of a simple set to filter the relevant substitutions from \(S^\text{all}\) with greater ease.
+The Zipperposition implementation of this function makes use of a map from type variables to substitutions in place of a set to filter the relevant substitutions from \(S^\text{all}\) more efficiently.
 The \textcolor{ourblueviolet}{max\_substs} exists for two main reasons:
@@
    \item This function is the most explosive part of the algorithm.
-If \(S^\text{all}\) contains \(n\) substitutions which each instantiate exactly one of \(v\) different type variable, up to \(n^v\) monomorphising substititions may be generated.
-This should be considered in light of the fact that the total number of type variables (used to generate these substitutions) is already doubly exponential in the number of loop iterations.
+If \(S\) contains \(n\) substitutions which each instantiate exactly one of \(v\) different type variable, up to \(n^v\) monomorphising substititions may be generated.
+This should be considered in light of the fact that the total number of type argument tuples, which  re used to generate \(S\), is doubly exponential in the number of loop iterations.
 \end{enumerate}
@@
 Once the newly generated monomorphic formulae are output, an extra post-processing step may be required.
-Indeed, monomorphic provers, such as E do not necessarily support type constructors.
+Indeed, monomorphic provers, such as E, do not necessarily support type constructors.
 Consequently, transforming the monomorphic formulae into equivalent ones without type constructors may be necessary as it was for the Zipperposition implementation.
-This transformation is largely syntactic, it is documented in [insert names] 's paper [insert citation].
+This transformation is largely syntactic and relatively simple.
+It is documented in [insert names] 's paper [insert citation, and section because it isn't the main topic].
 

### 2024-03-12 iterative_monomorphisation.tex
added precise definitions of functions assumed to have been implemented for the low level algorithm

@@
 \begin{enumerate}
-   \item[\labelitemi] extracting the initial type argument maps from the input problem
-   \item[\labelitemi] gathering type variables from a type tuple
-   \item[\labelitemi] type matching
-   \item[\labelitemi] returning the type variables instantiated by a substitution
-   \item[\labelitemi] testing the compatibility of two substitutions
+   \item[\labelitemi] \emph{initialisation\((\Phi)\)} with \(\Phi\) a set of (polymorphic) formulae extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\)
+   \item[\labelitemi] \emph{type\_variables\(((\tau_1, \dots, \tau_n))\)} with \((\tau_1, \dots,\tau_n)\) a type tuple gathers all the type variables from each of \(\tau_1, \dots, \tau_n\) into a set
+   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)} with \(\upsilon\) and \(\tau\) types will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)} where \(\sigma\) is the substitution resulting from the matching
+   \item[\labelitemi] \emph{domain\((\sigma)\)} where \(\sigma\) is a substitution returns the set of type variables instantiated by \(\sigma\)
+   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)} where \(\sigma_1\) and \(\sigma_2\) are substitutions tests the compatibility between \(\sigma_1\) and \(\sigma_2\)
    \item[\labelitemi] substitution composition

### 2024-03-15 iterative_monomorphisation.tex
improvements, corrections, ect...

@@
 \usepackage{amsmath}
+\usepackage{enumitem}
 
@@
 \newdefinition{example}[definition]{Example}
+
+\setlist[itemize]{topsep=0pt}
+\setlist[enumerate]{topsep=3pt}
 
@@
 A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
-It is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S^\text{all}\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S^\text{all}\) which instantiates at least one of the type variables in \(V\).
+It is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S\) which instantiates at least one of the type variables in \(V\).
 It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they may be composed and the function recursively called to instantiate the remaining type variables.
@@
 \end{enumerate}
+%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]
 
-%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]
+Finally, once all monomorphising substitutions have been generated, they are applied to the polymorphic formula of the input problem that they instantiate (\ref{gen_formulae}).
+The fact that the substitutions are monomorphising relative to the formula they applied to ensures that the resulting formulae are monomorphic.
+A final bound here is used to control the total number of new formulae.
+It largely overlaps in use with the bound limiting the number of monomorphising substitutions per clause, but can be useful as a means to set an absolute cap on the size of the final problem.
 
@@
 Indeed, monomorphic provers, such as E, do not necessarily support type constructors.
-Consequently, transforming the monomorphic formulae into equivalent ones without type constructors may be necessary as it was for the Zipperposition implementation.
+Consequently, transforming the monomorphic formulae into equivalent ones without type constructors may be necessary as was the case for the Zipperposition implementation.
 This transformation is largely syntactic and relatively simple.
-It is documented in [insert names] 's paper [insert citation, and section because it isn't the main topic].
+It is documented in [insert names] 's paper \cite[Sections 4 and 5]{mono-trans}.
 

### 2024-03-18 iterative_monomorphisation.tex
added a paragraph on the monomorphisation as preprocessr subsection

@@
 
+To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, several competitive higher-order polymorphic provers were run on the set of TPTP test problems.
+The total number of problems solved the ``Native'' results shown in the table.
+Evaluation of the monomorphisation approach is done in two steps, first each problem is monomorphised using the [insert configuration] configurations [options?].
+Secondly, each prover is ran on the monomorphised problem and the results are tallied in the ``Mono'' column of the table.
+In addition to the polymorphic provers used in the ``Native'' tests, two monomorphic provers are run on the monomorphised problems.
+Note that instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time is subtracted in order to compare fairly against the ``Native'' column which does not have a similar preprocessing phase.
+
+The ``Union'' column, adds up the total number of problems solved by both tests. It serves as an indicator of the usefulness of monomorphisation as a tool in a portfolio configuration.
+
 \begin{table}[ht]

### 2024-03-20 iterative_monomorphisation.tex
more editing on section 4

@@
 
-\setlist[itemize]{topsep=0pt}
-\setlist[enumerate]{topsep=3pt}
+\setlist[itemize]{topsep=3pt, itemsep=3pt}
+\setlist[enumerate]{topsep=3pt, itemsep=3pt}
 
@@
 
-The initial phase of each iteration consists of computing two maps, \(M\) and \(P\), from the input problem~$\Phi$.
+The initial phase of each iteration consists of computing two maps, \(M\) and \(N\), from the input problem~$\Phi$.
 %
@@
 
-\item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(P(\varphi)(f)\) consists of all tuples of type arguments to which \(f\) is applied in \(\varphi\) and which contains a type variable.
-For example, if \(\sym{foldl}\langle \ty{nat}, \ty{list}(\alpha)\rangle\) occurs in \(\varphi\), then \((\ty{nat}, \ty{list}(\alpha)) \in P(\varphi)(\sym{foldl}) \).
+\item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(N(\varphi)(f)\) consists of all tuples of type arguments to which \(f\) is applied in \(\varphi\) and which contains a type variable.
+For example, if \(\sym{foldl}\langle \ty{nat}, \ty{list}(\alpha)\rangle\) occurs in \(\varphi\), then \((\ty{nat}, \ty{list}(\alpha)) \in N(\varphi)(\sym{foldl}) \).
 \end{enumerate}
@@
 These definitions depend on the precise form of the input formulae, which has been left undefined.
-It is important to parametrise \(P\) with \(\varphi\) because type variables are implicitly quantified at the formula level.
+It is important to parametrise \(N\) with \(\varphi\) because type variables are implicitly quantified at the formula level.
 The formula indicates the scope of type variables.
@@
 
-Once the maps \(M\) and \(P\) are initialised, each iteration performs the following steps to create new instances of formulae:
+Once the maps \(M\) and \(N\) are initialised, each iteration performs the following steps to create new instances of formulae:
 
@@
    \begin{enumerate}
-    \item[2.1.] For each tuple \((\tau_1, \dots, \tau_n) \in  M(f)\)        and each tuple \((\upsilon_1, \dots, \upsilon_n) \in P(\varphi)(f)\),
+    \item[2.1.] For each tuple \((\tau_1, \dots, \tau_n) \in  M(f)\) and each tuple \((\upsilon_1, \dots, \upsilon_n) \in N(\varphi)(f)\),
      for each \(i\), match \(\upsilon_i\) against \(\tau_i\), yielding the substitution \(\sigma_i\) in case of success.
@@
 
-\textbf{ TODO rename non-monomorphic maps from \(P\) to \(N\) in high-level description section}
-
 The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a lower level algorithm with the following features.
 \begin{enumerate}
-    \item[\labelitemi] Bounds are presented at several stages of the algorithm to curb potentially explosive enumerations.
+    \item[\labelitemi] Bounds are presented at several stages to curb potentially explosive enumerations.
 
@@
 Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace \(N\) the map from formulae to symbols to non-monomorphic type argument tuples.
-Finally, a set \(S\) of the substitutions generated by the matchings is kept track of throughout the iterations, until it is used in the final phase.
+Finally, we keep track of a set \(S\) of substitutions generated by the matchings, until it is used to generate new formulae in the final phase.
 
@@
 \begin{enumerate}
-   \item[\labelitemi] \emph{initialisation\((\Phi)\)} with \(\Phi\) a set of (polymorphic) formulae extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\)
-   \item[\labelitemi] \emph{type\_variables\(((\tau_1, \dots, \tau_n))\)} with \((\tau_1, \dots,\tau_n)\) a type tuple gathers all the type variables from each of \(\tau_1, \dots, \tau_n\) into a set
-   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)} with \(\upsilon\) and \(\tau\) types will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)} where \(\sigma\) is the substitution resulting from the matching
-   \item[\labelitemi] \emph{domain\((\sigma)\)} where \(\sigma\) is a substitution returns the set of type variables instantiated by \(\sigma\)
-   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)} where \(\sigma_1\) and \(\sigma_2\) are substitutions tests the compatibility between \(\sigma_1\) and \(\sigma_2\)
-   \item[\labelitemi] substitution composition
-   \item[\labelitemi] substitution application to formulae 
+   \item[\labelitemi] \emph{initialisation\((\Phi)\)} with \(\Phi\) a set of (polymorphic) formulae extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
+   \item[\labelitemi] \emph{type\_variables\(((\tau_1, \dots, \tau_n))\)} with \((\tau_1, \dots,\tau_n)\) a type tuple gathers all the type variables from each of \(\tau_1, \dots, \tau_n\) into a set.
+   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)} with \(\upsilon\) and \(\tau\) types will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)} where \(\sigma\) is the substitution resulting from the matching.
+Note that the pseudocode only matches non-monomorphic types against monomorphic types, in practice it is sufficient for \emph{match} to only cover this case.
+   \item[\labelitemi] \emph{domain\((\sigma)\)} where \(\sigma\) is a substitution returns the set of type variables instantiated by \(\sigma\).
+   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)} where \(\sigma_1\) and \(\sigma_2\) are substitutions tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
+   \item[\labelitemi] Substitution composition and application are noted mathematically but still need to be implemented.
 \end{enumerate}
@@
 
-The iterative monomorphisation algorithm \ref{iter_mono}, as implemented in Zipperposition, has two main phases each consisting of a separate loop.
-The first loop applies a \emph{monomorphisation step} to each formula \ref{mono_step} for each iteration of the outermost for loop until the user-set limit it reached.
-This bound is the only limit which is necessary for the algorithm to terminate.
+The iterative monomorphisation algorithm (\ref{iter_mono}), as implemented in Zipperposition, has two main phases.
+The first phase applies a \emph{monomorphisation step} (\ref{mono_step}) to each formula in \(\Phi\) until the user-set limit it reached.
+This limit is the only bound necessary for the algorithm to terminate.
 At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
-%Updating the maps progressively over the course of an iteration would be feasible but would result in an uneven treatment of formulae.
-Type arguments in a formula processed later could be matched against monomorphic type arguments generated at an earlier time in the same iteration.
-As a result, the later a formula is processed the more new type arguments it could generate.
-
-Once the first phase is completed, the substitutions used to create new type arguments are combined to form \emph{monomorphising substitutions} (\ref{gen_formulae}).
-These substitutions are subsequently applied to the input problem's polymorphic formulae.
-The possible number of new (monomorphic) formulae resulting from this process is high but can be limited by setting a maximal number of new formulae.
-In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of clauses it is given.
+Once the first phase is completed, the substitutions used to create new type arguments are passed to the formula generation function (\ref{gen_formulae}) for the second phase.
 
@@
 The first step generates substitutions (\ref{subst_gen}) by matching the non-monomorphic type arguments against monomorphic ones.
-The separation of the type arguments into ``old'' and ``new'' maps is exploited here.
+The separation of type arguments into ``old'' and ``new'' maps is put to use here.
 Pairs of type argument tuples must contain at least one new type argument in order to be matched.
-This avoids re-computing some matchings already processed in previous iterations.
-Not all such redundant computations can be avoided because the same substitution can be obtained from matchings between entirely different pairs of types.
+This avoids re-computing some of the matchings processed in previous iterations.
+Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.
 
 The set of substitutions generated thusly is then iterated over and each substitution is applied to all polymorphic type arguments.
-The application of a substitution to a tuple of polymorphic type argument only takes place if least one type variable of a type in the tuple is instantiated by the substitution.
-This condition filters out all substitution application that would have otherwise left the type arguments unchanged.
+The application of a substitution to a tuple of polymorphic type arguments only takes place if least one type variable in the tuple is instantiated by the substitution.
+This condition filters out all substitution application that would have otherwise left all type arguments unchanged.
 
@@
 \begin{enumerate}
-   \item \textcolor{ourblueviolet}{\MonoCap} is an absolute limit on the total number of new type arguments
-   \item \textcolor{ourblueviolet}{\MonoMult} is used to allow the total number of type arguments to increase by a certain proportion
-   \item \textcolor{ourblueviolet}{\MonoFloor} balances out \MonoMult when the number of type arguments is too low, and guarantees that a minimum number of new type arguments
+   \item \textcolor{ourblueviolet}{\MonoCap} is an absolute limit on the total number of new type arguments.
+   \item \textcolor{ourblueviolet}{\MonoMult} is used to allow the total number of type arguments to increase by a certain proportion of the current number of type arguments.
+   \item \textcolor{ourblueviolet}{\MonoFloor} balances out \textcolor{ourblueviolet}{\MonoMult}, it guarantees that a minimum number of new type arguments even if the total number of type arguments is low.
 \end{enumerate}
 
-Additionally, it is possible to either generalise or further refine these bounds.
-Indeed, the maximum number of newly generated type arguments is fixed per formula, per iteration.
+It is possible to either generalise or further refine these bounds.
+The maximum number of newly generated type arguments is fixed per formula, per iteration.
 This algorithm can easily be modified such that the limit on new type arguments is fixed per iteration.
-Conversely, it can be made to limit new type arguments per symbol, per formula, per iteration.
+Conversely, new type arguments can be limited per symbol, per formula, per iteration.
 
@@
 In the pseudocode, this is checked by making sure that the while loop has iterated over all elements of the type argument tuple.
-If the substitutions are incompatible, the matchings are discarded.
-[Note that the composition of two compatible substitutions is commutative and that the order of composition is therefore irrelevant].
-[The Zipperposition implementation is written in a functional style and does not directly map onto the pseudocode].
+If any substitutions are incompatible, the matchings are discarded.
+The composition of two compatible substitutions is commutative and that the order of composition is therefore irrelevant.
 
 The various bounds presented here are, to some extent, redundant.
-For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type arguments, making the related bounds superfluous.
+For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type arguments, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous.
 Each bound nonetheless has its use cases.
@@
 The last phase uses these substitutions to instantiate the type variables in the input problem's polymorphic formuale.
+The ordering of the set \(S\) of substitutions can have a significant impact on the formulae resulting from the last phase.
+As discussed in parameter optimisation \ref{param_opti}, the best results were obtained when substitutions generated in earlier iterations are ordered first.
 
@@
 A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
-It is a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S\) which instantiates at least one of the type variables in \(V\).
+Such substitutions are computing using a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S\) which instantiates at least one of the type variables in \(V\).
 It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they may be composed and the function recursively called to instantiate the remaining type variables.
@@
 A final bound here is used to control the total number of new formulae.
-It largely overlaps in use with the bound limiting the number of monomorphising substitutions per clause, but can be useful as a means to set an absolute cap on the size of the final problem.
+It overlaps in role with the bound limiting the number of monomorphising substitutions per clause, but can be useful as a means to set an absolute cap on the size of the final problem.
+In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of clauses it is given.
 
@@
 \subsection{Parameter optimisation}
+\label{param_opti}
 

### 2024-03-20 iterative_monomorphisation.tex
edits

@@
 
-\setlist[itemize]{topsep=3pt, itemsep=3pt}
-\setlist[enumerate]{topsep=3pt, itemsep=3pt}
+\setlist[itemize]{topsep=5pt, itemsep=0pt}
+\setlist[enumerate]{topsep=5pt, itemsep=0pt}
 
@@
 The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a lower level algorithm with the following features.
-\begin{enumerate}
-    \item[\labelitemi] Bounds are presented at several stages to curb potentially explosive enumerations.
+First, numeric bounds are introduced to stop explosive enumerations.
+Second, type arguments are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
+Third, substitutions are directly applied to the type arguments instead of the formulae.
+This dispenses from having to re-extract the type arguments from the formulae at each iteration.
+New formulae are generated only once all iterations are complete, in a separate, final step.
 
-    \item[\labelitemi] Type arguments are separated into old and new sets to avoid re-computing some of the same matchings in successive iterations.
-
-    \item[\labelitemi] Substitutions are directly applied to the type arguments instead of the formulae.
-This dispenses from having to re-extract the type arguments from the formulae at each iteration.
-New formulae are only generated once all iterations are complete, in a separate, final step.
-
-\end{enumerate}
-
-The data structures used in the pseudocode are based on the ones used in the high-level description.
-Instead of \(M\) a map from symbols to monomorphic type argument tuples, we have \(M_\text{old}\) and \(M_\text{new}\) which play the same role whilst also distinguishing between type arguments having already been matched against and those that have not.
-Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace \(N\) the map from formulae to symbols to non-monomorphic type argument tuples.
-Finally, we keep track of a set \(S\) of substitutions generated by the matchings, until it is used to generate new formulae in the final phase.
+The data structures used in the pseudocode are based on the ones used in the high level description.
+Instead of a map \(M\) from symbols to monomorphic type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type arguments having already been matched against and those that have not.
+Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples.
+Finally, we keep track of a set \(S\) of substitutions generated by the matchings.
+It is used to generate new formulae in the final phase.
 
 All sets referenced in the pseudocode are assumed to be finite.
-Additionally, this algorithm relies on ``primitives'' whose implementation depend on the specifics of the grammar and logic used.
+Additionally, this algorithm relies on primitives whose implementation depends on the specifics of the grammar and logic used.
 They will therefore not be expanded upon.
@@
 \begin{enumerate}
-   \item[\labelitemi] \emph{initialisation\((\Phi)\)} with \(\Phi\) a set of (polymorphic) formulae extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
-   \item[\labelitemi] \emph{type\_variables\(((\tau_1, \dots, \tau_n))\)} with \((\tau_1, \dots,\tau_n)\) a type tuple gathers all the type variables from each of \(\tau_1, \dots, \tau_n\) into a set.
-   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)} with \(\upsilon\) and \(\tau\) types will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)} where \(\sigma\) is the substitution resulting from the matching.
-Note that the pseudocode only matches non-monomorphic types against monomorphic types, in practice it is sufficient for \emph{match} to only cover this case.
-   \item[\labelitemi] \emph{domain\((\sigma)\)} where \(\sigma\) is a substitution returns the set of type variables instantiated by \(\sigma\).
-   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)} where \(\sigma_1\) and \(\sigma_2\) are substitutions tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
-   \item[\labelitemi] Substitution composition and application are noted mathematically but still need to be implemented.
+   \item[\labelitemi] \emph{initialisation\((\Phi)\)}, where \(\Phi\) is a set of (polymorphic) formulae, extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
+   \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each of \(\tau_1, \dots, \tau_n\) into a set.
+   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)}, where \(\sigma\) is the substitution resulting from the matching.
+The pseudocode matches only non-monomorphic types against monomorphic types.
+   \item[\labelitemi] \emph{domain\((\sigma)\)}, where \(\sigma\) is a substitution, returns the set of type variables \(\alpha\) such that \(\sigma(
+   \alpha) \not= \alpha\).
+   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)}, where \(\sigma_1\) and \(\sigma_2\) are substitutions, tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
+   \item[\labelitemi] Composition and application of substitutions are written using mathematical notations.
 \end{enumerate}
@@
 
-The iterative monomorphisation algorithm (\ref{iter_mono}), as implemented in Zipperposition, has two main phases.
-The first phase applies a \emph{monomorphisation step} (\ref{mono_step}) to each formula in \(\Phi\) until the user-set limit it reached.
+The iterative monomorphisation algorithm (Figure~\ref{iter_mono}) has two phases.
+The first phase applies a \emph{monomorphisation step} (Figure~\ref{mono_step}) to each formula in \(\Phi\) until the user-set limit is reached.
 This limit is the only bound necessary for the algorithm to terminate.
 At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
-Once the first phase is completed, the substitutions used to create new type arguments are passed to the formula generation function (\ref{gen_formulae}) for the second phase.
+Once these iterations are completed, the first phase is completed and the substitutions used to create
+new type arguments are passed to the formula generation function
+(Figure~\ref{gen_formulae}) for the second phase.
 
@@
 
-Formula monomorphisation (\ref{mono_step}) computes new type arguments for a single formula, it is the core of the monomorphisation process.
-The first step generates substitutions (\ref{subst_gen}) by matching the non-monomorphic type arguments against monomorphic ones.
-The separation of type arguments into ``old'' and ``new'' maps is put to use here.
-Pairs of type argument tuples must contain at least one new type argument in order to be matched.
+Formula monomorphisation (Figure~\ref{mono_step}) computes new type arguments for a single formula.
+It forms the core of the monomorphisation process.
+The first step generates substitutions (Figure~\ref{subst_gen}) by matching the non-monomorphic type arguments against monomorphic ones.
+The separation of type arguments into old and new maps is put to use here.
+Pairs of type argument tuples must contain at least one new type argument to be matched.
 This avoids re-computing some of the matchings processed in previous iterations.
@@
 
-The set of substitutions generated thusly is then iterated over and each substitution is applied to all polymorphic type arguments.
-The application of a substitution to a tuple of polymorphic type arguments only takes place if least one type variable in the tuple is instantiated by the substitution.
+The set of substitutions generated in this way is then iterated over, and each substitution is tentatively applied to all polymorphic type arguments.
+The application of a substitution to a tuple of polymorphic type arguments takes place only if least one type variable in the tuple is instantiated by the substitution.
 This condition filters out all substitution application that would have otherwise left all type arguments unchanged.
@@
 
-Substitution generation (\ref{subst_gen}) consists in computing all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions.
+Substitution generation (Figure~\ref{subst_gen}) consists in computing all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions.
 For each pair of type argument tuples, the non-monomorphic types are matched against their monomorphic counterparts and the resulting substitutions are composed if they are compatible.
@@
 
-To monomorphise a polymorphic formula, we first compute its \emph{monomorphising substitutions} (\ref{mono_substs}).
+To monomorphise a polymorphic formula, we first compute its \emph{monomorphising substitutions} (Figure~\ref{mono_substs}).
 A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
@@
 
-Finally, once all monomorphising substitutions have been generated, they are applied to the polymorphic formula of the input problem that they instantiate (\ref{gen_formulae}).
+Finally, once all monomorphising substitutions have been generated, they are applied to the polymorphic formula of the input problem that they instantiate (Figure~\ref{gen_formulae}).
 The fact that the substitutions are monomorphising relative to the formula they applied to ensures that the resulting formulae are monomorphic.
@@
 In addition to the polymorphic provers used in the ``Native'' tests, two monomorphic provers are run on the monomorphised problems.
-Note that instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time is subtracted in order to compare fairly against the ``Native'' column which does not have a similar preprocessing phase.
+Note that instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time is subtracted to compare fairly against the ``Native'' column which does not have a similar preprocessing phase.
 

### 2024-03-20 iterative_monomorphisation.tex
edits

@@
 
-The set of substitutions generated in this way is then iterated over, and each substitution is tentatively applied to all polymorphic type arguments.
-The application of a substitution to a tuple of polymorphic type arguments takes place only if least one type variable in the tuple is instantiated by the substitution.
-This condition filters out all substitution application that would have otherwise left all type arguments unchanged.
-
-%[If we assume there are after \(k\) iterations \(N_k\) total type arguments split evenly between monomorphic and polymorphic type arguments, all in a single formula.
-There can be up to \((\frac{N_k}{2})^2\) successful matches yielding as many substitutions.
-Each substitution is then applied to each polymorphic type argument for a total of \((\frac{N_k}{2})^3\) new type arguments.
-If half of these new type arguments are monomorphic and the other half is polymorphic, then the next iteration will begin with \(N_{k+1} = (\frac{N_k}{2})^{3}\) evenly split type arguments.
-Meaning that the total number of type arguments on the \(k^\text{th}\) iteration can reach \(N_k = (\frac{N_0}{2})^{3^k}\).
-This paragraph is wrong on several points, it can be fixed but it would be annoying, the main idea is that it can be shown relatively easily in a semi-formal way that the total number of type arguments is doubly exponential in the number of iterations.]
+The set of substitutions generated in this way is then iterated over.
+Each substitution is applied to all polymorphic type argument tuples such that at least one type variable in the tuple is instantiated by the substitution.
+%This condition filters out all substitution application that would have otherwise left all type arguments unchanged.
 
 The total number of type arguments can increase cubicly in the number of type arguments at each iteration and can therefore grow doubly exponentially in the number of iterations.
-Depending on the shape and size of the input problem and the number of iterations performed, this may be problematic.
-Introducing bounds can help solve this potential issue.
-The limit on the number of new monomorphic type arguments (in \textcolor{ourblueviolet}{blue}) has three components (similar bounds are used for the non-monomorphic type arguments):
+We sketch an example of such growth.
+%
+If we assume that there are after \(k\) iterations \(N_k\) total type arguments split evenly between monomorphic and non-monomorphic type arguments, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
+Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples.
+If half of these new type arguments are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = \bigl(\frac{N_k}{2}\bigr)^{3}\) evenly split type argument tuples.
+Therefore, the total number of type arguments on the \(k\)th iteration can reach \(N_k = \bigl(\smash{\frac{N_0}{2}\bigr)^{3^k}} \cdot \bigl(\frac{1}{2}\bigl)^{3k}\).
+
+Depending on the shape and size of the input problem and the number of iterations performed, the doubly exponential growth may be problematic.
+Introducing bounds addresses this potential issue.
+The limit on the number of new monomorphic type arguments (in \textcolor{ourblueviolet}{blue}) has three components:
 \begin{enumerate}
-   \item \textcolor{ourblueviolet}{\MonoCap} is an absolute limit on the total number of new type arguments.
-   \item \textcolor{ourblueviolet}{\MonoMult} is used to allow the total number of type arguments to increase by a certain proportion of the current number of type arguments.
-   \item \textcolor{ourblueviolet}{\MonoFloor} balances out \textcolor{ourblueviolet}{\MonoMult}, it guarantees that a minimum number of new type arguments even if the total number of type arguments is low.
+   \item \textcolor{ourblueviolet}{\MonoCap} is an absolute limit on the total number of new type argument tuples.
+   \item \textcolor{ourblueviolet}{\MonoMult} is used to allow the total number of type argument tuples to increase by a certain proportion of the current number of type argument tuples.
+   \item \textcolor{ourblueviolet}{\MonoFloor} balances out \textcolor{ourblueviolet}{\MonoMult}.
+It prevents \textcolor{ourblueviolet}{\MonoMult} from inhibiting new type argument tuple generation if the total number of type argument tuples is low.
 \end{enumerate}
 
+Similar bounds are used for the non-monomorphic type arguments.
 It is possible to either generalise or further refine these bounds.

### 2024-03-20 iterative_monomorphisation.tex
hopefully made a couple things clearer

@@
 
-\SetKwData{MonoCap}{mono\_cap}\SetKwData{MonoMult}{mono\_mult}\SetKwData{MonoFloor}{mono\_floor}
+\SetKwData{MonoCap}{\textcolor{ourblueviolet}{mono\_cap}}
+\SetKwData{MonoMult}{\textcolor{ourblueviolet}{mono\_mult}}
+\SetKwData{MonoFloor}{\textcolor{ourblueviolet}{mono\_floor}}
 \SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
@@
 Each substitution is applied to all polymorphic type argument tuples such that at least one type variable in the tuple is instantiated by the substitution.
-%This condition filters out all substitution application that would have otherwise left all type arguments unchanged.
 
@@
 Introducing bounds addresses this potential issue.
-The limit on the number of new monomorphic type arguments (in \textcolor{ourblueviolet}{blue}) has three components:
+The limit on the number of new monomorphic type argument tuples is \(\emph{min}(\emph{max}(\MonoMult\cdot n, \MonoFloor), \MonoCap)\), where \(n\) is the total number of monomorphic type argument tuples.
+The components of this limit are:
 \begin{enumerate}
-   \item \textcolor{ourblueviolet}{\MonoCap} is an absolute limit on the total number of new type argument tuples.
-   \item \textcolor{ourblueviolet}{\MonoMult} is used to allow the total number of type argument tuples to increase by a certain proportion of the current number of type argument tuples.
-   \item \textcolor{ourblueviolet}{\MonoFloor} balances out \textcolor{ourblueviolet}{\MonoMult}.
-It prevents \textcolor{ourblueviolet}{\MonoMult} from inhibiting new type argument tuple generation if the total number of type argument tuples is low.
+   \item \MonoCap, an absolute limit on the total number of new type argument tuples.
+   \item \MonoMult, it is used to allow the total number of (monomorphic) type argument tuples to grow by a certain proportion of the current number \(n\) of monomorphic type argument tuples.
+   \item \MonoFloor, which balances out \MonoMult. It prevents \MonoMult from inhibiting new type argument tuple generation if \(n\) is too low.
 \end{enumerate}

### 2024-03-21 iterative_monomorphisation.tex
edits

@@
 The iterative monomorphisation algorithm (Figure~\ref{iter_mono}) has two phases.
-The first phase applies a \emph{monomorphisation step} (Figure~\ref{mono_step}) to each formula in \(\Phi\) until the user-set limit is reached.
+The first phase applies a \emph{monomorphisation step} (Figure~\ref{mono_step}) to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached.
 This limit is the only bound necessary for the algorithm to terminate.
+We use the colour blue to identify bounds and for all code that is related to bounds.
 At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
@@
 \SetKwData{MonoFloor}{\textcolor{ourblueviolet}{mono\_floor}}
-\SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
+\SetKwData{PolyCap}{\textcolor{ourblueviolet}{nonmono\_cap}}
+\SetKwData{PolyMult}{\textcolor{ourblueviolet}{nonmono\_mult}}
+\SetKwData{PolyFloor}{\textcolor{ourblueviolet}{nonmono\_floor}}
+\SetKwData{MaxMono}{\textcolor{ourblueviolet}{max\_mono\_args}}
+\SetKwData{MaxPoly}{\textcolor{ourblueviolet}{max\_nonmono\_args}}
 \SetKwData{TypeVars}{type\_variables}
-\SetKwData{MaxMono}{max\_mono\_args}
-\SetKwData{MaxPoly}{max\_poly\_args}
 \SetKwData{UsedSubst}{\(S\)}
@@
 Introducing bounds addresses this potential issue.
-The limit on the number of new monomorphic type argument tuples is \(\emph{min}(\emph{max}(\MonoMult\cdot n, \MonoFloor), \MonoCap)\), where \(n\) is the total number of monomorphic type argument tuples.
-The components of this limit are:
+The limit on the number of new monomorphic type argument tuples is \(\min(\max(\MonoMult\cdot m, \MonoFloor), \MonoCap)\), where \(m\) is the total number of monomorphic type argument tuples.
+The components of this limit are
 \begin{enumerate}
-   \item \MonoCap, an absolute limit on the total number of new type argument tuples.
-   \item \MonoMult, it is used to allow the total number of (monomorphic) type argument tuples to grow by a certain proportion of the current number \(n\) of monomorphic type argument tuples.
-   \item \MonoFloor, which balances out \MonoMult. It prevents \MonoMult from inhibiting new type argument tuple generation if \(n\) is too low.
+   \item \MonoCap, an absolute limit on the total number of new type argument tuples;
+   \item \MonoMult, which is used to allow the total number of (monomorphic) type argument tuples to grow by a certain proportion of the current number \(m\) of monomorphic type argument tuples;
+   \item \MonoFloor, which balances out \MonoMult, preventing \MonoMult from inhibiting new type argument tuple generation if \(m\) is too low.
 \end{enumerate}
 
-Similar bounds are used for the non-monomorphic type arguments.
-It is possible to either generalise or further refine these bounds.
-The maximum number of newly generated type arguments is fixed per formula, per iteration.
-This algorithm can easily be modified such that the limit on new type arguments is fixed per iteration.
-Conversely, new type arguments can be limited per symbol, per formula, per iteration.
+Similar bounds are used for the non-monomorphic type arguments:
+The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula.
+Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.
+%This algorithm can easily be modified such that the limit on new type arguments is fixed per iteration. Conversely, new type arguments can be limited per symbol, per formula, per iteration.
 
@@
 
-Substitution generation (Figure~\ref{subst_gen}) consists in computing all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions.
-For each pair of type argument tuples, the non-monomorphic types are matched against their monomorphic counterparts and the resulting substitutions are composed if they are compatible.
-In the pseudocode, this is checked by making sure that the while loop has iterated over all elements of the type argument tuple.
+Substitution generation (Figure~\ref{subst_gen}) consists of computing all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions.
+For each pair of type argument tuples, the non-monomorphic types are matched against their monomorphic counterparts, and the resulting substitutions are composed if they are compatible.
+In the pseudocode, this is checked by making sure that the \textbf{while} loop has iterated over all elements of the type argument tuple.
 If any substitutions are incompatible, the matchings are discarded.
-The composition of two compatible substitutions is commutative and that the order of composition is therefore irrelevant.
+Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant.
 
-The various bounds presented here are, to some extent, redundant.
-For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type arguments, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous.
+The various bounds presented here overlap to some extent.
+For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous.
 Each bound nonetheless has its use cases.
-For example, sets of problems which lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst sets of problems where substitution generation is more explosive might consider a limit on the number of generated substitution to be more useful.
+For example, problems that lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst sets of problems where substitution generation is more explosive might consider a limit on the number of generated substitution to be more useful.
 

### 2024-03-21 iterative_monomorphisation.tex
more

@@
 Each bound nonetheless has its use cases.
-For example, problems that lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst sets of problems where substitution generation is more explosive might consider a limit on the number of generated substitution to be more useful.
+For example, problems that lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.
 

### 2024-03-21 iterative_monomorphisation.tex
messing around with paragraphs, the result is a bit iffy

@@
 
+% some keywords
+\SetKwFunction{GenFormulae}{generate\_mono\_formulae}
+\SetKw{ST}{such that}
+
+\SetKwData{MonoCap}{\textcolor{ourblueviolet}{mono\_cap}}
+\SetKwData{MonoMult}{\textcolor{ourblueviolet}{mono\_mult}}
+\SetKwData{MonoFloor}{\textcolor{ourblueviolet}{mono\_floor}}
+\SetKwData{PolyCap}{\textcolor{ourblueviolet}{nonmono\_cap}}
+\SetKwData{PolyMult}{\textcolor{ourblueviolet}{nonmono\_mult}}
+\SetKwData{PolyFloor}{\textcolor{ourblueviolet}{nonmono\_floor}}
+\SetKwData{MaxMono}{\textcolor{ourblueviolet}{max\_mono\_args}}
+\SetKwData{MaxPoly}{\textcolor{ourblueviolet}{max\_nonmono\_args}}
+\SetKwData{TypeVars}{type\_variables}
+\SetKwData{UsedSubst}{\(S\)}
+
+\SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
+
+\SetKwFunction{Max}{max}\SetKwFunction{Min}{min}
+\SetKwFunction{Domain}{subst\_domain}\SetKwFunction{Len}{len}
+\SetKwFunction{SubstGen}{matches}\SetKwFunction{TyVars}{type\_variables}
+\SetKwFunction{FMonoStep}{formula\_mono\_step}
 \SetKwProg{Fn}{Function}{}{end}\SetKwFunction{FRecurs}{FnRecursive}%
@@
 
-\SetKwFunction{GenFormulae}{generate\_mono\_formulae}
-\SetKw{ST}{such that}
-
-\begin{figure}
-\begin{quote}
-\begin{algorithm}[H]
-\Fn(){\(\GenFormulae(\Phi, S)\)}{
-   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{To}{to}
-
-   \SetKwData{Loop}{num\_loops}\SetKwData{NewFormulae}{\(\Psi\)}
-   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(S\)}
-   \SetKwData{Limit}{total\_new\_formulae}
-
-   \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step} 
-   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{mono\_substs}
-   \SetKwFunction{GenFormulae}{generate\_formulae}
-
-   \KwData{\begin{minipage}[t]{.8\textwidth}
-     \strut set \(\Phi\) of polymorphic formulae \\
-     set $S$ of substitutions\strut
-     \end{minipage}}
-   \KwResult{set of monomorphic formulae}
-
-   \BlankLine
-
-   \NewFormulae \(\leftarrow\emptyset\)\;
-
-   \BlankLine
-
-   \ForEach{\(\varphi\in\Phi\) \ST $\varphi$ is non-monomorphic}{
-      \ForEach{\(\sigma\in \MonoSubst(\AllSubst, \TyVars{\(\varphi\)}, \emptyset, \{\})\)}{
-         \textcolor{ourblueviolet}{
-         \eIf{\(|\NewFormulae|<\Limit\)}{
-            \textcolor{black}{
-            \(\NewFormulae\leftarrow\NewFormulae\cup \{\sigma(\varphi)\}\)\;}
-         }
-         {\Return \NewFormulae}
-      }}
-   }
-
-   \BlankLine
-
-   \Return \NewFormulae
-}
-\end{algorithm}
-\end{quote}
-\caption{Pseudocode for monomorphic formula generation}
-\label{gen_formulae}
-\end{figure}
-
 The iterative monomorphisation algorithm (Figure~\ref{iter_mono}) has two phases.
-The first phase applies a \emph{monomorphisation step} (Figure~\ref{mono_step}) to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached.
+The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached.
 This limit is the only bound necessary for the algorithm to terminate.
@@
 At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
-Once these iterations are completed, the first phase is completed and the substitutions used to create
-new type arguments are passed to the formula generation function
-(Figure~\ref{gen_formulae}) for the second phase.
-
-\SetKwData{MonoCap}{\textcolor{ourblueviolet}{mono\_cap}}
-\SetKwData{MonoMult}{\textcolor{ourblueviolet}{mono\_mult}}
-\SetKwData{MonoFloor}{\textcolor{ourblueviolet}{mono\_floor}}
-\SetKwData{PolyCap}{\textcolor{ourblueviolet}{nonmono\_cap}}
-\SetKwData{PolyMult}{\textcolor{ourblueviolet}{nonmono\_mult}}
-\SetKwData{PolyFloor}{\textcolor{ourblueviolet}{nonmono\_floor}}
-\SetKwData{MaxMono}{\textcolor{ourblueviolet}{max\_mono\_args}}
-\SetKwData{MaxPoly}{\textcolor{ourblueviolet}{max\_nonmono\_args}}
-\SetKwData{TypeVars}{type\_variables}
-\SetKwData{UsedSubst}{\(S\)}
-
-\SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
-
-\SetKwFunction{Max}{max}\SetKwFunction{Min}{min}
-\SetKwFunction{Domain}{subst\_domain}\SetKwFunction{Len}{len}
-\SetKwFunction{SubstGen}{matches}\SetKwFunction{TyVars}{type\_variables}
-\SetKwFunction{FMonoStep}{formula\_mono\_step}
+Once these iterations are completed, the first phase is completed and the substitutions used to create new type arguments are passed to the formula generation function for the second phase.
 
@@
 It forms the core of the monomorphisation process.
-The first step generates substitutions (Figure~\ref{subst_gen}) by matching the non-monomorphic type arguments against monomorphic ones.
+A set of substitutions obtained from matching type argument tuples is iterated over.
 The separation of type arguments into old and new maps is put to use here.
-Pairs of type argument tuples must contain at least one new type argument to be matched.
+\emph{matches} (Figure~\ref{subst_gen}) is only called with at least one new type argument map.
 This avoids re-computing some of the matchings processed in previous iterations.
 Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.
-
-The set of substitutions generated in this way is then iterated over.
 Each substitution is applied to all polymorphic type argument tuples such that at least one type variable in the tuple is instantiated by the substitution.
@@
 We sketch an example of such growth.
-%
 If we assume that there are after \(k\) iterations \(N_k\) total type arguments split evenly between monomorphic and non-monomorphic type arguments, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
 Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples.
-If half of these new type arguments are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = \bigl(\frac{N_k}{2}\bigr)^{3}\) evenly split type argument tuples.
-Therefore, the total number of type arguments on the \(k\)th iteration can reach \(N_k = \bigl(\smash{\frac{N_0}{2}\bigr)^{3^k}} \cdot \bigl(\frac{1}{2}\bigl)^{3k}\).
+If half of these new type arguments are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples.
+Therefore, the total number of type arguments on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).
 
@@
 
+The substitutions used for generating new type arguments are obtained thanks to the \emph{matches} function (Figure~\ref{subst_gen}).
+Each symbol instances from \(N(\varphi)\) is matched against all symbol instances from \(M\) of the same symbol.
+For two such symbol instances, types from the non-monomorphic type argument tuple are matched against the corresponding types from the monomorphic type argument tuple.
+The resulting substitutions are composed if they are compatible.
+In the pseudocode, this is checked by making sure that the \textbf{while} loop has iterated over all elements of the type argument tuple.
+If any substitutions are incompatible, the matchings are discarded.
+Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant.
+
 \begin{figure}
@@
 \SetKwFunction{SubstGen}{matches}
-\Fn(){\(\SubstGen(M, P(\varphi))\)}{
+\Fn(){\(\SubstGen(M, N(\varphi))\)}{
    \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
@@
       \strut monomorphic type argument map \(M\) \\
-      non-monomorphic type argument map \(P(\varphi)\)\strut
+      non-monomorphic type argument map \(N(\varphi)\)\strut
       \end{minipage}
@@
 
-   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
+   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N(\varphi)\)}{
       \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
@@
 
-Substitution generation (Figure~\ref{subst_gen}) consists of computing all possible matchings between the type arguments of the given type argument maps and returns the resulting substitutions.
-For each pair of type argument tuples, the non-monomorphic types are matched against their monomorphic counterparts, and the resulting substitutions are composed if they are compatible.
-In the pseudocode, this is checked by making sure that the \textbf{while} loop has iterated over all elements of the type argument tuple.
-If any substitutions are incompatible, the matchings are discarded.
-Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant.
+\begin{figure}
+\begin{quote}
+\begin{algorithm}[H]
+\Fn(){\(\GenFormulae(\Phi, S)\)}{
+   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{To}{to}
+
+   \SetKwData{Loop}{num\_loops}\SetKwData{NewFormulae}{\(\Psi\)}
+   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(S\)}
+   \SetKwData{Limit}{total\_new\_formulae}
+
+   \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step} 
+   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{mono\_substs}
+   \SetKwFunction{GenFormulae}{generate\_formulae}
+
+   \KwData{\begin{minipage}[t]{.8\textwidth}
+     \strut set \(\Phi\) of polymorphic formulae \\
+     set $S$ of substitutions\strut
+     \end{minipage}}
+   \KwResult{set of monomorphic formulae}
+
+   \BlankLine
+
+   \NewFormulae \(\leftarrow\emptyset\)\;
+
+   \BlankLine
+
+   \ForEach{\(\varphi\in\Phi\) \ST $\varphi$ is non-monomorphic}{
+      \ForEach{\(\sigma\in \MonoSubst(\AllSubst, \TyVars{\(\varphi\)}, \emptyset, \{\})\)}{
+         \textcolor{ourblueviolet}{
+         \eIf{\(|\NewFormulae|<\Limit\)}{
+            \textcolor{black}{
+            \(\NewFormulae\leftarrow\NewFormulae\cup \{\sigma(\varphi)\}\)\;}
+         }
+         {\Return \NewFormulae}
+      }}
+   }
+
+   \BlankLine
+
+   \Return \NewFormulae
+}
+\end{algorithm}
+\end{quote}
+\caption{Pseudocode for monomorphic formula generation}
+\label{gen_formulae}
+\end{figure}
 

### 2024-03-22 iterative_monomorphisation.tex
small amount of editing

@@
 This limit is the only bound necessary for the algorithm to terminate.
-We use the colour blue to identify bounds and for all code that is related to bounds.
+We use the colour blue to identify bounds and code related to bounds.
 At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
-Once these iterations are completed, the first phase is completed and the substitutions used to create new type arguments are passed to the formula generation function for the second phase.
+Once these iterations are completed, the first phase is complete and the substitutions used to create new type arguments are passed to the formula generation function for the second phase.
 
@@
 A set of substitutions obtained from matching type argument tuples is iterated over.
-The separation of type arguments into old and new maps is put to use here.
-\emph{matches} (Figure~\ref{subst_gen}) is only called with at least one new type argument map.
+The separation of type arguments into old and new maps is put to use here as \emph{matches} (Figure~\ref{subst_gen}) is not called on type argument maps which are both old.
 This avoids re-computing some of the matchings processed in previous iterations.

### 2024-03-22 iterative_monomorphisation.tex
more edits

@@
 
-The iterative monomorphisation algorithm (Figure~\ref{iter_mono}) has two phases.
+The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}.
+It has two phases.
 The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached.
@@
 At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
-Once these iterations are completed, the first phase is complete and the substitutions used to create new type arguments are passed to the formula generation function for the second phase.
+Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.
 
@@
 
-Formula monomorphisation (Figure~\ref{mono_step}) computes new type arguments for a single formula.
+The formula monomorphisation algorithm is given in Figure~\ref{mono_step}.
 It forms the core of the monomorphisation process.
+It computes new type arguments for a single formula.
 A set of substitutions obtained from matching type argument tuples is iterated over.
-The separation of type arguments into old and new maps is put to use here as \emph{matches} (Figure~\ref{subst_gen}) is not called on type argument maps which are both old.
+The separation of type arguments into old and new maps is put to use to ensure that only combinations involving at least one new map are considered.
 This avoids re-computing some of the matchings processed in previous iterations.
@@
 
-The total number of type arguments can increase cubicly in the number of type arguments at each iteration and can therefore grow doubly exponentially in the number of iterations.
+The total number of type argument tuples can increase cubicly in the number of type arguments at each iteration and can therefore grow doubly exponentially in the number of iterations.
 We sketch an example of such growth.
-If we assume that there are after \(k\) iterations \(N_k\) total type arguments split evenly between monomorphic and non-monomorphic type arguments, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
+If we assume that there are after \(k\) iterations \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type arguments, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
 Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples.
-If half of these new type arguments are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples.
+If half of these new type argument tuples are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples.
 Therefore, the total number of type arguments on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).
@@
 
-The substitutions used for generating new type arguments are obtained thanks to the \emph{matches} function (Figure~\ref{subst_gen}).
-Each symbol instances from \(N(\varphi)\) is matched against all symbol instances from \(M\) of the same symbol.
-For two such symbol instances, types from the non-monomorphic type argument tuple are matched against the corresponding types from the monomorphic type argument tuple.
+The \emph{matches} function, which computes the substitutions used for generating new type arguments, is given in Figure~\ref{subst_gen}.
+Each symbol instance from \(N(\varphi)\) is matched against all corresponding symbol instances from \(M\).
+For two such symbol instances, types from the non-monomorphic type argument tuple are matched component-wise against the types from the monomorphic type argument tuple.
 The resulting substitutions are composed if they are compatible.
-In the pseudocode, this is checked by making sure that the \textbf{while} loop has iterated over all elements of the type argument tuple.
+In the pseudocode, this is checked by making sure that the \textbf{while} loop has successfully iterated over all elements of the type argument tuple.
 If any substitutions are incompatible, the matchings are discarded.
@@
 For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous.
-Each bound nonetheless has its use cases.
+Each bound nonetheless has its uses.
 For example, problems that lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.
@@
 The last phase uses these substitutions to instantiate the type variables in the input problem's polymorphic formuale.
-The ordering of the set \(S\) of substitutions can have a significant impact on the formulae resulting from the last phase.
-As discussed in parameter optimisation \ref{param_opti}, the best results were obtained when substitutions generated in earlier iterations are ordered first.
+In the presence of bounds, the order in which the elements of the set \(S\) of substitutions is traversed affects the formulae resulting from the last phase.
+As we will discuss in Section~\ref{param_opti}, the best results were obtained when substitutions generated in earlier iterations are ordered first.
 
-To monomorphise a polymorphic formula, we first compute its \emph{monomorphising substitutions} (Figure~\ref{mono_substs}).
-A substitution \(\sigma\) is a monomorphising substitution for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
-Such substitutions are computing using a recursive function, given a set of type variables \(V\) and a pool of substitutions \(S\), it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S\) which instantiates at least one of the type variables in \(V\).
-It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they may be composed and the function recursively called to instantiate the remaining type variables.
+To monomorphise a polymorphic formula, we first compute its {monomorphising substitutions} using the \emph{mono\_substs} function given in Figure~\ref{mono_substs}.
+A substitution \(\sigma\) is \emph{monomorphising} for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
+Such substitutions are computed using a recursive function.
+Given a set \(V\) of type variables and a set \(S\) of substitutions, it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S\) that instantiates at least one of the type variables in \(V\).
+It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they can be composed and the function recursively called to instantiate the remaining type variables.
 
-The Zipperposition implementation of this function makes use of a map from type variables to substitutions in place of a set to filter the relevant substitutions from \(S^\text{all}\) more efficiently.
-The \textcolor{ourblueviolet}{max\_substs} exists for two main reasons:
+The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S^\text{all}\) more efficiently.
+The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
 \begin{enumerate}
    \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic clause per initial polymorphic clause.
-Generating an excessive number of new clauses can lead to the downstream prover being flooded.
-   \item This function is the most explosive part of the algorithm.
-If \(S\) contains \(n\) substitutions which each instantiate exactly one of \(v\) different type variable, up to \(n^v\) monomorphising substititions may be generated.
-This should be considered in light of the fact that the total number of type argument tuples, which  re used to generate \(S\), is doubly exponential in the number of loop iterations.
+Generating an excessive number of new clauses can flood the downstream prover.
+   \item The \emph{mono\_substs} function is the most explosive part of the algorithm.
+If \(S\) contains \(n\) substitutions that each instantiate exactly one of \(v\) different type variables, up to \(n^v\) monomorphising substititions may be generated.
+Recall that the total number of type argument tuples that are used to generate \(S\) is doubly exponential in the number of loop iterations.
 \end{enumerate}

### 2024-03-22 iterative_monomorphisation.tex
more

@@
    \SetKwFunction{GenFormulae}{generate\_mono\_formulae}
+   \SetKwFunction{Mangle}{mangle}
 
@@
 
-   \Return \(\GenFormulae(\Phi, S)\)
+   \Return \(\Mangle(\GenFormulae(\Phi, S))\)
 }
@@
 The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}.
-It has two phases.
+It has three phases.
 The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached.
@@
 Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.
+The third phase
+mangles composite types, so that we can target a simply typed logic with no general support for $n$-ary type constructors.
+% \cite[Sections 4 and 5]{mono-trans}.
 
@@
 
+The \emph{generate\_mono\_formulae} function is given in Figure~\ref{gen_formulae}.
+It generates monomorphising substitutions and applies them to the polymorphic formula of the input problem that they instantiate.
+Since the substitutions are monomorphising relative to the formula they applied to, the resulting formulae are monomorphic.
+The \textcolor{ourblueviolet}{total\_new\_formulae} bound is used to control the total number of new formulae.
+It overlaps with \textcolor{ourblueviolet}{max\_substs} but can be useful to set an absolute limit on the size of the final problem.
+%In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of clauses it is given.
+
 To monomorphise a polymorphic formula, we first compute its {monomorphising substitutions} using the \emph{mono\_substs} function given in Figure~\ref{mono_substs}.
@@
 
-The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S^\text{all}\) more efficiently.
+The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S\) more efficiently.
 The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
@@
 %[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]
-
-Finally, once all monomorphising substitutions have been generated, they are applied to the polymorphic formula of the input problem that they instantiate (Figure~\ref{gen_formulae}).
-The fact that the substitutions are monomorphising relative to the formula they applied to ensures that the resulting formulae are monomorphic.
-A final bound here is used to control the total number of new formulae.
-It overlaps in role with the bound limiting the number of monomorphising substitutions per clause, but can be useful as a means to set an absolute cap on the size of the final problem.
-In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of clauses it is given.
-
-Once the newly generated monomorphic formulae are output, an extra post-processing step may be required.
-Indeed, monomorphic provers, such as E, do not necessarily support type constructors.
-Consequently, transforming the monomorphic formulae into equivalent ones without type constructors may be necessary as was the case for the Zipperposition implementation.
-This transformation is largely syntactic and relatively simple.
-It is documented in [insert names] 's paper \cite[Sections 4 and 5]{mono-trans}.
 

### 2024-03-22 iterative_monomorphisation.tex
added reference and added mangle as a primitve

@@
 \usepackage{multirow}
-\usepackage[noend]
-   {algorithm2e}
+\usepackage[noend]{algorithm2e}
 \usepackage{amsmath}
@@
 \begin{enumerate}
-\item[\labelitemi] Given a symbol \(f\) occurring in \(\Phi\), the set \(M(f)\) consists of all tuples of monomorphic type arguments to which \(f\) is applied in \(\Phi\).
+\item[\labelitemi] Given a symbol \(f\) occurring in \(\Phi\), the set \(M(f)\) consists of all monomorphic type arguments tuples to which \(f\) is applied in \(\Phi\).
 For example, if \(\sym{foldl}\langle \ty{nat}, \ty{int}\rangle\) occurs in \(\Phi\), then \((\ty{nat}, \ty{int}) \in M(\sym{foldl}) \).
 
-\item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(N(\varphi)(f)\) consists of all tuples of type arguments to which \(f\) is applied in \(\varphi\) and which contains a type variable.
+\item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(N(\varphi)(f)\) consists of all type arguments tuples to which \(f\) is applied in \(\varphi\) and which contains a type variable.
 For example, if \(\sym{foldl}\langle \ty{nat}, \ty{list}(\alpha)\rangle\) occurs in \(\varphi\), then \((\ty{nat}, \ty{list}(\alpha)) \in N(\varphi)(\sym{foldl}) \).
@@
 First, numeric bounds are introduced to stop explosive enumerations.
-Second, type arguments are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
+Second, type arguments tuples are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
 Third, substitutions are directly applied to the type arguments instead of the formulae.
@@
 The data structures used in the pseudocode are based on the ones used in the high level description.
-Instead of a map \(M\) from symbols to monomorphic type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type arguments having already been matched against and those that have not.
+Instead of a map \(M\) from symbols to monomorphic type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type arguments tuples having already been matched against and those that have not.
 Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples.
@@
    \item[\labelitemi] Composition and application of substitutions are written using mathematical notations.
+   \item[\labelitemi] \emph{mangle\((\Phi)\)}, where \(\Phi\) is a set of monomorphic formulae returns the same set of formulae where all types have been mangled.
 \end{enumerate}
@@
 At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
-Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.
-The third phase
-mangles composite types, so that we can target a simply typed logic with no general support for $n$-ary type constructors.
+Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.
+The third phase mangles the composite types of the newly monomorphised formulae.
+This allows targeting a simply typed logic with no general support for $n$-ary type constructors \cite[Sections 4 and 5]{monotonicity}.
 % \cite[Sections 4 and 5]{mono-trans}.
+% Paper name is gt
 
@@
 It forms the core of the monomorphisation process.
-It computes new type arguments for a single formula.
+It computes new type arguments tuples for a single formula.
 A set of substitutions obtained from matching type argument tuples is iterated over.
-The separation of type arguments into old and new maps is put to use to ensure that only combinations involving at least one new map are considered.
+The separation of type arguments tuples into old and new maps is put to use to ensure that only combinations involving at least one new map are considered.
 This avoids re-computing some of the matchings processed in previous iterations.
@@
 
-The total number of type argument tuples can increase cubicly in the number of type arguments at each iteration and can therefore grow doubly exponentially in the number of iterations.
+The total number of type argument tuples can increase cubicly in the number of type arguments tuples at each iteration and can therefore grow doubly exponentially in the number of iterations.
 We sketch an example of such growth.
-If we assume that there are after \(k\) iterations \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type arguments, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
+If we assume that there are after \(k\) iterations \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type arguments tuples, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
 Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples.
 If half of these new type argument tuples are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples.
-Therefore, the total number of type arguments on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).
+Therefore, the total number of type arguments tuples on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).
 
@@
 
-Similar bounds are used for the non-monomorphic type arguments:
+Similar bounds are used for the non-monomorphic type arguments tuples:
 The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula.
 Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.
-%This algorithm can easily be modified such that the limit on new type arguments is fixed per iteration. Conversely, new type arguments can be limited per symbol, per formula, per iteration.
 
@@
 Each bound nonetheless has its uses.
-For example, problems that lead to few successful matches but many type arguments may benefit from a limit on the number of new type arguments whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.
+For example, problems that lead to few successful matches but many type arguments tuples may benefit from a limit on the number of new type arguments tuples whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.
 
@@
 
-The first two groups of options are natural as they affect how many new monomorphic and polymorphic type arguments are generated respectively for each symbol at each iteration.
+The first two groups of options are natural as they affect how many new monomorphic and polymorphic type arguments tuples are generated respectively for each symbol at each iteration.
 These options are closely linked and must be set together if they are to be relevant.
 
-During an iteration, for each symbol (for each clause in the polymorphic case), the number of newly generated type arguments is determined by this formula:
+During an iteration, for each symbol (for each clause in the polymorphic case), the number of newly generated type arguments tuples is determined by this formula:
 \( \min(\text{cap}, \max(\text{floor}, \text{multiplier} \cdot \text{number of type arguments})) \)

### 2024-03-26 iterative_monomorphisation.tex
final-ish editing touches on section 4, fixed a mistake in one of the pseudocode algos

@@
 \SetKwData{MonoFloor}{\textcolor{ourblueviolet}{mono\_floor}}
-\SetKwData{PolyCap}{\textcolor{ourblueviolet}{nonmono\_cap}}
-\SetKwData{PolyMult}{\textcolor{ourblueviolet}{nonmono\_mult}}
-\SetKwData{PolyFloor}{\textcolor{ourblueviolet}{nonmono\_floor}}
+\SetKwData{PolyCap}{\textcolor{ourblueviolet}{nmono\_cap}}
+\SetKwData{PolyMult}{\textcolor{ourblueviolet}{nmono\_mult}}
+\SetKwData{PolyFloor}{\textcolor{ourblueviolet}{nmono\_floor}}
 \SetKwData{MaxMono}{\textcolor{ourblueviolet}{max\_mono\_args}}
-\SetKwData{MaxPoly}{\textcolor{ourblueviolet}{max\_nonmono\_args}}
+\SetKwData{MaxPoly}{\textcolor{ourblueviolet}{max\_nmono\_args}}
 \SetKwData{TypeVars}{type\_variables}
@@
    \item[\labelitemi] \emph{initialisation\((\Phi)\)}, where \(\Phi\) is a set of (polymorphic) formulae, extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
-   \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each of \(\tau_1, \dots, \tau_n\) into a set.
+   \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set.
    \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)}, where \(\sigma\) is the substitution resulting from the matching.
@@
 Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.
-The third phase mangles the composite types of the newly monomorphised formulae.
-This allows targeting a simply typed logic with no general support for $n$-ary type constructors \cite[Sections 4 and 5]{monotonicity}.
+The third phase mangles \cite[Sections 4 and 5]{monotonicity} the composite types of the newly monomorphised formulae.
+This allows targeting a simply typed logic with no general support for $n$-ary type constructors.
 % \cite[Sections 4 and 5]{mono-trans}.
@@
    \cup \SubstGen(M_{\text{old}}, N_{\text{new}}(\varphi))\)}{
-      \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N_{\text{old}}(\varphi)\)}{
+   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)\)}{
             \eIf{\(\TyVars((\upsilon_1, \dots, \upsilon_n)) \subseteq \Domain(\sigma)\)}{
@@
 The formula monomorphisation algorithm is given in Figure~\ref{mono_step}.
-It forms the core of the monomorphisation process.
-It computes new type arguments tuples for a single formula.
-A set of substitutions obtained from matching type argument tuples is iterated over.
+It forms the core of the monomorphisation process by computing new type arguments tuples for a single formula.
+Type argument tuples are matched against each other to obtain a set of substitutions which is iterated over in the outermost loop.
 The separation of type arguments tuples into old and new maps is put to use to ensure that only combinations involving at least one new map are considered.
@@
 Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.
-Each substitution is applied to all polymorphic type argument tuples such that at least one type variable in the tuple is instantiated by the substitution.
+New type argument tuples are obtained by applying each substitution to each non-monomorphic type argument tuple such that at least one type variable in the tuple is instantiated by the substitution.
 
 The total number of type argument tuples can increase cubicly in the number of type arguments tuples at each iteration and can therefore grow doubly exponentially in the number of iterations.
-We sketch an example of such growth.
-If we assume that there are after \(k\) iterations \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type arguments tuples, all in a single formula, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
+We sketch an example of such growth, for a single formula.
+If we assume that after \(k\) iterations, there are \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type arguments tuples, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
 Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples.
@@
 The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula.
+A significant difference with the monomorphic case is that \(n\) depends on the current formula being processed whilst \(m\) does not.
 Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.
@@
 The resulting substitutions are composed if they are compatible.
-In the pseudocode, this is checked by making sure that the \textbf{while} loop has successfully iterated over all elements of the type argument tuple.
+In the pseudocode, this is checked by making sure the \textbf{while} loop has successfully iterated over all elements of the type argument tuple.
 If any substitutions are incompatible, the matchings are discarded.
 Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant.
+The total number of substitutions generated is capped by \textcolor{ourblueviolet}{substitution\_limit}.
 
@@
 The various bounds presented here overlap to some extent.
-For instance, limiting the number of substitutions generated at this step may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous.
+For instance, having at most \textcolor{ourblueviolet}{substitution\_limit} substitutions generated by \emph{matches} may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous.
 Each bound nonetheless has its uses.
@@
 
-Once all monomorphisation iterations have been completed, we are left with a set of substitutions.
-The last phase uses these substitutions to instantiate the type variables in the input problem's polymorphic formuale.
+Once all monomorphisation iterations have been completed, we are left with a set of the substitutions that have been used to generate new type arguments.
+The last phase uses these substitutions to instantiate the type variables in the input problem's non-monomrophic formuale.
 In the presence of bounds, the order in which the elements of the set \(S\) of substitutions is traversed affects the formulae resulting from the last phase.
@@
 Generating an excessive number of new clauses can flood the downstream prover.
+The final number of output formulae is ensured to be at most \(|\Phi|\cdot \textcolor{ourblueviolet}{max\_substs}\).
    \item The \emph{mono\_substs} function is the most explosive part of the algorithm.

### 2024-03-28 iterative_monomorphisation.tex
nonm
fetched all citations from dblp and removed minor mistake
added paragraph to parameter optimisation

@@
 \SetKwData{MonoFloor}{\textcolor{ourblueviolet}{mono\_floor}}
-\SetKwData{PolyCap}{\textcolor{ourblueviolet}{nmono\_cap}}
-\SetKwData{PolyMult}{\textcolor{ourblueviolet}{nmono\_mult}}
-\SetKwData{PolyFloor}{\textcolor{ourblueviolet}{nmono\_floor}}
+\SetKwData{PolyCap}{\textcolor{ourblueviolet}{nonm\_cap}}
+\SetKwData{PolyMult}{\textcolor{ourblueviolet}{nonm\_mult}}
+\SetKwData{PolyFloor}{\textcolor{ourblueviolet}{nonm\_floor}}
 \SetKwData{MaxMono}{\textcolor{ourblueviolet}{max\_mono\_args}}
-\SetKwData{MaxPoly}{\textcolor{ourblueviolet}{max\_nmono\_args}}
+\SetKwData{MaxPoly}{\textcolor{ourblueviolet}{max\_nonm\_args}}
 \SetKwData{TypeVars}{type\_variables}
@@
 Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.
-The third phase mangles \cite[Sections 4 and 5]{monotonicity} the composite types of the newly monomorphised formulae.
+The third phase mangles the composite types of the newly monomorphised formulae.
 This allows targeting a simply typed logic with no general support for $n$-ary type constructors.
@@
    \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)\)}{
-            \eIf{\(\TyVars((\upsilon_1, \dots, \upsilon_n)) \subseteq \Domain(\sigma)\)}{
+            \eIf{\(\TyVars(\upsilon_1, \dots, \upsilon_n) \subseteq \Domain(\sigma)\)}{
                \textcolor{ourblueviolet}{
@@
 Generating an excessive number of new clauses can flood the downstream prover.
-The final number of output formulae is ensured to be at most \(|\Phi|\cdot \textcolor{ourblueviolet}{max\_substs}\).
+The final number of output formulae is ensured to be at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
    \item The \emph{mono\_substs} function is the most explosive part of the algorithm.
@@
 
-Considering the total number of parameters, it would be impractical to exhaustively test all option combinations.
-Consequently, options have been grouped together into sets of 2 or 3 options that will be more thouroughly tested.
-Two options will be part of the same group if it is deemed likely that a change in one of the options will significantly impact how much a change in the second option will affect the overall results.
-
-The first two groups of options are natural as they affect how many new monomorphic and polymorphic type arguments tuples are generated respectively for each symbol at each iteration.
-These options are closely linked and must be set together if they are to be relevant.
-
-During an iteration, for each symbol (for each clause in the polymorphic case), the number of newly generated type arguments tuples is determined by this formula:
-\( \min(\text{cap}, \max(\text{floor}, \text{multiplier} \cdot \text{number of type arguments})) \)
-
-The multiplier allows for a controlled increase of the number of type arguments, relative to the size of the problem whilst the floor prevents some symbols seeing no new type arguments if the multiplier is too low.
-The cap acts as a final limit in the case the explosion is too fast despite the multiplier bound.
+Each bound presented presents a tradeoff, a higher bound will allow for a more exhaustive monomorphisation at the cost of taking longer to run.
+Our main goal being higher performance automatic theorem provers, we chose the number of problems solved in a 30 second time-frame as a measure of success of the monomorphisation procedure.
+To find the best bounds for this, it would be ideal to exhaustively test all possible combinations of values each bound could take.
+Given the number of bounds, this is not realistic.
+Instead, closely linked bounds are grouped together and their possible values are more thouroughly searched.
+Once the best performing set of values for a group of bounds is found, these values are chosen as a default and the search for the next group begins.
 

### 2024-04-17 iterative_monomorphisation.tex
some progress on evaluation section

@@
 
-% Provide complete list of problems explain which ones were excluded and why
-% We ignore the following bounds: ty var limit per clause, subst per ty var, limiting the new type args per clause
+In order to measure the performance of an instance of an automatic theorem prover, we chose as a measure of success the number of probelms said instance managed to solve in under 30 seconds.
+This setup reflects practical use cases of automatic provers in interactive provers.
+The problems used for such tests are taken from the TPTP library.
+The set of selected problems is constituted of all problems in the TF1 and TH1 grammars, ie: all first and higher-order problems with rank-1 polymorphism.
+Because Zipperposition does not support reasoning on real numbers, all problems including real numbers or relying on axioms including real numbers were removed.
+This likely skews comparisons between provers dissavantageously for provers which do support real numbers.
+However, our evaluations aim to study the effect of monomorphisation, not compare different provers, this bias is therefore irrelevant.
 
@@
 
-Each bound presented presents a tradeoff, a higher bound will allow for a more exhaustive monomorphisation at the cost of taking longer to run.
-Our main goal being higher performance automatic theorem provers, we chose the number of problems solved in a 30 second time-frame as a measure of success of the monomorphisation procedure.
-To find the best bounds for this, it would be ideal to exhaustively test all possible combinations of values each bound could take.
+Each bound presented presents a tradeoff, a higher bound will allow for more exhaustive instantiations of type variables at the cost of taking more time.
+We chose the number of problems solved in under 30 seconds as a measure of success of the monomorphisation procedure.
+Ideally, all possible combinations of values each bound could be exhaustively tested to find the best compromise between completeness and speed.
 Given the number of bounds, this is not realistic.
-Instead, closely linked bounds are grouped together and their possible values are more thouroughly searched.
-Once the best performing set of values for a group of bounds is found, these values are chosen as a default and the search for the next group begins.
+Instead, closely related bounds are grouped together.
+For a single group combinations of different values each bound can take are tested.
+Once the best performing set of values for a group of bounds is found, these values are set for these bounds and the search for the next group begins.
 
+This method requires an initial default set of bounds, the values chosen for this are reasonable defaults based on preliminary tests during development and informal application of this method.
+A pitfall of this approach is to overfit the bounds to the problems used as measure of performance.
+To mitigate this, 500 test problems were taken at random from our set of 1754 selected TPTP problems.
+Tests whose results are used to inform the value of bounds or parameters make use of the 500 problems set.
+
+The number of new monomorphic (and non-monomorphic) type arguments that can be generated at each iteration can be limited at the clause level or the function symbol level.
+Bounds limiting the number of new type arguments at the function symbol level are more fine-grained and are therefore those that will be kept.
+The limit of new type arguments depends on three closely related bounds (ref to explanation).
 \begin{table}[th]
@@
 
-If the floor and multiplier are set to 0 and 0, no new polymorphic type arguments will be generated, as a result the cap will be irrelevant.
-
 \begin{table}[th]
-\caption{Evaluation of bounds for polymorphic type argument generation}
+\caption{Evaluation of bounds for non-monomorphic type argument generation}
 \centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}

### 2024-04-22 iterative_monomorphisation.tex
commented a table (preemptively)

--- 2024-03-28 iterative_monomorphisation.tex
@@
 
-% Provide complete list of problems explain which ones were excluded and why
-% We ignore the following bounds: ty var limit per clause, subst per ty var, limiting the new type args per clause
+To measure the performance of a Zipperposition instance, we chose as a measure of success the number of problems solved by that instance in under 30 seconds.
+This setup reflects practical use cases of automatic provers in the context of interactive provers.
+The problems used for these tests are taken from the TPTP library.
+The set of selected problems is made up of all problems in the TF1 and TH1 grammars, ie: all first-order and higher-order problems with rank-1 polymorphism.
+Because Zipperposition does not support reasoning with real numbers, all problems with real numbers or importing axioms with real numbers were removed.
 
@@
 
-Each bound presented presents a tradeoff, a higher bound will allow for a more exhaustive monomorphisation at the cost of taking longer to run.
-Our main goal being higher performance automatic theorem provers, we chose the number of problems solved in a 30 second time-frame as a measure of success of the monomorphisation procedure.
-To find the best bounds for this, it would be ideal to exhaustively test all possible combinations of values each bound could take.
-Given the number of bounds, this is not realistic.
-Instead, closely linked bounds are grouped together and their possible values are more thouroughly searched.
-Once the best performing set of values for a group of bounds is found, these values are chosen as a default and the search for the next group begins.
+Each bound of the monomorphisation process presents a tradeoff: a higher bound will allow for a more exhaustive instantiations of type variables at the cost of taking more time.
+Ideally, all possible combinations of values for each bound could be exhaustively tested to find the best compromise between completeness and speed.
+Given the number of bounds, however, this is not feasible.
+Instead, closely related bounds are grouped together and combinations of values for bounds in these groups are tested.
+Once the best performing set of values for a group of bounds is found, these values are assigned to the corresponding bounds as the search for the next group begins.
+This method requires an initial default set of bounds, the values chosen for these are based on preliminary tests during development.
 
+A potential pitfall of this approach is to overfit the bounds to the problems used as tests.
+To mitigate this, 500 test problems were taken at random from our set of 1754 selected TPTP problems.
+Tests whose results are used to inform the value of bounds or parameters always make use of the 500 problems set.
+
+Table [ref] groups bounds that influence the maximum number of newly generated monomorphic type arguments per clause per iteration.
+It is calculated using three bounds [backwards ref] which therefore form a natural group.
+Our results show that in this configuration, there seems to be certain range within which the values of the bounds have little effect on performance.
+Beyond this range, however, the number of problems solved drops-off sharply.
 \begin{table}[th]
 \caption{Evaluation of bounds for monomorphic type argument generation}
-\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
+\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.5em}}@{}}
    \toprule
    & &&& \multicolumn{3}{c}{floor} \\
-   & \multicolumn{3}{c}{1} & \multicolumn{3}{c}{3} & \multicolumn{3}{c}{9}\\
+   & \multicolumn{3}{c}{50} & \multicolumn{3}{c}{100} & \multicolumn{3}{c}{200}\\
    \cmidrule(l){2-10}
@@
     \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
-    25   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    50   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    100  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    \(\infty\) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    250   & 164 & 164 & 164 & 164 & 164 & 163 & 163 & 164 & 164 \\
+    500   & 164 & 163 & 164 & 164 & 165 & 163 & 164 & 164 & 164 \\
+    750   & 164 & 164 & 165 & 164 & 163 & 164 & 164 & 164 & 164 \\
+    %\(\infty\) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
     \bottomrule
@@
 
-If the floor and multiplier are set to 0 and 0, no new polymorphic type arguments will be generated, as a result the cap will be irrelevant.
+Table [ref 2] fills the same role except it is for bounds limiting the number of new non-monomorphic type arguments.
+The bound values are lower because non-monomorphic type arguments tend to be more explosive.
+The results are similar to that of table 1 [ref ].
 
 \begin{table}[th]
-\caption{Evaluation of bounds for polymorphic type argument generation}
+\caption{Evaluation of bounds for non-monomorphic type argument generation}
 \centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
@@
   * Böhme
-  * similar algo implemented in Why3?
 

### 2024-04-23 iterative_monomorphisation.tex
filled in first table

@@
 In the presence of bounds, the order in which the elements of the set \(S\) of substitutions is traversed affects the formulae resulting from the last phase.
-As we will discuss in Section~\ref{param_opti}, the best results were obtained when substitutions generated in earlier iterations are ordered first.
+As we will discuss in Section~\ref{param_opti}, the best results were obtained when substitutions generated in the same iteration are grouped together.
 
@@
 \caption{Evaluation of bounds for monomorphic type argument generation}
-\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.5em}}@{}}
+\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
    \toprule
-   & &&& \multicolumn{3}{c}{floor} \\
-   & \multicolumn{3}{c}{50} & \multicolumn{3}{c}{100} & \multicolumn{3}{c}{200}\\
-   \cmidrule(l){2-10}
-   & &&& \multicolumn{3}{c}{multiplier} \\
-   \multirow{1}{2em}{cap} & 0 & 1 & 2 & 0 & 1 & 2 & 0 & 1 & 2\\
-    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
-    250   & 164 & 164 & 164 & 164 & 164 & 163 & 163 & 164 & 164 \\
-    500   & 164 & 163 & 164 & 164 & 165 & 163 & 164 & 164 & 164 \\
-    750   & 164 & 164 & 165 & 164 & 163 & 164 & 164 & 164 & 164 \\
-    %\(\infty\) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & &&& \multicolumn{6}{c}{cap} \\
+   & \multicolumn{4}{c}{500} &\multicolumn{4}{c}{1000} & \multicolumn{4}{c}{\(\infty\)}\\
+   \cmidrule(l){2-13}
+   & &&& \multicolumn{6}{c}{floor} \\
+   \multirow{1}{2em}{mult} & 0 & 50 & 100 & 200& 0 & 50 & 100 & 200& 0 & 50 & 100 & 200\\
+    \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
+    0         & 170 & 151 & 146 & 136 & 170 & 151 & 144 & 137 & 170 & 170 & 170 & 170\\
+    1         & 130 & 130 & 133 & 132 & 131 & 131 & 131 & 131 & 132 & 131 & 131 & 132\\
+    2         & 130 & 131 & 131 & 132 & 131 & 131 & 132 & 130 & 132 & 132 & 131 & 130\\
+    \(\infty\)& 131 & 132 & 131 & 131 & 131 & 131 & 130 & 131 & 148 & 148 & 146 & 149\\
     \bottomrule
@@
 Table [ref 2] fills the same role except it is for bounds limiting the number of new non-monomorphic type arguments.
-The bound values are lower because non-monomorphic type arguments tend to be more explosive.
+The bound values are lower because non-monomorphic type arguments are more explosive.
 The results are similar to that of table 1 [ref ].
@@
 
+The substitution generation phase occurs once all type arguments have been generated.
+The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related as the construction of monomorphising substitution makes use of all available substitutions.
+Additionnally, the substitution ordering heuristic has a significant impact on monomorphising substitution generation.
+The ``age'' ordering of substitution orders substitutions based on which iteration they were generated in, earlier substitutions are ordered first.
+The ``random'' ordering randomly shuffles substitutions, preliminary tests clearly established this ordering to be garbage.
+Finally, the ``segregation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently.
 \begin{table}[th]
 \caption{Evaluation of bounds for substitution generation}
-\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
+\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1em}}@{}}
    \toprule
-   & \multicolumn{4}{c}{mono subst}\\
-   \multirow{1}{4em}{subst cap} & 2 & 5 & 7 & 10\\
-   \midrule
-   50   & 0 & 0 & 0 & 0\\
-   100  & 0 & 0 & 0 & 0\\
-   500  & 0 & 0 & 0 & 0\\
-   \(\infty\) & 0 & 0 & 0 & 0\\
+   & &&&& \multicolumn{4}{c}{ordering} \\
+   & \multicolumn{4}{c}{age} & \multicolumn{4}{c}{random} & \multicolumn{4}{c}{segregation}\\
+   \cmidrule(l){2-13}
+   & &&&& \multicolumn{4}{c}{mono subst}\\
+   \multirow{1}{4em}{subst cap} & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 \\
+   \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(l){10-13}
+   50   & 0 & 0 & 0 & 0& 0 & 0 & 0 & 0& 0 & 0 & 0 & 0\\
+   100   & 0 & 0 & 0 & 0& 0 & 0 & 0 & 0& 0 & 0 & 0 & 0\\
+   200   & 0 & 0 & 0 & 0& 0 & 0 & 0 & 0& 0 & 0 & 0 & 0\\
+   \(\infty\) & 0 & 0 & 0 & 0& 0 & 0 & 0 & 0& 0 & 0 & 0 & 0\\
 
@@
 
-This table groups two bounds that directly affect the number of clauses generated, a first one that allows at most a number of new clauses equal to `multiplier' times the original number of clauses.
-And the cap that gives an absolute bound over the total number of clauses generated.
-
-The choice to also include e-timeout in this table is due to the fact that E being successful in solving a monomorphised problem in a certain time limit has been observed to be highly dependent on the number of clauses it is given.
+Table [ref] groups together the bounds related to the size of the output problem to be passed to the E prover.
+The absolute cap refers to the maximum total number of formulae passed to E.
+The multiplier instead limits the total number of newly generated formulae based on the problem's initial number of formuale.
+The performance of the E prover is highly dependent on the number of formulae it is given.
+Testing the amount of time allocated to E on the monomorphised problem against the size of the monomorphised problem is necessary to account for this.
 
@@
     10 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    \(\infty\)& 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+    30 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
     \bottomrule
@@
 \end{table}
-The next table gathers the bounds linked to substitutions generation, the first one limits the number of monomorphising substitutions that are generated per clause during the final phase.
-The second one gives an absolute cap on the overall number of substitutions generated at each clause during the type argument generation.
 
-This last table attempts to determine which are the best bounds related to the depth at which the monomorphisation problem is run as well as how long zipperposition is allowed to run before monomorphisation (and the subsequent call of E) is ran.
+The last table [ref] tests the amount of time that monomorphisation is allowed to run for before being interrupted and returning to the normal operation of Zipperposition against the number of iteration of the monomorphisation algorithm.
+A higher loop number will lead to longer monomorphisation time, we therefore expect better performance whenever loop number and monomorphisation time are coherent.
 
 \begin{table}[th]
-% not super happy with caption
-\caption{Evaluation of parameters indirectly related to the size of the output problem}
+\caption{Evaluation of parameters related to the depth of monomorphisation}
 \centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
    \toprule
-   & \multicolumn{4}{c}{loop nb} \\
-   \multirow{1}{4.5em}{E call step} & 2 & 3 & 4 & 5\\
+   & \multicolumn{4}{c}{mono time} \\
+   \multirow{1}{4.5em}{loop nb} & 5 & 10 & 20 & 30\\
    \midrule
-   0 & 0 & 0 & 0 & 0\\
-   15 & 0 & 0 & 0 & 0\\
-   45 & 0 & 0 & 0 & 0\\
-   90 & 0 & 0 & 0 & 0\\
+   2 & 0 & 0 & 0 & 0\\
+   3 & 0 & 0 & 0 & 0\\
+   4 & 0 & 0 & 0 & 0\\
+   5 & 0 & 0 & 0 & 0\\
 

### 2024-04-24 iterative_monomorphisation.tex
minor changes

@@
 
-To measure the performance of a Zipperposition instance, we chose as a measure of success the number of problems solved by that instance in under 30 seconds.
-This setup reflects practical use cases of automatic provers in the context of interactive provers.
+The monomorphisation algorithm is paramatrised by many bounds, the first step of the evaluation process is to find appropriate values for these bounds.
+The second step compares the performance of Zipperposition with and without monomorphisation.
+Finally, the third step tests the relative performance of different provers on polymorphic problems and their monomorphised counterparts.
+
 The problems used for these tests are taken from the TPTP library.
-The set of selected problems is made up of all problems in the TF1 and TH1 grammars, ie: all first-order and higher-order problems with rank-1 polymorphism.
-Because Zipperposition does not support reasoning with real numbers, all problems with real numbers or importing axioms with real numbers were removed.
+The TPTP library contains ????
+problems in the TF1 and TH1 grammars, they are first-order and higher-order problems with rank-1 polymorphism.
+Because Zipperposition does not support reasoning with real numbers, we removed all problems including real numbers or importing axioms including real numbers.
+In total our testing suite contains 1754 polymorphic problems.
+For the first step, to measure the performance of a Zipperposition instance, we chose as a measure of success the number of problems solved by that instance in under 30 seconds with a single thread.
+This setup mirrors the setting of interactive provers in which an instance of an automatic prover can typically be called for similar durations.
+As a precaution against overfitting, all tests of this phase are made on 500 problems randomly selected out of the 1754 total problems.
 
@@
 Once the best performing set of values for a group of bounds is found, these values are assigned to the corresponding bounds as the search for the next group begins.
-This method requires an initial default set of bounds, the values chosen for these are based on preliminary tests during development.
+This method requires an initial default set of bounds and reasonable ranges of values for each bound.
+We chose these values based on the results of some preliminary evaluations.
 
@@
 To mitigate this, 500 test problems were taken at random from our set of 1754 selected TPTP problems.
-Tests whose results are used to inform the value of bounds or parameters always make use of the 500 problems set.
+Tests whose results are used to inform the value of bounds or parameters are always limited to the 500 problems set.
 
-Table [ref] groups bounds that influence the maximum number of newly generated monomorphic type arguments per clause per iteration.
-It is calculated using three bounds [backwards ref] which therefore form a natural group.
-Our results show that in this configuration, there seems to be certain range within which the values of the bounds have little effect on performance.
-Beyond this range, however, the number of problems solved drops-off sharply.
+Table~\ref{mono_ty_args} groups bounds that influence the maximum number of newly generated monomorphic type arguments per clause per iteration.
+It is calculated using three bounds which form a natural group.
+Our results show that TODO
+
 \begin{table}[th]
@@
     \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
-    0         & 170 & 151 & 146 & 136 & 170 & 151 & 144 & 137 & 170 & 170 & 170 & 170\\
-    1         & 130 & 130 & 133 & 132 & 131 & 131 & 131 & 131 & 132 & 131 & 131 & 132\\
-    2         & 130 & 131 & 131 & 132 & 131 & 131 & 132 & 130 & 132 & 132 & 131 & 130\\
-    \(\infty\)& 131 & 132 & 131 & 131 & 131 & 131 & 130 & 131 & 148 & 148 & 146 & 149\\
+    0         & 170 & 151 & 146 & 136 & 170 & 151 & 144 & 137 & 171 & 149 & 137 & 132\\
+    1         & 130 & 130 & 133 & 132 & 131 & 131 & 131 & 131 & 131 & 131 & 131 & 130\\
+    2         & 130 & 131 & 131 & 132 & 131 & 131 & 132 & 130 & 129 & 130 & 130 & 130\\
+    \(\infty\)& 131 & 132 & 131 & 131 & 131 & 131 & 130 & 131 & 145 & 144 & 146 & 147\\
     \bottomrule
 \end{tabular}
+\label{mono_ty_args}
 \end{table}
 
-\bigskip
-\bigskip
-
-Table [ref 2] fills the same role except it is for bounds limiting the number of new non-monomorphic type arguments.
+Table~\ref{nmon_ty_args} fills the same role except it is for bounds limiting the number of new non-monomorphic type arguments.
 The bound values are lower because non-monomorphic type arguments are more explosive.
-The results are similar to that of table 1 [ref ].
+The results are similar to that of table \ref{mono_ty_args}.
 
@@
 \end{tabular}
+\label{nmon_ty_args}
 \end{table}
@@
 
-Table [ref] groups together the bounds related to the size of the output problem to be passed to the E prover.
+Table \ref{pb_size} groups together the bounds related to the size of the output problem to be passed to the E prover.
 The absolute cap refers to the maximum total number of formulae passed to E.
@@
 \end{tabular}
+\label{pb_size}
 \end{table}
 
-The last table [ref] tests the amount of time that monomorphisation is allowed to run for before being interrupted and returning to the normal operation of Zipperposition against the number of iteration of the monomorphisation algorithm.
+Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for before being interrupted and returning to the normal operation of Zipperposition against the number of iteration of the monomorphisation algorithm.
 A higher loop number will lead to longer monomorphisation time, we therefore expect better performance whenever loop number and monomorphisation time are coherent.
@@
    5 & 0 & 0 & 0 & 0\\
-
    \bottomrule
 \end{tabular}
+\label{mono_time}
 \end{table}

### 2024-04-26 iterative_monomorphisation.tex
removed passive, filled table 2 and made other rephrasings

@@
 The problems used for these tests are taken from the TPTP library.
-The TPTP library contains ????
-problems in the TF1 and TH1 grammars, they are first-order and higher-order problems with rank-1 polymorphism.
+The TPTP library contains 1765 problems in the TF1 and TH1 grammars, they are the first-order and higher-order problems with rank-1 polymorphism.
 Because Zipperposition does not support reasoning with real numbers, we removed all problems including real numbers or importing axioms including real numbers.
 In total our testing suite contains 1754 polymorphic problems.
-For the first step, to measure the performance of a Zipperposition instance, we chose as a measure of success the number of problems solved by that instance in under 30 seconds with a single thread.
-This setup mirrors the setting of interactive provers in which an instance of an automatic prover can typically be called for similar durations.
-As a precaution against overfitting, all tests of this phase are made on 500 problems randomly selected out of the 1754 total problems.
+For the first step, we chose as a measure of success the number of problems solved by that instance in under 30 seconds each with a single thread.
+This setup mirrors the setting of interactive provers in which an instance of an automatic prover will typically be called for similar durations.
+As a precaution against overfitting, all tests of this phase are made on 500 problems randomly taken out of the 1754 total problems.
 
@@
 
+Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options with which Zipperposition will be instantiated.
+Zipperposition has a portfolio mode with several different configurations, we evaluated all base configurations which called E against our test set of 500 problems.
+The best performing configurationis the basis upon which we test the different monomorphisation options.
+Our evaluation found that the ``40 e lift'' configuration performed best by solving 136 problems in under 30 seconds each.
+
 Each bound of the monomorphisation process presents a tradeoff: a higher bound will allow for a more exhaustive instantiations of type variables at the cost of taking more time.
@@
 Given the number of bounds, however, this is not feasible.
-Instead, closely related bounds are grouped together and combinations of values for bounds in these groups are tested.
-Once the best performing set of values for a group of bounds is found, these values are assigned to the corresponding bounds as the search for the next group begins.
+Instead, we group closely related bounds together and test combinations of values for bounds in these groups.
+Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group.
 This method requires an initial default set of bounds and reasonable ranges of values for each bound.
-We chose these values based on the results of some preliminary evaluations.
-
-A potential pitfall of this approach is to overfit the bounds to the problems used as tests.
-To mitigate this, 500 test problems were taken at random from our set of 1754 selected TPTP problems.
-Tests whose results are used to inform the value of bounds or parameters are always limited to the 500 problems set.
+We chose these values based on the results of preliminary evaluations.
 
@@
 It is calculated using three bounds which form a natural group.
-Our results show that TODO
+Our results show that generating no new monomorphic type arguments yields significantly better results.
+This is only possible because generating non-monomorphic type arguments can generate substitutions which instantiate one or more type variables.
 
@@
     \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
-    0         & 170 & 151 & 146 & 136 & 170 & 151 & 144 & 137 & 171 & 149 & 137 & 132\\
+    0         & 170 & 151 & 146 & 136 & 170 & 151 & 144 & 137 &\bf{171}&149&137 & 132\\
     1         & 130 & 130 & 133 & 132 & 131 & 131 & 131 & 131 & 131 & 131 & 131 & 130\\
@@
 Table~\ref{nmon_ty_args} fills the same role except it is for bounds limiting the number of new non-monomorphic type arguments.
-The bound values are lower because non-monomorphic type arguments are more explosive.
-The results are similar to that of table \ref{mono_ty_args}.
+The bound values are lower because in preliminary evaluation we found non-monomorphic type arguments to be explosive.
+The results confirm that it is non-monomorphic argument generation which drives the creation of useful non-monomorphic clauses.
+This is indicatied by the very low number of problems solved when no new non-monomorphic type arguments are allowed.
+Additionally, we observe that removing all bounds considerably reduces performance.
+TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with a few extra problems that are easily monomorphised)
 
@@
 \caption{Evaluation of bounds for non-monomorphic type argument generation}
-\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
+\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
    \toprule
-   & &&& \multicolumn{3}{c}{floor} \\
-   & \multicolumn{3}{c}{0} & \multicolumn{3}{c}{3} & \multicolumn{3}{c}{6}\\
-   \cmidrule(l){2-10}
-   & &&& \multicolumn{3}{c}{multiplier} \\
-    \multirow{1}{2em}{cap} & 0 & 0.5 & 1 & 0 & 0.5 & 1 & 0 & 0.5 & 1 \\
-    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
-    10 &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    20 &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    40 &   & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    \(\infty\) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & &&& \multicolumn{6}{c}{cap} \\
+   & \multicolumn{4}{c}{500} &\multicolumn{4}{c}{1000} & \multicolumn{4}{c}{\(\infty\)}\\
+   \cmidrule(l){2-13}
+   & &&& \multicolumn{6}{c}{floor} \\
+   \multirow{1}{2em}{mult} & 0 & 10 & 50 & 100& 0 & 10 & 50 & 100& 0 & 10 & 50 & 100\\
+    \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
+    0          & 124 & 170 & 171 & 170 & 124 & 170 & 171 & 171 & 124 & 170 & 171 & 170\\
+    0.5        &165 &171&\bf{172}& 170 & 165 & 171 & 171 & 170 & 165 & 171 & 171 & 170\\
+    1          & 170 & 171 & 171 & 171 & 170 & 171 & 171 & 171 & 170 & 171 & 171 & 171\\
+    \(\infty\) & 171 & 171 & 171 & 171 & 168 & 169 & 167 & 168 & 124 & 124 & 124 & 124\\
     \bottomrule
@@
 The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related as the construction of monomorphising substitution makes use of all available substitutions.
-Additionnally, the substitution ordering heuristic has a significant impact on monomorphising substitution generation.
+The substitution ordering heuristic has a significant impact on monomorphising substitution generation.
 The ``age'' ordering of substitution orders substitutions based on which iteration they were generated in, earlier substitutions are ordered first.
@@
 The performance of the E prover is highly dependent on the number of formulae it is given.
-Testing the amount of time allocated to E on the monomorphised problem against the size of the monomorphised problem is necessary to account for this.
+To account for this, we test the effects of the amount of time allocated to E on the monomorphised problem against the size of the monomorphised problem.
 
@@
 
-Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for before being interrupted and returning to the normal operation of Zipperposition against the number of iteration of the monomorphisation algorithm.
-A higher loop number will lead to longer monomorphisation time, we therefore expect better performance whenever loop number and monomorphisation time are coherent.
+Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for before being interrupted and returning to the unmodified operation of Zipperposition against the number of iteration of the monomorphisation algorithm.
+A higher loop number will lead to longer monomorphisation time, we therefore expect better performance whenever loop number and monomorphisation time have coherent values.
 

### 2024-04-30 iterative_monomorphisation.tex
filled in next table

@@
 Zipperposition has a portfolio mode with several different configurations, we evaluated all base configurations which called E against our test set of 500 problems.
-The best performing configurationis the basis upon which we test the different monomorphisation options.
-Our evaluation found that the ``40 e lift'' configuration performed best by solving 136 problems in under 30 seconds each.
+The best performing configuration became the basis upon which we tested the different monomorphisation options.
+Our evaluation found that the ``40 e lift'' configuration performed best by solving 136 problems in under 30 seconds each.%[TODO recheck this]
 
-Each bound of the monomorphisation process presents a tradeoff: a higher bound will allow for a more exhaustive instantiations of type variables at the cost of taking more time.
-Ideally, all possible combinations of values for each bound could be exhaustively tested to find the best compromise between completeness and speed.
+Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables at the cost of taking more time.
+Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed.
 Given the number of bounds, however, this is not feasible.
@@
 Table~\ref{mono_ty_args} groups bounds that influence the maximum number of newly generated monomorphic type arguments per clause per iteration.
-It is calculated using three bounds which form a natural group.
-Our results show that generating no new monomorphic type arguments yields significantly better results.
-This is only possible because generating non-monomorphic type arguments can generate substitutions which instantiate one or more type variables.
+The limit of newly generated type arguments per clause is determined by three components which form a natural group of bounds.
+Our results show that generating no new monomorphic type arguments seems to be the best approach.
+It is possible to monorphise formuale without generating monomorphic type arguments because non-monomorphic type argument generation can generate substitutions which instantiate one or more type variables and it is these substitutions which are used to monomorphise formulae.
 
@@
 Additionally, we observe that removing all bounds considerably reduces performance.
-TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with a few extra problems that are easily monomorphised)
+%TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with a few extra problems that are easily monomorphised)
 
@@
    & &&& \multicolumn{6}{c}{floor} \\
-   \multirow{1}{2em}{mult} & 0 & 10 & 50 & 100& 0 & 10 & 50 & 100& 0 & 10 & 50 & 100\\
+   \multirow{1}{3em}{mult} & 0 & 10 & 50 & 100& 0 & 10 & 50 & 100& 0 & 10 & 50 & 100\\
     \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
     0          & 124 & 170 & 171 & 170 & 124 & 170 & 171 & 171 & 124 & 170 & 171 & 170\\
-    0.5        &165 &171&\bf{172}& 170 & 165 & 171 & 171 & 170 & 165 & 171 & 171 & 170\\
+    0.5        & 165&171&\bf{172}& 170 & 165 & 171 & 171 & 170 & 165 & 171 & 171 & 170\\
     1          & 170 & 171 & 171 & 171 & 170 & 171 & 171 & 171 & 170 & 171 & 171 & 171\\
@@
 The substitution generation phase occurs once all type arguments have been generated.
-The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related as the construction of monomorphising substitution makes use of all available substitutions.
+The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related as the construction of monomorphising substitutions makes use of all available substitutions.
 The substitution ordering heuristic has a significant impact on monomorphising substitution generation.
 The ``age'' ordering of substitution orders substitutions based on which iteration they were generated in, earlier substitutions are ordered first.
-The ``random'' ordering randomly shuffles substitutions, preliminary tests clearly established this ordering to be garbage.
+The ``random'' ordering randomly shuffles substitutions.
 Finally, the ``segregation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently.
+Table \ref{subst_gen_table} shows that the specific choice of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
 \begin{table}[th]
 \caption{Evaluation of bounds for substitution generation}
-\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1em}}@{}}
+\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
    \toprule
@@
    & &&&& \multicolumn{4}{c}{mono subst}\\
-   \multirow{1}{4em}{subst cap} & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 \\
+   \multirow{1}{5.5em}{subst cap} & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 \\
    \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(l){10-13}
-   50   & 0 & 0 & 0 & 0& 0 & 0 & 0 & 0& 0 & 0 & 0 & 0\\
-   100   & 0 & 0 & 0 & 0& 0 & 0 & 0 & 0& 0 & 0 & 0 & 0\\
-   200   & 0 & 0 & 0 & 0& 0 & 0 & 0 & 0& 0 & 0 & 0 & 0\\
-   \(\infty\) & 0 & 0 & 0 & 0& 0 & 0 & 0 & 0& 0 & 0 & 0 & 0\\
+   50         & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
+   100        & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
+   200        & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 171 & 170 & 171\\
+   \(\infty\) & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
 
@@
 \end{tabular}
+\label{subst_gen_table}
 \end{table}

### 2024-05-02 iterative_monomorphisation.tex
filled table and improved text

@@
 If \(S\) contains \(n\) substitutions that each instantiate exactly one of \(v\) different type variables, up to \(n^v\) monomorphising substititions may be generated.
-Recall that the total number of type argument tuples that are used to generate \(S\) is doubly exponential in the number of loop iterations.
+Recall that the total number of type argument tuples that are used to generate \(S\) can be doubly exponential in the number of loop iterations.
 \end{enumerate}
@@
 
-The monomorphisation algorithm is paramatrised by many bounds, the first step of the evaluation process is to find appropriate values for these bounds.
-The second step compares the performance of Zipperposition with and without monomorphisation.
-Finally, the third step tests the relative performance of different provers on polymorphic problems and their monomorphised counterparts.
+The monomorphisation algorithm is paramatrised by many bounds, the first part of the evaluation process is to find appropriate values for these bounds.
+The second part compares the performance of Zipperposition with and without monomorphisation.
+The third part tests the relative performance of different provers on polymorphic problems and their monomorphised counterparts.
 
-The problems used for these tests are taken from the TPTP library.
+The test problems are taken from the TPTP library.
 The TPTP library contains 1765 problems in the TF1 and TH1 grammars, they are the first-order and higher-order problems with rank-1 polymorphism.
@@
 In total our testing suite contains 1754 polymorphic problems.
-For the first step, we chose as a measure of success the number of problems solved by that instance in under 30 seconds each with a single thread.
+We chose as a measure of success the number of problems that could each be solved by an instance of a prover in under 30 seconds with a single thread.
 This setup mirrors the setting of interactive provers in which an instance of an automatic prover will typically be called for similar durations.
-As a precaution against overfitting, all tests of this phase are made on 500 problems randomly taken out of the 1754 total problems.
 
@@
 
+As a precaution against overfitting, all tests related to parameter optimisation were made on 500 randomly chosen problems out of 1754.
 Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options with which Zipperposition will be instantiated.
-Zipperposition has a portfolio mode with several different configurations, we evaluated all base configurations which called E against our test set of 500 problems.
+Zipperposition has a portfolio mode with several different configurations, we evaluated all base configurations which called E against our test set of 500 problems without using monomorphisation.
 The best performing configuration became the basis upon which we tested the different monomorphisation options.
@@
 Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed.
-Given the number of bounds, however, this is not feasible.
+Given the number of bounds however, this is not feasible.
 Instead, we group closely related bounds together and test combinations of values for bounds in these groups.
 Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group.
+If several values for a given bound solve the same number of problems, the values solving the greater number of problems with monomorphisation will be selected, if these numbers are also equal, the most constraining bounds will be selected.
 This method requires an initial default set of bounds and reasonable ranges of values for each bound.
-We chose these values based on the results of preliminary evaluations.
+We determined these values based on the results of preliminary evaluations.
 
@@
 It is possible to monorphise formuale without generating monomorphic type arguments because non-monomorphic type argument generation can generate substitutions which instantiate one or more type variables and it is these substitutions which are used to monomorphise formulae.
+This table also illustrates the impact that the choice of bounds can have on performance.
+The choice values of only three bounds can result in an increase of up to 30\% in the number of problems solved.
 
@@
 
-Table~\ref{nmon_ty_args} fills the same role except it is for bounds limiting the number of new non-monomorphic type arguments.
-The bound values are lower because in preliminary evaluation we found non-monomorphic type arguments to be explosive.
-The results confirm that it is non-monomorphic argument generation which drives the creation of useful non-monomorphic clauses.
+Table~\ref{nmon_ty_args} fills the same role as table~\ref{mono_ty_args} except it is for bounds limiting the number of new non-monomorphic type arguments.
+The bound values are lower because we found non-monomorphic type arguments to be explosive in preliminary evaluations.
+These results confirm that non-monomorphic argument generation drive the creation of useful non-monomorphic clauses.
 This is indicatied by the very low number of problems solved when no new non-monomorphic type arguments are allowed.
-Additionally, we observe that removing all bounds considerably reduces performance.
-%TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with a few extra problems that are easily monomorphised)
+Performance of the monomorphisation algorithm seems largely unaffected by bounds' values when within a certain range but drops off significantly beyond this range.
+%TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with all additionnal problems that are easily monomorphised)
 
@@
 The substitution generation phase occurs once all type arguments have been generated.
-The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related as the construction of monomorphising substitutions makes use of all available substitutions.
+The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related through the role they play in the construction of monomorphising substitutions.
 The substitution ordering heuristic has a significant impact on monomorphising substitution generation.
-The ``age'' ordering of substitution orders substitutions based on which iteration they were generated in, earlier substitutions are ordered first.
+The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first.
 The ``random'' ordering randomly shuffles substitutions.
 Finally, the ``segregation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently.
-Table \ref{subst_gen_table} shows that the specific choice of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
+Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
 \begin{table}[th]
@@
    \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(l){10-13}
+   %2          & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
+   5          & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 &\bf{171}&170&171\\
    50         & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
    100        & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
-   200        & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 171 & 170 & 171\\
+   %200        & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 &\bf{171}&170&171\\
    \(\infty\) & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
@@
 Table \ref{pb_size} groups together the bounds related to the size of the output problem to be passed to the E prover.
-The absolute cap refers to the maximum total number of formulae passed to E.
-The multiplier instead limits the total number of newly generated formulae based on the problem's initial number of formuale.
+The absolute cap is the maximum number of formulae passed to E.
+The multiplier limits the total number of newly generated formulae based on the problem's initial number of formulae.
 The performance of the E prover is highly dependent on the number of formulae it is given.
 To account for this, we test the effects of the amount of time allocated to E on the monomorphised problem against the size of the monomorphised problem.
+We expect degraded performance when E is given many formulae but little time and few formulae but a lot of time.
 
@@
 
-Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for before being interrupted and returning to the unmodified operation of Zipperposition against the number of iteration of the monomorphisation algorithm.
-A higher loop number will lead to longer monomorphisation time, we therefore expect better performance whenever loop number and monomorphisation time have coherent values.
+For larger problems, the monomorphisation algorithm may timeout despite the bounds.
+In these cases neither Zipperposition nor E will have had a chance to attempt to solve the problem.
+To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation.
+Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for against the number of iteration of the monomorphisation algorithm.
 

### 2024-05-02 iterative_monomorphisation.tex
edits together

@@
 
-The monomorphisation algorithm is paramatrised by many bounds, the first part of the evaluation process is to find appropriate values for these bounds.
-The second part compares the performance of Zipperposition with and without monomorphisation.
-The third part tests the relative performance of different provers on polymorphic problems and their monomorphised counterparts.
+The monomorphisation algorithm is parametrised by many bounds.
+The first part of the evaluation process is to find appropriate values for these bounds.
+The second part compares the performance of Zipperposition without and with monomorphisation.
+The third part compares the performance of different provers on polymorphic problems and their monomorphised counterparts.
 
-The test problems are taken from the TPTP library.
-The TPTP library contains 1765 problems in the TF1 and TH1 grammars, they are the first-order and higher-order problems with rank-1 polymorphism.
-Because Zipperposition does not support reasoning with real numbers, we removed all problems including real numbers or importing axioms including real numbers.
-In total our testing suite contains 1754 polymorphic problems.
-We chose as a measure of success the number of problems that could each be solved by an instance of a prover in under 30 seconds with a single thread.
-This setup mirrors the setting of interactive provers in which an instance of an automatic prover will typically be called for similar durations.
+The benchmarks are taken from the TPTP library \cite{tptp}.
+The TPTP library version TODO contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism.
+Because Zipperposition does not support reasoning with real numbers, we removed all problems that include real numbers.
+In total, our benchmark suite contains 1754 polymorphic problems.
+We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in under 30 seconds with a single thread.
+This corresponds to the default time limit in Sledgehammer.
 
@@
 
-As a precaution against overfitting, all tests related to parameter optimisation were made on 500 randomly chosen problems out of 1754.
-Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options with which Zipperposition will be instantiated.
-Zipperposition has a portfolio mode with several different configurations, we evaluated all base configurations which called E against our test set of 500 problems without using monomorphisation.
-The best performing configuration became the basis upon which we tested the different monomorphisation options.
-Our evaluation found that the ``40 e lift'' configuration performed best by solving 136 problems in under 30 seconds each.%[TODO recheck this]
+As a precaution against overfitting, the part of the evaluation related to parameter optimisation was carried out on 500 randomly chosen problems out of the 1754 selected problems, and the rest of the evaluation was carried out on the remaining 1254 problems.
+Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options used to run Zipperposition.
+Zipperposition has a portfolio mode with several configurations.
+In a preliminary experiment, we informally evaluated all portfolio configurations that called E against our benchmark suite of 500 problems without using monomorphisation.
+The best performing configuration became the base configuration, which we used as a basis to evaluate the different monomorphisation options.
+Our informal evaluation found that the so-called \verb|40_e_lift| configuration performed best by proving 136 problems.%[TODO recheck this]
 
@@
 Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed.
-Given the number of bounds however, this is not feasible.
+Given the number of bounds, however, this is not feasible.
 Instead, we group closely related bounds together and test combinations of values for bounds in these groups.
 Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group.
-If several values for a given bound solve the same number of problems, the values solving the greater number of problems with monomorphisation will be selected, if these numbers are also equal, the most constraining bounds will be selected.
+If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected,
+and shown in bold in Tables \ref{mono_ty_args} to \ref{xxx}.
 This method requires an initial default set of bounds and reasonable ranges of values for each bound.
-We determined these values based on the results of preliminary evaluations.
+We determined these values based on the results of preliminary experiments on the benchmark suite of 500 problems.
 
 Table~\ref{mono_ty_args} groups bounds that influence the maximum number of newly generated monomorphic type arguments per clause per iteration.
-The limit of newly generated type arguments per clause is determined by three components which form a natural group of bounds.
-Our results show that generating no new monomorphic type arguments seems to be the best approach.
-It is possible to monorphise formuale without generating monomorphic type arguments because non-monomorphic type argument generation can generate substitutions which instantiate one or more type variables and it is these substitutions which are used to monomorphise formulae.
-This table also illustrates the impact that the choice of bounds can have on performance.
-The choice values of only three bounds can result in an increase of up to 30\% in the number of problems solved.
+The limit on newly generated type arguments is determined by three components that form a natural group of bounds.
+The table shows that generating no new monomorphic type arguments seems to be the best approach.
+This result may seem counterintuitive, but it is possible to monomorphise formulae without generating monomorphic type arguments.
+This is because non-monomorphic type argument generation can generate substitutions that instantiate one or more type variables, and it is these substitutions that are used to monomorphise formulae.
 
@@
 
-Table~\ref{nmon_ty_args} fills the same role as table~\ref{mono_ty_args} except it is for bounds limiting the number of new non-monomorphic type arguments.
+Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments.
 The bound values are lower because we found non-monomorphic type arguments to be explosive in preliminary evaluations.
-These results confirm that non-monomorphic argument generation drive the creation of useful non-monomorphic clauses.
-This is indicatied by the very low number of problems solved when no new non-monomorphic type arguments are allowed.
-Performance of the monomorphisation algorithm seems largely unaffected by bounds' values when within a certain range but drops off significantly beyond this range.
+The table confirms that non-monomorphic type argument generation drives the creation of useful non-monomorphic formulae.
+This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed.
+Performance of the monomorphisation algorithm seems largely unaffected by the bounds' values when within a certain range but drops off significantly beyond this range.
 %TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with all additionnal problems that are easily monomorphised)

### 2024-05-07 iterative_monomorphisation.tex
finished first draft of 5.1

@@
 \usepackage{soul}
+\usepackage{multicol}
 \usepackage[dvipsnames]{xcolor}
@@
 
+% bounds keywords
 \SetKwData{MonoCap}{\textcolor{ourblueviolet}{mono\_cap}}
@@
 \SetKwData{MaxPoly}{\textcolor{ourblueviolet}{max\_nonm\_args}}
+\SetKwData{Limit}{\textcolor{ourblueviolet}{total\_new\_formulae}}
+\SetKwData{MonoSubstsLimit}{\textcolor{ourblueviolet}{max\_substs}}
+\SetKwData{SubstLimit}{\textcolor{ourblueviolet}{substitution\_limit}}
+\SetKwData{Loop}{\textcolor{ourblueviolet}{num\_loops}}
 \SetKwData{TypeVars}{type\_variables}
@@
 %   - correctly spelled TPTP, Leo-III, Zipperposition, E, Vampire
-%   - use of formula vs clause
 %   - polymorphic and monomorphic
@@
 
-   \SetKwData{Loop}{num\_loops}\SetKwData{NewFormulae}{\(\Psi\)}
    \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(S\)}
-   \SetKwData{Limit}{total\_new\_formulae}
 
@@
    \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
-   \SetKwData{SubstLimit}{substitution\_limit}
    \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
@@
 
-   \SetKwData{Loop}{num\_loops}\SetKwData{NewFormulae}{\(\Psi\)}
+   \SetKwData{NewFormulae}{\(\Psi\)}
    \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(S\)}
@@
    \SetKwFunction{Domain}{domain}
-   \SetKwData{Limit}{max\_substs}
    \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
@@
          \textcolor{ourblueviolet}{
-         \eIf{\(|S_{\text{res}}|<\Limit\)}{
+         \eIf{\(|S_{\text{res}}|<\MonoSubstsLimit\)}{
             \textcolor{black}{
@@
 It overlaps with \textcolor{ourblueviolet}{max\_substs} but can be useful to set an absolute limit on the size of the final problem.
-%In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of clauses it is given.
+%In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of formuale it is given.
 
@@
 \begin{enumerate}
-   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic clause per initial polymorphic clause.
-Generating an excessive number of new clauses can flood the downstream prover.
+   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic formula per initial polymorphic formula.
+Generating an excessive number of new formulae can flood the downstream prover.
 The final number of output formulae is ensured to be at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
@@
 The benchmarks are taken from the TPTP library \cite{tptp}.
-The TPTP library version TODO contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism.
+The TPTP library, in version 8.2.0, contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism.
 Because Zipperposition does not support reasoning with real numbers, we removed all problems that include real numbers.
@@
 
-As a precaution against overfitting, the part of the evaluation related to parameter optimisation was carried out on 500 randomly chosen problems out of the 1754 selected problems, and the rest of the evaluation was carried out on the remaining 1254 problems.
-Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options used to run Zipperposition.
-Zipperposition has a portfolio mode with several configurations.
-In a preliminary experiment, we informally evaluated all portfolio configurations that called E against our benchmark suite of 500 problems without using monomorphisation.
-The best performing configuration became the base configuration, which we used as a basis to evaluate the different monomorphisation options.
-Our informal evaluation found that the so-called \verb|40_e_lift| configuration performed best by proving 136 problems.%[TODO recheck this]
-
 Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables at the cost of taking more time.
@@
 Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group.
-If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected,
-and shown in bold in Tables \ref{mono_ty_args} to \ref{xxx}.
-This method requires an initial default set of bounds and reasonable ranges of values for each bound.
-We determined these values based on the results of preliminary experiments on the benchmark suite of 500 problems.
+If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected, and shown in bold in Tables \ref{mono_ty_args} to \ref{mono_time}.
 
-Table~\ref{mono_ty_args} groups bounds that influence the maximum number of newly generated monomorphic type arguments per clause per iteration.
+As a precaution against overfitting, the part of the evaluation related to parameter optimisation and all related preliminary evaluations were carried out on 500 randomly chosen problems out of the 1754 selected problems.
+The rest of the evaluation was carried out on the remaining 1254 problems.
+Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options used to run Zipperposition.
+Zipperposition has a portfolio mode with several configurations.
+Because the space of possible base configurations is too large to evaluate, we informally evaluated, in a preliminary experiment, all portfolio configurations that called E (against our benchmark suite of 500 problems)?
+without using monomorphisation.
+The best performing configuration became the base configuration, which we used as a basis to evaluate the different monomorphisation options.
+Our informal evaluation found that the so-called \verb|40_e_lift| configuration performed best by proving 136 problems.
+%[TODO recheck this]
+We conducted additionnal informal preliminary evaluations (on the benchmark suite of 500 problems)?
+to find appropriate default values and approriate test ranges for the monomorphisation bounds.
+We started with the base configuration \verb|40_e_lift|, with the following values for the monomorphisation bounds:
+
+\begin{multicols}{2}% 3 columns?
+\begin{itemize}
+   \item \MonoCap: irrelevant
+   \item \MonoMult: irrelevant
+   \item \MonoFloor: irrelevant
+   \item \PolyCap: \(\infty\)
+   \item \PolyMult: \(1\)
+   \item \PolyFloor: \(50\)
+   \item substitution ordering: segregation
+   \item \SubstLimit: \(\infty\)
+   \item \MonoSubstsLimit: \(10\)
+   \item \Limit: \(2000\)
+   \item new formulae limit multiplier: \(0\)
+   \item E timeout: \(30\)
+   \item monomorphisation timeout: \(20\)
+   \item \Loop: \(4\)
+\end{itemize}
+\end{multicols}
+Monomorphistion (and the subsequent call to E) occurs at the beginning of the invocation of Zipperposition.
+Departing from the original \verb|40_e_lift| configuration, which would otherwise call E after 8\% of the total alloted time for Zipperposition has passed.
+
+Table~\ref{mono_ty_args} groups bounds that influence the maximum number of newly generated monomorphic type arguments per formuala per iteration.
 The limit on newly generated type arguments is determined by three components that form a natural group of bounds.
@@
 This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed.
-Performance of the monomorphisation algorithm seems largely unaffected by the bounds' values when within a certain range but drops off significantly beyond this range.
+Performance of the monomorphisation algorithm seems to plateau for a large range of the tested values for these bounds and drops off sharply beyond.
 %TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with all additionnal problems that are easily monomorphised)
@@
 To account for this, we test the effects of the amount of time allocated to E on the monomorphised problem against the size of the monomorphised problem.
-We expect degraded performance when E is given many formulae but little time and few formulae but a lot of time.
+We observe that the timeout of E has the biggest impact on performance whilst the precise limits on problem size seem to have little effect.
 
@@
 \caption{Evaluation of bounds directly related to the size of the output problem}
-\centering\begin{tabular}{@{}l*{9}{>{\centering\arraybackslash}p{1.1em}}@{}}
+\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
    \toprule
-   & &&& \multicolumn{3}{c}{cap} \\
-   & \multicolumn{3}{c}{500} & \multicolumn{3}{c}{2000} & \multicolumn{3}{c}{\(\infty\)}\\
-   \cmidrule(l){2-10}
-   & &&& \multicolumn{3}{c}{multiplier} \\
-    \multirow{1}{5.4em}{E timeout (s)} & 1 & 2 & 3 & 1 & 2 & 3 & 1 & 2 & 3 \\
-    \cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(l){8-10}
-    2  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    5  & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    10 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-    30 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+   & &&&& \multicolumn{4}{c}{formula cap} \\
+   & \multicolumn{4}{c}{500} & \multicolumn{4}{c}{2000} & \multicolumn{4}{c}{\(\infty\)}\\
+   \cmidrule(l){2-13}
+   & &&&& \multicolumn{4}{c}{formula multiplier}\\
+   \multirow{1}{5.5em}{E timeout (s)} & 1 & 2 & 3 & \(\infty\)& 1 & 2 & 3 & \(\infty\)& 1 & 2 & 3 & \(\infty\)\\
+   \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(l){10-13}
+   2          & 155 & 154 & 155 & 155 & 154 & 154 & 155 & 154 & 155 & 154 & 156 & 137\\
+   5          & 163 & 164 & 164 & 163 & 163 & 164 & 163 & 163 & 163 & 162 & 163 & 137\\
+   10         & 166 & 167 & 166 & 166 & 166 & 166 & 166 & 166 & 165 & 166 & 167 & 137\\
+   30         &169&\bf{171}&170 & 169 & 170 & 170 & 170 & 170 & 163 & 169 & 170 & 124\\
    \bottomrule
@@
 Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for against the number of iteration of the monomorphisation algorithm.
+The number of iterations and timeout seem to have almost no effect on the performance of the monomorphisation algorithm.
 
@@
 \caption{Evaluation of parameters related to the depth of monomorphisation}
-\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1em}}@{}}
+\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
    \toprule
    & \multicolumn{4}{c}{mono time} \\
-   \multirow{1}{4.5em}{loop nb} & 5 & 10 & 20 & 30\\
+   \multirow{1}{4em}{loop nb} & 5 & 10 & 20 & 30\\
    \midrule
-   2 & 0 & 0 & 0 & 0\\
-   3 & 0 & 0 & 0 & 0\\
-   4 & 0 & 0 & 0 & 0\\
-   5 & 0 & 0 & 0 & 0\\
+   2     &\bf{171}&171&171 & 170\\
+   3     & 170 & 171 & 170 & 169\\
+   4     & 170 & 171 & 170 & 169\\
+   5     & 169 & 169 & 169 & 168\\
+   6     & 169 & 169 & 169 & 168\\
    \bottomrule

### 2024-05-17 iterative_monomorphisation.tex
filled out table 6

@@
 
+\subsection{E as Zipperposition backend}
+
+We conducted two evaluations of Zipperpostion with the monomorphisation algorithm and E as a backend.
+The first was on the 500 problems we used for option optimisatio and the second on the remaining 1034 problems.
+This version of Zipperposition was compared against the base Zipperposition prover.
+The base Zipperposition instance was run wiht the sequential portfolio mode \verb|portfolio.sequential.py|.
+
+Note that Zipperposition's portfolio mode seems to consider timeouts to be more of a polite suggestion than a hard limit, we consequently added a manual timeout of 30 seconds.
+
+\begin{table}[ht]
+\caption{Evaluation of Zipperposition without E vs. with E}
+\centering\begin{tabular}{@{}lccc@{}}
+   \toprule
+   & without E & with E & Union \\
+   \midrule
+   500 problems benchmark   & 152 & 185 & 195 (-10 +43) \\
+   fresh problems benchmark & 337 & 385 & 419 (-34 +82)\\
+   \bottomrule
+\end{tabular}
+\end{table}
+
 \subsection{Monomorphisation as preprocessor}
@@
    E  &   & 0 & 0 \\
-   Leo-III & 0 & 0 & 0 \\
+   Leo-III & 0 & 0 & 0 \\ % Leo-III's performance is comparatively very poor, not coherent with casc results
    Satallax &  & 0 & 0 \\
-   Vampire & 0 & 0 & 0 \\
-   Zipperposition & 0 & 0 & 0 \\[1.5\jot]
+   Vampire & 0 & 0 & 0 \\ % Vampire's parser fails on too many polymorphic problems (about 1/3 of them)
+   Zipperposition & 335 & 0 & 0 \\[1.5\jot]
    Total & 0 & 0 & 0 \\
-   \bottomrule
-\end{tabular}
-\end{table}
-
-\subsection{E as Zipperposition backend}
-
-\begin{table}[ht]
-\caption{Evaluation of Zipperposition without E vs. with E}
-\centering\begin{tabular}{@{}lccc@{}}
-   \toprule
-   & without E & with E & Union \\
-   \midrule
-   Zipperposition & 0 & 0 & 0 \\
    \bottomrule

### 2024-06-18 iterative_monomorphisation.tex
minor corrections

@@
 Because Zipperposition does not support reasoning with real numbers, we removed all problems that include real numbers.
-In total, our benchmark suite contains 1754 polymorphic problems.
+In total, our benchmark suite contains 1535 polymorphic problems.
 We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in under 30 seconds with a single thread.
@@
 
-As a precaution against overfitting, the part of the evaluation related to parameter optimisation and all related preliminary evaluations were carried out on 500 randomly chosen problems out of the 1754 selected problems.
-The rest of the evaluation was carried out on the remaining 1254 problems.
+As a precaution against overfitting, the part of the evaluation related to parameter optimisation and all related preliminary evaluations were carried out on 500 randomly chosen problems out of the 1535 selected problems.
+The rest of the evaluation was carried out on the remaining 1034 problems.
 Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options used to run Zipperposition.
 Zipperposition has a portfolio mode with several configurations.
-Because the space of possible base configurations is too large to evaluate, we informally evaluated, in a preliminary experiment, all portfolio configurations that called E (against our benchmark suite of 500 problems)?
-without using monomorphisation.
+Because the space of possible base configurations is too large to evaluate, we informally evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems without using monomorphisation.
 The best performing configuration became the base configuration, which we used as a basis to evaluate the different monomorphisation options.
-Our informal evaluation found that the so-called \verb|40_e_lift| configuration performed best by proving 136 problems.
+Our informal evaluation found that the so-called \verb|40_e_lift| configuration performed best by proving xxx problems.
 %[TODO recheck this]
-We conducted additionnal informal preliminary evaluations (on the benchmark suite of 500 problems)?
-to find appropriate default values and approriate test ranges for the monomorphisation bounds.
-We started with the base configuration \verb|40_e_lift|, with the following values for the monomorphisation bounds:
+We conducted additionnal informal preliminary evaluations (on the 500 problems benchmark suite) to find appropriate default values and test ranges for the monomorphisation bounds.
+We started with the base configuration \verb|40_e_lift|, with the following default values for the monomorphisation bounds:
 
@@
 Departing from the original \verb|40_e_lift| configuration, which would otherwise call E after 8\% of the total alloted time for Zipperposition has passed.
+All others options either appear explicitely in the following tables or have been left unmodified.
 
@@
 This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed.
-Performance of the monomorphisation algorithm seems to plateau for a large range of the tested values for these bounds and drops off sharply beyond.
+Performance of the monomorphisation algorithm seems to plateau for a large range of the tested values drops off sharply beyond.
 %TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with all additionnal problems that are easily monomorphised)
@@
 The multiplier limits the total number of newly generated formulae based on the problem's initial number of formulae.
-The performance of the E prover is highly dependent on the number of formulae it is given.
+The performance of the E prover depends on the number of formulae it is given.
 To account for this, we test the effects of the amount of time allocated to E on the monomorphised problem against the size of the monomorphised problem.
@@
 We conducted two evaluations of Zipperpostion with the monomorphisation algorithm and E as a backend.
-The first was on the 500 problems we used for option optimisatio and the second on the remaining 1034 problems.
+The first was on the 500 problems we used for option optimisation and the second on the remaining 1034 problems.
 This version of Zipperposition was compared against the base Zipperposition prover.
 The base Zipperposition instance was run wiht the sequential portfolio mode \verb|portfolio.sequential.py|.
+The results of these evaluation seem to suggest that some amount of overfitting did take place during the option optimisation phase because the number of problems solved only by the base version of Zipperposition grows disproportionately relative to the total number of problems.
 
@@
 To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, several competitive higher-order polymorphic provers were run on the set of TPTP test problems.
-The total number of problems solved the ``Native'' results shown in the table.
 Evaluation of the monomorphisation approach is done in two steps, first each problem is monomorphised using the [insert configuration] configurations [options?].
@@
 In addition to the polymorphic provers used in the ``Native'' tests, two monomorphic provers are run on the monomorphised problems.
-Note that instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time is subtracted to compare fairly against the ``Native'' column which does not have a similar preprocessing phase.
+Note that instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time (rounded up to the nearest second) is subtracted to compare fairly against the ``Native'' column which does not have a similar preprocessing phase.
 
@@
    E  &   & 0 & 0 \\
-   Leo-III & 0 & 0 & 0 \\ % Leo-III's performance is comparatively very poor, not coherent with casc results
+   Leo-III & 157 with E & 0 & 0 \\ % Leo-III's performance is comparatively very poor, not coherent with casc results
+   Leo-III &  with cvc4 & 0 & 0 \\ % Leo-III's performance is comparatively very poor, not coherent with casc results
    Satallax &  & 0 & 0 \\

### 2024-06-25 iterative_monomorphisation.tex
filled in tables 1-3 with new results, removed old results from tables 4-5
filled in tables with new values and edited a couple of paragraphs accordingly

@@
 Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group.
-If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected, and shown in bold in Tables \ref{mono_ty_args} to \ref{mono_time}.
-
-As a precaution against overfitting, the part of the evaluation related to parameter optimisation and all related preliminary evaluations were carried out on 500 randomly chosen problems out of the 1535 selected problems.
-The rest of the evaluation was carried out on the remaining 1034 problems.
+If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected, they are shown in bold in Tables \ref{mono_ty_args} to \ref{mono_time}.
+
+As a precaution against overfitting, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems.
+We carried out the rest of the evaluation on the remaining 1034 problems.
 Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options used to run Zipperposition.
 Zipperposition has a portfolio mode with several configurations.
-Because the space of possible base configurations is too large to evaluate, we informally evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems without using monomorphisation.
+Because the space of possible base configurations is too large to evaluate, we informally?
+evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, without using monomorphisation.
 The best performing configuration became the base configuration, which we used as a basis to evaluate the different monomorphisation options.
-Our informal evaluation found that the so-called \verb|40_e_lift| configuration performed best by proving xxx problems.
-%[TODO recheck this]
+Our informal?
+evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
+%[TODO discuss with Jasmin]
 We conducted additionnal informal preliminary evaluations (on the 500 problems benchmark suite) to find appropriate default values and test ranges for the monomorphisation bounds.
-We started with the base configuration \verb|40_e_lift|, with the following default values for the monomorphisation bounds:
+We started the option evaluation process with base configuration \verb|40_b.comb|, with the following default values for the monomorphisation bounds:
 
@@
 Monomorphistion (and the subsequent call to E) occurs at the beginning of the invocation of Zipperposition.
-Departing from the original \verb|40_e_lift| configuration, which would otherwise call E after 8\% of the total alloted time for Zipperposition has passed.
+Departing from the original \verb|40_b.comb| configuration, which would otherwise call E after 15\% of the total alloted time for Zipperposition has passed.
 All others options either appear explicitely in the following tables or have been left unmodified.
@@
     \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
-    0         & 170 & 151 & 146 & 136 & 170 & 151 & 144 & 137 &\bf{171}&149&137 & 132\\
-    1         & 130 & 130 & 133 & 132 & 131 & 131 & 131 & 131 & 131 & 131 & 131 & 130\\
-    2         & 130 & 131 & 131 & 132 & 131 & 131 & 132 & 130 & 129 & 130 & 130 & 130\\
-    \(\infty\)& 131 & 132 & 131 & 131 & 131 & 131 & 130 & 131 & 145 & 144 & 146 & 147\\
+    0       &\bf{183}& 167 & 166 & 163 & 183 & 168 & 165 & 163 & 183 & 167 & 166 & 164 \\
+    1          & 160 & 163 & 162 & 164 & 158 & 159 & 160 & 162 & 160 & 161 & 160 & 160 \\
+    2          & 161 & 160 & 161 & 160 & 159 & 159 & 160 & 161 & 160 & 160 & 161 & 159 \\
+    \(\infty\) & 159 & 160 & 162 & 159 & 157 & 160 & 159 & 160 & 169 & 168 & 167 & 169 \\
     \bottomrule
@@
 This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed.
-Performance of the monomorphisation algorithm seems to plateau for a large range of the tested values drops off sharply beyond.
+Performance of the monomorphisation algorithm seems to plateau for a large range of the tested values drops off beyond.
 %TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with all additionnal problems that are easily monomorphised)
@@
     \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
-    0          & 124 & 170 & 171 & 170 & 124 & 170 & 171 & 171 & 124 & 170 & 171 & 170\\
-    0.5        & 165&171&\bf{172}& 170 & 165 & 171 & 171 & 170 & 165 & 171 & 171 & 170\\
-    1          & 170 & 171 & 171 & 171 & 170 & 171 & 171 & 171 & 170 & 171 & 171 & 171\\
-    \(\infty\) & 171 & 171 & 171 & 171 & 168 & 169 & 167 & 168 & 124 & 124 & 124 & 124\\
+    0         &125&\bf{188}& 184 & 183 & 125 & 188 & 184 & 183 & 125 & 188 & 184 & 183 \\
+    0.5        & 181 & 188 & 184 & 183 & 181 & 188 & 184 & 183 & 181 & 188 & 184 & 183 \\
+    1          & 186 & 185 & 183 & 183 & 186 & 185 & 183 & 183 & 186 & 185 & 183 & 183 \\
+    \(\infty\) & 177 & 177 & 177 & 177 & 176 & 177 & 177 & 177 & 125 & 125 & 125 & 125 \\
     \bottomrule
@@
 
-The substitution generation phase occurs once all type arguments have been generated.
+%The substitution generation phase occurs once all type arguments have been generated.
 The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related through the role they play in the construction of monomorphising substitutions.
@@
 Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
+The substitution generation phase occurs once all type arguments have been generated.
+The bound limiting the number of monomorphising substiutions per formula is directly related to the ordering which dictates how such monomorphising substitutions are generated.
+The substitution ordering heuristic has a significant impact on monomorphising substitution generation.
+The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first.
+The ``random'' ordering randomly shuffles substitutions.
+Finally, the ``segregation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently.
+Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seems to affect performance when the ``segregation'' heuristic is used.
+% TODO find a better term than "segregation heuristic"
 \begin{table}[th]
 \caption{Evaluation of bounds for substitution generation}
-\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
+\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{5em}}@{}}
    \toprule
-   & &&&& \multicolumn{4}{c}{ordering} \\
-   & \multicolumn{4}{c}{age} & \multicolumn{4}{c}{random} & \multicolumn{4}{c}{segregation}\\
-   \cmidrule(l){2-13}
-   & &&&& \multicolumn{4}{c}{mono subst}\\
-   \multirow{1}{5.5em}{subst cap} & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 & 2 & 5 & 7 & 10 \\
-   \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(l){10-13}
-   %2          & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
-   5          & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 &\bf{171}&170&171\\
-   50         & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
-   100        & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
-   %200        & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 &\bf{171}&170&171\\
-   \(\infty\) & 158 & 158 & 158 & 158& 169 & 169 & 169 & 169& 167 & 170 & 170 & 171\\
-
+   & & ordering \\
+   \multirow{1}{5em}{mono subst} & age & random & segregation\\
+   \cmidrule(lr){2-2}\cmidrule(lr){3-3}\cmidrule(l){4-4}
+   2       & 162 & 183 & 179 \\
+   5       & 162 & 183 & 186 \\
+   7       & 162 & 183 & 186 \\
+   10      & 162 & 183 & \bf{188} \\
    \bottomrule
@@
    \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(l){10-13}
-   2          & 155 & 154 & 155 & 155 & 154 & 154 & 155 & 154 & 155 & 154 & 156 & 137\\
-   5          & 163 & 164 & 164 & 163 & 163 & 164 & 163 & 163 & 163 & 162 & 163 & 137\\
-   10         & 166 & 167 & 166 & 166 & 166 & 166 & 166 & 166 & 165 & 166 & 167 & 137\\
-   30         &169&\bf{171}&170 & 169 & 170 & 170 & 170 & 170 & 163 & 169 & 170 & 124\\
+   2          & 182 & 181 & 183 & 180 & 184 & 184 & 182 & 183 & 176 & 180 & 181 & 149 \\
+   5          & 189 & 189 & 189 & 188&\bf{190}&190& 190 & 190 & 180 & 184 & 187 & 149 \\
+   10         & 189 & 189 & 189 & 188 & 190 & 190 & 190 & 190 & 180 & 184 & 187 & 148 \\
+   30         & 187 & 187 & 187 & 186 & 188 & 188 & 188 & 188 & 175 & 180 & 185 & 125 \\
    \bottomrule
@@
    \midrule
-   2     &\bf{171}&171&171 & 170\\
-   3     & 170 & 171 & 170 & 169\\
-   4     & 170 & 171 & 170 & 169\\
-   5     & 169 & 169 & 169 & 168\\
-   6     & 169 & 169 & 169 & 168\\
+   2     & 191 & 191&\bf{192}&190\\
+   3     & 191 & 191 & 191 & 190 \\
+   4     & 190 & 190 & 190 & 189 \\
+   5     & 190 & 190 & 190 & 189 \\
+   6     & 189 & 190 & 189 & 189 \\
    \bottomrule
@@
    \midrule
-   500 problems benchmark   & 152 & 185 & 195 (-10 +43) \\
-   fresh problems benchmark & 337 & 385 & 419 (-34 +82)\\
+   500 problems benchmark   & 157 & 213 & 219 (-6 +62) \\
+   fresh problems benchmark & 341 & 442 & 467 (-25 +126)\\
    \bottomrule

### 2024-06-26 iterative_monomorphisation.tex
a couple new numbers

@@
    E  &   & 0 & 0 \\
-   Leo-III & 157 with E & 0 & 0 \\ % Leo-III's performance is comparatively very poor, not coherent with casc results
-   Leo-III &  with cvc4 & 0 & 0 \\ % Leo-III's performance is comparatively very poor, not coherent with casc results
+   Leo-III &  157 with E & 0 & 0 \\ 
+   Leo-III &  68 with cvc4 & 0 & 0 \\ % same as without anything for the native case
    Satallax &  & 0 & 0 \\
-   Vampire & 0 & 0 & 0 \\ % Vampire's parser fails on too many polymorphic problems (about 1/3 of them)
-   Zipperposition & 335 & 0 & 0 \\[1.5\jot]
+   %Vampire & 0 & 0 & 0 \\ % Vampire's parser fails on too many polymorphic problems (about 1/3 of them)
+   Zipperposition & 339 & 0 & 0 \\[1.5\jot]
    Total & 0 & 0 & 0 \\

### 2024-07-08 iterative_monomorphisation.tex
edited tables and comments

@@
 The first part of the evaluation process is to find appropriate values for these bounds.
-The second part compares the performance of Zipperposition without and with monomorphisation.
+The second part compares the performance of Zipperposition with and without monomorphisation.
 The third part compares the performance of different provers on polymorphic problems and their monomorphised counterparts.
@@
 Because Zipperposition does not support reasoning with real numbers, we removed all problems that include real numbers.
-In total, our benchmark suite contains 1535 polymorphic problems.
+In total, our benchmark suite contains 1534 polymorphic problems.
 We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in under 30 seconds with a single thread.
@@
 Zipperposition has a portfolio mode with several configurations.
-Because the space of possible base configurations is too large to evaluate, we informally?
-evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, without using monomorphisation.
+Because the space of possible base configurations is too large to evaluate, we (informally?) evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, without using monomorphisation.
 The best performing configuration became the base configuration, which we used as a basis to evaluate the different monomorphisation options.
-Our informal?
-evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
+Our (informal?) evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
 %[TODO discuss with Jasmin]
-We conducted additionnal informal preliminary evaluations (on the 500 problems benchmark suite) to find appropriate default values and test ranges for the monomorphisation bounds.
+We conducted additionnal informal preliminary evaluations on the 500 problems benchmark suite to find appropriate default values and test ranges for the monomorphisation bounds.
 We started the option evaluation process with base configuration \verb|40_b.comb|, with the following default values for the monomorphisation bounds:
@@
 
-Table~\ref{mono_ty_args} groups bounds that influence the maximum number of newly generated monomorphic type arguments per formuala per iteration.
+Table~\ref{mono_ty_args} groups bounds that control the maximum number of newly generated monomorphic type arguments per formuala per iteration.
 The limit on newly generated type arguments is determined by three components that form a natural group of bounds.
@@
     \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
-    0       &\bf{183}& 167 & 166 & 163 & 183 & 168 & 165 & 163 & 183 & 167 & 166 & 164 \\
-    1          & 160 & 163 & 162 & 164 & 158 & 159 & 160 & 162 & 160 & 161 & 160 & 160 \\
-    2          & 161 & 160 & 161 & 160 & 159 & 159 & 160 & 161 & 160 & 160 & 161 & 159 \\
-    \(\infty\) & 159 & 160 & 162 & 159 & 157 & 160 & 159 & 160 & 169 & 168 & 167 & 169 \\
+    0       &\bf{178}& 161 & 161 & 156 & 178 & 160 & 160 & 156 & 178 & 161 & 160 & 156 \\
+    1          & 155 & 155 & 155 & 158 & 153 & 154 & 154 & 156 & 154 & 154 & 155 & 155 \\
+    2          & 154 & 154 & 153 & 154 & 153 & 153 & 154 & 152 & 154 & 153 & 154 & 154 \\
+    \(\infty\) & 153 & 154 & 153 & 155 & 155 & 153 & 154 & 156 & 159 & 160 & 161 & 161 \\
     \bottomrule
@@
 This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed.
-Performance of the monomorphisation algorithm seems to plateau for a large range of the tested values drops off beyond.
-%TODO check that 124 is what we get if we just run Zipperposition for 10 seconds (maybe with all additionnal problems that are easily monomorphised)
+Performance of the monomorphisation algorithm seems to plateau for certain ranges of the tested values and drops off beyond.
+%TODO check that 125 is what we get if we just run Zipperposition for 10 seconds (maybe with all additionnal problems that are easily monomorphised)
 
@@
     \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
-    0         &125&\bf{188}& 184 & 183 & 125 & 188 & 184 & 183 & 125 & 188 & 184 & 183 \\
-    0.5        & 181 & 188 & 184 & 183 & 181 & 188 & 184 & 183 & 181 & 188 & 184 & 183 \\
-    1          & 186 & 185 & 183 & 183 & 186 & 185 & 183 & 183 & 186 & 185 & 183 & 183 \\
-    \(\infty\) & 177 & 177 & 177 & 177 & 176 & 177 & 177 & 177 & 125 & 125 & 125 & 125 \\
+    0         &125&\bf{184}& 182 & 177 & 125 & 184 & 182 & 177 & 125 & 184 & 182 & 177 \\
+    0.5        & 176 & 184 & 182 & 177 & 176 & 184 & 182 & 177 & 176 & 184 & 182 & 177 \\
+    1          & 182 & 181 & 178 & 177 & 182 & 181 & 178 & 177 & 182 & 181 & 178 & 177 \\
+    \(\infty\) & 173 & 174 & 174 & 174 & 174 & 174 & 173 & 173 & 125 & 125 & 125 & 125 \\
     \bottomrule
@@
 The substitution generation phase occurs once all type arguments have been generated.
-The bound limiting the number of monomorphising substiutions per formula is directly related to the ordering which dictates how such monomorphising substitutions are generated.
+The bound limiting the number of monomorphising substitutions per formula is directly related to the ordering which dictates how such monomorphising substitutions are generated.
 The substitution ordering heuristic has a significant impact on monomorphising substitution generation.
@@
 The ``random'' ordering randomly shuffles substitutions.
-Finally, the ``segregation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently.
+Finally, the ``segregation'' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently.
 Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seems to affect performance when the ``segregation'' heuristic is used.
-% TODO find a better term than "segregation heuristic"
+% TODO find a better term than "segregation heuristic" maybe separation ?
 \begin{table}[th]
@@
    \toprule
-   & & ordering \\
+   & \multicolumn{3}{c}{substitution ordering} \\
    \multirow{1}{5em}{mono subst} & age & random & segregation\\
    \cmidrule(lr){2-2}\cmidrule(lr){3-3}\cmidrule(l){4-4}
-   2       & 162 & 183 & 179 \\
-   5       & 162 & 183 & 186 \\
-   7       & 162 & 183 & 186 \\
-   10      & 162 & 183 & \bf{188} \\
+   2  & 161 & 178 & 175 \\
+   5  & 161 & 178 & 180 \\
+   7  & 161 & 178 & 182 \\
+   10 & 161 & 178 &\bf{184} \\
    \bottomrule
@@
 The absolute cap is the maximum number of formulae passed to E.
-The multiplier limits the total number of newly generated formulae based on the problem's initial number of formulae.
+The multiplier limits the total number of newly generated formulae based on the problem's initial number of formulae if it is bigger than the absolute cap.
 The performance of the E prover depends on the number of formulae it is given.
 To account for this, we test the effects of the amount of time allocated to E on the monomorphised problem against the size of the monomorphised problem.
-We observe that the timeout of E has the biggest impact on performance whilst the precise limits on problem size seem to have little effect.
+We observe that performance seems to be similar across configurations where the size of the output problem is sufficiently constrained.
+The effect of the the E timeout seems to be small beyond a minimal thershold.
 
@@
    \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(l){10-13}
-   2          & 182 & 181 & 183 & 180 & 184 & 184 & 182 & 183 & 176 & 180 & 181 & 149 \\
-   5          & 189 & 189 & 189 & 188&\bf{190}&190& 190 & 190 & 180 & 184 & 187 & 149 \\
-   10         & 189 & 189 & 189 & 188 & 190 & 190 & 190 & 190 & 180 & 184 & 187 & 148 \\
-   30         & 187 & 187 & 187 & 186 & 188 & 188 & 188 & 188 & 175 & 180 & 185 & 125 \\
+   2    & 178 & 178 & 179 & 179 & 178 & 178 & 179 & 178 & 169 & 176 & 177 & 149 \\
+   5 &\bf{186}& 186 & 186 & 185 & 186 & 186 & 186 & 186 & 173 & 182 & 185 & 149 \\
+   10   & 186 & 186 & 186 & 185 & 186 & 186 & 186 & 186 & 173 & 182 & 185 & 148 \\
+   30   & 184 & 184 & 184 & 183 & 184 & 184 & 184 & 184 & 168 & 178 & 183 & 125 \\
    \bottomrule
@@
 The number of iterations and timeout seem to have almost no effect on the performance of the monomorphisation algorithm.
+% TODO numbers are not definitive
 
@@
    Leo-III &  157 with E & 0 & 0 \\ 
-   Leo-III &  68 with cvc4 & 0 & 0 \\ % same as without anything for the native case
+   %Leo-III &  68 with cvc4 & 0 & 0 \\ % same as without anything for the native case
    Satallax &  & 0 & 0 \\

### 2024-07-08 iterative_monomorphisation.tex
more

@@
    \item \PolyFloor: \(50\)
-   \item substitution ordering: segregation
+   \item substitution ordering: separation
    \item \SubstLimit: \(\infty\)
@@
 
-Table~\ref{mono_ty_args} groups bounds that control the maximum number of newly generated monomorphic type arguments per formuala per iteration.
+Table~\ref{mono_ty_args} groups bounds that control the maximum number of newly generated monomorphic type arguments per formula per iteration.
 The limit on newly generated type arguments is determined by three components that form a natural group of bounds.
@@
 The ``random'' ordering randomly shuffles substitutions.
-Finally, the ``segregation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently.
+Finally, the ``separation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently.
 Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
@@
 The ``random'' ordering randomly shuffles substitutions.
-Finally, the ``segregation'' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently.
-Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seems to affect performance when the ``segregation'' heuristic is used.
-% TODO find a better term than "segregation heuristic" maybe separation ?
+Finally, the ``separation'' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently.
+Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seems to affect performance when the ``separation'' heuristic is used.
+% TODO find a better term than "separation heuristic" maybe separation ?
 \begin{table}[th]
@@
    & \multicolumn{3}{c}{substitution ordering} \\
-   \multirow{1}{5em}{mono subst} & age & random & segregation\\
+   \multirow{1}{5em}{mono subst} & age & random & separation\\
    \cmidrule(lr){2-2}\cmidrule(lr){3-3}\cmidrule(l){4-4}
@@
 
+  * monomorphisation:
+    Isabelle for \textit{smt} tactic (Böhme), \textit{metis}, Sledgehammer
+    Meson of HOL Light
+
   * explicitly quantified formulas a la TPTP, existentials, how to eliminate

### 2024-07-08 iterative_monomorphisation.tex
related work

@@
 The alternative is to translate polymorphic problems to monomorphic problems.
-One approach is to encode polymorphism using dedicated function or predicate symbols \cite{mono-trans} in a monomorphic logic.
-Another approach is to encode polymorphism using {iterative monomorphisation}, as introduced by B\"ohme \cite[Section 2.2.1]{sb-phd}.
+One approach is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic;
+we refer to Blanchette et al.\ \cite[Section~9]{mono-trans} for an overview of this approach.
+Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme \cite[Section 2.2.1]{sb-phd}.
 This method relies on heuristically instantiating the formulae's type variables with concrete types.
@@
 However, it is is documented only as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.
+A similar algorithm appears to be implemented in the \texttt{MESON} tactic \cite{harrison-1996} of HOL Light, but it is undocumented.
 
@@
 % target page number: 0.5-1
-\section{Related work}
-\label{sec:related-work}
-
-  * monomorphisation:
-    Isabelle for \textit{smt} tactic (Böhme), \textit{metis}, Sledgehammer
-    Meson of HOL Light
-
-  * explicitly quantified formulas a la TPTP, existentials, how to eliminate
-
-  * Böhme
+%\section{Related work}
+%\label{sec:related-work}
+%
+%  * monomorphisation:
+%    Isabelle for \textit{smt} tactic (Böhme), \textit{metis}, Sledgehammer
+%  * a precursor for this work is the \verb|POLY_ASSUME_TAC| preprocessor of
+%    MESON for HOL Light. It is unfortunately not documented beyond the comment
+%    ``Push duplicated copies of poly theorems to match existing assumptions''
+%    but it appears to perform at least partial monomorphisation.
+%
+%  * alternative: encode polymorphism
+%    * many approaches to encode monomorphic and polymorphic types, starting
+%      with Enderton \cite{xxx}, Stickel \cite{xxx}, and Wick and McCune
+%      \cite{xxx}
+%    * Blanchette et al.~\cite{xxx} exploit ``monotonicity'' to erase type
+%      information
+%    * see Blanchette et al.~\cite[Section~9]{xxx} for a more detailed coverage
+%      of encodings of polymorphism
+%    
+%%\li{https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette/enc_types_article.pdf}
 

### 2024-07-08 iterative_monomorphisation.tex
more
concl

@@
 Our evaluation demonstrates that this approach increases Zipperposition's success rate.
-Moreover, we find iterative monomorphisation to be competitive with native implementations of polymorphism.
+Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.
 \end{abstract}
@@
 
-In this paper, we present an algorithm based on our understanding of his description and implementation (Section~\ref{sec:high level-algorithm}).
+In this paper, we present an algorithm based on our understanding of Böhme's description and implementation (Section~\ref{sec:high level-algorithm}).
 We also provide a more detailed description to help future implementers.
@@
 
-% summary
-
-% Future avenues
+We described a procedure for iteratively instantiating polymorphic types to
+produce monomorphic problems as a preprocessor. Our primary motivation was to
+improve the success rate of Zipperposition and its monomorphic E backend, and
+indeed our evaluation shows a clear improvement. We also saw that even with
+automatic provers that support polymorphism, iterative monomorphisation is a
+better alternative in practice.
+
+We see the following avenues for future work. First, iterative monomorphisation
+blindly enumerates candidate instantiations, without exploiting any knowledge
+about the logical structure of the formulas in which symbols occur. For
+example, a lemma $\sym{p}\langle\alpha\rangle$ cannot be used to prove the
+conjecture $\lnot \sym{p}\langle\ty{nat}\rangle$ because of the incompatible
+polarities, but our procedure instantiates $\alpha$ with $\ty{nat}$ regardless.
+Second, some automatic provers as well as tools like Sledgehammer include
+a relevance filter that heuristically select a subset of the available axioms;
+filters such as SInE \cite{xxx} and MePo \cite{xxx} are iterative and could be
+interleaved with monomorphisation. Third, although one would expect native
+implementations of polymorphism to outperform any preprocessor, our evaluation
+clearly shows that this is not the case, therefore indicating that there is
+considerable room for improvement on the native front.
 
@@
 
-* Sascha Böhme
+We thank Sascha Böhme for fruitful discussions.
+
+This research is co-funded by the European Union (ERC, Nekoka, 101083038).
+Views and opinions expressed are however those of the authors only and do not
+necessarily reflect those of the European Union or the European Research
+Council. Neither the European Union nor the granting authority can be held
+responsible for them.
 

### 2024-07-08 iterative_monomorphisation.tex
more

@@
 Zipperposition has a portfolio mode with several configurations.
-Because the space of possible base configurations is too large to evaluate, we (informally?) evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, without using monomorphisation.
+Because the space of possible base configurations is too large to evaluate, we evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, without using monomorphisation.
 The best performing configuration became the base configuration, which we used as a basis to evaluate the different monomorphisation options.
-Our (informal?) evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
-%[TODO discuss with Jasmin]
+Our preliminary evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
 We conducted additionnal informal preliminary evaluations on the 500 problems benchmark suite to find appropriate default values and test ranges for the monomorphisation bounds.

### 2024-07-08 iterative_monomorphisation.tex
more

@@
 polarities, but our procedure instantiates $\alpha$ with $\ty{nat}$ regardless.
-Second, some automatic provers as well as tools like Sledgehammer include
-a relevance filter that heuristically select a subset of the available axioms;
-filters such as SInE \cite{xxx} and MePo \cite{xxx} are iterative and could be
-interleaved with monomorphisation. Third, although one would expect native
-implementations of polymorphism to outperform any preprocessor, our evaluation
-clearly shows that this is not the case, therefore indicating that there is
-considerable room for improvement on the native front.
+Second, some automatic provers as well as tools like Sledgehammer include a
+relevance filter that heuristically selects a subset of the available axioms;
+filters such as MePo \cite{meng-paulson-2009} and SInE
+\cite{hoder-voronkov-2011} are iterative and could be interleaved with
+monomorphisation. Third, although one would expect native implementations of
+polymorphism to outperform any preprocessor, our evaluation clearly shows that
+this is not the case, therefore indicating that there is considerable room for
+improvement on the native front.
 

### 2024-07-08 iterative_monomorphisation.tex
more
more edits and filled in table 5

@@
 It overlaps with \textcolor{ourblueviolet}{max\_substs} but can be useful to set an absolute limit on the size of the final problem.
-%In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of formuale it is given.
+%In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of formulae it is given.
 
@@
 Performance of the monomorphisation algorithm seems to plateau for certain ranges of the tested values and drops off beyond.
-%TODO check that 125 is what we get if we just run Zipperposition for 10 seconds (maybe with all additionnal problems that are easily monomorphised)
+%TODO check that 125 is what we get if we just run Zipperposition for 10 seconds (maybe with all additional problems that are easily monomorphised)
 
@@
 Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for against the number of iteration of the monomorphisation algorithm.
-The number of iterations and timeout seem to have almost no effect on the performance of the monomorphisation algorithm.
-% TODO numbers are not definitive
+Neither the number of iterations nor the monomorphisation timeout seem to have any significant effect on the performance of the monomorphisation algorithm.
 
@@
    \midrule
-   2     & 191 & 191&\bf{192}&190\\
-   3     & 191 & 191 & 191 & 190 \\
-   4     & 190 & 190 & 190 & 189 \\
-   5     & 190 & 190 & 190 & 189 \\
-   6     & 189 & 190 & 189 & 189 \\
+   2  &\bf{186}& 186 & 186 & 185 \\
+   3     & 186 & 186 & 186 & 185 \\
+   4     & 186 & 186 & 186 & 185 \\
+   5     & 185 & 185 & 186 & 184 \\
+   6     & 186 & 185 & 185 & 184 \\
    \bottomrule
@@
 
-We conducted two evaluations of Zipperpostion with the monomorphisation algorithm and E as a backend.
-The first was on the 500 problems we used for option optimisation and the second on the remaining 1034 problems.
-This version of Zipperposition was compared against the base Zipperposition prover.
-The base Zipperposition instance was run wiht the sequential portfolio mode \verb|portfolio.sequential.py|.
-The results of these evaluation seem to suggest that some amount of overfitting did take place during the option optimisation phase because the number of problems solved only by the base version of Zipperposition grows disproportionately relative to the total number of problems.
-
-Note that Zipperposition's portfolio mode seems to consider timeouts to be more of a polite suggestion than a hard limit, we consequently added a manual timeout of 30 seconds.
+We compared the performance of two instances of Zipperposition.
+The first instance is run in the portfolio mode \verb|portfolio.sequential.py| which attempts to prove the problem with various configuration in succession.
+Of these configuration, some may attempt to call E as a backend when the given problem is monomorphic.
+Because all given problems are non-monomorphic, this instance is labeled ``without E''.
+The second instance, is a modification of \verb|portfolio.sequential.py| where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase.
+This instance can successfully call E as a backend because it is able to provide E with a monomorphised problem.
+Each of the two instances is evaluated on the set of the 500 previously used problems and the set of the remaining 1034 problems separately.
+The results of these evaluations seem to suggest that some amount of overfitting took place during the option optimisation phase because the number of problems solved only by the base version of Zipperposition is disproportionately higher in the 1034 problems benchmark relative to the 500 problems benchmark.
 
@@
 \end{table}
-
-\break
 

### 2024-07-09 iterative_monomorphisation.tex
another table done

@@
 Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for against the number of iteration of the monomorphisation algorithm.
-Neither the number of iterations nor the monomorphisation timeout seem to have any significant effect on the performance of the monomorphisation algorithm.
+Neither parameter seems to have a significant effect on the performance of the monomorphisation algorithm.
 
@@
 Each of the two instances is evaluated on the set of the 500 previously used problems and the set of the remaining 1034 problems separately.
-The results of these evaluations seem to suggest that some amount of overfitting took place during the option optimisation phase because the number of problems solved only by the base version of Zipperposition is disproportionately higher in the 1034 problems benchmark relative to the 500 problems benchmark.
+The results of these evaluations seem to suggest that some amount of overfitting took place during the option optimisation phase because the number of problems by the version without E but not by the version with E is disproportionately higher in the 1034 problems benchmark relative to the 500 problems benchmark.
 
@@
    \midrule
-   500 problems benchmark   & 157 & 213 & 219 (-6 +62) \\
-   fresh problems benchmark & 341 & 442 & 467 (-25 +126)\\
+   500 problems benchmark   & 160 & 198 & 207 \\ % -9 +47
+   fresh problems benchmark & 337 & 410 & 434 \\ % -24 +97
    \bottomrule
@@
 To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, several competitive higher-order polymorphic provers were run on the set of TPTP test problems.
-Evaluation of the monomorphisation approach is done in two steps, first each problem is monomorphised using the [insert configuration] configurations [options?].
+Evaluation of the monomorphisation approach is done in two steps, first each problem is monomorphised using the options obtained from the first evaluation phase except for the monomorphisation timeout option which is removed.
 Secondly, each prover is ran on the monomorphised problem and the results are tallied in the ``Mono'' column of the table.
 In addition to the polymorphic provers used in the ``Native'' tests, two monomorphic provers are run on the monomorphised problems.
-Note that instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time (rounded up to the nearest second) is subtracted to compare fairly against the ``Native'' column which does not have a similar preprocessing phase.
-
-The ``Union'' column, adds up the total number of problems solved by both tests. It serves as an indicator of the usefulness of monomorphisation as a tool in a portfolio configuration.
+Instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time (rounded up to the nearest second) is subtracted to compare fairly against the ``Native'' column which does not have a similar preprocessing phase.
 

### 2024-07-09 iterative_monomorphisation.tex
edits

@@
 The first part of the evaluation process is to find appropriate values for these bounds.
-The second part compares the performance of Zipperposition with and without monomorphisation.
+The second part compares the performance of Zipperposition with and without E on polymorphic problems.
 The third part compares the performance of different provers on polymorphic problems and their monomorphised counterparts.
@@
 The benchmarks are taken from the TPTP library \cite{tptp}.
-The TPTP library, in version 8.2.0, contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism.
-Because Zipperposition does not support reasoning with real numbers, we removed all problems that include real numbers.
+Version 8.2.0 of the library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism.
+Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them.
 In total, our benchmark suite contains 1534 polymorphic problems.
@@
 Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed.
-Given the number of bounds, however, this is not feasible.
+Given the number of bounds, however, this is unfeasible.
 Instead, we group closely related bounds together and test combinations of values for bounds in these groups.
 Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group.
-If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected, they are shown in bold in Tables \ref{mono_ty_args} to \ref{mono_time}.
+If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected.
+Winners are shown in bold in Tables \ref{mono_ty_args} to \ref{mono_time}.
 
@@
 We carried out the rest of the evaluation on the remaining 1034 problems.
+
 Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options used to run Zipperposition.
-Zipperposition has a portfolio mode with several configurations.
-Because the space of possible base configurations is too large to evaluate, we evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, without using monomorphisation.
-The best performing configuration became the base configuration, which we used as a basis to evaluate the different monomorphisation options.
+The prover has a portfolio mode with several configurations.
+Because the space of possible base configurations is too large to evaluate
+exhaustively, we evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, but we disabled E.
 Our preliminary evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
-We conducted additionnal informal preliminary evaluations on the 500 problems benchmark suite to find appropriate default values and test ranges for the monomorphisation bounds.
-We started the option evaluation process with base configuration \verb|40_b.comb|, with the following default values for the monomorphisation bounds:
+This configuration became the base configuration, which we used as a basis to evaluate the monomorphisation options.
+
+We conducted additional informal evaluations on the 500 problems benchmark suite to find appropriate default values and test ranges for the monomorphisation bounds.
+We started the option evaluation process with the base configuration and the following default values for the monomorphisation bounds
+and the E invocation timeout:
 
@@
 \begin{itemize}
-   \item \MonoCap: irrelevant
-   \item \MonoMult: irrelevant
-   \item \MonoFloor: irrelevant
    \item \PolyCap: \(\infty\)
@@
    \item new formulae limit multiplier: \(0\)
-   \item E timeout: \(30\)
    \item monomorphisation timeout: \(20\)
    \item \Loop: \(4\)
+   \item E timeout: \(30\)
+   \item[] ~
 \end{itemize}
 \end{multicols}
-Monomorphistion (and the subsequent call to E) occurs at the beginning of the invocation of Zipperposition.
+\noindent
+The initial values for the \MonoCap, \MonoMult and \MonoFloor options are
+irrelevant because the options' values are set when computing
+Table~\ref{mono_ty_args} and used for subsequent tables.
+
+In all our experiments, monomorphisation and the subsequent call to E occur at the beginning of the invocation of Zipperposition.
 Departing from the original \verb|40_b.comb| configuration, which would otherwise call E after 15\% of the total alloted time for Zipperposition has passed.
@@
    & \multicolumn{4}{c}{mono time} \\
-   \multirow{1}{4em}{loop nb} & 5 & 10 & 20 & 30\\
+   \multirow{1}{5em}{num.\ loops} & 5 & 10 & 20 & 30\\
    \midrule

### 2024-07-12 iterative_monomorphisation.tex
added a couple of tables

@@
 
-Table \ref{pb_size} groups together the bounds related to the size of the output problem to be passed to the E prover.
+Table \ref{pb_size} groups the bounds related to the size of the output problem to be passed to the E prover.
 The absolute cap is the maximum number of formulae passed to E.
-The multiplier limits the total number of newly generated formulae based on the problem's initial number of formulae if it is bigger than the absolute cap.
-The performance of the E prover depends on the number of formulae it is given.
-To account for this, we test the effects of the amount of time allocated to E on the monomorphised problem against the size of the monomorphised problem.
-We observe that performance seems to be similar across configurations where the size of the output problem is sufficiently constrained.
-The effect of the the E timeout seems to be small beyond a minimal thershold.
+The multiplier limits the total number of newly generated formulae based on the problem's initial number of formulae.
+We find that the E prover tends to perform better when given a limited number of formulae.
 
@@
 \caption{Evaluation of bounds directly related to the size of the output problem}
-\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
+\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
    \toprule
-   & &&&& \multicolumn{4}{c}{formula cap} \\
-   & \multicolumn{4}{c}{500} & \multicolumn{4}{c}{2000} & \multicolumn{4}{c}{\(\infty\)}\\
-   \cmidrule(l){2-13}
-   & &&&& \multicolumn{4}{c}{formula multiplier}\\
-   \multirow{1}{5.5em}{E timeout (s)} & 1 & 2 & 3 & \(\infty\)& 1 & 2 & 3 & \(\infty\)& 1 & 2 & 3 & \(\infty\)\\
-   \cmidrule(lr){2-5}\cmidrule(lr){6-9}\cmidrule(l){10-13}
-   2    & 178 & 178 & 179 & 179 & 178 & 178 & 179 & 178 & 169 & 176 & 177 & 149 \\
-   5 &\bf{186}& 186 & 186 & 185 & 186 & 186 & 186 & 186 & 173 & 182 & 185 & 149 \\
-   10   & 186 & 186 & 186 & 185 & 186 & 186 & 186 & 186 & 173 & 182 & 185 & 148 \\
-   30   & 184 & 184 & 184 & 183 & 184 & 184 & 184 & 184 & 168 & 178 & 183 & 125 \\
+   & \multicolumn{4}{c}{formula multiplier}\\
+   \multirow{1}{5.2em}{formula cap} & 1 & 2 & 3 & \(\infty\)\\
+   \midrule
+   500       &\bf{184}& 184 & 184 & 183 \\
+   2000         & 184 & 184 & 184 & 184 \\
+   \(\infty\)   & 168 & 178 & 183 & 125 \\
    \bottomrule
@@
    \midrule
+   1     & 183 & 184 & 183 & 183  \\
    2  &\bf{186}& 186 & 186 & 185  \\
@@
    4     & 186 & 186 & 186 & 185  \\
-   5     & 185 & 185 & 186 & 184 \\
-   6     & 186 & 185 & 185 & 184 \\
+   5     & 185 & 185 & 185 & 184  \\
    \bottomrule
@@
 \label{mono_time}
+\end{table}
+
+Table \ref{e_settings} evaluates the options with which we call the E prover.
+The call point corresponds to a fraction of the total time alloted to Zipperposition.
+Zipperposition is run for \(\text{timeout} \times \text{call point}\) seconds before monomorphisation and the subsequent call to E occur.
+The E timeout limits how lond E is run before being interrupted and resuming the normal operation of Zipperposition.
+The longer Zipperposition is run, the more formulae are generated, when taking into account the explosive nature of monomorphisation ...
+
+\begin{table}[th]
+\caption{Evaluation of parameters related to the E prover}
+\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
+   \toprule
+   & \multicolumn{4}{c}{E call point} \\
+   \multirow{1}{4em}{E timeout} & 0.0 & 0.1 & 0.2 & 0.3\\
+   \midrule
+   2      & 180 & 143 & 132 & 124 \\
+   5   &\bf{185}& 142 & 134 & 125 \\
+   10     & 184 & 143 & 132 & 125 \\
+   20     & 184 & 137 & 133 & 125 \\
+   30     & 182 & 133 & 134 & 125 \\
+   \bottomrule
+\end{tabular}
+\label{e_settings}
 \end{table}

### 2024-07-12 iterative_monomorphisation.tex
more numbers

@@
 Given the number of bounds, however, this is unfeasible.
-Instead, we group closely related bounds together and test combinations of values for bounds in these groups.
+Instead, we group closely related bounds together and test combinations of values for the bounds in these groups.
 Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group.
 If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected.
-Winners are shown in bold in Tables \ref{mono_ty_args} to \ref{mono_time}.
+Winners are shown in bold in Tables \ref{mono_ty_args} to \ref{e_settings}.
 
 As a precaution against overfitting, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems.
-We carried out the rest of the evaluation on the remaining 1034 problems.
+We carried out the rest of the evaluations on the remaining 1034 problems.
 
@@
 The prover has a portfolio mode with several configurations.
-Because the space of possible base configurations is too large to evaluate
-exhaustively, we evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, but we disabled E.
+Because the space of possible base configurations is too large to evaluate exhaustively, we evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, but we disabled E.
 Our preliminary evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
@@
 We conducted additional informal evaluations on the 500 problems benchmark suite to find appropriate default values and test ranges for the monomorphisation bounds.
-We started the option evaluation process with the base configuration and the following default values for the monomorphisation bounds
-and the E invocation timeout:
+We started the option evaluation process with the base configuration and the following default values for monomorphisation bounds and parameters for E:
 
@@
    \item E timeout: \(30\)
-   \item[] ~
+   \item E call point: \(0\)
 \end{itemize}
@@
 \noindent
-The initial values for the \MonoCap, \MonoMult and \MonoFloor options are
-irrelevant because the options' values are set when computing
-Table~\ref{mono_ty_args} and used for subsequent tables.
-
-In all our experiments, monomorphisation and the subsequent call to E occur at the beginning of the invocation of Zipperposition.
-Departing from the original \verb|40_b.comb| configuration, which would otherwise call E after 15\% of the total alloted time for Zipperposition has passed.
-All others options either appear explicitely in the following tables or have been left unmodified.
+The initial values for the \MonoCap, \MonoMult and \MonoFloor options are irrelevant because the options' values are set when computing Table~\ref{mono_ty_args}.% and used for subsequent tables.
 
@@
 Table \ref{e_settings} evaluates the options with which we call the E prover.
-The call point corresponds to a fraction of the total time alloted to Zipperposition.
-Zipperposition is run for \(\text{timeout} \times \text{call point}\) seconds before monomorphisation and the subsequent call to E occur.
+The call point corresponds to the fraction of the total timeout of Zipperposition's timeout after which the normal operation of Zipperposition is interrupted to begin the monomorphisation process.
 The E timeout limits how lond E is run before being interrupted and resuming the normal operation of Zipperposition.
-The longer Zipperposition is run, the more formulae are generated, when taking into account the explosive nature of monomorphisation ...
+The longer Zipperposition is run, the more formulae are generated this and the explosive nature of iterative monomorphisation likely explains the poor performance of greater values of E call point.
 
@@
    & \multicolumn{4}{c}{E call point} \\
-   \multirow{1}{4em}{E timeout} & 0.0 & 0.1 & 0.2 & 0.3\\
+   \multirow{1}{4em}{E timeout} & 0 & 0.1 & 0.2 & 0.3\\
    \midrule
@@
 Each of the two instances is evaluated on the set of the 500 previously used problems and the set of the remaining 1034 problems separately.
-The results of these evaluations seem to suggest that some amount of overfitting took place during the option optimisation phase because the number of problems by the version without E but not by the version with E is disproportionately higher in the 1034 problems benchmark relative to the 500 problems benchmark.
+The results of these evaluations seem to suggest that no significant overfitting took place during the option optimisation phase.
+
+Table \ref{zipp_eval} shows that the use of E as a backend markedly improves the performance of Zipperposition.
+It is not a strict improvement as a number of problems are solved without E and not with E.
 
@@
    \bottomrule
+   \label{zipp_eval}
 \end{tabular}
@@
    \midrule
-   E  &   & 0 & 0 \\
-   Leo-III &  157 with E & 0 & 0 \\ 
+   E              &   & 340 & 340 \\
+   Leo-III with E &  157 & 231 & 0 \\ 
+   Zipperposition & 339 & 351 & 404 \\[1.5\jot]
+   Total          & 0 & 0 & 0 \\
+   % Satallax       &  & 0 & 0 \\ % doesn't understand negated conjectures
    %Leo-III &  68 with cvc4 & 0 & 0 \\ % same as without anything for the native case
-   Satallax &  & 0 & 0 \\
    %Vampire & 0 & 0 & 0 \\ % Vampire's parser fails on too many polymorphic problems (about 1/3 of them)
-   Zipperposition & 339 & 0 & 0 \\[1.5\jot]
-   Total & 0 & 0 & 0 \\
    \bottomrule

### 2024-07-12 iterative_monomorphisation.tex
more

@@
 
-The benchmarks are taken from the TPTP library \cite{tptp}.
-Version 8.2.0 of the library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism.
+The benchmarks are taken from version 8.2.0 of the TPTP library \cite{tptp}.
+The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism.
 Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them.
@@
 If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected.
-Winners are shown in bold in Tables \ref{mono_ty_args} to \ref{e_settings}.
+Winning entries are shown in bold in Tables \ref{mono_ty_args} to \ref{e_settings}.
 
@@
 
-We conducted additional informal evaluations on the 500 problems benchmark suite to find appropriate default values and test ranges for the monomorphisation bounds.
+We conducted additional informal evaluations on the 500 problem benchmark suite to find appropriate default values and test ranges for the monomorphisation bounds.
 We started the option evaluation process with the base configuration and the following default values for monomorphisation bounds and parameters for E:
@@
 
-For larger problems, the monomorphisation algorithm may timeout despite the bounds.
+For larger problems, the monomorphisation algorithm may time out despite the bounds.
 In these cases neither Zipperposition nor E will have had a chance to attempt to solve the problem.
@@
 
-Table \ref{e_settings} evaluates the options with which we call the E prover.
+Table \ref{e_settings} shows the impact of the options with which we call the E prover.
 The call point corresponds to the fraction of the total timeout of Zipperposition's timeout after which the normal operation of Zipperposition is interrupted to begin the monomorphisation process.
-The E timeout limits how lond E is run before being interrupted and resuming the normal operation of Zipperposition.
-The longer Zipperposition is run, the more formulae are generated this and the explosive nature of iterative monomorphisation likely explains the poor performance of greater values of E call point.
+The E timeout limits for how long E is run before being interrupted, at which point Zipperposition resumes normal operation.
+The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more iterative monomorphisation is explosive.
+This likely explains the poor performance for greater values of E call point.
 
@@
 
-We compared the performance of two instances of Zipperposition.
-The first instance is run in the portfolio mode \verb|portfolio.sequential.py| which attempts to prove the problem with various configuration in succession.
+We compare the performance of two instances of Zipperposition.
+The first instance is run in the portfolio mode \verb|portfolio.sequential.py|, which attempts to prove the problem with various configuration in succession.
 Of these configuration, some may attempt to call E as a backend when the given problem is monomorphic.
@@
 Each of the two instances is evaluated on the set of the 500 previously used problems and the set of the remaining 1034 problems separately.
-The results of these evaluations seem to suggest that no significant overfitting took place during the option optimisation phase.
-
-Table \ref{zipp_eval} shows that the use of E as a backend markedly improves the performance of Zipperposition.
-It is not a strict improvement as a number of problems are solved without E and not with E.
+The results of these evaluations do not suggest that any overfitting took place during the option optimisation phase.
 
@@
    \midrule
-   500 problems benchmark   & 160 & 198 & 207 \\ % -9 +47
-   fresh problems benchmark & 337 & 410 & 434 \\ % -24 +97
+   500 problem benchmark   & 160 & 198 & 207 \\ % -9 +47
+   1034 problem benchmark & 337 & 410 & 434 \\ % -24 +97
    \bottomrule
@@
 
+Table \ref{zipp_eval} shows that the use of E as a backend markedly improves the performance of Zipperposition. It is not a strict improvement, since some problems are solved without E and not with E.
+
 \subsection{Monomorphisation as preprocessor}
-
-To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, several competitive higher-order polymorphic provers were run on the set of TPTP test problems.
-Evaluation of the monomorphisation approach is done in two steps, first each problem is monomorphised using the options obtained from the first evaluation phase except for the monomorphisation timeout option which is removed.
-Secondly, each prover is ran on the monomorphised problem and the results are tallied in the ``Mono'' column of the table.
-In addition to the polymorphic provers used in the ``Native'' tests, two monomorphic provers are run on the monomorphised problems.
-Instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time (rounded up to the nearest second) is subtracted to compare fairly against the ``Native'' column which does not have a similar preprocessing phase.
 
@@
    \midrule
-   E              &   & 340 & 340 \\
+   E              & -- & 340 & 340 \\
    Leo-III with E &  157 & 231 & 0 \\ 
-   Zipperposition & 339 & 351 & 404 \\[1.5\jot]
-   Total          & 0 & 0 & 0 \\
+   Zipperposition & 339 & 351 & 404 \\%[1.5\jot]
+%   Total          & 0 & 0 & 0 \\
    % Satallax       &  & 0 & 0 \\ % doesn't understand negated conjectures
@@
    \bottomrule
+   \label{native_vs_mono_eval}
 \end{tabular}
 \end{table}
+
+To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher-order polymorphic provers were run on the 1034 problem benchmark.
+We chose Leo-III and Zipperposition.
+We unfortunately had to exclude Vampire because of parsing issues.
+
+Table~\ref{native_vs_mono_eval} shows the results.
+ Evaluation of the monomorphisation approach is done in two steps.
+First, each problem is monomorphised using the options obtained from the first evaluation phase except for the monomorphisation timeout option which is removed.
+Second, each prover is ran on the monomorphised problem and the results are tallied in the table's ``Mono'' column.
+In addition to the polymorphic provers used in the ``Native'' tests, the monomorphic prover E is run on the monomorphised problems to provide an additional yardstick.
+Instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time (rounded up to the nearest second) is subtracted to compare fairly against the ``Native'' column, which does not have a similar preprocessing phase.
 

### 2024-09-11 iterative_monomorphisation.tex
fixing things
put table captios at the bottom of tables
fixed error in algorithms
iterative_monomorphisation.tex

@@
 This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
-Although the example may seem artificial [contrived?], similar problems frequently arise in practice.
+Although the example may seem artificial, similar problems frequently arise in practice.
 For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
@@
 Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples.
-Finally, we keep track of a set \(S\) of substitutions generated by the matchings.
+Finally, we keep track of a map \(S\) from formulae to the substitutions generated by the matchings.
 It is used to generate new formulae in the final phase.
@@
       \ForEach{\(\varphi \in \Problem\)} {
-         \((M_{\Delta}, N_{\Delta}(\varphi), S_{\Delta})\leftarrow \FStep(M_{\text{old}}, N_{\text{old}}(\varphi), M_{\text{new}}, N_{\text{new}}(\varphi))\)\;
-         \(\AllSubst\leftarrow\AllSubst\cup S_{\Delta}\)\;
+         \((M_{\Delta}, N_{\Delta}, S_{\Delta})\leftarrow \FStep(M_{\text{old}}, N_{\text{old}}(\varphi), M_{\text{new}}, N_{\text{new}}(\varphi))\)\;
+         \(\AllSubst(\varphi)\leftarrow\AllSubst(\varphi)\cup S_{\Delta}\)\;
          \(M_{\text{next}}\leftarrow M_{\text{next}}\cup M_{\Delta}\)\;
-         \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\cup N_{\Delta}(\varphi)\)\;
+         \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\cup N_{\Delta}\)\;
       }
@@
 
-   \Fn(){\(\FMonoStep(\varphi, M_{\text{old}}, M_{\text{new}}, N_{\text{old}}(\varphi), N_{\text{new}}(\varphi)\)}{
+   \Fn(){\(\FMonoStep(\varphi, M_{\text{old}}, M_{\text{new}}, N_{\text{old}}(\varphi), N_{\text{new}}(\varphi))\)}{
 
@@
      \strut set \(\Phi\) of polymorphic formulae \\
-     set $S$ of substitutions\strut
+     substitution map $S$\strut
      \end{minipage}}
@@
    \ForEach{\(\varphi\in\Phi\) \ST $\varphi$ is non-monomorphic}{
-      \ForEach{\(\sigma\in \MonoSubst(\AllSubst, \TyVars{\(\varphi\)}, \emptyset, \{\})\)}{
+      \ForEach{\(\sigma\in \MonoSubst(\AllSubst(\varphi), \TyVars{\(\varphi\)}, \emptyset, \{\})\)}{
          \textcolor{ourblueviolet}{
@@
 The last phase uses these substitutions to instantiate the type variables in the input problem's non-monomrophic formuale.
-In the presence of bounds, the order in which the elements of the set \(S\) of substitutions is traversed affects the formulae resulting from the last phase.
+In the presence of bounds, the order in which the elements of the sets of substitutions of \(S\) are traversed affects the formulae resulting from the last phase.
 As we will discuss in Section~\ref{param_opti}, the best results were obtained when substitutions generated in the same iteration are grouped together.
@@
 Such substitutions are computed using a recursive function.
-Given a set \(V\) of type variables and a set \(S\) of substitutions, it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S\) that instantiates at least one of the type variables in \(V\).
+Given a set \(V\) of type variables and a set \(S(\varphi)\) of substitutions, it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S(\varphi)\) that instantiates at least one of the type variables in \(V\).
 It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they can be composed and the function recursively called to instantiate the remaining type variables.
@@
    \item The \emph{mono\_substs} function is the most explosive part of the algorithm.
-If \(S\) contains \(n\) substitutions that each instantiate exactly one of \(v\) different type variables, up to \(n^v\) monomorphising substititions may be generated.
-Recall that the total number of type argument tuples that are used to generate \(S\) can be doubly exponential in the number of loop iterations.
+For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) different type variables, up to \(n^v\) monomorphising substititions may be generated.
+Recall that the total number of type argument tuples that are used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations.
 \end{enumerate}
@@
 \begin{table}[th]
-\caption{Evaluation of bounds for monomorphic type argument generation}
 \centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
@@
 \label{mono_ty_args}
+\caption{Evaluation of bounds for monomorphic type argument generation}
 \end{table}
@@
 \begin{table}[th]
-\caption{Evaluation of bounds for non-monomorphic type argument generation}
 \centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
@@
 \label{nmon_ty_args}
+\caption{Evaluation of bounds for non-monomorphic type argument generation}
 \end{table}
@@
 \begin{table}[th]
-\caption{Evaluation of bounds for substitution generation}
 \centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{5em}}@{}}
@@
 \label{subst_gen_table}
+\caption{Evaluation of bounds for substitution generation}
 \end{table}
@@
 \begin{table}[th]
-\caption{Evaluation of bounds directly related to the size of the output problem}
 \centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
@@
 \label{pb_size}
+\caption{Evaluation of bounds directly related to the size of the output problem}
 \end{table}
@@
 \begin{table}[th]
-\caption{Evaluation of parameters related to the depth of monomorphisation}
 \centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
@@
 \label{mono_time}
+\caption{Evaluation of parameters related to the depth of monomorphisation}
 \end{table}
@@
 \begin{table}[th]
-\caption{Evaluation of parameters related to the E prover}
 \centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
@@
 \label{e_settings}
+\caption{Evaluation of parameters related to the E prover}
 \end{table}
@@
 
-We compare the performance of two instances of Zipperposition.
-The first instance is run in the portfolio mode \verb|portfolio.sequential.py|, which attempts to prove the problem with various configuration in succession.
-Of these configuration, some may attempt to call E as a backend when the given problem is monomorphic.
-Because all given problems are non-monomorphic, this instance is labeled ``without E''.
-The second instance, is a modification of \verb|portfolio.sequential.py| where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase.
-This instance can successfully call E as a backend because it is able to provide E with a monomorphised problem.
-Each of the two instances is evaluated on the set of the 500 previously used problems and the set of the remaining 1034 problems separately.
-The results of these evaluations do not suggest that any overfitting took place during the option optimisation phase.
-
 \begin{table}[ht]
-\caption{Evaluation of Zipperposition without E vs. with E}
 \centering\begin{tabular}{@{}lccc@{}}
@@
 \end{tabular}
+\caption{Evaluation of Zipperposition without E vs. with E}
 \end{table}
 
+We compare the performance of two instances of Zipperposition.
+The first instance is run in the portfolio mode \verb|portfolio.sequential.py|, which attempts to prove the problem with various configuration in succession.
+Of these configuration, some may attempt to call E as a backend when the given problem is monomorphic.
+Because all given problems are non-monomorphic, this instance is labeled ``without E''.
+The second instance, is a modification of \verb|portfolio.sequential.py| where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase.
+This instance can successfully call E as a backend because it is able to provide E with a monomorphised problem.
+Each of the two instances is evaluated on the set of the 500 previously used problems and the set of the remaining 1034 problems separately.
+The results of these evaluations do not suggest that any overfitting took place during the option optimisation phase.
+
 Table \ref{zipp_eval} shows that the use of E as a backend markedly improves the performance of Zipperposition. It is not a strict improvement, since some problems are solved without E and not with E.
@@
 \begin{table}[ht]
-\caption{Evaluation of native polymorphism vs.\ monomorphisation}
 \centering\begin{tabular}{@{}lccc@{}}
@@
    E              & -- & 340 & 340 \\
-   Leo-III with E &  157 & 231 & 0 \\ 
+   Leo-III with E &  157 & 231 & 274 \\ 
    Zipperposition & 339 & 351 & 404 \\%[1.5\jot]
@@
 \end{tabular}
+\caption{Evaluation of native polymorphism vs.\ monomorphisation}
 \end{table}

### 2025-01-22 iterative_monomorphisation_lncs.tex
changed typo in git file name

--- 2024-09-11 iterative_monomorphisation.tex
@@
-\documentclass[]{ceurart}
+\documentclass[runningheads]{llncs}
+%
+\usepackage[T1]{fontenc}
+% T1 fonts will be used to generate the final print and online PDFs,
+% so please use T1 fonts in your manuscript whenever possible.
+% Other font encondings may result in incorrect characters.
+%
+\usepackage{graphicx}
 
@@
 
+\begin{document}
 \SetNlSty{bfseries}{\color{black}}{}
@@
 
-\newdefinition{definition}{Definition}
-\newdefinition{example}[definition]{Example}
+%\newtheorem{definition}{Definition}
+%\newtheorem{example}[definition]{Example}
 
@@
 
-\begin{document}
-
-\copyrightyear{2024}
-\copyrightclause{Copyright for this paper by its authors. Use permitted under Creative Commons License Attribution 4.0 International (CC BY 4.0)}
-
-\conference{Submitted draft} % TODO update
+%\copyrightyear{2024}
+%\copyrightclause{Copyright for this paper by its authors. Use permitted under Creative Commons License Attribution 4.0 International (CC BY 4.0)}
+%
+%\conference{Submitted draft} % TODO update
 
@@
 
-\author[1,2]{Tanguy Bozec}[%
-email=tanguy.bozec@ens-paris-saclay.fr,
-]
-\author[2]{Jasmin Blanchette}[%
-email=jasmin.blanchette@ifi.lmu.de,
-%url=https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette_de.html,
-]
-\address[1]{ENS Paris-Saclay, Université Paris-Saclay, France}
-\address[2]{Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}
+\author{Tanguy Bozec\inst{1,2} \and
+Jasmin Blanchette\inst{2}}
+
+\institute{ENS Paris-Saclay, Université Paris-Saclay, France \and
+Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}
+
+\maketitle
 
 \begin{abstract}
-Monomorphisation can be used to extend monomorphic provers to support polymorphic logics.
+Monomorphisation can be used to extend some monomorphic provers to support polymorphic logics.
 We propose an iterative approach, which is necessarily incomplete but which works well in practice.
@@
 Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.
+
+\keywords{Polymorphism  \and Monomorphism    \and Automatic theorem proving}
 \end{abstract}
-
-\begin{keywords}
-   Polymorphism\sep
-   monomorphism\sep
-   automatic theorem proving
-\end{keywords}
-
-\maketitle
 
@@
 One of the main applications of automatic theorem provers is to provide automation to users of proof assistants.
-Many proof assistants, such as HOL4 \cite{slind-norrish-2008}, HOL Light \cite{harrison-2009}, and
-Isabelle/HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed only at the top level of formulae.
+Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009}, and
+Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed only at the top level of formulae.
 On the other hand, many automatic provers operate only on monomorphic logics.
-One approach to close this gap is to extend provers to natively support polymorphism, as has been done for Vampire \cite{bhayat-reger-2020}.
+One approach to close this gap is to extend provers to natively support polymorphism, as has been done for Vampire~\cite{bhayat-reger-2020}.
 This, however, entails a lot of work that needs to be redone for every prover.
@@
 One approach is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic;
-we refer to Blanchette et al.\ \cite[Section~9]{mono-trans} for an overview of this approach.
-Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme \cite[Section 2.2.1]{sb-phd}.
+we refer to Blanchette et al.~\cite[Section~9]{mono-trans} for an overview of this approach.
+Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}.
 This method relies on heuristically instantiating the formulae's type variables with concrete types.
@@
 By a type version of the compactness theorem, we have that in first order logic, given a polymorphic formula \(\varphi\), there exist an equisatisfiable finite set of monomorphic instances of \(\varphi\).
-However, such a set cannot be computed \cite[Theorem 1]{expr-poly-types}.
+However, such a set cannot be computed~\cite[Theorem 1]{expr-poly-types}.
 As a result, any monomorphisation method based on instantiation of type variables is
@@
 
-B\"ohme's iterative approach is implemented as part of the SMT (satisfiability modulo theories) integration \cite[Chapter 2]{sb-phd} in Isabelle/HOL.
-This implementation is also used by Sledgehammer \cite{judgement, hammer} to interface with superposition based automatic theorem provers.
-However, it is is documented only as a single subsection in his PhD thesis \cite[Section 2.2.1]{sb-phd}.
-A similar algorithm appears to be implemented in the \texttt{MESON} tactic \cite{harrison-1996} of HOL Light, but it is undocumented.
+B\"ohme's iterative approach is implemented as part of the SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd} in Isabelle/HOL.
+This implementation is also used by Sledgehammer~\cite{judgement,hammer} to interface with superposition based automatic theorem provers.
+However, it is is documented only as a single subsection in his PhD thesis~\cite[Section 2.2.1]{sb-phd}.
+A similar algorithm appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
 
@@
 
-We implemented iterative monomorphisation in Zipperposition \cite{zipp}, a higher order prover written in OCaml.
-Although Zipperposition is polymorphic, it uses the monomorphic prover E \cite{e} as a backend.
+We implemented iterative monomorphisation in Zipperposition~\cite{zipp}, a higher order prover written in OCaml.
+Although Zipperposition is polymorphic, it uses the monomorphic prover E~\cite{e} as a backend.
 This means that E can now be used with polymorphic problems.
@@
 
-Our empirical evaluation on TPTP \cite{tptp} problems attempts to answer three questions (Section~\ref{sec:evaluation}):
+Our empirical evaluation on TPTP~\cite{tptp} problems attempts to answer three questions (Section~\ref{sec:evaluation}):
 \begin{enumerate}
@@
 
-\item For Leo-III \cite{leo-iii} and Vampire \cite{vamp}, we find that monomorphisation is indeed more effective than native polymorphism.
+\item For Leo-III~\cite{leo-iii} and Vampire~\cite{vamp}, we find that monomorphisation is indeed more effective than native polymorphism.
 \end{enumerate}
@@
 Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic.
-In particular, it can operate in the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.
+In particular, it can operate in the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.
 
@@
 \begin{definition}
-Two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_m\mapsto\tau_m\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_n\mapsto\upsilon_n\}\) are \emph{compatible} if \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\) for all \(i, j\).
+Given two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_m\mapsto\tau_m\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_n\mapsto\upsilon_n\}\) they are said to be \emph{compatible} if \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\) for all \(i, j\).
 \end{definition}
@@
    \begin{enumerate}
-    \item[2.1.] For each tuple \((\tau_1, \dots, \tau_n) \in  M(f)\) and each tuple \((\upsilon_1, \dots, \upsilon_n) \in N(\varphi)(f)\),
+    \item[2.1.] For each tuple \((\tau_1, \dots, \tau_n) \in  M(f)\) and \((\upsilon_1, \dots, \upsilon_n) \in N(\varphi)(f)\),
      for each \(i\), match \(\upsilon_i\) against \(\tau_i\), yielding the substitution \(\sigma_i\) in case of success.
@@
 \begin{enumerate}
-   \item[5.] \(\forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,}\; \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
+   \item[5.] 
+         \( \forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,} \\ \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
 \end{enumerate}
@@
 This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
-Although the example may seem artificial, similar problems frequently arise in practice.
+Although the example may seem contrived, similar problems frequently arise in practice.
 For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
@@
 This allows targeting a simply typed logic with no general support for $n$-ary type constructors.
-% \cite[Sections 4 and 5]{mono-trans}.
+%~\cite[Sections 4 and 5]{mono-trans}.
 % Paper name is gt
@@
 Introducing bounds addresses this potential issue.
-The limit on the number of new monomorphic type argument tuples is \(\min(\max(\MonoMult\cdot m, \MonoFloor), \MonoCap)\), where \(m\) is the total number of monomorphic type argument tuples.
+The limit on the number of newly generated monomorphic type argument tuples is \(\min(\max(\MonoMult\cdot m, \MonoFloor), \MonoCap)\), where \(m\) is the total number of monomorphic type argument tuples.
 The components of this limit are
@@
 It generates monomorphising substitutions and applies them to the polymorphic formula of the input problem that they instantiate.
-Since the substitutions are monomorphising relative to the formula they applied to, the resulting formulae are monomorphic.
+Since the substitutions are monomorphising relative to the formula they applied to, the resulting formulae will be monomorphic.
 The \textcolor{ourblueviolet}{total\_new\_formulae} bound is used to control the total number of new formulae.
@@
 
-The benchmarks are taken from version 8.2.0 of the TPTP library \cite{tptp}.
+The benchmarks are taken from version 8.2.0 of the TPTP library~\cite{tptp}.
 The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism.
@@
 \begin{table}[th]
-\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
+\caption{Evaluation of bounds for monomorphic type argument generation}
+\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{2.5em}}@{}}
    \toprule
@@
    & &&& \multicolumn{6}{c}{floor} \\
-   \multirow{1}{2em}{mult} & 0 & 50 & 100 & 200& 0 & 50 & 100 & 200& 0 & 50 & 100 & 200\\
+   \multirow{1}{2.5em}{mult} & 0 & 50 & 100 & 200& 0 & 50 & 100 & 200& 0 & 50 & 100 & 200\\
     \cmidrule(lr){2-5} \cmidrule(lr){6-9} \cmidrule(l){10-13} 
@@
 \label{mono_ty_args}
-\caption{Evaluation of bounds for monomorphic type argument generation}
 \end{table}
@@
 \begin{table}[th]
-\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{1.5em}}@{}}
+\caption{Evaluation of bounds for non-monomorphic type argument generation}
+\centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{2.5em}}@{}}
    \toprule
@@
 \label{nmon_ty_args}
-\caption{Evaluation of bounds for non-monomorphic type argument generation}
 \end{table}
@@
 \begin{table}[th]
-\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{5em}}@{}}
+\caption{Evaluation of bounds for substitution generation}
+\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
    \toprule
    & \multicolumn{3}{c}{substitution ordering} \\
-   \multirow{1}{5em}{mono subst} & age & random & separation\\
+   \multirow{1}{6em}{mono subst} & age & random & separation\\
    \cmidrule(lr){2-2}\cmidrule(lr){3-3}\cmidrule(l){4-4}
@@
 \label{subst_gen_table}
-\caption{Evaluation of bounds for substitution generation}
 \end{table}
@@
 \begin{table}[th]
-\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
+\caption{Evaluation of bounds directly related to the size of the output problem}
+\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{3em}}@{}}
    \toprule
@@
 \label{pb_size}
-\caption{Evaluation of bounds directly related to the size of the output problem}
 \end{table}
@@
 \begin{table}[th]
-\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
+\caption{Evaluation of parameters related to the depth of monomorphisation}
+\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{2.5em}}@{}}
    \toprule
    & \multicolumn{4}{c}{mono time} \\
-   \multirow{1}{5em}{num.\ loops} & 5 & 10 & 20 & 30\\
+   \multirow{1}{6em}{num.\ loops} & 5 & 10 & 20 & 30\\
    \midrule
@@
 \label{mono_time}
-\caption{Evaluation of parameters related to the depth of monomorphisation}
 \end{table}
@@
 \begin{table}[th]
-\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{1.5em}}@{}}
+\caption{Evaluation of parameters related to the E prover}
+\centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{3em}}@{}}
    \toprule
    & \multicolumn{4}{c}{E call point} \\
-   \multirow{1}{4em}{E timeout} & 0 & 0.1 & 0.2 & 0.3\\
+   \multirow{1}{5em}{E timeout} & 0 & 0.1 & 0.2 & 0.3\\
    \midrule
@@
 \label{e_settings}
-\caption{Evaluation of parameters related to the E prover}
 \end{table}
@@
 \begin{table}[ht]
-\centering\begin{tabular}{@{}lccc@{}}
+\caption{Evaluation of Zipperposition without E vs. with E}
+\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
    \toprule
-   & without E & with E & Union \\
+   & without E & with E & union \\
    \midrule
@@
 \end{tabular}
-\caption{Evaluation of Zipperposition without E vs. with E}
 \end{table}
@@
 Because all given problems are non-monomorphic, this instance is labeled ``without E''.
-The second instance, is a modification of \verb|portfolio.sequential.py| where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase.
+The second instance, is a modification of the \verb|portfolio.sequential.py| file where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase.
 This instance can successfully call E as a backend because it is able to provide E with a monomorphised problem.
@@
 \begin{table}[ht]
-\centering\begin{tabular}{@{}lccc@{}}
+\caption{Evaluation of native polymorphism vs.\ monomorphisation}
+\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
    \toprule
@@
 \end{tabular}
-\caption{Evaluation of native polymorphism vs.\ monomorphisation}
 \end{table}
@@
 %    * many approaches to encode monomorphic and polymorphic types, starting
-%      with Enderton \cite{xxx}, Stickel \cite{xxx}, and Wick and McCune
-%      \cite{xxx}
-%    * Blanchette et al.~\cite{xxx} exploit ``monotonicity'' to erase type
+%      with Enderton~\cite{xxx}, Stickel~\cite{xxx}, and Wick and McCune
+%     ~\cite{xxx}
+%    * Blanchette et al.~~\cite{xxx} exploit ``monotonicity'' to erase type
 %      information
-%    * see Blanchette et al.~\cite[Section~9]{xxx} for a more detailed coverage
+%    * see Blanchette et al.~~\cite[Section~9]{xxx} for a more detailed coverage
 %      of encodings of polymorphism
@@
 relevance filter that heuristically selects a subset of the available axioms;
-filters such as MePo \cite{meng-paulson-2009} and SInE
-\cite{hoder-voronkov-2011} are iterative and could be interleaved with
+filters such as MePo~\cite{meng-paulson-2009} and SInE
+~\cite{hoder-voronkov-2011} are iterative and could be interleaved with
 monomorphisation. Third, although one would expect native implementations of
@@
 
-\section*{Acknowledgements}
+\begin{credits}
+   \subsubsection{\ackname}
 
@@
    This research is co-funded by the European Union (ERC, Nekoka, 101083038).
-Views and opinions expressed are however those of the authors only and do not
-necessarily reflect those of the European Union or the European Research
-Council. Neither the European Union nor the granting authority can be held
-responsible for them.
-
+Views and opinions expressed are however those of the authors only and do not necessarily reflect those of the European Union or the European Research Council.
+Neither the European Union nor the granting authority can be held responsible for them.
+\end{credits}
+
+\bibliographystyle{splncs04}
 \bibliography{citations}

### 2025-02-13 iterative_monomorphisation.tex
contrived
rename

@@
 
-\author[1,2]{Tanguy Bozec}[%
-email=tanguy.bozec@ens-paris-saclay.fr,
-]
-\author[2]{Jasmin Blanchette}[%
-email=jasmin.blanchette@ifi.lmu.de,
-%url=https://www.tcs.ifi.lmu.de/mitarbeiter/jasmin-blanchette_de.html,
-]
-\address[1]{ENS Paris-Saclay, Université Paris-Saclay, France}
-\address[2]{Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}
+\author{Tanguy Bozec\inst{1,2} \and
+Jasmin Blanchette\inst{2}\orcidID{0000-0002-8367-0936}}
+
+\institute{ENS Paris-Saclay, Université Paris-Saclay, France \and
+Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}
+
+\maketitle
 
 \begin{abstract}
-Monomorphisation can be used to extend monomorphic provers to support polymorphic logics.
+Monomorphisation can be used to extend some monomorphic provers to support polymorphic logics.
 We propose an iterative approach, which is necessarily incomplete but which works well in practice.

### 2025-02-13 iterative_monomorphisation.tex
tune
tuning

@@
 \usepackage{enumitem}
+\usepackage{hyperref}
 
@@
 
-\author{Tanguy Bozec\inst{1,2} \and
-Jasmin Blanchette\inst{2}\orcidID{0000-0002-8367-0936}}
+\def\orcid#1{{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}
+
+\author{Tanguy Bozec\inst{1,2}\orcid{TODO} \and
+Jasmin Blanchette\inst{2}\orcid{0000-0002-8367-0936}}
+
+\authorrunning{T. Bozec and J. Blanchette}
 
 \institute{ENS Paris-Saclay, Université Paris-Saclay, France \and
-Institute of Informatics, Ludwig-Maximilians-Universität München, Germany}
+%Institute of Informatics,
+Ludwig-Maximilians-Universität München, Germany}
 
@@
 
+\begin{sloppy}
 \begin{abstract}
-Monomorphisation can be used to extend some monomorphic provers to support polymorphic logics.
+Monomorphisation can be used to extend monomorphic provers to support polymorphic logics.
 We propose an iterative approach, which is necessarily incomplete but which works well in practice.
-It is implemented in the Zipperposition prover, where it can be used to translate away polymorphism before invoking the monomorphic prover E as a backend.
+We implemented it in the Zipperposition prover, where it can be used to translate away polymorphism before invoking the monomorphic prover E as a backend.
 Our evaluation demonstrates that this approach increases Zipperposition's success rate.
@@
 
-\keywords{Polymorphism  \and Monomorphism    \and Automatic theorem proving}
+\keywords{Polymorphism  \and monomorphism \and first-order logic
+  \and higher-order logic}
 \end{abstract}
+\end{sloppy}
 
@@
 Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009}, and
-Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed only at the top level of formulae.
+Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae.
 On the other hand, many automatic provers operate only on monomorphic logics.
@@
 The alternative is to translate polymorphic problems to monomorphic problems.
-One approach is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic;
+One approach to achieve this is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic;
 we refer to Blanchette et al.~\cite[Section~9]{mono-trans} for an overview of this approach.
@@
 \SetKwFunction{IterMono}{iterative\_monomorphisation}
-\begin{figure}[b]
-\begin{quote}
+\begin{figure}[t]
+%\begin{quote}
 \begin{algorithm}[H]
@@
 \end{algorithm}
-\end{quote}
+%\end{quote}
 \caption{Pseudocode for iterative monomorphisation}
@@
 
-\begin{figure}
-\begin{quote}
+\begin{figure}[t]
+%\begin{quote}
 \begin{algorithm}[H]
@@
 \end{algorithm}
-\end{quote}
+%\end{quote}
 \caption{Pseudocode for formula monomorphisation step}
@@
 \end{algorithm}
-\end{quote}
+%\end{quote}
 \caption{Pseudocode for match generation}
@@
 \end{algorithm}
-\end{quote}
+%\end{quote}
 \caption{Pseudocode for monomorphic formula generation}
@@
 
-\begin{figure}
-\begin{quote}
+\begin{figure}[t]
+%\begin{quote}
    \begin{algorithm}[H]
@@
 \end{algorithm}
-\end{quote}
+%\end{quote}
 \caption{Pseudocode for monomorphising substitution generation}

### 2025-02-13 iterative_monomorphisation.tex
tables
more
shorten and edit

@@
 
-\keywords{Polymorphism  \and monomorphism \and first-order logic
-  \and higher-order logic}
+\keywords{Polymorphism  \and monomorphism \and automated reasoning}
 \end{abstract}
@@
 
-One of the main applications of automatic theorem provers is to provide automation to users of proof assistants.
-Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009}, and
-Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae.
+Automatic theorem provers are employed to provide automation to users of proof assistants.
+Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009}, and Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae.
 On the other hand, many automatic provers operate only on monomorphic logics.
@@
 The alternative is to translate polymorphic problems to monomorphic problems.
-One approach to achieve this is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic;
-we refer to Blanchette et al.~\cite[Section~9]{mono-trans} for an overview of this approach.
+One approach \cite[Section~9]{mono-trans} to achieve this is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic.
 Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}.
@@
 
-B\"ohme's iterative approach is implemented as part of the SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd} in Isabelle/HOL.
+B\"ohme's iterative approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}.
 This implementation is also used by Sledgehammer~\cite{judgement,hammer} to interface with superposition based automatic theorem provers.
-However, it is is documented only as a single subsection in his PhD thesis~\cite[Section 2.2.1]{sb-phd}.
+However, it is is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
 A similar algorithm appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
@@
 We also provide a more detailed description to help future implementers.
-In addition, this description shows some of the ways in which an implementation can be made more efficient and avoid explosions (Section~\ref{sec:low-level-algorithm}).
-
-The algorithm works as follows.
+In addition, our description shows some of the ways in which an implementation can be made more efficient and avoid explosions (Section~\ref{sec:low-level-algorithm}).
+
+Briefly, the algorithm works as follows.
 We assume problems to be sets of formulae.
@@
 If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, it may be possible to generate the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc.
-However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has already been generated.
+However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has been generated.
 This example also makes it clear that the set of instances that can be generated is infinite.
@@
 After the iterations are completed, the new monomorphic symbol instances are used to instantiate the polymorphic symbols in the problem's formulae, generating new monomorphic formulae.
-Finally, because monomorphic provers generally do not support $n$-ary type constructors, types must be `mangled'; for example, the compound type $\ty{list}(\ty{int})$ could be mangled to the constant $\ty{list\_int}$.
+Finally, because monomorphic provers do not support $n$-ary type constructors, types must be `mangled'; for example, the compound type $\ty{list}(\ty{int})$ could be mangled to $\ty{list\_int}$.
 
@@
 Although Zipperposition is polymorphic, it uses the monomorphic prover E~\cite{e} as a backend.
-This means that E can now be used with polymorphic problems.
-Moreover, our implementation of the algorithm in Zipperposition can be used as a preprocessor to interface with other stand-alone provers.
+Thanks to our work, E can now be used with polymorphic problems.
+Moreover, our implementation can be used as a preprocessor to interface with other stand-alone provers.
 
@@
 
+\smallskip
+
 \item How competitive are monomorphic provers on monomorphised polymorphic problems?
+
+\smallskip
 
@@
 
-Our findings are as follows:
-\begin{enumerate}
-\item Zipperposition benefits substantially from the E backend.
-
-\item E with monomorphisation comes close second to the polymorphic prover Vampire.
-
-\item For Leo-III~\cite{leo-iii} and Vampire~\cite{vamp}, we find that monomorphisation is indeed more effective than native polymorphism.
-\end{enumerate}
+%Our findings are as follows:
+%\begin{enumerate}
+%\item Zipperposition benefits substantially from the E backend.
+%
+%\smallskip
+%
+%\item E with monomorphisation comes close second to the polymorphic prover Vampire.
+%
+%\smallskip
+%
+%\item For Leo-III~\cite{leo-iii} and Vampire~\cite{vamp}, we find that monomorphisation is indeed more effective than native polymorphism.
+%\end{enumerate}
 
@@
 \SetKwFunction{IterMono}{iterative\_monomorphisation}
-\begin{figure}[t]
+\begin{figure}[t!]
 %\begin{quote}
@@
 
-\begin{figure}[t]
+\begin{figure}[t!]
 %\begin{quote}
@@
 \caption{Evaluation of bounds for monomorphic type argument generation}
+
+\medskip
+
 \centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{2.5em}}@{}}
@@
 \caption{Evaluation of bounds for non-monomorphic type argument generation}
+
+\medskip
+
 \centering\begin{tabular}{@{}l*{12}{>{\centering\arraybackslash}p{2.5em}}@{}}
@@
 \caption{Evaluation of bounds for substitution generation}
+
+\medskip
+
 \centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
@@
 \caption{Evaluation of bounds directly related to the size of the output problem}
+
+\medskip
+
 \centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{3em}}@{}}
@@
 \caption{Evaluation of parameters related to the depth of monomorphisation}
+
+\medskip
+
 \centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{2.5em}}@{}}
@@
 \caption{Evaluation of parameters related to the E prover}
+
+\medskip
+
 \centering\begin{tabular}{@{}l*{4}{>{\centering\arraybackslash}p{3em}}@{}}
@@
 \begin{table}[ht]
-\caption{Evaluation of Zipperposition without E vs. with E}
+\caption{Evaluation of Zipperposition without E vs.\ with E}
+
+\medskip
+
 \centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
@@
    \bottomrule
+\end{tabular}
 \label{zipp_eval}
-\end{tabular}
 \end{table}
@@
 \caption{Evaluation of native polymorphism vs.\ monomorphisation}
+
+\medskip
+
 \centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
@@
 Table~\ref{native_vs_mono_eval} shows the results.
- Evaluation of the monomorphisation approach is done in two steps.
+The monomorphisation approach is evaluated in two steps.
 First, each problem is monomorphised using the options obtained from the first evaluation phase except for the monomorphisation timeout option which is removed.
@@
 
+\looseness=-1
 We see the following avenues for future work. First, iterative monomorphisation
@@
 polarities, but our procedure instantiates $\alpha$ with $\ty{nat}$ regardless.
-Second, some automatic provers as well as tools like Sledgehammer include a
-relevance filter that heuristically selects a subset of the available axioms;
+Second, some automatic provers as well as tools such as Sledgehammer include
+relevance filters that heuristically select a subset of the available axioms;
 filters such as MePo~\cite{meng-paulson-2009} and SInE
@@
 monomorphisation. Third, although one would expect native implementations of
-polymorphism to outperform any preprocessor, our evaluation clearly shows that
-this is not the case, therefore indicating that there is considerable room for
-improvement on the native front.
+polymorphism to outperform any preprocessor, this is not the case now, therefore indicating that there is considerable room for improvement on the native front.
 
 \begin{credits}
-   \subsubsection{\ackname}
+   \subsubsection{Acknowledgements.}
 

### 2025-02-13 iterative_monomorphisation.tex
shorten

@@
 %\newtheorem{definition}{Definition}
-%\newtheorem{example}[definition]{Example}
+\newtheorem{examplex}[definition]{Example}
 
@@
 
-\item How competitive are monomorphic provers on monomorphised polymorphic problems?
+\item How competitive are monomorphic provers on monomorphised %polymorphic
+problems?
 
@@
 
-This algorithm works independently of the structure of the problem's formulae.
+Our algorithm works independently of the structure of the problem's formulae.
 It relies exclusively on the formulae's monomorphic and polymorphic symbol instances.
@@
 Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic.
-In particular, it can operate in the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.
+In particular, it can work with the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.
 
@@
 
-\begin{definition}
+\begin{definition}\rm
 A (\emph{polymorphic}) \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
@@
 
-\begin{definition}
+\begin{definition}\rm
 A type is \emph{monomorphic} if it contains no type variables.
@@
 
-\begin{definition}
-A (function or predicate) {symbol} \(f\) has a \emph{type arity} that defines the number of type arguments it takes.
+\begin{definition}\rm
+A (function or predicate) {symbol} \(f\) has a \emph{type arity} that specifies the number of type arguments it takes.
 A \emph{symbol instance} is a symbol applied to type arguments listed between angle brackets: \(f\langle \tau_1, \dots, \tau_n\rangle\), where each $\tau_i$ is a type.
-If $n = 0$, we omit the brackets (e.g.\ $f$).
+If $n = 0$, we omit the angle brackets (e.g.\ $f$).
 \end{definition}
 
-\begin{definition}
+\begin{definition}\rm
 A (\emph{type}) \emph{substitution} is a partial function mapping a finite number of type variables to corresponding types. Substitutions are written as
@@
 
-\begin{definition}
-Given two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_m\mapsto\tau_m\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_n\mapsto\upsilon_n\}\) they are said to be \emph{compatible} if \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\) for all \(i, j\).
+\begin{definition}\rm
+Two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_m\mapsto\tau_m\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots,\allowbreak \beta_n\mapsto\upsilon_n\}\) are said to be \emph{compatible} if \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\) for all \(i, j\).
 \end{definition}
 
-\begin{definition}
+\begin{definition}\rm
 Given two types \(\tau, \upsilon\), \emph{matching} \(\upsilon\) against \(\tau\) will either fail or yield a substitution \(\sigma\) such that \(\sigma(\upsilon) = \tau\).
@@
 
-In the following sections, we will always match a polymorphic type $\upsilon$ against a monomorphic type $\tau$.
+%In the following sections, we will always match a polymorphic type $\upsilon$ against a monomorphic type $\tau$.
 
@@
 
+\smallskip
+
 \item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(N(\varphi)(f)\) consists of all type arguments tuples to which \(f\) is applied in \(\varphi\) and which contains a type variable.
@@
 
-These definitions depend on the precise form of the input formulae, which has been left undefined.
-It is important to parametrise \(N\) with \(\varphi\) because type variables are implicitly quantified at the formula level.
+%These definitions depend on the precise form of the input formulae, which has been left undefined.
+It is important to parametrise \(N\) with \(\varphi\), because type variables are implicitly quantified at the formula level.
 The formula indicates the scope of type variables.
@@
 
-The algorithm is clearly sound because the newly generated formulae are instances of the initial problem's formulae, where type variables have been instantiated with monomorphic types.
+The algorithm is sound because the newly generated formulae are instances of the initial problem's formulae. %, where type variables have been instantiated with monomorphic types.
 It is, however, not guaranteed to be complete.
 
-\begin{example}Consider the following problem:
+\begin{examplex}Consider the following problem:
 \begin{enumerate}
@@
 
-\end{example}
+\end{examplex}
 
 This example illustrates how an infinite number of new formulae can be generated from a simple initial problem.
-Although the example may seem contrived, similar problems frequently arise in practice.
+%Although the example may seem contrived, similar problems frequently arise in practice.
 For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
-Any reasonable implementation requires bounds limiting the number of new type arguments, substitutions and formulae to be useful.
+Any reasonable implementation requires bounds limiting the number of new type arguments, substitutions and formulae.
 
@@
 
-All sets referenced in the pseudocode are assumed to be finite.
-Additionally, this algorithm relies on primitives whose implementation depends on the specifics of the grammar and logic used.
-They will therefore not be expanded upon.
+All sets referenced in the pseudocode are %assumed to be
+finite. Moreover, the algorithm relies on primitives whose implementation depends on the specifics of the grammar and logic used. %They will therefore not be expanded upon.
 Functions computing the following are assumed to be available: 
@@
    \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set.
-   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)}, where \(\sigma\) is the substitution resulting from the matching.
+   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)}, where \(\sigma\) results from the matching.
 The pseudocode matches only non-monomorphic types against monomorphic types.
-   \item[\labelitemi] \emph{domain\((\sigma)\)}, where \(\sigma\) is a substitution, returns the set of type variables \(\alpha\) such that \(\sigma(
+   \item[\labelitemi] \emph{domain\((\sigma)\)} %, where \(\sigma\) is a substitution,
+   returns the set of type variables \(\alpha\) such that \(\sigma(
    \alpha) \not= \alpha\).
-   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)}, where \(\sigma_1\) and \(\sigma_2\) are substitutions, tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
-   \item[\labelitemi] Composition and application of substitutions are written using mathematical notations.
-   \item[\labelitemi] \emph{mangle\((\Phi)\)}, where \(\Phi\) is a set of monomorphic formulae returns the same set of formulae where all types have been mangled.
+   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)} %, where \(\sigma_1\) and \(\sigma_2\) are substitutions,
+   tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
+%   \item[\labelitemi] Composition and application of substitutions are written using mathematical notations.
+   \item[\labelitemi] \emph{mangle\((\Phi)\)}, where \(\Phi\) is a set of monomorphic formulae, returns the same set of formulae where all types have been mangled.
 \end{enumerate}
@@
 
-\begin{table}[th]
+\begin{table}[t!]
 \caption{Evaluation of bounds for monomorphic type argument generation}
@@
 
-\begin{table}[th]
+\begin{table}[t!]
 \caption{Evaluation of bounds for non-monomorphic type argument generation}
@@
 % TODO find a better term than "separation heuristic" maybe separation ?
-\begin{table}[th]
+
+\begin{table}[t!]
 \caption{Evaluation of bounds for substitution generation}
@@
 
-\begin{table}[th]
+\begin{table}[t!]
 \caption{Evaluation of bounds directly related to the size of the output problem}
@@
 
-\begin{table}[th]
+\begin{table}[t!]
 \caption{Evaluation of parameters related to the depth of monomorphisation}
@@
 
-\begin{table}[th]
+\begin{table}[t!]
 \caption{Evaluation of parameters related to the E prover}

### 2025-02-13 iterative_monomorphisation.tex
it fits

@@
 
-\smallskip
-
 \item How competitive are monomorphic provers on monomorphised %polymorphic
 problems?
-
-\smallskip
 
@@
 %
-%\smallskip
-%
 %\item E with monomorphisation comes close second to the polymorphic prover Vampire.
-%
-%\smallskip
 %
@@
 
-\smallskip
-
 \item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(N(\varphi)(f)\) consists of all type arguments tuples to which \(f\) is applied in \(\varphi\) and which contains a type variable.
@@
 
-\begin{examplex}Consider the following problem:
+\begin{examplex}\rm
+Consider the following problem:
 \begin{enumerate}
@@
 This avoids re-computing some of the matchings processed in previous iterations.
-Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.
+%Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.
 New type argument tuples are obtained by applying each substitution to each non-monomorphic type argument tuple such that at least one type variable in the tuple is instantiated by the substitution.
@@
 \begin{enumerate}
-   \item \MonoCap, an absolute limit on the total number of new type argument tuples;
+   \item \MonoCap, a %n absolute
+   limit on the total number of new type argument tuples;
    \item \MonoMult, which is used to allow the total number of (monomorphic) type argument tuples to grow by a certain proportion of the current number \(m\) of monomorphic type argument tuples;
@@
 Once all monomorphisation iterations have been completed, we are left with a set of the substitutions that have been used to generate new type arguments.
-The last phase uses these substitutions to instantiate the type variables in the input problem's non-monomrophic formuale.
-In the presence of bounds, the order in which the elements of the sets of substitutions of \(S\) are traversed affects the formulae resulting from the last phase.
-As we will discuss in Section~\ref{param_opti}, the best results were obtained when substitutions generated in the same iteration are grouped together.
+The last phase uses this set to instantiate the type variables in the input problem's non-monomrophic formuale.
+In the presence of bounds, the order in which the elements of \(S\) are traversed affects the formulae resulting from the last phase.
+%As we will discuss in Section~\ref{param_opti}, the best results were obtained when substitutions generated in the same iteration are grouped together.
 
@@
 
-The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S\) more efficiently.
+The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S\) efficiently.
 The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
@@
    \item The \emph{mono\_substs} function is the most explosive part of the algorithm.
-For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) different type variables, up to \(n^v\) monomorphising substititions may be generated.
+For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) %different
+type variables, up to \(n^v\) monomorphising substititions may be generated.
 Recall that the total number of type argument tuples that are used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations.
@@
 
-\begin{table}[ht]
+\begin{table}[t!]
 \caption{Evaluation of Zipperposition without E vs.\ with E}

### 2025-02-13 iterative_monomorphisation.tex
tune

@@
 
-We described a procedure for iteratively instantiating polymorphic types to
+We described an algorithm for iteratively instantiating polymorphic types to
 produce monomorphic problems as a preprocessor. Our primary motivation was to
@@
 conjecture $\lnot \sym{p}\langle\ty{nat}\rangle$ because of the incompatible
-polarities, but our procedure instantiates $\alpha$ with $\ty{nat}$ regardless.
+polarities, but our algorithm instantiates $\alpha$ with $\ty{nat}$ regardless.
 Second, some automatic provers as well as tools such as Sledgehammer include

### 2025-02-13 iterative_monomorphisation.tex
edits

@@
 The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula.
-A significant difference with the monomorphic case is that \(n\) depends on the current formula being processed whilst \(m\) does not.
+An important difference with the monomorphic case is that \(n\) depends on the current formula being processed whilst \(m\) does not.
 Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.
@@
 The third part compares the performance of different provers on polymorphic problems and their monomorphised counterparts.
+Our raw evaluation data is available online.%
+\footnote{\url{https://nekoka-project.github.io/pubs/monomorphization_data.zip}}
 
@@
 In total, our benchmark suite contains 1534 polymorphic problems.
-We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in under 30 seconds with a single thread.
-This corresponds to the default time limit in Sledgehammer.
+We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in at most 30 seconds with a single thread.
+%This corresponds to the default time limit in Sledgehammer.
 
@@
 
-We conducted additional informal evaluations on the 500 problem benchmark suite to find appropriate default values and test ranges for the monomorphisation bounds.
+We conducted additional informal evaluations on the 500 problem suite to find appropriate default values and test ranges for the monomorphisation bounds.
 We started the option evaluation process with the base configuration and the following default values for monomorphisation bounds and parameters for E:
-
-\begin{multicols}{2}% 3 columns?
+%
+\begin{minipage}[t]{.5\textwidth}
+\vskip0\topsep
+
 \begin{itemize}
@@
    \item \MonoSubstsLimit: \(10\)
+\end{itemize}
+
+\vskip0\topsep
+\end{minipage}\begin{minipage}[t]{.5\textwidth}
+\vskip0\topsep
+
+\begin{itemize}
    \item \Limit: \(2000\)
@@
 \end{itemize}
-\end{multicols}
-\noindent
+
+\vskip0\topsep
+\end{minipage}
+%
 The initial values for the \MonoCap, \MonoMult and \MonoFloor options are irrelevant because the options' values are set when computing Table~\ref{mono_ty_args}.% and used for subsequent tables.
@@
 
+\looseness=-1
 Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments.
@@
 This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed.
-Performance of the monomorphisation algorithm seems to plateau for certain ranges of the tested values and drops off beyond.
+Performance of the monomorphisation algorithm seems to plateau for some ranges of values and drops off beyond.
 %TODO check that 125 is what we get if we just run Zipperposition for 10 seconds (maybe with all additional problems that are easily monomorphised)
@@
 The bound limiting the number of monomorphising substitutions per formula is directly related to the ordering which dictates how such monomorphising substitutions are generated.
-The substitution ordering heuristic has a significant impact on monomorphising substitution generation.
+The substitution ordering heuristic has a substantial impact on monomorphising substitution generation.
 The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first.
@@
 For larger problems, the monomorphisation algorithm may time out despite the bounds.
-In these cases neither Zipperposition nor E will have had a chance to attempt to solve the problem.
+In these cases neither Zipperposition nor E will have had a chance to try to solve the problem.
 To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation.
-Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for against the number of iteration of the monomorphisation algorithm.
-Neither parameter seems to have a significant effect on the performance of the monomorphisation algorithm.
+Table~\ref{mono_time} tests the amount of time that monomorphisation is allowed to run for against the number of iteration of the monomorphisation algorithm.
+Neither parameter seems to have much of an effect on the performance of the algorithm.
 
@@
    \midrule
-   500 problem benchmark   & 160 & 198 & 207 \\ % -9 +47
-   1034 problem benchmark & 337 & 410 & 434 \\ % -24 +97
+   500 problem suite   & 160 & 198 & 207 \\ % -9 +47
+   1034 problem suite & 337 & 410 & 434 \\ % -24 +97
    \bottomrule
@@
 
-\begin{table}[ht]
+\begin{table}[t!]
 \caption{Evaluation of native polymorphism vs.\ monomorphisation}
@@
 
-To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher-order polymorphic provers were run on the 1034 problem benchmark.
+To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher-order polymorphic provers were run on the 1034 problem suite.
 We chose Leo-III and Zipperposition.

### 2025-02-13 iterative_monomorphisation.tex
fix url
too big for my server
cites

@@
 B\"ohme's iterative approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}.
-This implementation is also used by Sledgehammer~\cite{judgement,hammer} to interface with superposition based automatic theorem provers.
+This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers.
 However, it is is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
@@
 
-We implemented iterative monomorphisation in Zipperposition~\cite{zipp}, a higher order prover written in OCaml.
+We implemented iterative monomorphisation in Zipperposition~\cite{vukmirovic-et-al-2021}, a higher order prover written in OCaml.
 Although Zipperposition is polymorphic, it uses the monomorphic prover E~\cite{e} as a backend.
@@
 Our raw evaluation data is available online.%
-\footnote{\url{https://nekoka-project.github.io/pubs/monomorphization_data.zip}}
+\footnote{\url{TODO}}
 

### 2025-02-13 iterative_monomorphisation.tex
spellcheck

@@
 Once all monomorphisation iterations have been completed, we are left with a set of the substitutions that have been used to generate new type arguments.
-The last phase uses this set to instantiate the type variables in the input problem's non-monomrophic formuale.
+The last phase uses this set to instantiate the type variables in the input problem's non-monomorphic formuale.
 In the presence of bounds, the order in which the elements of \(S\) are traversed affects the formulae resulting from the last phase.
@@
 For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) %different
-type variables, up to \(n^v\) monomorphising substititions may be generated.
+type variables, up to \(n^v\) monomorphising substitutions may be generated.
 Recall that the total number of type argument tuples that are used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations.

### 2025-02-14 iterative_monomorphisation.tex
avoid anticipating eval results
comma

@@
 Monomorphisation can be used to extend monomorphic provers to support polymorphic logics.
-We propose an iterative approach, which is necessarily incomplete but which works well in practice.
+We propose an iterative approach that is incomplete but pragmatic.
 We implemented it in the Zipperposition prover, where it can be used to translate away polymorphism before invoking the monomorphic prover E as a backend.
@@
 Automatic theorem provers are employed to provide automation to users of proof assistants.
-Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009}, and Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae.
+Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae.
 On the other hand, many automatic provers operate only on monomorphic logics.

### 2025-02-17 iterative_monomorphisation.tex
edits

@@
 
-\author{Tanguy Bozec\inst{1,2}\orcid{TODO} \and
+\author{Tanguy Bozec\inst{1,2}\orcid{0009-0005-9497-4168} \and
 Jasmin Blanchette\inst{2}\orcid{0000-0002-8367-0936}}
@@
 The substitution ordering heuristic has a substantial impact on monomorphising substitution generation.
-The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first.
-The ``random'' ordering randomly shuffles substitutions.
-Finally, the ``separation'' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently.
-Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seems to affect performance when the ``separation'' heuristic is used.
+The `age' ordering of substitution orders substitutions generated in earlier iterations are ordered first.
+The `random' ordering randomly shuffles substitutions.
+Finally, the `separation' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently.
+Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seems to affect performance when the `separation' heuristic is used.
 % TODO find a better term than "separation heuristic" maybe separation ?
@@
 Of these configuration, some may attempt to call E as a backend when the given problem is monomorphic.
-Because all given problems are non-monomorphic, this instance is labeled ``without E''.
+Because all given problems are non-monomorphic, this instance is labeled `without E'.
 The second instance, is a modification of the \verb|portfolio.sequential.py| file where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase.
@@
 The monomorphisation approach is evaluated in two steps.
-First, each problem is monomorphised using the options obtained from the first evaluation phase except for the monomorphisation timeout option which is removed.
-Second, each prover is ran on the monomorphised problem and the results are tallied in the table's ``Mono'' column.
-In addition to the polymorphic provers used in the ``Native'' tests, the monomorphic prover E is run on the monomorphised problems to provide an additional yardstick.
-Instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time (rounded up to the nearest second) is subtracted to compare fairly against the ``Native'' column, which does not have a similar preprocessing phase.
+First, each problem is monomorphised using the options obtained from the first evaluation phase except for the monomorphisation timeout option which is increased to 30~seconds.
+For 149~problems, monomorphisation times out.
+Second, each prover is ran on the remaining monomorphised problems, and the results are tallied in the table's `Mono' column.
+In addition to the polymorphic provers used in the `Native' tests, the monomorphic prover E is run on the monomorphised problems to provide an additional yardstick.
+Instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time (rounded up to the nearest second) is subtracted to compare fairly against the `Native' column, which does not have a similar preprocessing phase.
+
+Despite the monomorphisation timeouts, monomorphisation is more effective than Leo-III's and Zipperposition's native polymorphism on the benchmarks.
 

### 2025-02-19 iterative_monomorphisation.tex
mmore
added links

@@
 Moreover, our implementation can be used as a preprocessor to interface with other stand-alone provers.
+Our source code is available online.%
+\footnote{\url{https://github.com/nartannt/iterative_monomorphisation}}
 
@@
 \end{enumerate}
+%
+Our raw evaluation data is available online.%
+\footnote{\url{https://zenodo.org/records/14881532}}
 
@@
 The third part compares the performance of different provers on polymorphic problems and their monomorphised counterparts.
-Our raw evaluation data is available online.%
-\footnote{\url{TODO}}
 
 The benchmarks are taken from version 8.2.0 of the TPTP library~\cite{tptp}.
-The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism.
+The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first order and higher order logic with rank-1 polymorphism.
 Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them.
@@
 
-To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher-order polymorphic provers were run on the 1034 problem suite.
+To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher order polymorphic provers were run on the 1034 problem suite.
 We chose Leo-III and Zipperposition.

### 2025-02-19 iterative_monomorphisation.tex
shorten

@@
 
-Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables at the cost of taking more time.
+Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables but takes more time.
 Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed.
@@
 Instead, we group closely related bounds together and test combinations of values for the bounds in these groups.
-Once we find the best performing set of values for a group of bounds, we assign these values to the corresponding bounds as we begin the search for the next group.
-If several groups of bounds result in the same number of proved problems, the most constraining bound values are selected.
+Once we find the best performing set of values for a group, we assign these values to the corresponding bounds as we begin the search for the next group.
+If several groups result in the same number of proved problems, we select the most constraining values.
 Winning entries are shown in bold in Tables \ref{mono_ty_args} to \ref{e_settings}.
@@
 The bound limiting the number of monomorphising substitutions per formula is directly related to the ordering which dictates how such monomorphising substitutions are generated.
-The substitution ordering heuristic has a substantial impact on monomorphising substitution generation.
-The `age' ordering of substitution orders substitutions generated in earlier iterations are ordered first.
+The %substitution ordering
+heuristic greatly affects monomorphising substitution generation.
+The `age' ordering of substitution orders substitutions generated in earlier iterations first.
 The `random' ordering randomly shuffles substitutions.

### 2025-02-19 iterative_monomorphisation.tex
shorten

@@
 All symbols of the problem are collected, and the polymorphic instances of symbols are matched against the monomorphic ones.
-This yields new instances of symbols, both polymorphic and monomorphic.
-The process is then iterated
-a number of times, making use of the newly generated instances.
+This yields new symbol instances, both polymorphic and monomorphic.
+The process is then iterated, making use of the newly generated instances.
 
@@
 Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic.
-In particular, it can work with the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.
+In particular, it can work with the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}.
+% implemented by several automatic provers.
 
@@
 If $n = 0$, we omit the parentheses (e.g.\ \(\ty{int}\)).
-\end{definition}
-
-\begin{definition}\rm
 A type is \emph{monomorphic} if it contains no type variables.
@@
 The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem.
-It operates by applying an arbitrary number of iterations.
-Each iteration takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae.
+It operates by applying a fixed number of iterations, each of which taking a polymorphic problem as argument and returning a problem with additional partially instantiated formulae.
 A single iteration consists of a collection phase and an instantiation phase.
@@
 
-The algorithm presented in the previous section is simple but too naive for an implementation. In this section, we present a lower level algorithm with the following features.
+\looseness=-1
+The algorithm presented above is too naive for an implementation. In this section, we present a lower level algorithm with the following features.
 First, numeric bounds are introduced to stop explosive enumerations.
@@
 The data structures used in the pseudocode are based on the ones used in the high level description.
-Instead of a map \(M\) from symbols to monomorphic type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type arguments tuples having already been matched against and those that have not.
+Instead of a map \(M\) from symbols to monomorphic\pagebreak[2] type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type arguments tuples having already been matched against and those that have not.
 Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples.
@@
 The formula monomorphisation algorithm is given in Figure~\ref{mono_step}.
-It forms the core of the monomorphisation process by computing new type arguments tuples for a single formula.
+It forms the core of the process by computing new type arguments tuples for a single formula.
 Type argument tuples are matched against each other to obtain a set of substitutions which is iterated over in the outermost loop.
 The separation of type arguments tuples into old and new maps is put to use to ensure that only combinations involving at least one new map are considered.
-This avoids re-computing some of the matchings processed in previous iterations.
+This avoids re-computing some matchings processed in previous iterations.
 %Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.
-New type argument tuples are obtained by applying each substitution to each non-monomorphic type argument tuple such that at least one type variable in the tuple is instantiated by the substitution.
+New tuples are obtained by applying each substitution to each non-monomorphic type argument tuple such that at least one tuple component is instantiated by the substitution.
 
@@
 The final number of output formulae is ensured to be at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
-   \item The \emph{mono\_substs} function is the most explosive part of the algorithm.
+   \item The \emph{mono\_substs} function is the algorithm's most explosive part.
 For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) %different
 type variables, up to \(n^v\) monomorphising substitutions may be generated.
-Recall that the total number of type argument tuples that are used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations.
+Recall that the total number of type argument tuples used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations.
 \end{enumerate}

### 2025-02-21 iterative_monomorphisation.tex
Jannis edits

@@
 Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae.
-On the other hand, many automatic provers operate only on monomorphic logics.
-One approach to close this gap is to extend provers to natively support polymorphism, as has been done for Vampire~\cite{bhayat-reger-2020}.
+By contrast, many automatic provers operate only on monomorphic logics.
+One approach to closing this gap is to extend provers to natively support polymorphism, as has been done for Vampire~\cite{bhayat-reger-2020}.
 This, however, entails a lot of work that needs to be redone for every prover.
@@
 The alternative is to translate polymorphic problems to monomorphic problems.
-One approach \cite[Section~9]{mono-trans} to achieve this is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic.
+One approach \cite[Section~9]{mono-trans} is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic.
 Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}.
@@
 
-By a type version of the compactness theorem, we have that in first order logic, given a polymorphic formula \(\varphi\), there exist an equisatisfiable finite set of monomorphic instances of \(\varphi\).
-However, such a set cannot be computed~\cite[Theorem 1]{expr-poly-types}.
-As a result, any monomorphisation method based on instantiation of type variables is
+Any translation method based on instantiation is
 bound to be incomplete.
+By a type version of the compactness theorem, we have that for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of \(\varphi\).
+However, this set cannot be computed~\cite[Theorem~1]{expr-poly-types}.
 
@@
 After the iterations are completed, the new monomorphic symbol instances are used to instantiate the polymorphic symbols in the problem's formulae, generating new monomorphic formulae.
-Finally, because monomorphic provers do not support $n$-ary type constructors, types must be `mangled'; for example, the compound type $\ty{list}(\ty{int})$ could be mangled to $\ty{list\_int}$.
+Finally, because monomorphic provers support only nullary type constructors, types must be `mangled'; for example, the compound type $\ty{list}(\ty{int})$ could be mangled to $\ty{list\_int}$.
 
@@
 \begin{definition}\rm
-A (\emph{polymorphic}) \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
+A (\emph{polymorphic}) \emph{type} \(\tau\) is a type variable (e.g.\ \(\alpha\)) or
 the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\), \(\ty{map}(\ty{int},\ty{string})\)).
@@
 \begin{definition}\rm
-Given two types \(\tau, \upsilon\), \emph{matching} \(\upsilon\) against \(\tau\) will either fail or yield a substitution \(\sigma\) such that \(\sigma(\upsilon) = \tau\).
+Given two types \(\tau, \upsilon\), \emph{matching} \(\upsilon\) against \(\tau\) either fails or yields a substitution \(\sigma\) such that \(\sigma(\upsilon) = \tau\).
 \end{definition}
@@
 The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem.
-It operates by applying a fixed number of iterations, each of which taking a polymorphic problem as argument and returning a problem with additional partially instantiated formulae.
+It operates by applying a fixed number of iterations, each of which takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae.
 A single iteration consists of a collection phase and an instantiation phase.
-Once the iterations are completed, a final step filters out all non-monomorphic formulae returned by the last iteration.
-
-The initial phase of each iteration consists of computing two maps, \(M\) and \(N\), from the input problem~$\Phi$.
+Once the iterations are completed, a final step discards all non-monomorphic formulae returned by the last iteration.
+
+The initial phase of each iteration computes two maps, \(M\) and \(N\), from the input problem~$\Phi$.
 %
 \begin{enumerate}
-\item[\labelitemi] Given a symbol \(f\) occurring in \(\Phi\), the set \(M(f)\) consists of all monomorphic type arguments tuples to which \(f\) is applied in \(\Phi\).
+\item[\labelitemi] Given a symbol \(f\) occurring in \(\Phi\), the set \(M(f)\) consists of all monomorphic type argument tuples to which \(f\) is applied in \(\Phi\).
 For example, if \(\sym{foldl}\langle \ty{nat}, \ty{int}\rangle\) occurs in \(\Phi\), then \((\ty{nat}, \ty{int}) \in M(\sym{foldl}) \).
 
-\item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(N(\varphi)(f)\) consists of all type arguments tuples to which \(f\) is applied in \(\varphi\) and which contains a type variable.
+\item[\labelitemi] Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(N(\varphi)(f)\) consists of all type argument tuples to which \(f\) is applied in \(\varphi\) and which contain a type variable.
 For example, if \(\sym{foldl}\langle \ty{nat}, \ty{list}(\alpha)\rangle\) occurs in \(\varphi\), then \((\ty{nat}, \ty{list}(\alpha)) \in N(\varphi)(\sym{foldl}) \).
@@
 %These definitions depend on the precise form of the input formulae, which has been left undefined.
-It is important to parametrise \(N\) with \(\varphi\), because type variables are implicitly quantified at the formula level.
+\(N\) is parameterised with \(\varphi\) because type variables are implicitly quantified at the formula level.
 The formula indicates the scope of type variables.
@@
    \begin{enumerate}
-    \item[2.1.] For each tuple \((\tau_1, \dots, \tau_n) \in  M(f)\) and \((\upsilon_1, \dots, \upsilon_n) \in N(\varphi)(f)\),
+    \item[2.1.] For each \((\tau_1, \dots, \tau_n) \in  M(f)\) and \((\upsilon_1, \dots, \upsilon_n) \in N(\varphi)(f)\) and
      for each \(i\), match \(\upsilon_i\) against \(\tau_i\), yielding the substitution \(\sigma_i\) in case of success.
@@
 The algorithm is sound because the newly generated formulae are instances of the initial problem's formulae. %, where type variables have been instantiated with monomorphic types.
-It is, however, not guaranteed to be complete.
+It is, however, not complete.
 
@@
 First, numeric bounds are introduced to stop explosive enumerations.
-Second, type arguments tuples are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
+Second, type argument tuples are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
 Third, substitutions are directly applied to the type arguments instead of the formulae.
@@
 The data structures used in the pseudocode are based on the ones used in the high level description.
-Instead of a map \(M\) from symbols to monomorphic\pagebreak[2] type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type arguments tuples having already been matched against and those that have not.
+Instead of a map \(M\) from symbols to monomorphic\pagebreak[2] type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type argument tuples having already been matched against and those that have not.
 Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples.
@@
 At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
-Once these iterations are completed, the first phase is done and the substitutions used to create new type arguments tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.
+Once these iterations are completed, the first phase is done and the substitutions used to create new type argument tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.
 The third phase mangles the composite types of the newly monomorphised formulae.
@@
 The formula monomorphisation algorithm is given in Figure~\ref{mono_step}.
-It forms the core of the process by computing new type arguments tuples for a single formula.
+It forms the core of the process by computing new type argument tuples for a single formula.
 Type argument tuples are matched against each other to obtain a set of substitutions which is iterated over in the outermost loop.
-The separation of type arguments tuples into old and new maps is put to use to ensure that only combinations involving at least one new map are considered.
+The separation of type argument tuples into old and new maps is put to use to ensure that only combinations involving at least one new map are considered.
 This avoids re-computing some matchings processed in previous iterations.
@@
 
-The total number of type argument tuples can increase cubicly in the number of type arguments tuples at each iteration and can therefore grow doubly exponentially in the number of iterations.
+The total number of type argument tuples can increase cubicly in the number of type argument tuples at each iteration and can therefore grow doubly exponentially in the number of iterations.
 We sketch an example of such growth, for a single formula.
-If we assume that after \(k\) iterations, there are \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type arguments tuples, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
+If we assume that after \(k\) iterations, there are \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type argument tuples, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
 Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples.
 If half of these new type argument tuples are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples.
-Therefore, the total number of type arguments tuples on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).
+Therefore, the total number of type argument tuples on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).
 
@@
 
-Similar bounds are used for the non-monomorphic type arguments tuples:
+Similar bounds are used for the non-monomorphic type argument tuples:
 The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula.
@@
 Each bound nonetheless has its uses.
-For example, problems that lead to few successful matches but many type arguments tuples may benefit from a limit on the number of new type arguments tuples whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.
+For example, problems that lead to few successful matches but many type argument tuples may benefit from a limit on the number of new type argument tuples whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.
 
@@
 
-   We thank Sascha Böhme for fruitful discussions.
+   We thank Sascha Böhme for fruitful discussions and
+   Jannis Limperg for suggesting textual improvements.
 

### 2025-02-21 iterative_monomorphisation.tex
Jannisg

@@
    \item \(\sym{p}\langle \ty{int}\rangle(0)\)
-   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){,}\; \sym{p}\langle\alpha\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
+   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){.}\; \sym{p}\langle\alpha\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
 \end{enumerate}
@@
 \begin{enumerate}
-   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){,}\; \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
+   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){.}\; \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
 \end{enumerate}
@@
 \begin{enumerate}
-   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){,}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
+   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){.}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
 \end{enumerate}
@@
    \item[5.] 
-         \( \forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,} \\ \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
+         \( \forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){.} \\ \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
 \end{enumerate}
@@
 Third, substitutions are directly applied to the type arguments instead of the formulae.
-This dispenses from having to re-extract the type arguments from the formulae at each iteration.
+This avoids having to re-extract the type arguments from the formulae at each iteration.
 New formulae are generated only once all iterations are complete, in a separate, final step.
 
-The data structures used in the pseudocode are based on the ones used in the high level description.
-Instead of a map \(M\) from symbols to monomorphic\pagebreak[2] type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type argument tuples having already been matched against and those that have not.
+The data structures used in the algorithm are based on the ones used in the high level description.
+Instead of a map \(M\) from symbols to monomorphic\pagebreak[2] type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between those type argument tuples that have already been matched against and those that have not.
 Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples.
-Finally, we keep track of a map \(S\) from formulae to the substitutions generated by the matchings.
+Finally, maintain a map \(S\) from formulae to the substitutions generated by the matchings.
 It is used to generate new formulae in the final phase.
 
-All sets referenced in the pseudocode are %assumed to be
+All sets referenced in the algorithm are %assumed to be
 finite. Moreover, the algorithm relies on primitives whose implementation depends on the specifics of the grammar and logic used. %They will therefore not be expanded upon.
@@
    \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set.
-   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)}, where \(\sigma\) results from the matching.
-The pseudocode matches only non-monomorphic types against monomorphic types.
+   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, matches \(\upsilon\) against \(\tau\) and either fails or returns \emph{Success\((\sigma)\)}, where \(\sigma\) results from the matching.
+The algorithm matches only non-monomorphic types against monomorphic types.
    \item[\labelitemi] \emph{domain\((\sigma)\)} %, where \(\sigma\) is a substitution,
@@
 %\end{quote}
-\caption{Pseudocode for iterative monomorphisation}
+\caption{Algorithm for iterative monomorphisation}
 \label{iter_mono}
@@
 The third phase mangles the composite types of the newly monomorphised formulae.
-This allows targeting a simply typed logic with no general support for $n$-ary type constructors.
+This allows targeting a simply typed logic with no support for $n$-ary type constructors.
 %~\cite[Sections 4 and 5]{mono-trans}.
@@
    \Fn(){\(\FMonoStep(\varphi, M_{\text{old}}, M_{\text{new}}, N_{\text{old}}(\varphi), N_{\text{new}}(\varphi))\)}{
+
+   \SetKw{Let}{let}
 
@@
 
-   \ForEach{\(\sigma\in \SubstGen(M_{\text{new}}, N_{\text{new}}(\varphi))
+   \Let $S' = \SubstGen(M_{\text{new}}, N_{\text{new}}(\varphi))
    \cup \SubstGen(M_{\text{new}}, N_{\text{old}}(\varphi))
-   \cup \SubstGen(M_{\text{old}}, N_{\text{new}}(\varphi))\)}{
-   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)\)}{
+   \cup \SubstGen(M_{\text{old}}, N_{\text{new}}(\varphi))$\;
+   \ForEach{\(\sigma\in S'\)}{
+   \ForEach{\((f\mapsto (\upsilon_1, \dots,\upsilon_n))\in N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)\)}{
             \eIf{\(\TyVars(\upsilon_1, \dots, \upsilon_n) \subseteq \Domain(\sigma)\)}{
@@
                   \textcolor{black}{
-                  \(M_{\text{next}}(f)\leftarrow M_{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
+                  \(M_{\text{next}}(f)\leftarrow M_{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\kern-300mm\}\)\;
                   \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
@@
                   \(M_{\text{next}} \leftarrow M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}})\)\;
-                  \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\setminus (N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi))\)\;
+                  \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\setminus (N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi))\kern-300mm\)\;
                   \Return \((M_{\text{next}}, N_{\text{next}}(\varphi), \UsedSubst)\)
@@
                \If{\(|N_{\text{next}}(\varphi)| < \MaxPoly\)}{ \textcolor{black} {
-               \(N_{\text{next}}(\varphi)(f)\leftarrow N_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
+               \(N_{\text{next}}(\varphi)(f)\leftarrow N_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\kern-300mm\)\;
                \(\UsedSubst \leftarrow \UsedSubst \cup\{\sigma\}\)\;
@@
 %\end{quote}
-\caption{Pseudocode for formula monomorphisation step}
+\caption{Algorithm for formula monomorphisation step}
 \label{mono_step}
@@
 Type argument tuples are matched against each other to obtain a set of substitutions which is iterated over in the outermost loop.
-The separation of type argument tuples into old and new maps is put to use to ensure that only combinations involving at least one new map are considered.
+The separation of type argument tuples into old and new maps is used to ensure that only combinations involving at least one new map are considered.
 This avoids re-computing some matchings processed in previous iterations.
@@
 
-The total number of type argument tuples can increase cubicly in the number of type argument tuples at each iteration and can therefore grow doubly exponentially in the number of iterations.
+The total number of type argument tuples can increase cubically in the number of type argument tuples at each iteration and can therefore grow doubly exponentially in the number of iterations.
 We sketch an example of such growth, for a single formula.
@@
 An important difference with the monomorphic case is that \(n\) depends on the current formula being processed whilst \(m\) does not.
-Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.
+Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and iteration.
 
@@
 The resulting substitutions are composed if they are compatible.
-In the pseudocode, this is checked by making sure the \textbf{while} loop has successfully iterated over all elements of the type argument tuple.
+In the algorithm, this is checked by making sure the \textbf{while} loop has successfully iterated over all elements of the type argument tuple.
 If any substitutions are incompatible, the matchings are discarded.
 Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant.
-The total number of substitutions generated is capped by \textcolor{ourblueviolet}{substitution\_limit}.
+The total number of substitutions generated is limited by \textcolor{ourblueviolet}{substitution\_limit}.
 
@@
          \While{\(i \leq n\)}{
-            \eIf{\(\Match(\upsilon_i, \tau_i)\) has the form \(\Success(\sigma_i)\) \And \(\Compatible(\sigma, \sigma_i)\)}{
+            \eIf{\(\Match(\upsilon_i, \tau_i)\) has the form \(\Success(\sigma_i)\) \And \(\Compatible(\sigma, \sigma_i)\kern-300mm\)}{
                \(\strut\sigma\leftarrow\sigma_i\circ\sigma\)
@@
 %\end{quote}
-\caption{Pseudocode for match generation}
+\caption{Algorithm for match generation}
 \label{subst_gen}
 \end{figure}
+
+%%% @TANGUY: Jannis wrote:
+% "I feel like this pseudocode could be simplified a lot by taking a
+% higher-level view of the while loop. E.g.
+
+% if for all v_i and t_i, match(v_i, t_i) = Success(σ_i) and the σ_i are all 
+% compatible then
+%    σ := σ_1 \comp ... \comp σ_n
+
+% The fact that you break early is imo trivial enough that it doesn't need to be
+% discussed explicitly."
 
@@
 
+%%% @TANGUY: Jannis wrote:
+% "All the other limits are named something like max_x or x_cap."
    \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step} 
@@
 %\end{quote}
-\caption{Pseudocode for monomorphic formula generation}
+\caption{Algorithm for monomorphic formula generation}
 \label{gen_formulae}
@@
 Each bound nonetheless has its uses.
-For example, problems that lead to few successful matches but many type argument tuples may benefit from a limit on the number of new type argument tuples whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.
+For example, problems that lead to few successful matches but many type argument tuples may benefit from a limit on the number of new type argument tuples whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitutions.
 
@@
 %\end{quote}
-\caption{Pseudocode for monomorphising substitution generation}
+\caption{Algorithm for monomorphising substitution generation}
 \label{mono_substs}
@@
 Once all monomorphisation iterations have been completed, we are left with a set of the substitutions that have been used to generate new type arguments.
-The last phase uses this set to instantiate the type variables in the input problem's non-monomorphic formuale.
+The last phase uses this set to instantiate the type variables in the input problem's non-monomorphic formulae.
 In the presence of bounds, the order in which the elements of \(S\) are traversed affects the formulae resulting from the last phase.
@@
 The \emph{generate\_mono\_formulae} function is given in Figure~\ref{gen_formulae}.
-It generates monomorphising substitutions and applies them to the polymorphic formula of the input problem that they instantiate.
-Since the substitutions are monomorphising relative to the formula they applied to, the resulting formulae will be monomorphic.
+It generates monomorphising substitutions and applies them to the polymorphic formulae of the input problem that they instantiate.
+A substitution \(\sigma\) is \emph{monomorphising} for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
+Since the substitutions are monomorphising relative to the formula they are applied to, the resulting formulae will be monomorphic.
 The \textcolor{ourblueviolet}{total\_new\_formulae} bound is used to control the total number of new formulae.
@@
 To monomorphise a polymorphic formula, we first compute its {monomorphising substitutions} using the \emph{mono\_substs} function given in Figure~\ref{mono_substs}.
-A substitution \(\sigma\) is \emph{monomorphising} for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
 Such substitutions are computed using a recursive function.
-Given a set \(V\) of type variables and a set \(S(\varphi)\) of substitutions, it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S(\varphi)\) that instantiates at least one of the type variables in \(V\).
+Given a set \(V\) of type variables and a set \(S(\varphi)\) of substitutions, it selects a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S(\varphi)\) that instantiates at least one of the type variables in \(V\).
 It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they can be composed and the function recursively called to instantiate the remaining type variables.
@@
 \begin{enumerate}
-   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic formula per initial polymorphic formula.
-Generating an excessive number of new formulae can flood the downstream prover.
-The final number of output formulae is ensured to be at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
+   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic formulae per initial polymorphic formula.
+Generating an excessive number of new formulae can overwhelm the downstream prover.
+The final number of output formulae is limited to at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
    \item The \emph{mono\_substs} function is the algorithm's most explosive part.
@@
 Recall that the total number of type argument tuples used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations.
+%%% @TANGUY: About the last sentence, Jannis writes:
+%%% "Not immediately clear to me what this entails."
 \end{enumerate}
@@
 The monomorphisation algorithm is parametrised by many bounds.
-The first part of the evaluation process is to find appropriate values for these bounds.
-The second part compares the performance of Zipperposition with and without E on polymorphic problems.
+The first part of the evaluation process seeks appropriate values for these bounds.
+The second part compares the performance of Zipperposition without E and with the new monormorphising E backend on polymorphic problems.
 The third part compares the performance of different provers on polymorphic problems and their monomorphised counterparts.
@@
 In total, our benchmark suite contains 1534 polymorphic problems.
-We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in at most 30 seconds with a single thread.
+We chose as a measure of success for a given prover (or prover configuration) the number of problems
+that could be solved by the prover in at most 30 seconds per problem with a single thread.
+
+%%% @TANGUY: Jannis writes:
+% "Also, does the 30-sec timeout include the monomorphisation pass itself?"
+
 %This corresponds to the default time limit in Sledgehammer.
@@
 
-Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables but takes more time.
-Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed.
-Given the number of bounds, however, this is unfeasible.
-Instead, we group closely related bounds together and test combinations of values for the bounds in these groups.
+Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiation of type variables but takes more time.
+Since we cannot test all possible combinations of values for all bounds to find the best compromise between completeness and speed,
+we group closely related bounds together and test combinations of values for the bounds in these groups.
 Once we find the best performing set of values for a group, we assign these values to the corresponding bounds as we begin the search for the next group.
@@
 
-Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options used to run Zipperposition.
+Before finding values to assign to the bounds of the monomorphisation algorithm, we must choose which base options to run Zipperposition with.
 The prover has a portfolio mode with several configurations.
@@
 
-We conducted additional informal evaluations on the 500 problem suite to find appropriate default values and test ranges for the monomorphisation bounds.
+We conducted additional informal evaluations on the 500 problems to find appropriate default values and test ranges for the monomorphisation bounds.
 We started the option evaluation process with the base configuration and the following default values for monomorphisation bounds and parameters for E:
@@
 
-Table~\ref{mono_ty_args} groups bounds that control the maximum number of newly generated monomorphic type arguments per formula per iteration.
+Table~\ref{mono_ty_args} groups bounds that control the maximum number of newly generated monomorphic type arguments per formula and iteration.
 The limit on newly generated type arguments is determined by three components that form a natural group of bounds.
@@
 Finally, the `separation' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently.
-Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seems to affect performance when the `separation' heuristic is used.
+Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seem to affect performance when the `separation' heuristic is used.
 % TODO find a better term than "separation heuristic" maybe separation ?
@@
 Table \ref{pb_size} groups the bounds related to the size of the output problem to be passed to the E prover.
-The absolute cap is the maximum number of formulae passed to E.
+The absolute limit is the maximum number of formulae passed to E.
 The multiplier limits the total number of newly generated formulae based on the problem's initial number of formulae.
@@
 To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation.
-Table~\ref{mono_time} tests the amount of time that monomorphisation is allowed to run for against the number of iteration of the monomorphisation algorithm.
-Neither parameter seems to have much of an effect on the performance of the algorithm.
+Table~\ref{mono_time} tests the amount of time that is allocated to monomorphisation against the number of iteration of the monomorphisation algorithm.
+Neither parameter seems to substantially affect the performance of the algorithm.
 
@@
 The E timeout limits for how long E is run before being interrupted, at which point Zipperposition resumes normal operation.
-The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more iterative monomorphisation is explosive.
-This likely explains the poor performance for greater values of E call point.
+The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more explosive iterative monomorphisation is.
+This likely explains the poor performance for greater values of the E call point.
 
@@
 
-\subsection{E as Zipperposition backend}
+\subsection{E as a Zipperposition backend}
 
@@
 We compare the performance of two instances of Zipperposition.
-The first instance is run in the portfolio mode \verb|portfolio.sequential.py|, which attempts to prove the problem with various configuration in succession.
-Of these configuration, some may attempt to call E as a backend when the given problem is monomorphic.
-Because all given problems are non-monomorphic, this instance is labeled `without E'.
-The second instance, is a modification of the \verb|portfolio.sequential.py| file where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase.
+The first instance is run in the portfolio mode \verb|portfolio.sequential.py|, which attempts to prove the problem with various configurations tried in succession.
+Of these configurations, some may attempt to call E as a backend when the given problem is monomorphic, but because all given problems are non-monomorphic, this instance is labeled `without E'.
+The second instance is a modification of the \verb|portfolio.sequential.py| file where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase.
 This instance can successfully call E as a backend because it is able to provide E with a monomorphised problem.
-Each of the two instances is evaluated on the set of the 500 previously used problems and the set of the remaining 1034 problems separately.
+Each of the two instances is evaluated on the set of the 500 previously used problems, and the set of the remaining 1034 problems is evaluated separately.
 The results of these evaluations do not suggest that any overfitting took place during the option optimisation phase.
@@
 
-\subsection{Monomorphisation as preprocessor}
+\subsection{Monomorphisation as a preprocessor}
 
@@
 The monomorphisation approach is evaluated in two steps.
-First, each problem is monomorphised using the options obtained from the first evaluation phase except for the monomorphisation timeout option which is increased to 30~seconds.
+First, each problem is monomorphised using the options obtained from the first evaluation phase except for the monomorphisation timeout option, which is increased to 30~seconds.
 For 149~problems, monomorphisation times out.
-Second, each prover is ran on the remaining monomorphised problems, and the results are tallied in the table's `Mono' column.
-In addition to the polymorphic provers used in the `Native' tests, the monomorphic prover E is run on the monomorphised problems to provide an additional yardstick.
-Instead of running each prover 30 seconds on the monomorphised problems, the monomorphisation time (rounded up to the nearest second) is subtracted to compare fairly against the `Native' column, which does not have a similar preprocessing phase.
+Second, each prover is run on the remaining monomorphised problems, and the results are tallied in the table's `Mono' column.
+In addition to the polymorphic provers used in the `Native' tests, the monomorphic prover E is run on the monomorphised problems to provide an additional point of comparison.
+Instead of running each prover for 30 seconds on the monomorphised problems, the monomorphisation time (rounded up to the nearest second) is subtracted to compare fairly against the `Native' column, which does not have a similar preprocessing phase.
 
@@
 blindly enumerates candidate instantiations, without exploiting any knowledge
-about the logical structure of the formulas in which symbols occur. For
+about the logical structure of the formulae in which symbols occur. For
 example, a lemma $\sym{p}\langle\alpha\rangle$ cannot be used to prove the
@@
 monomorphisation. Third, although one would expect native implementations of
-polymorphism to outperform any preprocessor, this is not the case now, therefore indicating that there is considerable room for improvement on the native front.
+polymorphism to outperform any preprocessor, this is not the case now, indicating that there is considerable room for improvement on the native front.
 

### 2025-02-24 iterative_monomorphisation.tex
tune

@@
 Automatic theorem provers are employed to provide automation to users of proof assistants.
-Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle/HOL~\cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae.
+Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle\slash HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae.
 By contrast, many automatic provers operate only on monomorphic logics.
@@
 bound to be incomplete.
-By a type version of the compactness theorem, we have that for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of \(\varphi\).
+By a type version of the compactness theorem, there exists for any first order polymorphic formula \(\varphi\) an equisatisfiable finite set of monomorphic instances of \(\varphi\).
 However, this set cannot be computed~\cite[Theorem~1]{expr-poly-types}.
@@
 Consider the unary type constructor \ty{list}.
-If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, it may be possible to generate the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc.
+If a formula contains $\ty{list}(\alpha)$, where $\alpha$ is a type variable, the types $\ty{list}(\ty{int})$, $\ty{list}(\ty{list}(\ty{int}))$, etc., can be generated.
 However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has been generated.

### 2025-02-26 iterative_monomorphisation.tex
adressing Janis' comments on the paper

@@
 \SetKwData{MaxPoly}{\textcolor{ourblueviolet}{max\_nonm\_args}}
-\SetKwData{Limit}{\textcolor{ourblueviolet}{total\_new\_formulae}}
+\SetKwData{Limit}{\textcolor{ourblueviolet}{max\_new\_formulae}}
 \SetKwData{MonoSubstsLimit}{\textcolor{ourblueviolet}{max\_substs}}
-\SetKwData{SubstLimit}{\textcolor{ourblueviolet}{substitution\_limit}}
+\SetKwData{SubstLimit}{\textcolor{ourblueviolet}{substitution\_cap}}
 \SetKwData{Loop}{\textcolor{ourblueviolet}{num\_loops}}
@@
    \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set.
-   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, matches \(\upsilon\) against \(\tau\) and either fails or returns \emph{Success\((\sigma)\)}, where \(\sigma\) results from the matching.
+   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, matches \(\upsilon\) against \(\tau\) and either fails or returns \emph{Some\((\sigma)\)}, where \(\sigma\) results from the matching.
 The algorithm matches only non-monomorphic types against monomorphic types.
@@
 Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant.
-The total number of substitutions generated is limited by \textcolor{ourblueviolet}{substitution\_limit}.
+The total number of substitutions generated is limited by \textcolor{ourblueviolet}{substitution\_cap}.
+
+%\begin{figure}[t!]
+%%\begin{quote}
+%\begin{algorithm}[H]
+%\SetKwFunction{SubstGen}{matches}
+%\Fn(){\(\SubstGen(M, N(\varphi))\)}{
+%   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
+%   \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
+%   \SetKwFunction{Match}{match}
+%   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
+%   \SetKw{Break}{break}
+%
+%   \KwData{\begin{minipage}[t]{.8\textwidth}
+%      \strut monomorphic type argument map \(M\) \\
+%      non-monomorphic type argument map \(N(\varphi)\)\strut
+%      \end{minipage}
+%   }
+%   \KwResult{set of substitutions}
+%   \BlankLine
+%
+%   \(S\leftarrow \emptyset\)\;
+%
+%   \BlankLine
+%
+%   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N(\varphi)\)}{
+%      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
+%         \(\sigma\leftarrow \{\}\)\;
+%         \(i \leftarrow 1\)\;
+%         \While{\(i \leq n\)}{
+%            \eIf{\(\Match(\upsilon_i, \tau_i)\) has the form \(\Success(\sigma_i)\) \And \(\Compatible(\sigma, \sigma_i)\kern-300mm\)}{
+%               \(\strut\sigma\leftarrow\sigma_i\circ\sigma\)
+%            }
+%            {\Break}
+%
+%            \(i\leftarrow i+1\)\;
+%         }
+%         
+%         \If{\(i > n\)}{
+%            \textcolor{ourblueviolet}{
+%               \eIf{\(|S| < \SubstLimit\)}{
+%                  \textcolor{black}{\(S\leftarrow S\cup \{\sigma\}\)\;}
+%               }
+%               {\Return \(S\)}
+%            }
+%         }
+%      }
+%
+%   }
+%
+%   \BlankLine
+%
+%   \Return \(S\)\;
+%
+%}
+%\end{algorithm}
+%%\end{quote}
+%\caption{Algorithm for match generation}
+%\label{subst_gen}
+%\end{figure}
 
@@
    \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}
-   \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
+   \SetKwFunction{Success}{Some}\SetKwFunction{Compatible}{compatible}
    \SetKwFunction{Match}{match}
@@
       \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
-         \(\sigma\leftarrow \{\}\)\;
-         \(i \leftarrow 1\)\;
-         \While{\(i \leq n\)}{
-            \eIf{\(\Match(\upsilon_i, \tau_i)\) has the form \(\Success(\sigma_i)\) \And \(\Compatible(\sigma, \sigma_i)\kern-300mm\)}{
-               \(\strut\sigma\leftarrow\sigma_i\circ\sigma\)
-            }
-            {\Break}
-
-            \(i\leftarrow i+1\)\;
-         }
-
-         \If{\(i > n\)}{
+         \If{for all \(i\), \(\Match(\upsilon_i, \tau_i) = \Success(\sigma_i)\) \And all \(\sigma_i\) are compatible}
+            {
+               \(\sigma\leftarrow\sigma_1\circ\dots\circ\sigma_n\)
+
+            }
+
             \textcolor{ourblueviolet}{
@@
                {\Return \(S\)}
-            }
             }
@@
 
-%%% @TANGUY: Jannis wrote:
-% "I feel like this pseudocode could be simplified a lot by taking a
-% higher-level view of the while loop. E.g.
-
-% if for all v_i and t_i, match(v_i, t_i) = Success(σ_i) and the σ_i are all 
-% compatible then
-%    σ := σ_1 \comp ... \comp σ_n
-
-% The fact that you break early is imo trivial enough that it doesn't need to be
-% discussed explicitly."
-
 \begin{figure}[t!]
@@
    \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(S\)}
-   \SetKwData{Limit}{total\_new\_formulae}
-
-%%% @TANGUY: Jannis wrote:
-% "All the other limits are named something like max_x or x_cap."
+   \SetKwData{Limit}{max\_new\_formulae}
+
    \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step} 
@@
 The various bounds presented here overlap to some extent.
-For instance, having at most \textcolor{ourblueviolet}{substitution\_limit} substitutions generated by \emph{matches} may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous.
+For instance, having at most \textcolor{ourblueviolet}{substitution\_cap} substitutions generated by \emph{matches} may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous.
 Each bound nonetheless has its uses.
@@
 Since the substitutions are monomorphising relative to the formula they are applied to, the resulting formulae will be monomorphic.
-The \textcolor{ourblueviolet}{total\_new\_formulae} bound is used to control the total number of new formulae.
+The \textcolor{ourblueviolet}{max\_new\_formulae} bound is used to control the total number of new formulae.
 It overlaps with \textcolor{ourblueviolet}{max\_substs} but can be useful to set an absolute limit on the size of the final problem.
@@
 Recall that the total number of type argument tuples used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations.
-%%% @TANGUY: About the last sentence, Jannis writes:
-%%% "Not immediately clear to me what this entails."
+The starting \(n\) may therefore already be very large.
 \end{enumerate}
@@
 We chose as a measure of success for a given prover (or prover configuration) the number of problems
-that could be solved by the prover in at most 30 seconds per problem with a single thread.
-
-%%% @TANGUY: Jannis writes:
-% "Also, does the 30-sec timeout include the monomorphisation pass itself?"
+that could be solved by the prover in at most 30 seconds%(including monomorphisation when relevant)
+per problem with a single thread.
 

### 2025-02-26 iterative_monomorphisation.tex
kind of fix line break
minor

@@
 The resulting substitutions are composed if they are compatible.
-In the algorithm, this is checked by making sure the \textbf{while} loop has successfully iterated over all elements of the type argument tuple.
+In the algorithm, compatibility is checked by making sure the \textbf{while} loop has successfully iterated over all elements of the type argument tuple.
 If any substitutions are incompatible, the matchings are discarded.
@@
 
-   \ForEach{\(\varphi\in\Phi\) \ST $\varphi$ is non-monomorphic}{
+   \ForEach{\(\varphi\in\Phi\) \textbf{s.t.} $\varphi$ is non-monomorphic}{
       \ForEach{\(\sigma\in \MonoSubst(\AllSubst(\varphi), \TyVars{\(\varphi\)}, \emptyset, \{\})\)}{
@@
    {
-      \Let \(\alpha \) \ST \(\alpha\in V\)\;
-      \ForEach{\(\sigma_{\negvthinspace\Delta}\in S\) \ST \(\alpha\in\Domain(\sigma_{\negvthinspace\Delta})\) \And \(\Compatible(\sigma,\sigma_{\negvthinspace\Delta})\)}{
+      \Let \(\alpha \) \textbf{s.t.} \(\alpha\in V\)\;
+      \ForEach{\(\sigma_{\negvthinspace\Delta}\in S\) \textbf{s.t.} \(\alpha\in\Domain(\sigma_{\negvthinspace\Delta})\) \And \(\Compatible(\sigma,\sigma_{\negvthinspace\Delta})\)}{
          \textcolor{ourblueviolet}{
@@
 We chose as a measure of success for a given prover (or prover configuration) the number of problems
-that could be solved by the prover in at most 30 seconds%(including monomorphisation when relevant)
+that could be solved by the prover in at most 30 seconds
+%(including monomorphisation when relevant)
 per problem with a single thread.

### 2025-02-26 iterative_monomorphisation.tex
fixed a couple of typos and changed algorithm for match generation to deal with line being to long

@@
 Each symbol instance from \(N(\varphi)\) is matched against all corresponding symbol instances from \(M\).
-For two such symbol instances, types from the non-monomorphic type argument tuple are matched component-wise against the types from the monomorphic type argument tuple.
+For two such symbol instances, the types from the non-monomorphic type argument tuple are matched component-wise against the types from the monomorphic type argument tuple.
 The resulting substitutions are composed if they are compatible.
@@
    \BlankLine
-
    \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N(\varphi)\)}{
       \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
-         \If{for all \(i\), \(\Match(\upsilon_i, \tau_i) = \Success(\sigma_i)\) \And all \(\sigma_i\) are compatible}
+         \If{for all \(0 \leq i\leq n\), \(\Match(\upsilon_i, \tau_i) = \Success(\sigma_i)\)}
             {
-               \(\sigma\leftarrow\sigma_1\circ\dots\circ\sigma_n\)
-
-            }
+               \If {\(\sigma_1,\dots,\sigma_n\) are compatible}{
+
+                  \(\sigma\leftarrow\sigma_1\circ\dots\circ\sigma_n\) \phantom{\big S}
 
@@
             }
+            }
+
+            }
+
       }

### 2025-03-03 iterative_monomorphisation.tex
fix
Mark Summerfield suggestions

@@
 
-Automatic theorem provers are employed to provide automation to users of proof assistants.
+Automatic theorem provers provide automation for proof assistant users.
 Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle\slash HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae.
 By contrast, many automatic provers operate only on monomorphic logics.
-One approach to closing this gap is to extend provers to natively support polymorphism, as has been done for Vampire~\cite{bhayat-reger-2020}.
-This, however, entails a lot of work that needs to be redone for every prover.
+One approach to closing this gap is to extend provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}.
+However, this entails a lot of work that must be redone for every prover.
 
@@
 
-Any translation method based on instantiation is
-bound to be incomplete.
+Any translation method based on instantiation will inevitably be incomplete.
 By a type version of the compactness theorem, there exists for any first order polymorphic formula \(\varphi\) an equisatisfiable finite set of monomorphic instances of \(\varphi\).
@@
 This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers.
-However, it is is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
+However, it is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
 A similar algorithm appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
@@
 We also provide a more detailed description to help future implementers.
-In addition, our description shows some of the ways in which an implementation can be made more efficient and avoid explosions (Section~\ref{sec:low-level-algorithm}).
+In addition, our description shows some of the ways in which an implementation can be made more efficient and avoid combinatorial explosions (Section~\ref{sec:low-level-algorithm}).
 
@@
 We assume problems to be sets of formulae.
-All symbols of the problem are collected, and the polymorphic instances of symbols are matched against the monomorphic ones.
+All the problem's symbols are collected, and the polymorphic symbol instances are matched against the monomorphic ones.
 This yields new symbol instances, both polymorphic and monomorphic.
@@
 However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has been generated.
-This example also makes it clear that the set of instances that can be generated is infinite.
+This example also shows that an infinite set of instances can be generated.
 
@@
 After the iterations are completed, the new monomorphic symbol instances are used to instantiate the polymorphic symbols in the problem's formulae, generating new monomorphic formulae.
-Finally, because monomorphic provers support only nullary type constructors, types must be `mangled'; for example, the compound type $\ty{list}(\ty{int})$ could be mangled to $\ty{list\_int}$.
+Finally, because monomorphic provers support only nullary type constructors, types must be `mangled'; for example, the compound type $\ty{list}(\ty{int})$ might be mangled to $\ty{list\_int}$.
 
@@
 The algorithm is sound because the newly generated formulae are instances of the initial problem's formulae. %, where type variables have been instantiated with monomorphic types.
-It is, however, not complete.
+However, it is not complete.
 
@@
 For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
-Any reasonable implementation requires bounds limiting the number of new type arguments, substitutions and formulae.
+/Any reasonable implementation must limit the number of new type arguments, substitutions and formulae.
 
@@
 
-\looseness=-1
-The algorithm presented above is too naive for an implementation. In this section, we present a lower level algorithm with the following features.
-First, numeric bounds are introduced to stop explosive enumerations.
+The algorithm presented above is too naïve for a practical implementation. In this section, we present a lower level algorithm with the following features.
+First, numeric bounds are introduced to stop combinatorially explosive enumerations.
 Second, type argument tuples are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
@@
 This avoids having to re-extract the type arguments from the formulae at each iteration.
-New formulae are generated only once all iterations are complete, in a separate, final step.
+New formulae are generated only once all iterations are completed, in a separate, final step.
 
 The data structures used in the algorithm are based on the ones used in the high level description.
-Instead of a map \(M\) from symbols to monomorphic\pagebreak[2] type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between those type argument tuples that have already been matched against and those that have not.
+Instead of a map \(M\) from symbols to monomorphic type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between those type argument tuples that have already been matched against and those that have not.
 Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples.
-Finally, maintain a map \(S\) from formulae to the substitutions generated by the matchings.
+Finally, we maintain a map \(S\) from formulae to the substitutions generated by the matchings.
 It is used to generate new formulae in the final phase.
@@
 At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
-Once these iterations are completed, the first phase is done and the substitutions used to create new type argument tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.
+Once these iterations are completed, the first phase is complete and the substitutions used to create new type argument tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.
 The third phase mangles the composite types of the newly monomorphised formulae.
@@
 An important difference with the monomorphic case is that \(n\) depends on the current formula being processed whilst \(m\) does not.
-Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and iteration.
+Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.
 
@@
 
-                  \(\sigma\leftarrow\sigma_1\circ\dots\circ\sigma_n\) \phantom{\big S}
+                  \(\sigma\leftarrow\sigma_1\circ\dots\circ\sigma_n\)\strut
 
@@
 For instance, having at most \textcolor{ourblueviolet}{substitution\_cap} substitutions generated by \emph{matches} may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous.
-Each bound nonetheless has its uses.
-For example, problems that lead to few successful matches but many type argument tuples may benefit from a limit on the number of new type argument tuples whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitutions.
+Nonetheless every bound has uses.
+For example, problems that lead to few successful matches but many type argument tuples may benefit from a limit on the number of new type argument tuples whilst problems for which substitution generation is more combinatorially explosive may benefit from a limit on the number of generated substitutions.
 
@@
    \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic formulae per initial polymorphic formula.
-Generating an excessive number of new formulae can overwhelm the downstream prover.
+Generating an excessive number of new formulae can overwhelm the prover.
 The final number of output formulae is limited to at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
-   \item The \emph{mono\_substs} function is the algorithm's most explosive part.
+   \item The \emph{mono\_substs} function is the algorithm's most combinatorially explosive part.
 For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) %different
@@
 
-Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiation of type variables but takes more time.
+Each bound of the monomorphisation process represents a tradeoff: a higher bound allows for a more exhaustive instantiation of type variables but takes more time.
 Since we cannot test all possible combinations of values for all bounds to find the best compromise between completeness and speed,
@@
 
-Table~\ref{mono_ty_args} groups bounds that control the maximum number of newly generated monomorphic type arguments per formula and iteration.
+Table~\ref{mono_ty_args} groups bounds that control the maximum number of newly generated monomorphic type arguments per formula and per iteration.
 The limit on newly generated type arguments is determined by three components that form a natural group of bounds.
@@
 Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments.
-The bound values are lower because we found non-monomorphic type arguments to be explosive in preliminary evaluations.
+The bound values are lower because we found non-monomorphic type arguments to be combinatorially explosive in preliminary evaluations.
 The table confirms that non-monomorphic type argument generation drives the creation of useful non-monomorphic formulae.
@@
 To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation.
-Table~\ref{mono_time} tests the amount of time that is allocated to monomorphisation against the number of iteration of the monomorphisation algorithm.
-Neither parameter seems to substantially affect the performance of the algorithm.
+Table~\ref{mono_time} tests the amount of time that is allocated to monomorphisation against the number of iterations of the monomorphisation algorithm.
+Neither parameter seems to substantially affect the algorithm's performance.
 
@@
 The call point corresponds to the fraction of the total timeout of Zipperposition's timeout after which the normal operation of Zipperposition is interrupted to begin the monomorphisation process.
-The E timeout limits for how long E is run before being interrupted, at which point Zipperposition resumes normal operation.
-The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more explosive iterative monomorphisation is.
+The E timeout limits how long E is run before being interrupted, at which point Zipperposition resumes normal operation.
+The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more combinatorially explosive iterative monomorphisation is.
 This likely explains the poor performance for greater values of the E call point.
@@
 We chose Leo-III and Zipperposition.
-We unfortunately had to exclude Vampire because of parsing issues.
+Unfortunately we needed to exclude Vampire because of parsing issues.
 
@@
 relevance filters that heuristically select a subset of the available axioms;
-filters such as MePo~\cite{meng-paulson-2009} and SInE
-~\cite{hoder-voronkov-2011} are iterative and could be interleaved with
+filters such as MePo~\cite{meng-paulson-2009} and SInE~\cite{hoder-voronkov-2011} are iterative and could be interleaved with
 monomorphisation. Third, although one would expect native implementations of
-polymorphism to outperform any preprocessor, this is not the case now, indicating that there is considerable room for improvement on the native front.
+polymorphism to outperform any preprocessor, currently this is not the case, suggesting that there is considerable room for improvement on the native front.
 
@@
 
-   We thank Sascha Böhme for fruitful discussions and
-   Jannis Limperg for suggesting textual improvements.
+   We thank Sascha Böhme for fruitful discussions.
+   We thank Jannis Limperg and Mark Summerfield for suggesting textual improvements.
 

### 2025-04-16 REBUTTAL.txt
more

Dear Jasmin,

Thank you for your submission to CADE-30. The CADE-30 rebuttal period
has now started. The reviews on your paper are attached to this e-mail.
You have the opportunity to respond to them until April 18, midnight
(AoE). To submit your response please log on the EasyChair Web page
for CADE-30 and select your submission on the menu.

Please keep in mind the following during this process:

* We are not enforcing any upper bound on the length of your response
 but please try to limit it to 1000 words. Anything over that will
 be optional for the reviewers to read.

* The response must focus on factual errors in the reviews and the
 questions posed by the reviewers, if any. Please do not provide new
 research or reformulate the presentation. Try to be as concise and
 to the point as possible.

* The response period is an opportunity to react to the reviews but
 there is no requirement to do so. Thus, if you feel the reviews are
 accurate and the reviewers have not asked any questions, you do not
 have to respond. On the other hand, reviewers may not look favorably
 on a paper if authors fail to answer their questions.

* The reviews are as submitted by the PC members. Please keep in mind
 that this is not their final version and no coordination among the
 PC members was required at this stage. The reviews may later be
 updated to take into account your response and the discussions among
 the PC members. Moreover, we may find it necessary to solicit other
 outside reviews after the rebuttal period.

* While the program committee will read your response carefully and
 take it into account during the discussions, it will not directly
 respond to your response, either through the PC chairs or in the
 final versions of the reviews.

* Your response will be seen by all PC members who have access to the
 discussion of your paper, so please be polite and constructive.

Best wishes,
 Clark and Uwe

----------------------- REVIEW 1 ---------------------
SUBMISSION: 17
TITLE: Iterative Monomorphisation
AUTHORS: Tanguy Bozec and Jasmin Blanchette

----------- Overall evaluation -----------
The paper discusses the process of monomorphisation, that is
heuristic instantiation of a polymorphic automated reasoning
problem and then using a monomorphic prover on it. The authors
re-implement the procedure in Zipperposition and evaluate it.

I have two main concerns about the paper:

First, the novelty. The authors admit that algorithms
doing the same task in a similar way already existed in HOL
Light since the ninties and later in Sledgehammer. A more
detailed comparison of these would be needed. Apart from the
already mentioned monomorhpisation is also needed in many
other contexts, for example interation of ITPs with Metis and
leancop or in HOLyHammer in
HOL4. I do not know if some of these algorithms are the same,
but in any case I would really need to see a comparison of
the current work with the existing ones. What do other tools
that call monomorhpic ATPs like Why3 do? Even if some tools
might fail it should be discussed.

Second, it is not clear what logic does the work support.
Quote: "any reasonable rank-1 polymorphic logic", what is
reasonable? In the second definition the authors define a
"type arity" of a symbol and actually it is based on this
definition that I infer some things that are possible.
But does the logic have subtyping or even just type classes
(as Isabelle/HOL's does)? Things should be defined more
precisely.

Minor:
Page 1 'there exists, for any ... ' can we do it in the normal
order as in 'for any ... there exists ...'

'appears to be implemented' please look at what is implemented
and describe what things do what in HOL Light, HOL4, and other
logic transformation tools, say what does why3 or leo's
transformation preprocessor do with polymorphic problems.

Page 2: 'instances can be generated is infinite' it would be
good to show an example problem where more iterations are
needed to solve.

Page 3: 'applying a fixed number of iterations' why fixed?

Fig 2: The algorithm in MESON looks much simpler in code. Can you
compare them?

Page 7: 'we sketch an example' this is not really an example.

Why are the complexities useful? Are the complexities of the
Sledgehammer / MESON / ... algorithms different?

Fig 4: This does not bring much, the actual code would look
shorter than this algorithmization... Could we write these
algorithms more compactly?

Results (mostly table 1 on page 11):
Just an idea, as monomorphisation is supposed to add most to the
already existing polymorphic procedure, maybe rather than optimize
it on its own it is better to optimize it on problems that
polymorhpism cannot solve?

Page 14/Sec 5.2: I am a bit surprised about the way E is used and
the reasons it is used. Isn't Zipperposition an already fully
capable monomorphic superposition prover? Can you justify this?

"exclude Vampire because of parsing issues". Vampire should be fully
capable of parsing any compliant TPTP files and its parser is one
of the most efficient TPTP parsers.

----------------------- REVIEW 2 ---------------------
SUBMISSION: 17
TITLE: Iterative Monomorphisation
AUTHORS: Tanguy Bozec and Jasmin Blanchette

----------- Overall evaluation -----------
The paper describes an incremental monomorphisation algorithm, which can be
used to translate away rank-1 polymorphism from HOL/FOL problems by iteratively
instantiating the formulae’s type variables with concrete types. These problems
thus become accessible to reasoning systems that do not support polymorphism.
This necessarily incomplete but pragmatic approach has been briefly described
in Boehme's PhD thesis [5] and implemented as part of Isabelle/HOL’s SMT
integration and also used by the Sledgehammer system. The authors expand what
can be derived from these sources into a detailed description and suggest
guidelines for efficient implementation. They also implemented the algorithm in
the Zipperposition prover and report on its empirical evaluation on polymorphic
TPTP problems.

I am glad for a paper like this, as the topic clearly deserves a canonical
reference. At the same time, the question whether such format brings enough
contribution for a full conference paper cannot be avoided.

I was glad to see the exposition split into the "High level" and "Low level"
algorithm. The High level is already presented in a formally precise way and
accompanied by a small example, so what's left for the Low level are some
optimizations (make sure you don't repeat work of previous iterations; store
substitutions instead of formulas and only apply them in the end) and the
introduction of various numeric bounds that prevent the run from exploding (in
various dimensions). While establishing (maybe one could even say inventing)
these bounds seems important for practice, the optimizations, on the other
hand, look quite routine and their appearance in the pseudocode, which drives
the Low level section, make it (perhaps unnecessarily?) quite hard to penetrate
to a casual reader.

The evaluation part uses the new Zipperposition implementation of the
monomorphisation algorithm and describes how the authors decided on some
reasonable default values for the various numeric bounds to optimize
performance of the prover on the monomorphised versions of the polymorphic
problems from the TPTP library. Given the large number of these bounds, this
was a rather tedious process and the numbers reported in the numerous tables
are perhaps a bit uninteresting in the grand scheme of things. This again gives
the paper a bit of a workshop-ish flavor, but could arguably help future
implementors for the algorithm guide their way of tuning the bounds. I was also
not entirely convinced by the arguments for/against overfitting to the problem
set (details below).

Overall, however, given also how clearly the paper is written, I think it's
worth considering its acceptance!

Particulars:

- Could miniscoping help (or change the output to begin with)? It seems
  important what the individual formulas \varphi are in the input. What if we
  get \forall \alpha \in $tType. (\varphi_1 \land \varphi_2), could the
  algorithm benefit from instead considering? (\forall \alpha \in $tType.
  \varphi_1) \land (\forall \alpha \in $tType. \varphi_2)?

- "As a precaution against overfitting" - I don't see how this in itself
  could work as a precaution. Couldn't you still simply have overfitted to the
  500 random problems?

- Table 6 + the corresponding paragraph - so the "E timeout" is in seconds?
  And the "E call point" is a fraction of the overall timeout? "The longer
  Zipperposition runs before E is invoked, the more formulae are generated, and
  the more explosive iterative monomorphisation is." - how long does it run for
  a fraction value 0?

- "The results of these evaluations do not suggest that any overfitting took
  place during the option optimisation phase." - What's the justification for
  this? Could you please spell it out?

- "We unfortunately had to exclude Vampire because of parsing issues." - Did
  you try contacting the Vampire team about the issue or how to best mitigate
  it? They generally respond quite quickly these days.

----------------------- REVIEW 3 ---------------------
SUBMISSION: 17
TITLE: Iterative Monomorphisation
AUTHORS: Tanguy Bozec and Jasmin Blanchette

----------- Overall evaluation -----------
Paper summary
-------------

The paper introduces a method to generate monomorphic formulae out of
polymorphic formulae by substituting type variables by monomorphic types. This
is interesting in cases where an automatic theorem prover is to be invoked
which does not offer native support for polymorphism. In order to limit the
number of newly created formulae due to the instantiation of type variables,
several bounds are adopted. The approach is implemented in Zipperposition, a
polymorphic higher order prover. Zipperposition uses the E prover as a backend
and to date it could use E exclusively for monomorphic problems. This work
enables solving polymorphic problems in Zipperposition using E. An experimental
evaluation was carried out to determine the parameter combination on which
Zipperposition performed best on polymorphic formulae.

Overall evaluation
------------------

Applying the low level algorithm to Example 6 was hard. The notation is unclear
in parts and assumptions had to be made which might be incorrect (see also
questions to the authors), e.g.,:
- Function iterative_monomorphisation: S ← \emptyset, but S is never updated,
  instead S(\phi) is computed; similarly for N_old and N_old(\phi), N_new and
  N_new(\phi). I assume S is defined as the union of the S(\phi), same for
  N_old and N_new
- Function formula_mono_step: "(f |→ (υ_1,...,υ_n)) \in N_old(\phi) \union
  N_new(\phi)". I assume f is called on the non-monomorphic argument
  (υ_1,...,υ_n).
- Function formula_mono_step: M_next and N_next(\phi) are referenced, but they
  are defined in the main function iterative_monomorphisation and not known in
  formula_mono_step. I assume they are implicitly considered global variables.
- Function formula_mono_step: N_next and N_next(\phi). I assume N_next is the
  union of all N_next(\phi).
- p.7: "limit on the number of newly generated monomorphic type argument
  tupes". I assume this is max_mono_args in formula_mono_steps; "Similar bounds
  are used for the non-monomorphic type argument tuples". I assume this is
  max_nonm_args. However, their definitions do not coincide with the ones in
  formula_mono_steps.

The presentation need be more careful/consistent in some parts: 
- Function iterative_monomorphisation: the call of formula_mono_step does not
  match its definition (one parameter is missing, the parameter are ordered
  differently
- p.5: "match(v, \tau), where v and \tau are types"; from my understanding, v
  is a type variable
- Function formula_mono_step: type_variables is never used again (left hand
  side of "type_variables ← type_variables(\phi)"
- Function formula_mono_step: type_variables(\phi) vs.
  type_variables(v_1,...,v_n): argument type does not fit
- p.7: "limit on the number of newly generated non-/monomorphic type argument
  tupes". If these are max_mono_args/max_nonm_args, their definitions do not
  coincide with the ones in formula_mono_steps.
- p.7, second-to-last line: while loop -> foreach loop?

Adding more textual explanations and some examples would further enhance clarity.

The repository on Github seems not to be public, so I could not reproduce the
experiments.

I consider this work relevant. The presented approach leads to a significantly
better performance of Zipperposition with E compared to Zipperposition without
E. However, in my opinion the presentation need be enhanced significantly for
the paper to be accepted.

Questions to the authors
------------------------

- Function iterative_monomorphisation: how are S and S(\phi) related, and N_old
  and N_old(\phi), N_new and N_new(\phi)?
- p. 5: "match(v, \tau), where v and \tau are types"; isn't v a type variable
  instead of a type?
- Function formula_mono_step: what do mean bey "(f |→ (υ_1,...,υ_n))"?
- M_next and N_next(\phi): are these global variables?
- Which limits for generating argument tuples are you referring to on p. 7?
- p.7, 2nd par.: As far as I understood, no new formulae containing
  non-monomorphic type arguments are generated. However, your explanations
  suggest otherwise. What am I missing? Can you give an example?
- Function matches: wouldn't it be more efficient to check |S| before executing
  match(v_i, \tau_i) and to save the execution of match in case |S| is already
  to large?

Minor comments
--------------

- On a black and white printout, the blue text is barely distinguishable from
  the black text. I suggest to use a lighter shade.

------------------------------------------------------

### 2025-04-16 rebuttal_answer.txt
rename

Common concerns and questions:

Novelty:
The main issue with this paper is that of novelty. We felt that an overview of this monomorphisation technique deserved its own paper for three reasons:
   - Genericity, we made an effort to avoid imposing conditions on the underlying logic, whereas other implementations or discussions often do so within a specific (albeit often generalisable) framework.
   - Possible applications, to the best of our knowledge, monomorphisation has not been used to enhance the performance of polymorphic provers. This is valuable as an indication of the benefits of re-evaluating the performance of native polymorphic provers.
   - Reference, this kind of algorithm is relatively widespread. It seemed necessary to provide a baseline reference of this algorithm against which future improvements or alternative methods could be compared.

Vampire does not feature among the evaluated provers. This decision was made because at the time of testing, Vampire could not parse a significant number of our problem set. Although this issue was adressed in the `master` branch, this was not the case in the `hol` branch, which featured the required higher-order reasoning. Regretfully, we did not contact the Vampire team.

Specific questions and comments:

Review 1 (the mean one):
Comparison with other implementations: [Not sure what to say, it is a lot of work to go through and understand other implementations for relatively little benefit, we know this algorithm is out there in some form and used in various provers]

Logic: The type system required to apply the monomorphisation algorithm is explicitely defined in the preliminairies. The underlying logic, is left purposefully vague because we need very few hypotheses for the algorithm to be applicable. The TF1 and TH1 logics are cited as being compatible with our work. Extensions to subtyping or typeclasses were not considered.

Minor:
 - The questions on pages 2 and 3 are adressed in example 6 (page 4).
 - Figure 2: [not successful in finding the relevant code so far, does he want me to compare them in the rebuttal or in the paper]
 - The complexities are not formal bounds on the algorithm but give an idea of the explosiveness of the procedure. Any systematic enumerations of all monomorphic types will encounter the same issue.
 - Figure 4: [There are two nested for loops for 8 lines of pseudo code]
 - Results: An important potential application of iterative monomorphisation would be extending monomorphic provers. Performance on problems that can already be solved by polymorphic provers is still relevant in this usecase.
 - page 14: The previous version of Zipperposition made use of E as a backend, because it is a very efficient prover. Zipperposition's strength was its implementation of novel algorithms. Delegating to E when the subproblems required optimised implementation of well-known procedures proved to be a successfull. This technique is used elsewhere in theorem proving, for instance Leo-III can be run with E or CVC4 as backend.

Review 2 (the nice one):
 - Microscoping: [it seems that this would be unsound, but if it can be done it would be more efficient]
 - Overfitting: The goal was to limit any overfitting to the 500 problems and ensure that resulting bound values still performed well on the fresh 1043 remaining problems. We estimated this to be successfull because the ration of solved problems was similar in both cases.
 - Timeout:
   + E timeout is in seconds and dictates how long does the E prover run for
   + E call point is the fraction of the overall timeout of Zipperposition after which we call the E prover
   + ex: For E timeout = 5s, E call point = 0.2 and Zippperposition timeout = 30s, we call the E prover for 5 seconds after having had Zipperposition run for 6 seconds

Review 3 (the technical one):

[Not much to say, mostly fixes to do, haven't had time to dig into the algos proper, but it is straightforward]

### 2025-04-16 rebuttal_answer.txt
edit answers

@@
-Common concerns and questions:
+We first want to thank the reviewers for their comments and suggestions. We first address the main concerns and questions shared by multiple reviewers, then we answer more specific questions.
 
-Novelty:
-The main issue with this paper is that of novelty. We felt that an overview of this monomorphisation technique deserved its own paper for three reasons:
+## Reviewer 1
+
+Concerning novelty, we felt that an overview of this monomorphisation technique deserved its own paper for these reasons:
+
+  - Reference: this kind of algorithm is relatively widespread.
+It seemed necessary to provide a baseline reference of this algorithm against which future improvements or alternative methods could be compared.
+
+  - Evaluation goes here TODO
+
+Kill, use as source for canibalization:
+
   - Genericity, we made an effort to avoid imposing conditions on the underlying logic, whereas other implementations or discussions often do so within a specific (albeit often generalisable) framework.
+
   - Possible applications, to the best of our knowledge, monomorphisation has not been used to enhance the performance of polymorphic provers.
 This is valuable as an indication of the benefits of re-evaluating the performance of native polymorphic provers.
-   - Reference, this kind of algorithm is relatively widespread.
-It seemed necessary to provide a baseline reference of this algorithm against which future improvements or alternative methods could be compared.
 
-Vampire does not feature among the evaluated provers.
-This decision was made because at the time of testing, Vampire could not parse a significant number of our problem set.
-Although this issue was adressed in the `master` branch, this was not the case in the `hol` branch, which featured the required higher-order reasoning.
-Regretfully, we did not contact the Vampire team.
+Concerning comparison with other implementations, this would be an interesting research project, but we consider it beyond the scope of the present paper.
+The algorithms are quite cryptic and poorly documented, so it would take quite some engineering to get them running and to understand their inner workings.
+In fact, we attempted to understand one of these algorithms, the one by Sascha Böhme, and we gave up after several hours of study.
 
-Specific questions and comments:
+Concerning the logic, the type system required to apply the monomorphisation algorithm is explicitly defined in the preliminairies.
+The underlying logic is left purposefully vague because we need very few hypotheses for the algorithm to be applicable.
+The TF1 and TH1 logics are cited as being compatible with our work.
+The algorithm does not support subtyping and type classes, but it could likely be extended to support them.
+By contrast, higher-rank polymorphism would not be compatible.
+We will clarify this in the text.
 
-Review 1 (the mean one):
-Comparison with other implementations: [Not sure what to say, it is a lot of work to go through and understand other implementations for relatively little benefit, we know this algorithm is out there in some form and used in various provers]
+Concerning the question about page 2, Example 6 on page 4 shows that monomorphisation can be infinite.
 
-Logic: The type system required to apply the monomorphisation algorithm is explicitely defined in the preliminairies.
-The underlying logic, is left purposefully vague because we need very few hypotheses for the algorithm to be applicable.
-The TF1 and TH1 logics are cited as being compatible with our work.
-Extensions to subtyping or typeclasses were not considered.
+Concerning the question about page 3, "fixed" is perhaps not the best word.
+The algorithm applies a *bounded* number of iterations.
+The bound is necessary to avoid divergence in cases such as Example 6.
 
-Minor:
- - The questions on pages 2 and 3 are adressed in example 6 (page 4).
- - Figure 2: [not successful in finding the relevant code so far, does he want me to compare them in the rebuttal or in the paper]
- - The complexities are not formal bounds on the algorithm but give an idea of the explosiveness of the procedure. Any systematic enumerations of all monomorphic types will encounter the same issue.
- - Figure 4: [There are two nested for loops for 8 lines of pseudo code]
- - Results: An important potential application of iterative monomorphisation would be extending monomorphic provers.
+Concerning the question "Why are the complexities useful?": The complexities are not formal bounds on the algorithm but give an idea of the explosiveness of the procedure.
+Any systematic enumerations of all monomorphic types will encounter the same issue.
+
+Concerning Figure 4, we do not understand the criticism. Tbe code is very short, and a few lines of it are instrumentalisation for the bounds.
+
+Concerning "just an idea", an important potential application of iterative monomorphisation would be extending monomorphic provers.
 Performance on problems that can already be solved by polymorphic provers is still relevant in this usecase.
- - page 14: The previous version of Zipperposition made use of E as a backend, because it is a very efficient prover.
+
+Concerning the question about page 14, the previous version of Zipperposition made use of E as a backend, because it is a very efficient prover.
 Zipperposition's strength was its implementation of novel algorithms.
@@
 This technique is used elsewhere in theorem proving, for instance Leo-III can be run with E or CVC4 as backend.
+See Section 8 of Vukmirovic et al., "Making Higher-Order Superposition Work".
+We will briefly explain this in the next version of the text.
 
-Review 2 (the nice one):
+Concerning Vampire's parser, we excluded Vampire because at the time of testing, Vampire could not parse a significant number of our problem set.
+Although this issue was addressed in the `master` branch, this was not the case in the `hol` branch, which featured the required higher-order reasoning.
+In retrospect, we should have contacted the Vampire team.
+
+## Reviewer 2
+
  - Microscoping: [it seems that this would be unsound, but if it can be done it would be more efficient]
@@
 
-Review 3 (the technical one):
+## Reviewer 3
 

### 2025-04-16 rebuttal_answer.txt
more

@@
 
-Concerning Figure 4, we do not understand the criticism. Tbe code is very short, and a few lines of it are instrumentalisation for the bounds.
+Concerning Figure 4, we do not understand the criticism. Tbe code is very short, and a few lines of it are for enforcing the bounds.
 
 Concerning "just an idea", an important potential application of iterative monomorphisation would be extending monomorphic provers.
-Performance on problems that can already be solved by polymorphic provers is still relevant in this usecase.
+Performance on problems that can already be solved by polymorphic provers is still relevant in this use case.
 
@@
 Zipperposition's strength was its implementation of novel algorithms.
-Delegating to E when the subproblems required optimised implementation of well-known procedures proved to be a successfull.
-This technique is used elsewhere in theorem proving, for instance Leo-III can be run with E or CVC4 as backend.
+Delegating to E when the subproblems required optimised implementation of well-known procedures proved to be a successful.
+This technique is used elsewhere in theorem proving, for instance, Leo-III can be run with E or CVC4 as backend.
 See Section 8 of Vukmirovic et al., "Making Higher-Order Superposition Work".
@@
 Concerning Vampire's parser, we excluded Vampire because at the time of testing, Vampire could not parse a significant number of our problem set.
-Although this issue was addressed in the `master` branch, this was not the case in the `hol` branch, which featured the required higher-order reasoning.
+Although this issue was addressed in the "master" branch, this was not the case in the "hol" branch, which featured the required higher-order reasoning.
 In retrospect, we should have contacted the Vampire team.
@@
 
- - Microscoping: [it seems that this would be unsound, but if it can be done it would be more efficient]
- - Overfitting: The goal was to limit any overfitting to the 500 problems and ensure that resulting bound values still performed well on the fresh 1043 remaining problems.
+Miniscoping could indeed help on some formulas, although top-level conjunctions tend to arise seldom in practice.
+This could easily be achieved by a preprocessor and would require no changes to our algorithm.
+
+Concerning overfitting, the goal was to limit any overfitting to the 500 problems and ensure that resulting bound values still performed well on the fresh 1043 remaining problems.
 We estimated this to be successfull because the ration of solved problems was similar in both cases.
- - Timeout:
-   + E timeout is in seconds and dictates how long does the E prover run for
-   + E call point is the fraction of the overall timeout of Zipperposition after which we call the E prover
-   + ex: For E timeout = 5s, E call point = 0.2 and Zippperposition timeout = 30s, we call the E prover for 5 seconds after having had Zipperposition run for 6 seconds
+We will clarify the text.
+
+Concerning the E timeout, the "E timeout" is in seconds and dictates how long does the E prover run for.
+The "E call point" is the fraction of the overall timeout of Zipperposition after which we call the E prover.
+For example, for E timeout = 5 s, E call point = 0.2 and Zippperposition timeout = 30 s, we call the E prover for 5 seconds after having had Zipperposition run for 6 seconds.
+We will clarify this in the text.
+
+Concerning novelty and Vampire's parser, see our answers to reviewer 1 above.
 
@@
 
-[Not much to say, mostly fixes to do, haven't had time to dig into the algos proper, but it is straightforward]
+We thank this reviewer for the extensive comments on the presentation and the code.
+We will consider their comments when revising the text.
+We will also use a slightly lighter shade of blue for the bound-related code.
+
+We answer the "questions for the authors" below. In each case, we will also clarify the text accordingly.
+
+> Function iterative_monomorphisation: how are S and S(\phi) related, and N_old and N_old(\phi), N_new and N_new(\phi)?
+
+TODO
+
+> p. 5: "match(v, \tau), where v and \tau are types"; isn't v a type variable instead of a type?
+
+TODO
+
+> Function formula_mono_step: what do mean bey "(f |→ (υ_1,...,υ_n))"?
+
+TODO
+
+> M_next and N_next(\phi): are these global variables?
+
+TODO
+
+> Which limits for generating argument tuples are you referring to on p. 7?
+
+TODO
+
+> p.7, 2nd par.: As far as I understood, no new formulae containing non-monomorphic type arguments are generated.
+However, your explanations suggest otherwise.
+What am I missing?
+Can you give an example?
+
+TODO
+
+> Function matches: wouldn't it be more efficient to check |S| before executing match(v_i, \tau_i) and to save the execution of match in case |S| is already to large?
+
+TODO

### 2025-04-16 rebuttal_answer.txt
finished todos and did some editing

@@
-We first want to thank the reviewers for their comments and suggestions. We first address the main concerns and questions shared by multiple reviewers, then we answer more specific questions.
+We first want to thank the reviewers for their comments and suggestions. We answered the questions and comments in separate part for each reviewer.
 
@@
 
-Concerning novelty, we felt that an overview of this monomorphisation technique deserved its own paper for these reasons:
+Concerning novelty, we felt that an overview of this monomorphisation technique deserved its own paper for two main reasons:
 
-  - Reference: this kind of algorithm is relatively widespread.
-It seemed necessary to provide a baseline reference of this algorithm against which future improvements or alternative methods could be compared.
+  - This algorithm and its variants have several implementations.
+It seemed necessary to provide a baseline reference of this algorithm against which improvements or alternatives can be compared.
+To this end, the paper takes a generic approach, wheras other implementations and discussions of iterative monomorphisation often operate within a specific framework.
 
-  - Evaluation goes here TODO
-
-Kill, use as source for canibalization:
-
-  - Genericity, we made an effort to avoid imposing conditions on the underlying logic, whereas other implementations or discussions often do so within a specific (albeit often generalisable) framework.
-
-  - Possible applications, to the best of our knowledge, monomorphisation has not been used to enhance the performance of polymorphic provers.
-This is valuable as an indication of the benefits of re-evaluating the performance of native polymorphic provers.
+  - We conducted an evaluation of the applicability of this algorithm to extending monomorphic provers.
+Its possible application in increasing the perfomance of native polymorphic provers is to the best of our knowledge a novel one.
 
 Concerning comparison with other implementations, this would be an interesting research project, but we consider it beyond the scope of the present paper.
-The algorithms are quite cryptic and poorly documented, so it would take quite some engineering to get them running and to understand their inner workings.
-In fact, we attempted to understand one of these algorithms, the one by Sascha Böhme, and we gave up after several hours of study.
+These algorithms are often quite cryptic and can be poorly documented, so it would take quite some engineering to understand their inner workings.
+In fact, we attempted to understand Sascha Böhme's implementation, and gave up after several hours.
 
@@
 
-Concerning the question about page 2, Example 6 on page 4 shows that monomorphisation can be infinite.
+Concerning the question about page 2, Example 6 on page 4 shows that monomorphisation can run for infinitely many iterations.
 
@@
 Concerning the question "Why are the complexities useful?": The complexities are not formal bounds on the algorithm but give an idea of the explosiveness of the procedure.
-Any systematic enumerations of all monomorphic types will encounter the same issue.
+Any systematic enumerations of all monomorphic instances of non-monomorphic types will encounter the same issue.
 
-Concerning Figure 4, we do not understand the criticism. Tbe code is very short, and a few lines of it are for enforcing the bounds.
+Concerning Figure 4, we do not understand the criticism. The code is short, and several lines of it are for enforcing the bounds.
 
-Concerning "just an idea", an important potential application of iterative monomorphisation would be extending monomorphic provers.
+Concerning the idea on table 1 page 11, an important potential application of iterative monomorphisation would be extending monomorphic provers.
 Performance on problems that can already be solved by polymorphic provers is still relevant in this use case.
 
-Concerning the question about page 14, the previous version of Zipperposition made use of E as a backend, because it is a very efficient prover.
+Concerning the question about page 14, the previous version of Zipperposition made use of E as a backend because it is a very efficient prover.
 Zipperposition's strength was its implementation of novel algorithms.
-Delegating to E when the subproblems required optimised implementation of well-known procedures proved to be a successful.
+Delegating to E when the subproblems required optimised implementation of well-known procedures proved to be successful.
 This technique is used elsewhere in theorem proving, for instance, Leo-III can be run with E or CVC4 as backend.
@@
 Concerning overfitting, the goal was to limit any overfitting to the 500 problems and ensure that resulting bound values still performed well on the fresh 1043 remaining problems.
-We estimated this to be successfull because the ration of solved problems was similar in both cases.
+We estimated this to be successfull because the ratio of solved problems was similar in both cases.
 We will clarify the text.
@@
 
-Concerning novelty and Vampire's parser, see our answers to reviewer 1 above.
+Concerning novelty and Vampire's parser, see our answers to reviewer 1 above (respectively the first and last paragraph).
 
@@
 
-TODO
+S is the collection of all S(\phi), the same goes for N_old and N_new
 
@@
 
-TODO
+The v is an \upsilon and is a type not a type variable, the match would not be usefull if it were a type variable.
 
@@
 
-TODO
+This notation was used in an attempt to concisely enumerate all the assignments of the form:
+N_old(\phi)(f) = (u_1, ..., u_n) (similarly for N_new)
 
@@
 
-TODO
+They should have been defined at the function's initialisation. M_next and N_next(\phi) are local to the function.
 
@@
 
-TODO
+These limits refers to max_mono_args and max_nmon_args at the beginning of fig. 2.
 
@@
 
-TODO
+The algorithm can be split into the type argument tuple generation phase and the formula instantiation phase.
+In the first phase, we generate non-monomorphic type arguments, which can grow doubly-exponentially.
+This is what is discussed in the second paragraph.
+In the formula instantiation phase (generate_mono_formulae in the algorithm), we ensure that all type variables have been instantiated.
+We are not subject to the same explosion as in the previous phase.
 
@@
 
-TODO
+This is correct, the check should be performed before computing the match.

### 2025-04-17 rebuttal_answer.txt
mor

@@
-We first want to thank the reviewers for their comments and suggestions. We answered the questions and comments in separate part for each reviewer.
+We first want to thank the reviewers for their comments and suggestions. We answered the questions and comments separately for each reviewer.
 
@@
 It seemed necessary to provide a baseline reference of this algorithm against which improvements or alternatives can be compared.
-To this end, the paper takes a generic approach, wheras other implementations and discussions of iterative monomorphisation often operate within a specific framework.
+To this end, the paper takes a generic approach, whereas other implementations and discussions of iterative monomorphisation often operate within a specific framework.
 
   - We conducted an evaluation of the applicability of this algorithm to extending monomorphic provers.
-Its possible application in increasing the perfomance of native polymorphic provers is to the best of our knowledge a novel one.
+Its possible application in increasing the performance of native polymorphic provers is, we believe, novel.
 
@@
 These algorithms are often quite cryptic and can be poorly documented, so it would take quite some engineering to understand their inner workings.
-In fact, we attempted to understand Sascha Böhme's implementation, and gave up after several hours.
+Indeed, we attempted to understand Sascha Böhme's code and gave up after several hours.
 
-Concerning the logic, the type system required to apply the monomorphisation algorithm is explicitly defined in the preliminairies.
+Concerning the logic, the type system required to apply the monomorphisation algorithm is explicitly defined in the preliminaries.
 The underlying logic is left purposefully vague because we need very few hypotheses for the algorithm to be applicable.
@@
 
-Concerning the question "Why are the complexities useful?": The complexities are not formal bounds on the algorithm but give an idea of the explosiveness of the procedure.
+Concerning the question of the usefulness of complexities, they are not formal bounds on the algorithm but give an idea of the explosiveness of the procedure.
 Any systematic enumerations of all monomorphic instances of non-monomorphic types will encounter the same issue.
@@
 
-Concerning the idea on table 1 page 11, an important potential application of iterative monomorphisation would be extending monomorphic provers.
+Concerning the idea in Table 1 on page 11, an important potential application of iterative monomorphisation would be extending monomorphic provers.
 Performance on problems that can already be solved by polymorphic provers is still relevant in this use case.
@@
 Concerning the question about page 14, the previous version of Zipperposition made use of E as a backend because it is a very efficient prover.
-Zipperposition's strength was its implementation of novel algorithms.
+Zipperposition's strength was its implementation of higher-orderness.
 Delegating to E when the subproblems required optimised implementation of well-known procedures proved to be successful.
-This technique is used elsewhere in theorem proving, for instance, Leo-III can be run with E or CVC4 as backend.
-See Section 8 of Vukmirovic et al., "Making Higher-Order Superposition Work".
+This technique is used elsewhere in theorem proving, for instance, Leo-III can be run with E or CVC4 as a backend.
+See Section 8 of Vukmirovic et al., "Making Higher-Order Superposition Work" for details.
 We will briefly explain this in the next version of the text.
@@
 
-Miniscoping could indeed help on some formulas, although top-level conjunctions tend to arise seldom in practice.
-This could easily be achieved by a preprocessor and would require no changes to our algorithm.
+Miniscoping could indeed help on some formulas, although top-level conjunctions
+tend to arise seldom in practice. Miniscoping could easily be achieved by a
+preprocessor and would require no changes to our algorithm.
 
 Concerning overfitting, the goal was to limit any overfitting to the 500 problems and ensure that resulting bound values still performed well on the fresh 1043 remaining problems.
-We estimated this to be successfull because the ratio of solved problems was similar in both cases.
+We estimated this to be successful because the ratio of solved problems was similar in both cases.
 We will clarify the text.
 
-Concerning the E timeout, the "E timeout" is in seconds and dictates how long does the E prover run for.
+Concerning the E backend, the "E timeout" is in seconds and dictates how long does the E prover run for.
 The "E call point" is the fraction of the overall timeout of Zipperposition after which we call the E prover.
@@
 
-Concerning novelty and Vampire's parser, see our answers to reviewer 1 above (respectively the first and last paragraph).
+Concerning novelty and Vampire's parser, see our answers to reviewer 1 above (respectively the first and the last paragraph).
 
@@
 
-S is the collection of all S(\phi), the same goes for N_old and N_new
+S is the collection of all S(\phi). The same applies for N_old and N_new.
 
@@
 
-The v is an \upsilon and is a type not a type variable, the match would not be usefull if it were a type variable.
+The v is an \upsilon and is a type not a type variable, the match would not be useful if it were a type variable.
 
@@
 
-This notation was used in an attempt to concisely enumerate all the assignments of the form:
+This notation was used in an attempt to concisely enumerate all the assignments of the form
 N_old(\phi)(f) = (u_1, ..., u_n) (similarly for N_new)
@@
 
-These limits refers to max_mono_args and max_nmon_args at the beginning of fig. 2.
+These limits refer to max_mono_args and max_nmon_args at the beginning of Figure 2.
 
@@
 The algorithm can be split into the type argument tuple generation phase and the formula instantiation phase.
-In the first phase, we generate non-monomorphic type arguments, which can grow doubly-exponentially.
+In the first phase, we generate non-monomorphic type arguments, which can grow doubly exponentially.
 This is what is discussed in the second paragraph.
@@
 
-This is correct, the check should be performed before computing the match.
+This is correct. The check should be performed before computing the match.

### 2025-04-17 rebuttal_answer.txt
joinded lines for first paragraph of reviewer 2 answer
minor edits

@@
 
-Concerning the question about page 14, the previous version of Zipperposition made use of E as a backend because it is a very efficient prover.
+Concerning the question about page 14, the previous version of Zipperposition used E as a backend because it is a very efficient prover.
 Zipperposition's strength was its implementation of higher-orderness.
@@
 This technique is used elsewhere in theorem proving, for instance, Leo-III can be run with E or CVC4 as a backend.
-See Section 8 of Vukmirovic et al., "Making Higher-Order Superposition Work" for details.
+See Section 8 of Vukmirovic et al.
+"Making Higher-Order Superposition Work" for details.
 We will briefly explain this in the next version of the text.
@@
 
-Miniscoping could indeed help on some formulas, although top-level conjunctions
-tend to arise seldom in practice. Miniscoping could easily be achieved by a
-preprocessor and would require no changes to our algorithm.
+Miniscoping could indeed help on some formulas, although top-level conjunctions tend to arise seldom in practice.
+Miniscoping could easily be achieved by a preprocessor and would require no changes to our algorithm.
 
-Concerning overfitting, the goal was to limit any overfitting to the 500 problems and ensure that resulting bound values still performed well on the fresh 1043 remaining problems.
+Concerning overfitting, the goal was to limit any overfitting to the 500 problems and ensure that the resulting bound values still performed well on the fresh 1043 remaining problems.
 We estimated this to be successful because the ratio of solved problems was similar in both cases.
@@
 
-Concerning the E backend, the "E timeout" is in seconds and dictates how long does the E prover run for.
+Concerning the E backend, the "E timeout" is in seconds and dictates how long the E prover runs for.
 The "E call point" is the fraction of the overall timeout of Zipperposition after which we call the E prover.

### 2025-05-04 iterative_monomorphisation.tex
edited paper to take into account changes from reviewers, forgot to rename files

@@
 Any translation method based on instantiation will inevitably be incomplete.
-By a type version of the compactness theorem, there exists for any first order polymorphic formula \(\varphi\) an equisatisfiable finite set of monomorphic instances of \(\varphi\).
+By a type version of the compactness theorem, for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of \(\varphi\).
 However, this set cannot be computed~\cite[Theorem~1]{expr-poly-types}.
@@
 However, it is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
-A similar algorithm appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
+\texttt{SMT-LIB} makes use of this monomorphisation approach \cite{smt-lib-mono}.
+A similar algorithm also appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
 
@@
 The precise form of formulae is left unspecified.
-Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic.
+Due to this generality, iterative monomorphisation can be used with any standard variant of rank-1 polymorphic logic.
 In particular, it can work with the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}.
@@
 The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem.
-It operates by applying a fixed number of iterations, each of which takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae.
+It operates by applying a bounded number of iterations, each of which takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae.
 A single iteration consists of a collection phase and an instantiation phase.
@@
    \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set.
+This function is overloaded to accept a formula \(\varphi\) as input in which case it returns the set of all type variables which occur in the formula.
    \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, matches \(\upsilon\) against \(\tau\) and either fails or returns \emph{Some\((\sigma)\)}, where \(\sigma\) results from the matching.
@@
       \ForEach{\(\varphi \in \Problem\)} {
-         \((M_{\Delta}, N_{\Delta}, S_{\Delta})\leftarrow \FStep(M_{\text{old}}, N_{\text{old}}(\varphi), M_{\text{new}}, N_{\text{new}}(\varphi))\)\;
+         \((M_{\Delta}, N_{\Delta}, S_{\Delta})\leftarrow \FStep(\varphi, M_{\text{old}}, M_{\text{new}}, N_{\text{old}}(\varphi), N_{\text{new}}(\varphi))\)\;
          \(\AllSubst(\varphi)\leftarrow\AllSubst(\varphi)\cup S_{\Delta}\)\;
@@
 
-   \(\TypeVars \leftarrow \TyVars(\varphi)\)\;
+   %\(\TypeVars \leftarrow \TyVars(\varphi)\)\;
 
    \textcolor{ourblueviolet}{
-   \(\MaxMono \leftarrow \Max(\Min(\MonoFloor, |M_{\text{old}}\cup M_{\text{new}}| \cdot \MonoMult), \MonoCap)\)\;
-   \(\MaxPoly \leftarrow \Max(\Min(\PolyFloor, |N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)| \cdot \PolyMult), \PolyCap)\)\;
+   \(\MaxMono \leftarrow \Min(\Max(\MonoFloor, |M_{\text{old}}\cup M_{\text{new}}| \cdot \MonoMult), \MonoCap)\)\;
+   \(\MaxPoly \leftarrow \Min(\Max(\PolyFloor, |N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)| \cdot \PolyMult), \PolyCap)\)\;
    }
@@
 
-   \Let $S' = \SubstGen(M_{\text{new}}, N_{\text{new}}(\varphi))
+   $S' \leftarrow \SubstGen(M_{\text{new}}, N_{\text{new}}(\varphi))
    \cup \SubstGen(M_{\text{new}}, N_{\text{old}}(\varphi))
@@
 The total number of type argument tuples can increase cubically in the number of type argument tuples at each iteration and can therefore grow doubly exponentially in the number of iterations.
-We sketch an example of such growth, for a single formula.
+We give a sketch of how such growth can occur for a single formula.
 If we assume that after \(k\) iterations, there are \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type argument tuples, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
@@
 The resulting substitutions are composed if they are compatible.
-In the algorithm, compatibility is checked by making sure the \textbf{while} loop has successfully iterated over all elements of the type argument tuple.
+In the algorithm, compatibility is checked by making sure the \textbf{foreach} loop has successfully iterated over all elements of the type argument tuple.
 If any substitutions are incompatible, the matchings are discarded.
@@
 
-As a precaution against overfitting, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems.
+To detect if overfitting took place, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems.
 We carried out the rest of the evaluations on the remaining 1034 problems.
@@
 Table \ref{e_settings} shows the impact of the options with which we call the E prover.
-The call point corresponds to the fraction of the total timeout of Zipperposition's timeout after which the normal operation of Zipperposition is interrupted to begin the monomorphisation process.
-The E timeout limits how long E is run before being interrupted, at which point Zipperposition resumes normal operation.
+The point at which Zipperposition interupts its normal operation and begins the monomorphisation process is determined by the `E call point' parameter.
+It is expressed as a fraction of the total time alloted to Zipperposition.
+The E timeout (in seconds) limits how long E is run before being interrupted, at which point Zipperposition resumes normal operation.
 The longer Zipperposition runs before E is invoked, the more formulae are generated, and the more combinatorially explosive iterative monomorphisation is.
@@
 We compare the performance of two instances of Zipperposition.
+By default, Zipperposition may call E as a backend when given a monomorphic problem~\cite{vukmirovic-et-al-2021}.
 The first instance is run in the portfolio mode \verb|portfolio.sequential.py|, which attempts to prove the problem with various configurations tried in succession.
-Of these configurations, some may attempt to call E as a backend when the given problem is monomorphic, but because all given problems are non-monomorphic, this instance is labeled `without E'.
+Since all given problems are non-monomorphic, these configurations can never invoke E as a backend, this instance is therefore labeled `without E'.
 The second instance is a modification of the \verb|portfolio.sequential.py| file where each configuration is modified analogously to \verb|40_b.comb| with the options obtained in the previous evaluation phase.
@@
 Each of the two instances is evaluated on the set of the 500 previously used problems, and the set of the remaining 1034 problems is evaluated separately.
-The results of these evaluations do not suggest that any overfitting took place during the option optimisation phase.
+The proportion of problems successfully solved on the 500 and 1034 problem suites are similar, suggesting that no overfitting took place during the option optimisation phase.
 
@@
 We chose Leo-III and Zipperposition.
-Unfortunately we needed to exclude Vampire because of parsing issues.
+Unfortunately we needed to exclude Vampire because of parsing issues\footnote{Specifically, Vampire failed to parse a significant number of the benchmarks problems.
+The fix for this issue was not available for the higher-order branch of Vampire at the time of the evaluation.}.
 

### 2025-05-06 iterative_monomorphisation.tex
added a couple of fixes in line with the reviews

@@
 \SetKwData{Loop}{\textcolor{ourblueviolet}{num\_loops}}
-\SetKwData{TypeVars}{type\_variables}
+\SetKwData{TypeVars}{type\_vars}
 \SetKwData{UsedSubst}{\(S\)}
@@
 \SetKwFunction{Max}{max}\SetKwFunction{Min}{min}
-\SetKwFunction{Domain}{subst\_domain}\SetKwFunction{Len}{len}
-\SetKwFunction{SubstGen}{matches}\SetKwFunction{TyVars}{type\_variables}
+\SetKwFunction{Domain}{subst\_dom}\SetKwFunction{Len}{len}
+\SetKwFunction{SubstGen}{matches}\SetKwFunction{TyVars}{type\_vars}
 \SetKwFunction{FMonoStep}{formula\_mono\_step}
@@
 
-\definecolor{ourblueviolet}{HTML}{0071BC}
+\definecolor{ourblueviolet}{rgb}{0.1, 0.5, 0.9}
 \definecolor{grey}{rgb}{0.8,0.8,0.8}
@@
    \item[\labelitemi] \emph{initialisation\((\Phi)\)}, where \(\Phi\) is a set of (polymorphic) formulae, extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
-   \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set.
+   \item[\labelitemi] \emph{type\_vars\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set.
 This function is overloaded to accept a formula \(\varphi\) as input in which case it returns the set of all type variables which occur in the formula.
@@
    \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step} 
-   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{mono\_substs}
+   \SetKwFunction{TyVars}{type\_vars}\SetKwFunction{MonoSubst}{mono\_substs}
    \SetKwFunction{GenFormulae}{generate\_mono\_formulae}
@@
 
-   \BlankLine
-
    $S' \leftarrow \SubstGen(M_{\text{new}}, N_{\text{new}}(\varphi))
@@
    \cup \SubstGen(M_{\text{old}}, N_{\text{new}}(\varphi))$\;
+   \((M_\text{next}, N_\text{next}(\varphi)) \leftarrow (\emptyset, \emptyset)\)\;
    \ForEach{\(\sigma\in S'\)}{
-   \ForEach{\((f\mapsto (\upsilon_1, \dots,\upsilon_n))\in N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)\)}{
+   \ForEach{\(f \in \text{dom}(N_\text{old}(\varphi)\cup N_\text{new}(\varphi))\)}{
+   \ForEach{\((\upsilon_1, \dots,\upsilon_n)\in N_{\text{old}}(\varphi)(f)\cup N_{\text{new}}(\varphi)(f)\)}{
             \eIf{\(\TyVars(\upsilon_1, \dots, \upsilon_n) \subseteq \Domain(\sigma)\)}{
@@
    }
+   }
 
@@
    \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step} 
-   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{mono\_substs}
+   \SetKwFunction{TyVars}{type\_vars}\SetKwFunction{MonoSubst}{mono\_substs}
    \SetKwFunction{GenFormulae}{generate\_formulae}
@@
 We chose Leo-III and Zipperposition.
-Unfortunately we needed to exclude Vampire because of parsing issues\footnote{Specifically, Vampire failed to parse a significant number of the benchmarks problems.
-The fix for this issue was not available for the higher-order branch of Vampire at the time of the evaluation.}.
+Unfortunately we needed to exclude Vampire because of parsing issues\footnote{Vampire failed to parse a significant number of the benchmarks problems.
+The fix was unavailable for the higher-order branch of Vampire at the time of the evaluation.}.
 

### 2025-05-11 iterative_monomorphisation.tex
added appendix example

@@
 At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
+No new formulae are generated at this stage, only new type arguments and substitutions.
 Once these iterations are completed, the first phase is complete and the substitutions used to create new type argument tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase.
+This is when the new formulae are generated.
 The third phase mangles the composite types of the newly monomorphised formulae.
@@
    \ForEach{\(\sigma\in S'\)}{
-   \ForEach{\(f \in \text{dom}(N_\text{old}(\varphi)\cup N_\text{new}(\varphi))\)}{
-   \ForEach{\((\upsilon_1, \dots,\upsilon_n)\in N_{\text{old}}(\varphi)(f)\cup N_{\text{new}}(\varphi)(f)\)}{
+   %\ForEach{\(f \in \text{dom}(N_\text{old}(\varphi)\cup N_\text{new}(\varphi))\)}{
+   \ForEach{\((f \mapsto (\upsilon_1, \dots,\upsilon_n))\in N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)\)}{
             \eIf{\(\TyVars(\upsilon_1, \dots, \upsilon_n) \subseteq \Domain(\sigma)\)}{
@@
       }
-   }
+   %}
    }
@@
       \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
+               \textcolor{ourblueviolet}{
+                  \If{\(|S| > \SubstLimit\)}
+                     {\Return \(S\)}}
          \If{for all \(0 \leq i\leq n\), \(\Match(\upsilon_i, \tau_i) = \Success(\sigma_i)\)}
             {
+
                \If {\(\sigma_1,\dots,\sigma_n\) are compatible}{
@@
 
-               \textcolor{ourblueviolet}{
-                  \eIf{\(|S| < \SubstLimit\)}{
                      \textcolor{black}{\(S\leftarrow S\cup \{\sigma\}\)\;}
-                  }
-                  {\Return \(S\)}
-            }
             }
@@
 
+\section{Appendix}
+
+\begin{examplex}
+This example runs an example through the detailed version of the iterative algorithm works. Consider the following initial problem:
+\begin{itemize}
+   \item[\(\langle1\rangle\)] \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
+   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{f}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{f}\langle \alpha, \alpha \rangle(x, x) \land \sym{f}\langle \alpha, \beta\rangle(x, z)\)
+\end{itemize}
+%
+\(M_{\text{new}}\) is initialised with \(\{f\mapsto (\ty{int}, \ty{nat}); f\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{f\mapsto (\ty{list}(\alpha), \alpha); f\mapsto (\alpha, \alpha); f\mapsto (\alpha, \beta)\}\}\).
+Then we enter the main loop (assuming \Loop is at least 1).
+We iterate over each formula and call \FStep for each of them.
+Nothing happens for formula \(\langle1\rangle\) because it contains no type variables.
+For formula \(\langle2\rangle\), \(M_{\text{new}}\) and \(N_{\text{new}}(\langle2\rangle)\) are passed as argument to \SubstGen.
+
+The three non-monomorphic type argument tuples are matched against their monomorphic counterparts in \(M_{\text{new}}\):
+\begin{itemize}
+   \item \((\ty{list}(\alpha), \alpha)\) fails in both cases because \(\ty{list}(\alpha)\) fails to match against \ty{int}.
+   \item \((\alpha, \alpha)\) fails to match against \((\ty{int}, \ty{nat})\) because the substitutions resulting from the match of the first and second element of the tuple are incompatible.
+The second match is successful and yields the substitution \(\sigma_1 = \{\alpha \mapsto \ty{int}\}\)
+   \item \((\alpha, \beta)\) succeeds in both cases and generates the substitutions \(\sigma_2 = \{\alpha \mapsto \ty{int}, \beta \mapsto \ty{int}\}\) and \(\sigma_3 = \{\alpha \mapsto \ty{int}, \beta \mapsto \ty{nat}\}\).
+\end{itemize}
+
+\SubstGen returns and the substitutions are used to generate new type argument tuples.
+Each substitution is applied to the type arguments of the function symbols of formula \(\langle2\rangle\) because they are the only function symbols with non-monomorphic type arguments.
+\begin{table}
+\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
+   \toprule
+   & \(\sigma_1\) &\(\sigma_2\)&\(\sigma_3\)\\
+   \cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(l){4-4} 
+   \((\ty{list}(\alpha), \alpha)\)&\((\ty{list}(\ty{int}), \ty{int})\)& \((\ty{list}(\ty{int}), \ty{int})\) & \((\ty{list}(\ty{int}), \ty{int})\) \\
+   \((\alpha, \alpha)\) &\((\ty{int}, \ty{int})\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{int})\)\\
+   \((\alpha, \beta)\)  &\((\ty{int}, \beta)\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{nat})\)\\
+    \bottomrule
+\end{tabular}
+\end{table}
+\vspace{-0.5em}
+
+With a very small and simple initial problem, there are already up to nine new type arguments tuples generated at this step in the first iteration alone, though only four are unique.
+Once the new type argument tuples are added to their respective maps, a new iteration is begun.
+We only consider one iteration and continue to the next phase of the algorithm.
+
+\GenFormulae is the next function to be called.
+This function iterates over all non-monomorphic formulae of the initial problem, in our case this will only be \(\langle2\rangle\).
+The set of type variable tuples of \(\langle2\rangle\) is passed to \MonoSubst along with the set of all previously generated substitutions.
+
+First, we instantiate \(\alpha\), the first type variable of \(\langle2\rangle\).
+If \(\sigma_1\) is selected to instantiate \(\alpha\), both \(\sigma_2\) and \(\sigma_3\) will in turn be selected to instantiate \(\beta\).
+This will generate two monomorphising substitutions: \(\sigma_1 \circ \sigma_2 = \sigma_2\) and \(\sigma_1 \circ \sigma_3 = \sigma_3\) which are added to the set of monomorphising substitutions \(S_{\text{res}}\).
+Now \(\sigma_2\) is selected, it instantiates both \(\alpha\) and \(\beta\) at once.
+In our case, \(\sigma_2\) is already in the set of monomorphising substitutions and is therefore ignored.
+\(\sigma_3\) is treated similarly.
+No more than \SubstLimit monomorphising substitutions can be generated this way.
+
+Our monomorphising substitutions have been generated.
+We only need to apply them to the formula they monomorphises.
+This is repeated for all non-monomorphic formulae or until \Limit is reached.
+The ordering of formulae will impact the output problem in the latter case.
+For some larger input problems, \MonoSubst could be sufficiently explosive to only allow a handful of different formulae to be monomorphised.
+The ability to determine \SubstLimit independently can help avoid this issue.
+
+Finally, all monomorphic formulae, both new and old, are mangled.
+This step is unecessary if the target language allows type constructors.
+In our example two new monomorphic formulae are generated.
+The algorithm outputs:
+\vspace{-0.5em}
+\begin{enumerate}
+   \item \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
+   \item[3.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(x, z)\)
+   \item[4.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{nat}\rangle(x, z)\)
+\end{enumerate}
+
+\end{examplex}
+
 \bibliographystyle{splncs04}

### 2025-05-12 iterative_monomorphisation.tex
move appendix etc.
mroe

@@
 For example, the \sym{concat} function of Isabelle/HOL, which is characterised in the base case by the equation \(\sym{concat}\langle\alpha\rangle\; \sym{Nil}\langle\ty{list}(\alpha)\rangle = \sym{Nil}\langle\alpha\rangle\), exhibits the same behavior.
-/Any reasonable implementation must limit the number of new type arguments, substitutions and formulae.
+Any reasonable implementation must limit the number of new type arguments, substitutions and formulae.
+A more elaborate example is provided in Appendix~\ref{app}.%
+\footnote{The appendix example will be included here in the camera-ready version if
+the program chairs give use an extra page or two, as is customary.}
 
@@
 
-\section{Appendix}
-
-\begin{examplex}
+\bibliographystyle{splncs04}
+\bibliography{citations}
+
+\appendix
+
+\section{Detailed Example}
+\label{app}
+
+\begin{examplex}\rm
 This example runs an example through the detailed version of the iterative algorithm works. Consider the following initial problem:
@@
 
-\bibliographystyle{splncs04}
-\bibliography{citations}
-
 \end{document}

### 2025-05-12 iterative_monomorphisation.tex
shorten and thank

@@
 \def\vthinspace{\kern+0.083333em}
+
+\hyphenation{mono-mor-phisa-tion}
+\hyphenation{mono-mor-phic}
+\hyphenation{non-mono-mor-phic}
 
@@
 Monomorphisation can be used to extend monomorphic provers to support polymorphic logics.
-We propose an iterative approach that is incomplete but pragmatic.
-We implemented it in the Zipperposition prover, where it can be used to translate away polymorphism before invoking the monomorphic prover E as a backend.
-Our evaluation demonstrates that this approach increases Zipperposition's success rate.
+We propose a pragmatic iterative approach.
+We implemented it in the Zipperposition prover, where it is used to translate away polymorphism before invoking the monomorphic prover E as a backend.
+Our evaluation shows that this approach increases Zipperposition's success rate.
 Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.
@@
 In this paper, we present an algorithm based on our understanding of Böhme's description and implementation (Section~\ref{sec:high level-algorithm}).
-We also provide a more detailed description to help future implementers.
-In addition, our description shows some of the ways in which an implementation can be made more efficient and avoid combinatorial explosions (Section~\ref{sec:low-level-algorithm}).
+We also provide a more detailed\pagebreak[2] \hbox{description} to help future implementers.
+In addition, our paper shows some of the ways in which an implementation can be made more efficient and avoid combinatorial explosions (Section~\ref{sec:low-level-algorithm}).
 
@@
 %\end{enumerate}
+
+\kern-1pt %%% TYPESETTING: Hack
 
@@
 
+\pagebreak[2]
+
 Our abstract framework relies on the following basic definitions.
@@
 The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem.
-It operates by applying a bounded number of iterations, each of which takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae.
+It works by applying a bounded number of iterations, each taking a polymorphic problem as argument and returning a problem with new partially instantiated formulae.
 A single iteration consists of a collection phase and an instantiation phase.
@@
    \item Create an empty set of formulae \(\Phi'\).
+
+\pagebreak[2]
 
@@
 
-The algorithm presented above is too naïve for a practical implementation. In this section, we present a lower level algorithm with the following features.
+\looseness=-1
+The algorithm presented above is too naïve in practice. In this section, we present a lower level algorithm with the following features.
 First, numeric bounds are introduced to stop combinatorially explosive enumerations.
@@
 We use the colour blue to identify bounds and code related to bounds.
-At the end of each of these iterations, the old and new type argument maps are updated with newly generated types.
+At the end of each of these iterations,\pagebreak[2]
+the old and new type argument maps are updated with newly generated types.
 No new formulae are generated at this stage, only new type arguments and substitutions.
@@
 %Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.
-New tuples are obtained by applying each substitution to each non-monomorphic type argument tuple such that at least one tuple component is instantiated by the substitution.
+New tuples are obtained by applying
+each substitution to each non-monomorphic type argument tuple such that at least one tuple component is instantiated by the substitution.
 
@@
 If several groups result in the same number of proved problems, we select the most constraining values.
-Winning entries are shown in bold in Tables \ref{mono_ty_args} to \ref{e_settings}.
+Winning entries are shown in bold in Tables \ref{mono_ty_args}~to~\ref{e_settings}.
 
@@
 The bound values are lower because we found non-monomorphic type arguments to be combinatorially explosive in preliminary evaluations.
-The table confirms that non-monomorphic type argument generation drives the creation of useful non-monomorphic formulae.
+The table confirms that non-monomorphic type argument generation drives the creation of useful non-mono\-morphic formulae.
 This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed.
@@
    We thank Sascha Böhme for fruitful discussions.
-   We thank Jannis Limperg and Mark Summerfield for suggesting textual improvements.
+   We thank Jannis Limperg, Mark Summerfield, and three anonymous reviewers for suggesting textual improvements.
 

### 2025-05-15 iterative_monomorphisation.tex
minor

@@
 By contrast, many automatic provers operate only on monomorphic logics.
-One approach to closing this gap is to extend provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}.
+One way to closing this gap is to extend provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}.
 However, this entails a lot of work that must be redone for every prover.
@@
 Any translation method based on instantiation will inevitably be incomplete.
-By a type version of the compactness theorem, for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of \(\varphi\).
-However, this set cannot be computed~\cite[Theorem~1]{expr-poly-types}.
+By a type version of the compactness theorem, for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of~\(\varphi\).
+However, this set cannot be computed in general~\cite[Theorem~1]{expr-poly-types}.
 
@@
 However, it is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
-\texttt{SMT-LIB} makes use of this monomorphisation approach \cite{smt-lib-mono}.
-A similar algorithm also appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
+\relax{SMT-LIB} makes use of this monomorphisation approach \cite{smt-lib-mono}.
+A similar algorithm also appears to be implemented in the \relax{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
 

### 2025-06-25 iterative_monomorphisation.tex
fix
more reviewers

@@
 Unfortunately we needed to exclude Vampire because of parsing issues\footnote{Vampire failed to parse a significant number of the benchmarks problems.
-The fix was unavailable for the higher-order branch of Vampire at the time of the evaluation.}.
+The fix was unavailable for the higher order branch of Vampire at the time of the evaluation.}.
 
@@
    We thank Sascha Böhme for fruitful discussions.
-   We thank Jannis Limperg, Mark Summerfield, and three anonymous reviewers for suggesting textual improvements.
+   We thank Jannis Limperg, Mark Summerfield, and the anonymous reviewers for suggesting textual improvements.
 

### 2025-06-25 iterative_monomorphisation.tex
shorten

@@
 Automatic theorem provers provide automation for proof assistant users.
-Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle\slash HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae.
-By contrast, many automatic provers operate only on monomorphic logics.
-One way to closing this gap is to extend provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}.
+Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle\slash HOL \cite{nipkow-et-al-2002}, support rank~1 polymorphism, where type quantification is allowed at the top level of formulae.
+By contrast, many automatic provers only work with monomorphic logics.
+One way to close this gap is to extend provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}.
 However, this entails a lot of work that must be redone for every prover.
-
+%
 The alternative is to translate polymorphic problems to monomorphic problems.
-One approach \cite[Section~9]{mono-trans} is to encode polymorphism using dedicated function or predicate symbols in a monomorphic logic.
+One approach \cite{mono-trans} is to encode a polymorphic type system using dedicated function or predicate symbols in a monomorphic logic.
 Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}.
-This method relies on heuristically instantiating the formulae's type variables with concrete types.
+This method heuristically instantiates the formulae's type variables with concrete types.
 
@@
 We also provide a more detailed\pagebreak[2] \hbox{description} to help future implementers.
-In addition, our paper shows some of the ways in which an implementation can be made more efficient and avoid combinatorial explosions (Section~\ref{sec:low-level-algorithm}).
-
-Briefly, the algorithm works as follows.
+In addition, we present some optimisations
+to curb the combinatorial explosions (Section~\ref{sec:low-level-algorithm}).
+% some of the ways in which an implementation can be made more efficient and avoid combinatorial explosions (Section~\ref{sec:low-level-algorithm}).
+
+The algorithm works as follows.
 We assume problems to be sets of formulae.
@@
 The process is then iterated, making use of the newly generated instances.
-
+%
 Consider the unary type constructor \ty{list}.
@@
 Thanks to our work, E can now be used with polymorphic problems.
-Moreover, our implementation can be used as a preprocessor to interface with other stand-alone provers.
+Moreover, our implementation can be used as a preprocessor for other stand-alone provers.
 Our source code is available online.%
 \footnote{\url{https://github.com/nartannt/iterative_monomorphisation}}
-
-Our empirical evaluation on TPTP~\cite{tptp} problems attempts to answer three questions (Section~\ref{sec:evaluation}):
+%
+Our evaluation on the TPTP~\cite{tptp} tries to answer three questions (Section~\ref{sec:evaluation}):
 \begin{enumerate}
@@
 \end{enumerate}
-%
-Our raw evaluation data is available online.%
-\footnote{\url{https://zenodo.org/records/14881532}}
 
@@
 The precise form of formulae is left unspecified.
-Due to this generality, iterative monomorphisation can be used with any standard variant of rank-1 polymorphic logic.
+Due to this generality, iterative monomorphisation can be used with any standard variant of rank~1 polymorphic logic.
 In particular, it can work with the polymorphic first and higher order logics embodied by TPTP's TF1 and TH1 syntaxes~\cite{blanchette-paskevich-2013,th1}.
 % implemented by several automatic provers.
-
-\pagebreak[2]
 
@@
 the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\), \(\ty{map}(\ty{int},\ty{string})\)).
-If $n = 0$, we omit the parentheses (e.g.\ \(\ty{int}\)).
+If
+$n = 0$, we omit the parentheses (e.g.\ \(\ty{int}\)).
 A type is \emph{monomorphic} if it contains no type variables.
@@
 Any reasonable implementation must limit the number of new type arguments, substitutions and formulae.
-A more elaborate example is provided in Appendix~\ref{app}.%
-\footnote{The appendix example will be included here in the camera-ready version if
-the program chairs give use an extra page or two, as is customary.}
 
@@
 
+\section{Detailed example}
+
+We run an example through the detailed version of the iterative algorithm. Consider this initial problem:
+\begin{itemize}
+   \item[\(\langle1\rangle\)] \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
+   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{f}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{f}\langle \alpha, \alpha \rangle(x, x) \land \sym{f}\langle \alpha, \beta\rangle(x, z)\)
+\end{itemize}
+%
+\(M_{\text{new}}\) is initialised with \(\{f\mapsto (\ty{int}, \ty{nat}); f\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{f\mapsto (\ty{list}(\alpha), \alpha); f\mapsto (\alpha, \alpha); f\mapsto (\alpha, \beta)\}\}\).
+Then we enter the main loop (assuming \Loop is at least 1).
+We iterate over each formula and call \FStep for each of them.
+Nothing happens for formula \(\langle1\rangle\) because it contains no type variables.
+For formula \(\langle2\rangle\), \(M_{\text{new}}\) and \(N_{\text{new}}(\langle2\rangle)\) are passed as argument to \SubstGen.
+
+The three non-monomorphic type argument tuples are matched against their monomorphic counterparts in \(M_{\text{new}}\):
+\begin{itemize}
+   \item \((\ty{list}(\alpha), \alpha)\) fails in both cases because \(\ty{list}(\alpha)\) fails to match against \ty{int}.
+   \item \((\alpha, \alpha)\) fails to match against \((\ty{int}, \ty{nat})\) because the substitutions resulting from the match of the first and second element of the tuple are incompatible.
+The second match is successful and yields the substitution \(\sigma_1 = \{\alpha \mapsto \ty{int}\}\)
+   \item \((\alpha, \beta)\) succeeds in both cases and generates the substitutions \(\sigma_2 = \{\alpha \mapsto \ty{int}, \beta \mapsto \ty{int}\}\) and \(\sigma_3 = \{\alpha \mapsto \ty{int}, \beta \mapsto \ty{nat}\}\).
+\end{itemize}
+
+\SubstGen returns and the substitutions are used to generate new type argument tuples.
+Each substitution is applied to the type arguments of the function symbols of formula \(\langle2\rangle\) because they are the only function symbols with non-monomorphic type arguments.
+\begin{table}
+\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
+   \toprule
+   & \(\sigma_1\) &\(\sigma_2\)&\(\sigma_3\)\\
+   \cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(l){4-4} 
+   \((\ty{list}(\alpha), \alpha)\)&\((\ty{list}(\ty{int}), \ty{int})\)& \((\ty{list}(\ty{int}), \ty{int})\) & \((\ty{list}(\ty{int}), \ty{int})\) \\
+   \((\alpha, \alpha)\) &\((\ty{int}, \ty{int})\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{int})\)\\
+   \((\alpha, \beta)\)  &\((\ty{int}, \beta)\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{nat})\)\\
+    \bottomrule
+\end{tabular}
+\end{table}
+
+With a very small and simple initial problem, there are already up to nine new type arguments tuples generated at this step in the first iteration alone, though only four are unique.
+Once the new type argument tuples are added to their respective maps, a new iteration is begun.
+We only consider one iteration and continue to the next phase of the algorithm.
+
+\GenFormulae is the next function to be called.
+This function iterates over all non-monomorphic formulae of the initial problem, in our case this will only be \(\langle2\rangle\).
+The set of type variable tuples of \(\langle2\rangle\) is passed to \MonoSubst along with the set of all previously generated substitutions.
+
+First, we instantiate \(\alpha\), the first type variable of \(\langle2\rangle\).
+If \(\sigma_1\) is selected to instantiate \(\alpha\), both \(\sigma_2\) and \(\sigma_3\) will in turn be selected to instantiate \(\beta\).
+This will generate two monomorphising substitutions: \(\sigma_1 \circ \sigma_2 = \sigma_2\) and \(\sigma_1 \circ \sigma_3 = \sigma_3\) which are added to the set of monomorphising substitutions \(S_{\text{res}}\).
+Now \(\sigma_2\) is selected, it instantiates both \(\alpha\) and \(\beta\) at once.
+In our case, \(\sigma_2\) is already in the set of monomorphising substitutions and is therefore ignored.
+\(\sigma_3\) is treated similarly.
+No more than \SubstLimit monomorphising substitutions can be generated in this way.
+
+Our monomorphising substitutions have been generated.
+We only need to apply them to the formula they monomorphises.
+This is repeated for all non-monomorphic formulae or until \Limit is reached.
+The ordering of formulae will impact the output problem in the latter case.
+For some larger input problems, \MonoSubst could be sufficiently explosive to only allow a handful of different formulae to be monomorphised.
+The ability to determine \SubstLimit independently can help avoid this issue.
+
+Finally, all monomorphic formulae, both new and old, are mangled.
+This step is unecessary if the target language allows type constructors.
+In our example two new monomorphic formulae are generated.
+The algorithm outputs:
+\begin{enumerate}
+   \item \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
+   \item[3.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(x, z)\)
+   \item[4.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{nat}\rangle(x, z)\)
+\end{enumerate}
+
 % target page number: 2-4
@@
 The benchmarks are taken from version 8.2.0 of the TPTP library~\cite{tptp}.
-The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first order and higher order logic with rank-1 polymorphism.
+The library contains 1765 problems in TF1 and TH1, corresponding respectively to first and higher order logic with rank~1 polymorphism.
 Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them.
@@
 per problem with a single thread.
+Our raw evaluation data is available online.%
+\footnote{\url{https://zenodo.org/records/14881532}}
 
@@
 
-\subsection{Parameter optimisation}
-\label{param_opti}
+\paragraph{\upshape\bfseries Parameter optimisation.}
 
@@
 
-\subsection{E as a Zipperposition backend}
+\paragraph{\upshape\bfseries E as a Zipperposition backend.}
 
@@
 
-\subsection{Monomorphisation as a preprocessor}
+\paragraph{\upshape\bfseries Monomorphisation as a preprocessor.}
 
@@
    We thank Sascha Böhme for fruitful discussions.
-   We thank Jannis Limperg, Mark Summerfield, and the anonymous reviewers for suggesting textual improvements.
-
+   We thank Jannis Limperg, Mark Summerfield, and the peer reviewers for suggesting textual improvements.
    This research is co-funded by the European Union (ERC, Nekoka, 101083038).
@@
 
-\appendix
-
-\section{Detailed Example}
-\label{app}
-
-\begin{examplex}\rm
-This example runs an example through the detailed version of the iterative algorithm works. Consider the following initial problem:
-\begin{itemize}
-   \item[\(\langle1\rangle\)] \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
-   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{f}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{f}\langle \alpha, \alpha \rangle(x, x) \land \sym{f}\langle \alpha, \beta\rangle(x, z)\)
-\end{itemize}
-%
-\(M_{\text{new}}\) is initialised with \(\{f\mapsto (\ty{int}, \ty{nat}); f\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{f\mapsto (\ty{list}(\alpha), \alpha); f\mapsto (\alpha, \alpha); f\mapsto (\alpha, \beta)\}\}\).
-Then we enter the main loop (assuming \Loop is at least 1).
-We iterate over each formula and call \FStep for each of them.
-Nothing happens for formula \(\langle1\rangle\) because it contains no type variables.
-For formula \(\langle2\rangle\), \(M_{\text{new}}\) and \(N_{\text{new}}(\langle2\rangle)\) are passed as argument to \SubstGen.
-
-The three non-monomorphic type argument tuples are matched against their monomorphic counterparts in \(M_{\text{new}}\):
-\begin{itemize}
-   \item \((\ty{list}(\alpha), \alpha)\) fails in both cases because \(\ty{list}(\alpha)\) fails to match against \ty{int}.
-   \item \((\alpha, \alpha)\) fails to match against \((\ty{int}, \ty{nat})\) because the substitutions resulting from the match of the first and second element of the tuple are incompatible.
-The second match is successful and yields the substitution \(\sigma_1 = \{\alpha \mapsto \ty{int}\}\)
-   \item \((\alpha, \beta)\) succeeds in both cases and generates the substitutions \(\sigma_2 = \{\alpha \mapsto \ty{int}, \beta \mapsto \ty{int}\}\) and \(\sigma_3 = \{\alpha \mapsto \ty{int}, \beta \mapsto \ty{nat}\}\).
-\end{itemize}
-
-\SubstGen returns and the substitutions are used to generate new type argument tuples.
-Each substitution is applied to the type arguments of the function symbols of formula \(\langle2\rangle\) because they are the only function symbols with non-monomorphic type arguments.
-\begin{table}
-\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
-   \toprule
-   & \(\sigma_1\) &\(\sigma_2\)&\(\sigma_3\)\\
-   \cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(l){4-4} 
-   \((\ty{list}(\alpha), \alpha)\)&\((\ty{list}(\ty{int}), \ty{int})\)& \((\ty{list}(\ty{int}), \ty{int})\) & \((\ty{list}(\ty{int}), \ty{int})\) \\
-   \((\alpha, \alpha)\) &\((\ty{int}, \ty{int})\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{int})\)\\
-   \((\alpha, \beta)\)  &\((\ty{int}, \beta)\)&\((\ty{int}, \ty{int})\) &\((\ty{int}, \ty{nat})\)\\
-    \bottomrule
-\end{tabular}
-\end{table}
-\vspace{-0.5em}
-
-With a very small and simple initial problem, there are already up to nine new type arguments tuples generated at this step in the first iteration alone, though only four are unique.
-Once the new type argument tuples are added to their respective maps, a new iteration is begun.
-We only consider one iteration and continue to the next phase of the algorithm.
-
-\GenFormulae is the next function to be called.
-This function iterates over all non-monomorphic formulae of the initial problem, in our case this will only be \(\langle2\rangle\).
-The set of type variable tuples of \(\langle2\rangle\) is passed to \MonoSubst along with the set of all previously generated substitutions.
-
-First, we instantiate \(\alpha\), the first type variable of \(\langle2\rangle\).
-If \(\sigma_1\) is selected to instantiate \(\alpha\), both \(\sigma_2\) and \(\sigma_3\) will in turn be selected to instantiate \(\beta\).
-This will generate two monomorphising substitutions: \(\sigma_1 \circ \sigma_2 = \sigma_2\) and \(\sigma_1 \circ \sigma_3 = \sigma_3\) which are added to the set of monomorphising substitutions \(S_{\text{res}}\).
-Now \(\sigma_2\) is selected, it instantiates both \(\alpha\) and \(\beta\) at once.
-In our case, \(\sigma_2\) is already in the set of monomorphising substitutions and is therefore ignored.
-\(\sigma_3\) is treated similarly.
-No more than \SubstLimit monomorphising substitutions can be generated this way.
-
-Our monomorphising substitutions have been generated.
-We only need to apply them to the formula they monomorphises.
-This is repeated for all non-monomorphic formulae or until \Limit is reached.
-The ordering of formulae will impact the output problem in the latter case.
-For some larger input problems, \MonoSubst could be sufficiently explosive to only allow a handful of different formulae to be monomorphised.
-The ability to determine \SubstLimit independently can help avoid this issue.
-
-Finally, all monomorphic formulae, both new and old, are mangled.
-This step is unecessary if the target language allows type constructors.
-In our example two new monomorphic formulae are generated.
-The algorithm outputs:
-\vspace{-0.5em}
-\begin{enumerate}
-   \item \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
-   \item[3.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(x, z)\)
-   \item[4.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{nat}\rangle(x, z)\)
-\end{enumerate}
-
-\end{examplex}
-
 \end{document}

### 2025-06-25 iterative_monomorphisation.tex
shorten

@@
 This will generate two monomorphising substitutions: \(\sigma_1 \circ \sigma_2 = \sigma_2\) and \(\sigma_1 \circ \sigma_3 = \sigma_3\) which are added to the set of monomorphising substitutions \(S_{\text{res}}\).
-Now \(\sigma_2\) is selected, it instantiates both \(\alpha\) and \(\beta\) at once.
-In our case, \(\sigma_2\) is already in the set of monomorphising substitutions and is therefore ignored.
-\(\sigma_3\) is treated similarly.
+Now \(\sigma_2\) is selected.
+It instantiates both \(\alpha\) and \(\beta\) at once.
+Here, \(\sigma_2\) is already in \(S_{\text{res}}\) and is therefore ignored, and \(\sigma_3\) is treated similarly.
 No more than \SubstLimit monomorphising substitutions can be generated in this way.
@@
 We chose as a measure of success for a given prover (or prover configuration) the number of problems
-that could be solved by the prover in at most 30 seconds
+that could be solved by the prover in at most 30~s
 %(including monomorphisation when relevant)
@@
 Performance of the monomorphisation algorithm seems to plateau for some ranges of values and drops off beyond.
-%TODO check that 125 is what we get if we just run Zipperposition for 10 seconds (maybe with all additional problems that are easily monomorphised)
+%TODO check that 125 is what we get if we just run Zipperposition for 10~s (maybe with all additional problems that are easily monomorphised)
 
@@
 We chose Leo-III and Zipperposition.
-Unfortunately we needed to exclude Vampire because of parsing issues\footnote{Vampire failed to parse a significant number of the benchmarks problems.
-The fix was unavailable for the higher order branch of Vampire at the time of the evaluation.}.
+Unfortunately we needed to exclude Vampire because of parsing issues.
+%Vampire failed to parse a significant number of the benchmarks problems.
+The fix was unavailable for Vampire's higher order branch at the time of the evaluation.
 
@@
 The monomorphisation approach is evaluated in two steps.
-First, each problem is monomorphised using the options obtained from the first evaluation phase except for the monomorphisation timeout option, which is increased to 30~seconds.
+First, each problem is monomorphised using the options obtained from the first evaluation phase except for the monomorphisation timeout option, which is increased to 30~s.
 For 149~problems, monomorphisation times out.
-Second, each prover is run on the remaining monomorphised problems, and the results are tallied in the table's `Mono' column.
+Second, each prover is run on the remaining monomorphised problems, and the results are tallied in the `Mono' column.
 In addition to the polymorphic provers used in the `Native' tests, the monomorphic prover E is run on the monomorphised problems to provide an additional point of comparison.
-Instead of running each prover for 30 seconds on the monomorphised problems, the monomorphisation time (rounded up to the nearest second) is subtracted to compare fairly against the `Native' column, which does not have a similar preprocessing phase.
+Instead of running each prover for 30~s on the monomorphised problems, the monomorphisation time (rounded up to the nearest second) is subtracted to compare fairly against the `Native' column, which does not have a similar preprocessing phase.
 

### 2025-06-25 iterative_monomorphisation.tex
avoid margin

@@
 Monomorphisation can be used to extend monomorphic provers to support polymorphic logics.
-We propose a pragmatic iterative approach.
+We describe a pragmatic iterative approach.
 We implemented it in the Zipperposition prover, where it is used to translate away polymorphism before invoking the monomorphic prover E as a backend.
@@
 
-We run an example through the detailed version of the iterative algorithm. Consider this initial problem:
+To illustrate the low level version of the iterative algorithm, we consider the following initial problem:
 \begin{itemize}
-   \item[\(\langle1\rangle\)] \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
-   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{f}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{f}\langle \alpha, \alpha \rangle(x, x) \land \sym{f}\langle \alpha, \beta\rangle(x, z)\)
+   \item[\(\langle1\rangle\)] \(\sym{p}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{p}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
+   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{p}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{p}\langle \alpha, \alpha \rangle(x, x) \land \sym{p}\langle \alpha, \beta\rangle(x, z)\)
 \end{itemize}
 %
-\(M_{\text{new}}\) is initialised with \(\{f\mapsto (\ty{int}, \ty{nat}); f\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{f\mapsto (\ty{list}(\alpha), \alpha); f\mapsto (\alpha, \alpha); f\mapsto (\alpha, \beta)\}\}\).
+\(M_{\text{new}}\) is initialised with \(\{\sym p\mapsto (\ty{int}, \ty{nat}); \sym p\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{\sym p\mapsto (\ty{list}(\alpha), \alpha); \sym p\mapsto (\alpha, \alpha); \sym p\mapsto (\alpha, \beta)\}\}\).
 Then we enter the main loop (assuming \Loop is at least 1).
@@
 
-\SubstGen returns and the substitutions are used to generate new type argument tuples.
+Then \SubstGen returns, and the substitutions are used to generate new type argument tuples.
 Each substitution is applied to the type arguments of the function symbols of formula \(\langle2\rangle\) because they are the only function symbols with non-monomorphic type arguments.
+The table below summarises the situation:
 \begin{table}
@@
 
-With a very small and simple initial problem, there are already up to nine new type arguments tuples generated at this step in the first iteration alone, though only four are unique.
+With a very small and simple initial problem, there are already up to nine new type arguments tuples generated at this step in the first iteration alone, although only four are unique.
 Once the new type argument tuples are added to their respective maps, a new iteration is begun.
@@
 
-\GenFormulae is the next function to be called.
-This function iterates over all non-monomorphic formulae of the initial problem, in our case this will only be \(\langle2\rangle\).
+The next function to be called is \GenFormulae.
+It iterates over all non-monomorphic formulae of the initial problem; in our case, this will only be \(\langle2\rangle\).
 The set of type variable tuples of \(\langle2\rangle\) is passed to \MonoSubst along with the set of all previously generated substitutions.
@@
 If \(\sigma_1\) is selected to instantiate \(\alpha\), both \(\sigma_2\) and \(\sigma_3\) will in turn be selected to instantiate \(\beta\).
-This will generate two monomorphising substitutions: \(\sigma_1 \circ \sigma_2 = \sigma_2\) and \(\sigma_1 \circ \sigma_3 = \sigma_3\) which are added to the set of monomorphising substitutions \(S_{\text{res}}\).
+This will generate two monomorphising substitutions, \(\sigma_1 \circ \sigma_2 = \sigma_2\) and \(\sigma_1 \circ \sigma_3 = \sigma_3\), which are added to the set of monomorphising substitutions \(S_{\text{res}}\).
 Now \(\sigma_2\) is selected.
-It instantiates both \(\alpha\) and \(\beta\) at once.
+It simultaneously instantiates \(\alpha\) and \(\beta\).
 Here, \(\sigma_2\) is already in \(S_{\text{res}}\) and is therefore ignored, and \(\sigma_3\) is treated similarly.
@@
 
-Our monomorphising substitutions have been generated.
-We only need to apply them to the formula they monomorphises.
+The monomorphising substitutions have been generated.
+We only need to apply them to the formula they monomorphise.
 This is repeated for all non-monomorphic formulae or until \Limit is reached.
@@
 For some larger input problems, \MonoSubst could be sufficiently explosive to only allow a handful of different formulae to be monomorphised.
-The ability to determine \SubstLimit independently can help avoid this issue.
-
-Finally, all monomorphic formulae, both new and old, are mangled.
-This step is unecessary if the target language allows type constructors.
-In our example two new monomorphic formulae are generated.
-The algorithm outputs:
+The ability to set \SubstLimit independently can help avoid this issue.
+
+Finally, if the target language supports only nullary type construtors, the types of all monomorphic formulae, both new and old, are mangled.
+%This step is unecessary if the target language allows type constructors.
+In our example, two new monomorphic formulae are generated. The algorithm outputs
 \begin{enumerate}
-   \item \(\sym{f}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
-   \item[3.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{int}\rangle(x, z)\)
-   \item[4.] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \sym{f}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{f}\langle \ty{int}, \ty{int} \rangle(x, x) \land \sym{f}\langle \ty{int}, \ty{nat}\rangle(x, z)\)
+   \item[\(\langle1\rangle\)] \(\sym{p}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{p}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
+   \item[\(\langle3\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \begin{aligned}[t]
+     & \sym{p}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{p}\langle \ty{int}, \ty{int} \rangle(x, x) \land {} \\[-\jot]
+     & \sym{p}\langle \ty{int}, \ty{int}\rangle(x, z)\end{aligned}\)
+   \item[\(\langle4\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \begin{aligned}[t]
+     & \sym{p}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{p}\langle \ty{int}, \ty{int} \rangle(x, x) \land {} \\[-\jot]
+     & \sym{p}\langle \ty{int}, \ty{nat}\rangle(x, z)\end{aligned}\)
 \end{enumerate}
@@
 \looseness=-1
-Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments.
+Table~\ref{nmon_ty_args} is similar to Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments.
 The bound values are lower because we found non-monomorphic type arguments to be combinatorially explosive in preliminary evaluations.

### 2025-06-26 iterative_monomorphisation.tex
cleanup
spellings

@@
 % so please use T1 fonts in your manuscript whenever possible.
-% Other font encondings may result in incorrect characters.
+% Other font encodings may result in incorrect characters.
 %
@@
 %These definitions depend on the precise form of the input formulae, which has been left undefined.
-\(N\) is parameterised with \(\varphi\) because type variables are implicitly quantified at the formula level.
+\(N\) is parametrised with \(\varphi\) because type variables are implicitly quantified at the formula level.
 The formula indicates the scope of type variables.
@@
 
-To illustrate the low level version of the iterative algorithm, we consider the following initial problem:
+To illustrate the low level version of the iterative algorithm, we consider the following (admittedly contrived) initial problem:
 \begin{itemize}
@@
 Finally, if the target language supports only nullary type construtors, the types of all monomorphic formulae, both new and old, are mangled.
-%This step is unecessary if the target language allows type constructors.
+%This step is unnecessary if the target language allows type constructors.
 In our example, two new monomorphic formulae are generated. The algorithm outputs
@@
 Table \ref{e_settings} shows the impact of the options with which we call the E prover.
-The point at which Zipperposition interupts its normal operation and begins the monomorphisation process is determined by the `E call point' parameter.
-It is expressed as a fraction of the total time alloted to Zipperposition.
+The point at which Zipperposition interrupts its normal operation and begins the monomorphisation process is determined by the `E call point' parameter.
+It is expressed as a fraction of the total time allotted to Zipperposition.
 The E timeout (in seconds) limits how long E is run before being interrupted, at which point Zipperposition resumes normal operation.

### 2025-06-26 iterative_monomorphisation.tex
work on intro

@@
 Our evaluation shows that this approach increases Zipperposition's success rate.
-Moreover, we find that iterative monomorphisation outperforms native implementations of polymorphism.
+Moreover, we find that iterative monomorphisation outperforms some native implementations of polymorphism.
 
@@
 The alternative is to translate polymorphic problems to monomorphic problems.
-One approach \cite{mono-trans} is to encode a polymorphic type system using dedicated function or predicate symbols in a monomorphic logic.
-Another approach is to encode polymorphism using {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}.
-This method heuristically instantiates the formulae's type variables with concrete types.
-
-Any translation method based on instantiation will inevitably be incomplete.
-By a type version of the compactness theorem, for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of~\(\varphi\).
-However, this set cannot be computed in general~\cite[Theorem~1]{expr-poly-types}.
-
-B\"ohme's iterative approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}.
+
+One approach \cite{mono-trans} is to encode a complete polymorphic type system, %using dedicated function or predicate symbols in a monomorphic logic,
+but this increases the size of the input problem substantially and slows down provers \cite{mono-trans}.
+%
+Another approach to encode polymorphism is based on {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}.
+%
+Iterative monomorphisation heuristically instantiates the formulae's type variables with concrete types.
+%
+However, any translation relying on a finite number of instantiations is inevitably incomplete.
+By a typed version of the compactness theorem, for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of~\(\varphi\), but
+it cannot be computed in general~\cite[Theorem~1]{expr-poly-types}.
+
+B\"ohme's approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}.
 This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers.
 However, it is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
-\relax{SMT-LIB} makes use of this monomorphisation approach \cite{smt-lib-mono}.
-A similar algorithm also appears to be implemented in the \relax{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
+\relax{SMT-LIB} also uses iterative monomorphisation \cite{smt-lib-mono}.
+Moreover, a similar algorithm appears to be implemented in the \relax{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
 
@@
 The algorithm works as follows.
-We assume problems to be sets of formulae.
+The input problem is a set of formulae.
 All the problem's symbols are collected, and the polymorphic symbol instances are matched against the monomorphic ones.
@@
 However, because new types emerge through matching, $\ty{list}(\ty{list}(\ty{int}))$ can be obtained only once the $\ty{list}(\ty{int})$ instance has been generated.
-This example also shows that an infinite set of instances can be generated.
+%This example also shows that an infinite set of instances could be produced if the algorithm kept going for ever.
 

### 2025-06-26 iterative_monomorphisation.tex
edits

@@
 
-The algorithm is sound because the newly generated formulae are instances of the initial problem's formulae. %, where type variables have been instantiated with monomorphic types.
+The algorithm is trivially sound because the newly generated formulae are instances of the initial problem's formulae. %, where type variables have been instantiated with monomorphic types.
 However, it is not complete.
@@
 \begin{enumerate}
-   \item \(\sym{p}\langle \ty{int}\rangle(0)\)
-   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){.}\; \sym{p}\langle\alpha\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
+   \item[$\langle1\rangle$] \(\sym{p}\langle \ty{int}\rangle(0)\)
+   \item[$\langle2\rangle$] \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){.}\; \sym{p}\langle\alpha\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
 \end{enumerate}
@@
 \begin{enumerate}
-   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){.}\; \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
+   \item[$\langle3\rangle$] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){.}\; \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
 \end{enumerate}
@@
 \begin{enumerate}
-   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){.}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
+   \item[$\langle4\rangle$] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){.}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
 \end{enumerate}
@@
 \begin{enumerate}
-   \item[5.] 
+   \item[$\langle5\rangle$]
          \( \forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){.} \\ \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
@@
    \item[\labelitemi] \emph{type\_vars\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set.
-This function is overloaded to accept a formula \(\varphi\) as input in which case it returns the set of all type variables which occur in the formula.
+This function is overloaded to accept a formula \(\varphi\) as input, in which case it returns the set of all type variables which occur in the formula.
    \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, matches \(\upsilon\) against \(\tau\) and either fails or returns \emph{Some\((\sigma)\)}, where \(\sigma\) results from the matching.
@@
 The formula monomorphisation algorithm is given in Figure~\ref{mono_step}.
-It forms the core of the process by computing new type argument tuples for a single formula.
+It forms the core of the process.
+Essentially, it computes new type argument tuples for a single formula.
 Type argument tuples are matched against each other to obtain a set of substitutions which is iterated over in the outermost loop.
@@
 We give a sketch of how such growth can occur for a single formula.
-If we assume that after \(k\) iterations, there are \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type argument tuples, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
+If we assume that after \(k\) iterations, there are \(N_k\) type argument tuples in total divided evenly between monomorphic and non-monomorphic type argument tuples, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions.
 Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples.
@@
    \item[\(\langle1\rangle\)] \(\sym{p}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{p}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
-   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{,}\; \sym{p}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{p}\langle \alpha, \alpha \rangle(x, x) \land \sym{p}\langle \alpha, \beta\rangle(x, z)\)
+   \item[\(\langle 2\rangle\)] \(\forall x: \alpha{,}\; y:\ty{list}(\alpha){,}\; z: \beta{.}\; \sym{p}\langle \ty{list}(\alpha), \alpha\rangle(y, x) \land \sym{p}\langle \alpha, \alpha \rangle(x, x) \land \sym{p}\langle \alpha, \beta\rangle(x, z)\)
 \end{itemize}
@@
    \item[\(\langle1\rangle\)] \(\sym{p}\langle \ty{int}, \ty{nat}\rangle(-1, 3) \land \sym{p}\langle \ty{int}, \ty{int}\rangle(-1, -2)\)
-   \item[\(\langle3\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{,}\; \begin{aligned}[t]
+   \item[\(\langle3\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{int}{.}\; \begin{aligned}[t]
      & \sym{p}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{p}\langle \ty{int}, \ty{int} \rangle(x, x) \land {} \\[-\jot]
      & \sym{p}\langle \ty{int}, \ty{int}\rangle(x, z)\end{aligned}\)
-   \item[\(\langle4\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{,}\; \begin{aligned}[t]
+   \item[\(\langle4\rangle\)] \(\forall x: \ty{int}{,}\; y:\ty{list}(\ty{int}){,}\; z: \ty{nat}{.}\; \begin{aligned}[t]
      & \sym{p}\langle \ty{list}(\ty{int}), \ty{int}\rangle(y, x) \land \sym{p}\langle \ty{int}, \ty{int} \rangle(x, x) \land {} \\[-\jot]

### 2025-06-26 iterative_monomorphisation.tex
pass on example

@@
 %
-\(M_{\text{new}}\) is initialised with \(\{\sym p\mapsto (\ty{int}, \ty{nat}); \sym p\mapsto (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{\sym p\mapsto (\ty{list}(\alpha), \alpha); \sym p\mapsto (\alpha, \alpha); \sym p\mapsto (\alpha, \beta)\}\}\).
+\(M_{\text{new}}\) is initialised with \(\{\sym p\mapsto \{(\ty{int}, \ty{nat}), (\ty{int}, \ty{int})\}\) and \(N_{\text{new}}\) with \(\{\langle 2\rangle\mapsto \{\sym p\mapsto \{(\ty{list}(\alpha), \alpha), (\alpha, \alpha), (\alpha, \beta)\}\}\}\).
 Then we enter the main loop (assuming \Loop is at least 1).
@@
 Nothing happens for formula \(\langle1\rangle\) because it contains no type variables.
-For formula \(\langle2\rangle\), \(M_{\text{new}}\) and \(N_{\text{new}}(\langle2\rangle)\) are passed as argument to \SubstGen.
+For formula \(\langle2\rangle\), \(M_{\text{new}}\) and \(N_{\text{new}}(\langle2\rangle)\) are passed as arguments to \SubstGen.
 
@@
 The table below summarises the situation:
-\begin{table}
-\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
+%
+\begin{center}
+\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
    \toprule
@@
 \end{tabular}
-\end{table}
-
-With a very small and simple initial problem, there are already up to nine new type arguments tuples generated at this step in the first iteration alone, although only four are unique.
+\end{center}
+%
+With a simple two-formula initial problem, there are already up to nine new type arguments tuples generated at this step in the first iteration alone, although only four are unique.
 Once the new type argument tuples are added to their respective maps, a new iteration is begun.
@@
 
-The monomorphising substitutions have been generated.
+Now, the monomorphising substitutions have been generated.
 We only need to apply them to the formula they monomorphise.
 This is repeated for all non-monomorphic formulae or until \Limit is reached.
-The ordering of formulae will impact the output problem in the latter case.
+The order of formulae will impact the output problem in the latter case.
 For some larger input problems, \MonoSubst could be sufficiently explosive to only allow a handful of different formulae to be monomorphised.
@@
 %This step is unnecessary if the target language allows type constructors.
-In our example, two new monomorphic formulae are generated. The algorithm outputs
+Assuming that mangling is not necessary, the algorithm outputs
 \begin{enumerate}
@@
 \end{enumerate}
+In this example, two new monomorphic formulae are generated.
 

### 2025-06-26 iterative_monomorphisation.tex
edits

@@
 
-To detect if overfitting took place, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems.
+To guard against overfitting took place, we carried out the part of the evaluation related to parameter optimisation and all preliminary evaluations on 500 randomly chosen problems out of the 1534 selected problems.
 We carried out the rest of the evaluations on the remaining 1034 problems.
@@
 Because the space of possible base configurations is too large to evaluate exhaustively, we evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, but we disabled E.
-Our preliminary evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
+The preliminary evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
 This configuration became the base configuration, which we used as a basis to evaluate the monomorphisation options.
@@
    \item \PolyFloor: \(50\)
-   \item substitution ordering: separation
+   \item substitution order: separation
    \item \SubstLimit: \(\infty\)
@@
 The limit on newly generated type arguments is determined by three components that form a natural group of bounds.
+
 The table shows that generating no new monomorphic type arguments seems to be the best approach.
 This result may seem counterintuitive, but it is possible to monomorphise formulae without generating monomorphic type arguments.
-This is because non-monomorphic type argument generation can generate substitutions that instantiate one or more type variables, and it is these substitutions that are used to monomorphise formulae.
+This is because non-monomorphic type argument generation can produce substitutions that instantiate one or more type variables, and it is these substitutions that are used to monomorphise formulae.
 
@@
 The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related through the role they play in the construction of monomorphising substitutions.
-The substitution ordering heuristic has a significant impact on monomorphising substitution generation.
-The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first.
-The ``random'' ordering randomly shuffles substitutions.
-Finally, the ``separation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently.
-Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
+The substitution order heuristic has a significant impact on monomorphising substitution generation.
+The ``age'' order of substitution orders substitutions generated in earlier iterations are ordered first.
+The ``random'' order randomly shuffles substitutions.
+Finally, the ``separation'' order separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently.
+Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the order heuristic.
 The substitution generation phase occurs once all type arguments have been generated.
-The bound limiting the number of monomorphising substitutions per formula is directly related to the ordering which dictates how such monomorphising substitutions are generated.
-The %substitution ordering
+The bound limiting the number of monomorphising substitutions per formula is directly related to the order which dictates how such monomorphising substitutions are generated.
+The %substitution order
 heuristic greatly affects monomorphising substitution generation.
-The `age' ordering of substitution orders substitutions generated in earlier iterations first.
-The `random' ordering randomly shuffles substitutions.
-Finally, the `separation' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently.
-Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seem to affect performance when the `separation' heuristic is used.
+The `age' order of substitution orders substitutions generated in earlier iterations first.
+The `random' order randomly shuffles substitutions.
+Finally, the `separation' order separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently.
+Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions seem to affect performance only when the `separation' heuristic is used.
 % TODO find a better term than "separation heuristic" maybe separation ?
@@
    \toprule
-   & \multicolumn{3}{c}{substitution ordering} \\
+   & \multicolumn{3}{c}{substitution order} \\
    \multirow{1}{6em}{mono subst} & age & random & separation\\
@@
 For larger problems, the monomorphisation algorithm may time out despite the bounds.
-In these cases neither Zipperposition nor E will have had a chance to try to solve the problem.
+In these cases, neither Zipperposition nor E will have had a chance to try to solve the problem.
 To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation.
@@
 %Vampire failed to parse a significant number of the benchmarks problems.
-The fix was unavailable for Vampire's higher order branch at the time of the evaluation.
+The fix for these issues was unavailable for Vampire's higher order branch at the time of the evaluation.
 
@@
 
-We described an algorithm for iteratively instantiating polymorphic types to
-produce monomorphic problems as a preprocessor. Our primary motivation was to
-improve the success rate of Zipperposition and its monomorphic E backend, and
-indeed our evaluation shows a clear improvement. We also saw that even with
-automatic provers that support polymorphism, iterative monomorphisation is a
-better alternative in practice.
+We described a translation algorithm that iteratively instantiates polymorphic types to
+produce monomorphic problems. Our primary motivation was to improve the success
+rate of Zipperposition and its monomorphic E backend, and indeed our evaluation
+shows a clear improvement. We also saw that even with automatic provers that
+support polymorphism, iterative monomorphisation is a better alternative in
+practice.
 

### 2025-07-01 iterative_monomorphisation.tex
letter
clarified portfolio config explanation and why we have separate type argument maps

@@
 \usepackage{array}
+\usepackage[misc]{ifsym}
 \usepackage{multirow}
@@
 \lstset{breaklines=true}
+
+\newcommand\CORR{(\Letter)}
 
@@
 
-\author{Tanguy Bozec\inst{1,2}\orcid{0009-0005-9497-4168} \and
+\author{Tanguy Bozec\inst{1,2}\textsuperscript{\CORR}\orcid{0009-0005-9497-4168} \and
 Jasmin Blanchette\inst{2}\orcid{0000-0002-8367-0936}}
@@
 The iterative monomorphisation algorithm takes a polymorphic problem as input and returns a monomorphic problem.
-It works by applying a bounded number of iterations, each taking a polymorphic problem as argument and returning a problem with new partially instantiated formulae.
-A single iteration consists of a collection phase and an instantiation phase.
-Once the iterations are completed, a final step discards all non-monomorphic formulae returned by the last iteration.
+It applies a bounded number of iterations, each taking a polymorphic problem as argument and returning a problem with new partially instantiated formulae.
+Once the iterations are completed, a final step discards all non-monomorphic formulae.% returned by the last iteration.
 
@@
 This is not necessary for \(M\) since all the types it contains are monomorphic.
+To avoid copying the monomorphic types for each formula, \(M\) and \(N\) are kept separate.
 
@@
 Before finding values to assign to the bounds of the monomorphisation algorithm, we must choose which base options to run Zipperposition with.
-The prover has a portfolio mode with several configurations.
-Because the space of possible base configurations is too large to evaluate exhaustively, we evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, but we disabled E.
-The preliminary evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
+Because the space of possible base configurations is too large to evaluate exhaustively, we evaluated, in a preliminary experiment, all pre-existing portfolio configurations that called E against our benchmark suite of 500 problems.
+Since E could not treat these non-monomorphic problems, it was disabled.
+The preliminary evaluation found that the \verb|40_b.comb| configuration performed best by proving 131 problems.
 This configuration became the base configuration, which we used as a basis to evaluate the monomorphisation options.

### 2025-07-02 iterative_monomorphisation.tex
did

@@
 
-One approach \cite{mono-trans} is to encode a complete polymorphic type system, %using dedicated function or predicate symbols in a monomorphic logic,
+\looseness=-1
+To achieve this, one approach \cite{mono-trans} is to encode a complete polymorphic type system, %using dedicated function or predicate symbols in a monomorphic logic,
 but this increases the size of the input problem substantially and slows down provers \cite{mono-trans}.
@@
 This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers.
-However, it is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
-\relax{SMT-LIB} also uses iterative monomorphisation \cite{smt-lib-mono}.
-Moreover, a similar algorithm appears to be implemented in the \relax{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
+However, it is documented only superficially~\cite[Section 2.2.1]{sb-phd}.
+An iterative monomorphisation approach is also described in the context of SMT-LIB \cite{smt-lib-mono}.
+Moreover, a similar algorithm appears to have been implemented in the \relax{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
 

### 2025-09-08 paper/iterative_monomorphisation.tex
reorganising files and adding old presentation (to be thouroghly rewritten)

--- 2025-07-02 iterative_monomorphisation.tex
@@
 By contrast, many automatic provers only work with monomorphic logics.
-One way to close this gap is to extend provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}.
+One way to close this gap is to extend automatic provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}.
 However, this entails a lot of work that must be redone for every prover.
@@
 An iterative monomorphisation approach is also described in the context of SMT-LIB \cite{smt-lib-mono}.
-Moreover, a similar algorithm appears to have been implemented in the \relax{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
+Moreover, a similar algorithm appears to be implemented in the \relax{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
 
